---
title: "3D Spatial Gene Expression Plotter"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(rgl)
library(Rvcg)
library(Seurat)
library(dplyr)
library(viridis)

# Initialize rgl for HTML knitting
setupKnitr(autoprint = TRUE)

# ==========================================
# CONFIGURATION
# ==========================================
# Change this variable to plot any gene of interest
target_gene <- "CG5767" 
```

## 1. Setup Directories
Load the Blender 3D models and SWC centrelines.

```{r define_dirs}
blender_dir <- "/home/mattia/Thesis/3D model"
swc_dir <- "/mnt/d/BlackieGasparEtAl2024-sourceData/BlackieGasparEtAl2024-sourceData/OregonR/Centrelines"
```

## 2. Load and Align 3D Meshes

We import the gut mesh and align the M4T centreline perfectly to its center of mass, utilizing the same uniform scaling factor as the established model.

```{r load_meshes}
# Load Meshes
gut_mesh    <- vcgImport(file.path(blender_dir, "gut.obj"))

# Load Centreline
swc_file <- file.path(swc_dir, "M4T.swc")
centreline <- read.table(swc_file, comment.char = "#")
colnames(centreline) <- c("id", "type", "x", "y", "z", "radius", "parent")

# Align Centreline to Gut Mesh mathematically
c_pts <- centreline[, c("x", "y", "z")]
gut_pts <- t(gut_mesh$vb[1:3, ])

c_center <- colMeans(c_pts)
gut_center <- colMeans(gut_pts)

# Shift to 0,0,0
c_aligned <- sweep(c_pts, 2, c_center, "-")

# Scale uniformly by 1.20 (based on previous optimal alignment)
c_aligned <- c_aligned * 1.20

# Shift to Gut's Center of Mass
c_aligned <- sweep(c_aligned, 2, gut_center, "+")

# Overwrite with perfectly aligned coordinates
centreline$x <- c_aligned[, "x"]
centreline$y <- c_aligned[, "y"]
centreline$z <- c_aligned[, "z"]

# Calculate cumulative percentage length along the centreline
pts <- as.matrix(centreline[, c("x", "y", "z")])
dists <- sqrt(rowSums((pts[-1,] - pts[-nrow(pts),])^2))
centreline$cum_dist <- c(0, cumsum(dists))
centreline$pct_length <- centreline$cum_dist / max(centreline$cum_dist)
```

## 3. Map Transcriptomics Data

We load the single-cell data with transferred spatial pseudotime and map it onto the 3D centreline coordinates.

```{r map_data}
seurat_path <- "../Random data/sc_with_spatial_scores.rds"
if (!file.exists(seurat_path)) stop("Seurat object not found.")

seurat_obj <- readRDS(seurat_path)

# Use Smoothed Pseudotime for positional mapping (scale 0-100)
pt_values <- seurat_obj$Smoothed_Pseudotime

# Filter cells lacking pseudotime
valid_cells <- !is.na(pt_values)
seurat_obj <- subset(seurat_obj, cells = Cells(seurat_obj)[valid_cells])
pt_values <- seurat_obj$Smoothed_Pseudotime

# Interpolation helper function
interpolate_coord <- function(target_pct, cline) {
  if (is.na(target_pct)) return(c(NA, NA, NA))
  
  idx_after <- which(cline$pct_length >= target_pct)[1]
  if (is.na(idx_after)) return(as.numeric(cline[nrow(cline), c("x", "y", "z")]))
  if (idx_after == 1) return(as.numeric(cline[1, c("x", "y", "z")]))
  
  idx_before <- idx_after - 1
  pct_before <- cline$pct_length[idx_before]
  pct_after <- cline$pct_length[idx_after]
  
  ratio <- ifelse(pct_after == pct_before, 0, (target_pct - pct_before) / (pct_after - pct_before))
  
  coord_before <- as.numeric(cline[idx_before, c("x", "y", "z")])
  coord_after <- as.numeric(cline[idx_after, c("x", "y", "z")])
  
  return(coord_before + ratio * (coord_after - coord_before))
}

# Map 0-100 pseudotime to 0.05-0.95 physical length (margins added for biological realism)
mapped_pcts <- 0.05 + 0.90 * (pt_values / 100)

mapped_coords <- t(sapply(mapped_pcts, interpolate_coord, cline = centreline))
seurat_obj$Mesh_X <- mapped_coords[, 1]
seurat_obj$Mesh_Y <- mapped_coords[, 2]
seurat_obj$Mesh_Z <- mapped_coords[, 3]
```

## 4. Map Gene Expression to Gut Mesh (Single-Cell)

To completely optimize the 3D rendering, we now project the gene expression directly onto the 3D gut mesh itself rather than drawing thousands of individual spheres. We do this by mathematically smoothing the expression along the pseudotime trajectory using LOESS, and then mapping each gut vertex to its closest point on that trajectory!

```{r extract_expression_and_map}
# Switch to SCT assay if available
if ("SCT" %in% Assays(seurat_obj)) {
  DefaultAssay(seurat_obj) <- "SCT"
}

if (!target_gene %in% rownames(seurat_obj)) {
  stop(paste("Target gene", target_gene, "not found in the Seurat object."))
}

expr_data <- FetchData(seurat_obj, vars = target_gene)[[1]]
expr_data_log <- log1p(expr_data)

# 1. Smooth expression along pseudotime to create a continuous profile
lo_sc <- loess(expr_data_log ~ pt_values, span = 0.2, control = loess.control(surface = "direct"))

# 2. Predict expression at each node of the centreline
# Reverse the mapping formula: mapped_pcts = 0.05 + 0.90 * (pt / 100) -> pt = (pct - 0.05) / 0.90 * 100
pt_nodes <- (centreline$pct_length - 0.05) / 0.90 * 100

# Clamp pt_nodes to the min/max of pt_values to prevent LOESS extrapolation crashes!
pt_nodes_sc_clamped <- pmax(min(pt_values), pmin(max(pt_values), pt_nodes))
node_expr_sc <- predict(lo_sc, newdata = pt_nodes_sc_clamped)
node_expr_sc[is.na(node_expr_sc)] <- 0
node_expr_sc[node_expr_sc < 0] <- 0

# 3. For each vertex on the gut mesh, find the closest centreline node
gut_verts <- t(gut_mesh$vb[1:3, ])
c_pts <- as.matrix(centreline[, c("x", "y", "z")])

cat("Mapping", nrow(gut_verts), "mesh vertices to centreline...\n")
nearest_node_idx <- integer(nrow(gut_verts))
for(i in 1:nrow(gut_verts)) {
  dsq <- (c_pts[,1] - gut_verts[i,1])^2 + 
         (c_pts[,2] - gut_verts[i,2])^2 + 
         (c_pts[,3] - gut_verts[i,3])^2
  nearest_node_idx[i] <- which.min(dsq)
}

# 4. Assign smoothed expression to each vertex and colorize
vert_expr_sc <- node_expr_sc[nearest_node_idx]

color_pal <- viridis(100, option = "magma") 
max_expr <- max(vert_expr_sc)
if (max_expr == 0) max_expr <- 1 
vert_scaled <- (vert_expr_sc - min(vert_expr_sc)) / (max_expr - min(vert_expr_sc))
vert_scaled[is.na(vert_scaled)] <- 0
color_idx <- round(vert_scaled * 99) + 1

# Clone mesh to avoid overwriting base for later plots
gut_mesh_sc <- gut_mesh
gut_mesh_sc$material$color <- color_pal[color_idx]
```

## 5. Render 3D Scene (Single-Cell)

Because we eliminated thousands of floating 3D spheres, this renders instantly and perfectly paints the model!

```{r render_scene, rgl=TRUE}
open3d()

# Base Mesh (now fully colored by gene expression!)
shade3d(gut_mesh_sc, alpha = 1.0) 

# Centreline
lines3d(centreline$x, centreline$y, centreline$z, color = "black", lwd = 2, alpha = 0.5)

# Adjust camera view
aspect3d("iso")
view3d(theta = 45, phi = 30, zoom = 0.8)

# Add floating title
text3d(x = min(gut_pts[,1]), y = max(gut_pts[,2]), z = max(gut_pts[,3]), 
       texts = paste("Gene:", target_gene, "(Single-Cell Projection)"), 
       col = "black", cex = 2)
```

## 6. Map Spatial Transcriptomics (ST) Data

We perform the exact same mapping logic for the Spatial Transcriptomics dataset.

```{r map_st_data}
st_path <- "../Random data/merged_guts.rds"
if (!file.exists(st_path)) stop("ST object not found.")

st_obj <- readRDS(st_path)
st_obj <- subset(st_obj, cells = grep("^Gut3_", Cells(st_obj), value = TRUE))

pt_values_st <- st_obj$Spatial_Pseudotime
valid_spots <- !is.na(pt_values_st)
st_obj <- subset(st_obj, cells = Cells(st_obj)[valid_spots])
pt_values_st <- st_obj$Spatial_Pseudotime

# Set correct assay (SCT or Segmented)
if ("SCT" %in% Assays(st_obj)) {
  DefaultAssay(st_obj) <- "SCT"
} else if ("Segmented" %in% Assays(st_obj)) {
  DefaultAssay(st_obj) <- "Segmented"
}

if (!target_gene %in% rownames(st_obj)) {
  warning(paste("Target gene", target_gene, "not found in the ST object. Plotting blank values."))
  expr_data_st <- rep(0, ncol(st_obj))
} else {
  expr_data_st <- FetchData(st_obj, vars = target_gene)[[1]]
}

expr_data_log_st <- log1p(expr_data_st)

# 1. Smooth expression along pseudotime
lo_st <- loess(expr_data_log_st ~ pt_values_st, span = 0.2, control = loess.control(surface = "direct"))

# 2. Predict expression at each centreline node
# Clamp to avoid ST LOESS extrapolation crashes
pt_nodes_st_clamped <- pmax(min(pt_values_st), pmin(max(pt_values_st), pt_nodes))
node_expr_st <- predict(lo_st, newdata = pt_nodes_st_clamped)
node_expr_st[is.na(node_expr_st)] <- 0
node_expr_st[node_expr_st < 0] <- 0

# 3. Assign smoothed expression to each vertex and colorize
# (We reuse the nearest_node_idx calculated earlier, saving computation time!)
vert_expr_st <- node_expr_st[nearest_node_idx]

max_expr_st <- max(vert_expr_st)
if (max_expr_st == 0) max_expr_st <- 1 
vert_scaled_st <- (vert_expr_st - min(vert_expr_st)) / (max_expr_st - min(vert_expr_st))
vert_scaled_st[is.na(vert_scaled_st)] <- 0
color_idx_st <- round(vert_scaled_st * 99) + 1

gut_mesh_st <- gut_mesh
gut_mesh_st$material$color <- color_pal[color_idx_st]
```

## 7. Render 3D Scene (ST)

```{r render_scene_st, rgl=TRUE}
open3d()

# Base Mesh
shade3d(gut_mesh_st, alpha = 1.0) 

# Centreline
lines3d(centreline$x, centreline$y, centreline$z, color = "black", lwd = 2, alpha = 0.5)

# Adjust camera view
aspect3d("iso")
view3d(theta = 45, phi = 30, zoom = 0.8)

# Add floating title
text3d(x = min(gut_pts[,1]), y = max(gut_pts[,2]), z = max(gut_pts[,3]), 
       texts = paste("Gene:", target_gene, "(Spatial Transcriptomics)"), 
       col = "black", cex = 2)
```
