---
title: "Topological Scanning Analysis (Starvation - 111125ST)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(Seurat)
library(dplyr)
library(clusterProfiler)
library(org.Dm.eg.db) 
library(GO.db) 

seurat_obj_path <- "/home/mattia/Thesis/111125ST_def/Random data/merged_guts.rds"
seurat_obj <- readRDS(seurat_obj_path)
```

## Generate Per-Cell GO Assay
We convert GO Biological Process terms into a per-cell activity matrix using `AddModuleScore`. This allows us to perform "Differential GO Expression" exactly like standard gene DGE.

```{r generate_go_assay}
# 1. Fetch all GO-BP terms and their corresponding genes for Drosophila
cat("Fetching GO-BP terms from org.Dm.eg.db...\n")
go_data <- as.data.frame(org.Dm.egGO2ALLEGS)
# Filter for Biological Process (BP)
go_data <- go_data[go_data$Ontology == "BP", ]

# Convert Entrez IDs to Gene Symbols
gene_symbols <- select(org.Dm.eg.db, keys = as.character(go_data$gene_id), columns = "SYMBOL", keytype = "ENTREZID")
go_data$SYMBOL <- gene_symbols$SYMBOL[match(go_data$gene_id, gene_symbols$ENTREZID)]
go_data <- go_data[!is.na(go_data$SYMBOL), ]

# Group genes by GO term
go_list <- split(go_data$SYMBOL, go_data$go_id)

# Filter GO terms by size (e.g., between 10 and 200 genes to avoid too broad or too narrow terms)
go_list <- go_list[sapply(go_list, length) >= 10 & sapply(go_list, length) <= 200]
cat(paste("Total filtered GO terms to score:", length(go_list), "\n"))

# Due to Seurat's AddModuleScore limitations with massive lists, we process in chunks
# (In a real massive run, UCell or escape packages might be faster, but AddModuleScore is native)
cat("Calculating per-cell GO module scores. This may take a few minutes...\n")

# To keep this script lightweight for the test, we could sample a subset of GO terms here, 
# but we run all filtered terms for the full exhaustive analysis.
go_scores <- list()
chunk_size <- 50
num_chunks <- ceiling(length(go_list) / chunk_size)

for(i in 1:num_chunks) {
  start_idx <- ((i-1) * chunk_size) + 1
  end_idx <- min(i * chunk_size, length(go_list))
  current_chunk <- go_list[start_idx:end_idx]
  
  # Calculate module scores
  seurat_obj <- AddModuleScore(seurat_obj, features = current_chunk, name = "GO_Score_", ctrl = 5)
  
  # Extract the newly added score columns
  new_cols <- paste0("GO_Score_", 1:length(current_chunk))
  scores_chunk <- seurat_obj@meta.data[, new_cols, drop=FALSE]
  colnames(scores_chunk) <- names(current_chunk)
  
  go_scores[[i]] <- scores_chunk
  
  # Remove the temporary columns from metadata to keep it clean
  seurat_obj@meta.data <- seurat_obj@meta.data[, !colnames(seurat_obj@meta.data) %in% new_cols]
}

# Combine all chunk scores into a single matrix
go_matrix <- do.call(cbind, go_scores)
go_matrix <- t(go_matrix) # Genes (GO Terms) x Cells

# Add as a new Assay to Seurat
seurat_obj[["GO"]] <- CreateAssayObject(data = go_matrix)
DefaultAssay(seurat_obj) <- "GO"
cat("Successfully added 'GO' Assay to the Seurat object!\n")
```

## Define Sliding Windows
We systematically divide the gut (10% to 90% pseudotime) into 10% overlapping windows.

```{r define_windows}
cell_pt <- seurat_obj$Spatial_Pseudotime

# Start: 10, End: 90, Step: 5, Window Size: 10
starts <- seq(10, 80, by = 5)
ends <- starts + 10

windows <- lapply(1:length(starts), function(i) c(starts[i], ends[i]))
names(windows) <- paste0("Win_", starts, "_", ends)

cat(paste("Generated", length(windows), "sliding windows.\n"))
```

## Exhaustive Pairwise Topological Scan
We iterate through all pairs of windows separated by at least 10% pseudotime and identify Differentially Expressed GO Terms.

```{r pairwise_scan}
results_list <- list()
summary_df <- data.frame(Pair = character(), Shared_GO_Terms = integer(), stringsAsFactors = FALSE)

# Generate pairs
for (i in 1:(length(windows) - 1)) {
  for (j in (i + 1):length(windows)) {
    
    # Check topological distance (must be >= 10% apart)
    # End of window 1 is windows[[i]][2]. Start of window 2 is windows[[j]][1].
    # Distance between them: windows[[j]][1] - windows[[i]][2]
    dist <- windows[[j]][1] - windows[[i]][2]
    
    if (dist >= 10) {
      pair_name <- paste0(names(windows)[i], "_AND_", names(windows)[j])
      
      # Define cells in each segment
      seg1_cells <- colnames(seurat_obj)[cell_pt >= windows[[i]][1] & cell_pt <= windows[[i]][2] & !is.na(cell_pt)]
      seg2_cells <- colnames(seurat_obj)[cell_pt >= windows[[j]][1] & cell_pt <= windows[[j]][2] & !is.na(cell_pt)]
      contact_cells <- c(seg1_cells, seg2_cells)
      
      if (length(contact_cells) > 20) {
        ident_vec <- rep("Rest", ncol(seurat_obj))
        names(ident_vec) <- colnames(seurat_obj)
        ident_vec[contact_cells] <- "Contact"
        Idents(seurat_obj) <- ident_vec
        
        # Run DGE on the GO Assay!
        # Because these are module scores (usually ranging -1 to 1), logfc.threshold needs to be smaller 
        # compared to raw RNA log counts. We look for highly specific GO terms.
        markers <- FindMarkers(seurat_obj, ident.1 = "Contact", ident.2 = "Rest", 
                               assay = "GO",
                               min.pct = 0, # Module scores are continuous, min.pct doesn't apply well
                               logfc.threshold = 0.05, 
                               return.thresh = 0.05, 
                               only.pos = TRUE)
        
        # Enforce expression in both sides (since module scores are continuous, we require a positive average score in both)
        score_data <- GetAssayData(seurat_obj, assay = "GO", layer = "data")[rownames(markers), , drop=FALSE]
        mean_seg1 <- rowMeans(score_data[, seg1_cells, drop=FALSE])
        mean_seg2 <- rowMeans(score_data[, seg2_cells, drop=FALSE])
        
        valid_terms <- rownames(markers)[mean_seg1 > 0.05 & mean_seg2 > 0.05]
        markers <- markers[valid_terms, , drop=FALSE]
        
        # Save results
        if (nrow(markers) > 0) {
          # Map GO IDs back to readable terms
          term_names <- Term(GOTERM[rownames(markers)])
          markers$Term_Name <- term_names
          
          results_list[[pair_name]] <- markers
          summary_df <- rbind(summary_df, data.frame(Pair = pair_name, Shared_GO_Terms = nrow(markers)))
        } else {
          summary_df <- rbind(summary_df, data.frame(Pair = pair_name, Shared_GO_Terms = 0))
        }
      }
    }
  }
}

cat(paste("Finished scanning", nrow(summary_df), "distant region pairs.\n"))
cat(paste("Found", sum(summary_df$Shared_GO_Terms > 0), "pairs with significant shared topological GO pathways!\n"))

# Save the summary matrix
write.csv(summary_df, "/home/mattia/Thesis/111125ST_def/Random data/Topological_Scanning_Summary.csv", row.names = FALSE)
```
