---
title: "Topological Association Analysis (ST) - New 3D Model"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(Seurat)
library(Rvcg)
library(rgl)
library(dplyr)
library(dbscan)
library(ggplot2)
```

## 1. Introduction
This script identifies topologically associated areas in the gut mesh—regions physically touching (low Euclidean distance) but topologically distant along the tube (high pseudotime distance). It utilizes the **new 3D modelling approach** directly painting the `gut.obj` mesh instead of the old Plotly scatter cache.

## 2. Load 3D Model and Data

```{r load_data}
blender_dir <- "/home/mattia/Thesis/3D model"
swc_dir <- "/mnt/d/BlackieGasparEtAl2024-sourceData/BlackieGasparEtAl2024-sourceData/OregonR/Centrelines"

# 1. Load Meshes
gut_mesh <- vcgImport(file.path(blender_dir, "gut.obj"))

# 2. Load Centreline
swc_file <- file.path(swc_dir, "M4T.swc")
centreline <- read.table(swc_file, comment.char = "#")
colnames(centreline) <- c("id", "type", "x", "y", "z", "radius", "parent")

# Align Centreline to Gut Mesh
c_pts <- centreline[, c("x", "y", "z")]
gut_pts <- t(gut_mesh$vb[1:3, ])

c_center <- colMeans(c_pts)
gut_center <- colMeans(gut_pts)

c_aligned <- sweep(c_pts, 2, c_center, "-")
c_aligned <- c_aligned * 1.20 # Scale
c_aligned <- sweep(c_aligned, 2, gut_center, "+")

centreline$x <- c_aligned[, "x"]
centreline$y <- c_aligned[, "y"]
centreline$z <- c_aligned[, "z"]

# Calculate cumulative percentage length
pts <- as.matrix(centreline[, c("x", "y", "z")])
dists <- sqrt(rowSums((pts[-1,] - pts[-nrow(pts),])^2))
centreline$cum_dist <- c(0, cumsum(dists))
centreline$pct_length <- centreline$cum_dist / max(centreline$cum_dist)

# Map pct_length to ST Pseudotime (0-100 scale)
pt_nodes <- (centreline$pct_length - 0.05) / 0.90 * 100

# Pre-calculate nearest node for each mesh vertex to paint the mesh later!
cat("Mapping", nrow(gut_pts), "mesh vertices to centreline...\n")
nearest_node_idx <- integer(nrow(gut_pts))
for(i in 1:nrow(gut_pts)) {
  dsq <- (pts[,1] - gut_pts[i,1])^2 + 
         (pts[,2] - gut_pts[i,2])^2 + 
         (pts[,3] - gut_pts[i,3])^2
  nearest_node_idx[i] <- which.min(dsq)
}

# 3. Load ST Object
st_path <- "../Random data/merged_guts.rds"
st_obj <- readRDS(st_path)
st_obj <- subset(st_obj, cells = grep("^Gut3_", Cells(st_obj), value = TRUE))

valid_spots <- !is.na(st_obj$Spatial_Pseudotime)
st_obj <- subset(st_obj, cells = Cells(st_obj)[valid_spots])

if ("SCT" %in% Assays(st_obj)) {
  DefaultAssay(st_obj) <- "SCT"
} else if ("Segmented" %in% Assays(st_obj)) {
  DefaultAssay(st_obj) <- "Segmented"
}
```

## 3. Find Topological Contacts

```{r find_contacts}
# 1. Euclidean vs Pseudotime Distances of the Centreline Nodes
euc_dist <- as.matrix(dist(pts))
pt_dist <- as.matrix(dist(pt_nodes))

# We use a strict Euclidean threshold (0.12) to ensure we only capture the most intimately touching segments 
# strictly confined to the highly coiled middle region, without bleeding along the tube.
contact_matrix <- (euc_dist < 0.12) & (pt_dist > 15)

contact_pairs <- which(contact_matrix, arr.ind = TRUE)
contact_pairs <- contact_pairs[contact_pairs[,1] < contact_pairs[,2], , drop = FALSE]

if(nrow(contact_pairs) == 0) {
  stop("No topological contacts found.")
}

pt_coords <- data.frame(
  pt1 = pt_nodes[contact_pairs[,1]],
  pt2 = pt_nodes[contact_pairs[,2]]
)

eps_val <- 1.5 
clusters <- dbscan::dbscan(pt_coords, eps = eps_val, minPts = 3)

contact_pairs_df <- data.frame(
  idx1 = contact_pairs[,1],
  idx2 = contact_pairs[,2],
  pt1 = pt_coords$pt1,
  pt2 = pt_coords$pt2,
  cluster = clusters$cluster
) %>% filter(cluster > 0)

num_clusters <- length(unique(contact_pairs_df$cluster))
cat(paste("Found", num_clusters, "unique topological associations.\n"))

# Extract regions
contact_regions_list <- list()
for (c in unique(contact_pairs_df$cluster)) {
  c_pairs <- contact_pairs_df %>% filter(cluster == c)
  contact_regions_list[[as.character(c)]] <- list(
    seg1 = c(min(c_pairs$pt1), max(c_pairs$pt1)),
    seg2 = c(min(c_pairs$pt2), max(c_pairs$pt2)),
    nodes1 = unique(c_pairs$idx1),
    nodes2 = unique(c_pairs$idx2)
  )
}
```

## 4. Visualize Global Topological Associations in 3D

Before running the heavy differential gene expression, we can instantly visualize all detected topological loops globally on the 3D gut mesh.

```{r plot_global_associations, rgl=TRUE}
# Base mesh in light semi-transparent grey
v_colors_global <- rep("grey90", nrow(gut_pts))

# Define a palette of bright distinct colors for the loops
loop_colors <- c("firebrick1", "dodgerblue1", "forestgreen", "darkorchid1", "darkorange1")

for (i in seq_along(names(contact_regions_list))) {
  c <- names(contact_regions_list)[i]
  bounds <- contact_regions_list[[c]]
  
  col_to_use <- loop_colors[(i %% length(loop_colors)) + 1]
  
  # Paint both interacting segments of the loop with the same distinct color
  v_colors_global[nearest_node_idx %in% bounds$nodes1] <- col_to_use
  v_colors_global[nearest_node_idx %in% bounds$nodes2] <- col_to_use
}

gut_mesh_global <- gut_mesh
gut_mesh_global$material$color <- v_colors_global

open3d()
shade3d(gut_mesh_global, alpha = 0.95)
lines3d(centreline$x, centreline$y, centreline$z, color = "black", lwd = 2, alpha = 0.5)
aspect3d("iso")
view3d(theta = 45, phi = 30, zoom = 0.8)
text3d(x = min(gut_pts[,1]), y = max(gut_pts[,2]), z = max(gut_pts[,3]), 
       texts = paste("Global Topological Associations (N =", num_clusters, ")"), 
       col = "black", cex = 2)
```

## 5. DGE and Local Visualization per Association

```{r dge_and_plot, rgl=TRUE, fig.width=8, fig.height=4}
cell_pt <- st_obj$Spatial_Pseudotime
window <- 2 

for (c in names(contact_regions_list)) {
  assoc_name <- paste0("Association_", c)
  cat(paste("\n\n###", assoc_name, "\n"))
  
  bounds <- contact_regions_list[[c]]
  
  # --- 1. RENDER 3D MESH ---
  # Color the mesh mapping to Seg1 (red) and Seg2 (blue)
  v_colors <- rep("grey80", nrow(gut_pts))
  
  v_colors[nearest_node_idx %in% bounds$nodes1] <- "firebrick1"
  v_colors[nearest_node_idx %in% bounds$nodes2] <- "dodgerblue1"
  
  # Create localized mesh
  gut_mesh_assoc <- gut_mesh
  gut_mesh_assoc$material$color <- v_colors
  
  open3d()
  shade3d(gut_mesh_assoc, alpha = 0.9)
  lines3d(centreline$x, centreline$y, centreline$z, color = "black", lwd = 2, alpha = 0.5)
  aspect3d("iso")
  view3d(theta = 45, phi = 30, zoom = 0.8)
  text3d(x = min(gut_pts[,1]), y = max(gut_pts[,2]), z = max(gut_pts[,3]), 
         texts = paste("Topological Loop:", assoc_name), col = "black", cex = 2)
  
  # --- 2. DIFFERENTIAL GENE EXPRESSION ---
  is_in_seg1 <- !is.na(cell_pt) & (cell_pt >= (bounds$seg1[1] - window) & cell_pt <= (bounds$seg1[2] + window))
  is_in_seg2 <- !is.na(cell_pt) & (cell_pt >= (bounds$seg2[1] - window) & cell_pt <= (bounds$seg2[2] + window))
  
  cells_in_seg1 <- colnames(st_obj)[is_in_seg1]
  cells_in_seg2 <- colnames(st_obj)[is_in_seg2]
  cells_in_contact <- unique(c(cells_in_seg1, cells_in_seg2))
  cells_out <- colnames(st_obj)[!colnames(st_obj) %in% cells_in_contact]
  
  if (length(cells_in_contact) > 10 && length(cells_out) > 10) {
    ident_vec <- rep("Rest", ncol(st_obj))
    names(ident_vec) <- colnames(st_obj)
    ident_vec[cells_in_contact] <- "Contact"
    Idents(st_obj) <- ident_vec
    
    markers <- FindMarkers(st_obj, ident.1 = "Contact", ident.2 = "Rest", 
                           min.pct = 0.20, logfc.threshold = 0.40, return.thresh = 0.05, only.pos = TRUE)
    markers <- markers %>% filter((pct.1 - pct.2) > 0.15)
    
    expr_data <- GetAssayData(st_obj, layer = "data")[rownames(markers), , drop=FALSE]
    pct_seg1 <- rowMeans(expr_data[, cells_in_seg1, drop=FALSE] > 0)
    pct_seg2 <- rowMeans(expr_data[, cells_in_seg2, drop=FALSE] > 0)
    
    valid_genes <- rownames(markers)[pct_seg1 >= 0.20 & pct_seg2 >= 0.20]
    markers <- markers[valid_genes, ]
    
    top_markers <- markers %>% filter(p_val_adj < 0.05) %>% head(3) # Plot top 3
    
    if(nrow(top_markers) > 0) {
      for (gene in rownames(top_markers)) {
        data_df <- FetchData(st_obj, vars = c(gene, "Spatial_Pseudotime"))
        colnames(data_df) <- c("Expression", "Pseudotime")
        data_df <- data_df[!is.na(data_df$Pseudotime), ]
        
        y_min <- min(data_df$Expression, na.rm = TRUE) - 0.5
        y_max <- max(data_df$Expression, na.rm = TRUE) + 0.5
        
        p <- ggplot(data_df, aes(x = Pseudotime, y = Expression)) +
          geom_point(alpha = 0.3, size = 1, color = "grey50") +
          geom_smooth(method = "loess", span = 0.3, color = "darkblue", se = FALSE) +
          geom_rect(aes(xmin = bounds$seg1[1] - window, xmax = bounds$seg1[2] + window, 
                        ymin = y_min, ymax = y_max), fill = "firebrick1", alpha = 0.02, inherit.aes = FALSE) +
          geom_rect(aes(xmin = bounds$seg2[1] - window, xmax = bounds$seg2[2] + window, 
                        ymin = y_min, ymax = y_max), fill = "dodgerblue1", alpha = 0.02, inherit.aes = FALSE) +
          theme_minimal() +
          labs(title = paste(gene, "Expression -", assoc_name),
               subtitle = paste("Log2FC:", round(top_markers[gene, "avg_log2FC"], 2)))
        
        print(p)
      }
    } else {
      cat("No significant co-expressed markers found.\n")
    }
  } else {
    cat("Not enough cells for DGE.\n")
  }
}
```
