---
title: "Topological Association Analysis (SC Dataset - 111125ST_def)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(Seurat)
library(dplyr)
library(plotly)
library(clusterProfiler)
library(org.Dm.eg.db) 
library(dbscan) 
library(ggplot2)
```

## Introduction
This script identifies topologically associated areas in the 3D midgut mesh—regions that are physically touching (low Euclidean distance) but topologically distant along the gut tube (high pseudotime distance, >15%). It clusters the contacts into unique pairs of touching segments, performs Differential Gene Expression (DGE) and Gene Ontology (GO) Enrichment, and visualizations the expression linearly.

```{r load_data}
seurat_obj_path <- "/home/mattia/Thesis/111125ST_def/Random data/sc_with_spatial_scores.rds"
mesh_cache_file <- "/home/mattia/Thesis/111125ST_def/figures/3D_mesh_plots_starvation/tube_mesh_cache.rds"

if (!file.exists(mesh_cache_file)) {
  stop("Mesh cache not found. Please run plot_3D_expression_Starvation.Rmd first to generate the mesh.")
}

seurat_obj <- readRDS(seurat_obj_path)
mesh_data <- readRDS(mesh_cache_file)
centerline_pts <- mesh_data$centerline_pts
```

## Find Unique Contact Associations
We look for pairs of centerline points separated by >15% of the gut length, but physically closer than the gut diameter. We cluster them in 2D pseudotime space to guarantee each association connects exactly two continuous segments.

```{r find_contacts}
# 1. Calculate pairwise Euclidean distance matrix
coords <- as.matrix(centerline_pts[, 1:3])
euc_dist <- as.matrix(dist(coords))

# 2. Calculate pairwise Pseudotime distance matrix
pt_dist <- as.matrix(dist(centerline_pts$pt_scale))

# Find the normal gut radius
normal_radius <- mean(centerline_pts$radius, na.rm = TRUE)

# 3. Define contact pairs
# Physically close (Euclidean distance < 2 * normal_radius)
# Topologically distant (Pseudotime distance > 15%)
contact_matrix <- (euc_dist < (normal_radius * 2)) & (pt_dist > 15)

contact_pairs <- which(contact_matrix, arr.ind = TRUE)
contact_pairs <- contact_pairs[contact_pairs[,1] < contact_pairs[,2], ] # Keep upper triangle

if(nrow(contact_pairs) == 0) {
  stop("No topological contacts found with current thresholds.")
}

# 4. Group into Unique Associations (2D Pseudotime Clustering)
# Each pair has a pt1 and pt2 coordinate. Clustering these 2D points ensures 
# we only group pairs that link the exact same two continuous segments.
pt_coords <- data.frame(
  pt1 = centerline_pts$pt_scale[contact_pairs[,1]],
  pt2 = centerline_pts$pt_scale[contact_pairs[,2]]
)

# Cluster the contact links into continuous areas (using a strict 1.5% pseudotime window to enforce true continuity)
eps_val <- 1.5 
clusters <- dbscan::dbscan(pt_coords, eps = eps_val, minPts = 3)

contact_pairs_df <- data.frame(
  idx1 = contact_pairs[,1],
  idx2 = contact_pairs[,2],
  pt1 = pt_coords$pt1,
  pt2 = pt_coords$pt2,
  cluster = clusters$cluster
)

# Filter out noise (cluster 0)
contact_pairs_df <- contact_pairs_df %>% filter(cluster > 0)
num_clusters <- length(unique(contact_pairs_df$cluster))
cat(paste("Found", num_clusters, "unique topological associations (pairs of segments).\n"))
```

## Map Cells to Contact Associations
```{r map_cells}
contact_regions_list <- list()

for (c in unique(contact_pairs_df$cluster)) {
  c_pairs <- contact_pairs_df %>% filter(cluster == c)
  # A unique association is composed of exactly two segment ranges
  seg1_range <- c(min(c_pairs$pt1), max(c_pairs$pt1))
  seg2_range <- c(min(c_pairs$pt2), max(c_pairs$pt2))
  contact_regions_list[[as.character(c)]] <- list(seg1 = seg1_range, seg2 = seg2_range)
}

cell_pt <- seurat_obj$Smoothed_Pseudotime
window <- 2 # Match cells within +/- 2% pseudotime of the boundaries
```

## Differential Gene Expression & Linear Visualization
```{r perform_dge_and_plot}
dge_results <- list()
top_genes_list <- list()
all_plots_list <- list()

for (c in names(contact_regions_list)) {
  assoc_name <- paste0("Association_", c)
  cat(paste("\n--- Analyzing", assoc_name, "---\n"))
  
  bounds <- contact_regions_list[[c]]
  
  # Identify cells in Seg1 and Seg2 separately
  is_in_seg1 <- sapply(cell_pt, function(x) {
    if(is.na(x)) return(FALSE)
    return(x >= (bounds$seg1[1] - window) && x <= (bounds$seg1[2] + window))
  })
  
  is_in_seg2 <- sapply(cell_pt, function(x) {
    if(is.na(x)) return(FALSE)
    return(x >= (bounds$seg2[1] - window) && x <= (bounds$seg2[2] + window))
  })
  
  cells_in_seg1 <- colnames(seurat_obj)[is_in_seg1]
  cells_in_seg2 <- colnames(seurat_obj)[is_in_seg2]
  cells_in_contact <- unique(c(cells_in_seg1, cells_in_seg2))
  cells_out <- colnames(seurat_obj)[!colnames(seurat_obj) %in% cells_in_contact]
  
  if (length(cells_in_contact) > 10 && length(cells_out) > 10) {
    # Create a temporary identity vector
    ident_vec <- rep("Rest", ncol(seurat_obj))
    names(ident_vec) <- colnames(seurat_obj)
    ident_vec[cells_in_contact] <- "Contact"
    Idents(seurat_obj) <- ident_vec
    
    # Strict parameters to eliminate noise
    markers <- FindMarkers(seurat_obj, ident.1 = "Contact", ident.2 = "Rest", 
                           min.pct = 0.20, logfc.threshold = 0.40, return.thresh = 0.05, only.pos = TRUE)
                           
    # Enforce spatial specificity: must be substantially more frequent in loop
    markers <- markers %>% filter((pct.1 - pct.2) > 0.15)
    
    # --- ENFORCE CO-EXPRESSION IN BOTH DOMAINS ---
    # We require the gene to be expressed in at least 20% of cells in Seg1 AND 20% in Seg2
    expr_data <- GetAssayData(seurat_obj, layer = "data")[rownames(markers), , drop=FALSE]
    pct_seg1 <- rowMeans(expr_data[, cells_in_seg1, drop=FALSE] > 0)
    pct_seg2 <- rowMeans(expr_data[, cells_in_seg2, drop=FALSE] > 0)
    
    valid_genes <- rownames(markers)[pct_seg1 >= 0.20 & pct_seg2 >= 0.20]
    markers <- markers[valid_genes, ]
    
    dge_results[[assoc_name]] <- markers
    
    # All significant markers for plotting
    top_markers <- markers %>% filter(p_val_adj < 0.05)
    
    if(nrow(top_markers) > 0) {
      top_genes_list[[assoc_name]] <- rownames(top_markers)
      
      # Prepare data for Linear Plot
      for (gene in rownames(top_markers)) {
        
        # Calculate Scores
        logfc_score <- round(top_markers[gene, "avg_log2FC"], 2)
        pct_in <- round(top_markers[gene, "pct.1"] * 100, 1)
        pct_out <- round(top_markers[gene, "pct.2"] * 100, 1)
        
        score_text <- paste0("Enrichment (Log2FC): ", logfc_score, 
                             " | Expressed: ", pct_in, "% (In Loop) vs ", pct_out, "% (Out Loop)")
        
        data_df <- FetchData(seurat_obj, vars = c(gene, "Smoothed_Pseudotime"))
        colnames(data_df) <- c("Expression", "Pseudotime")
        data_df <- data_df[!is.na(data_df$Pseudotime), ]
        
        y_min <- min(data_df$Expression, na.rm = TRUE) - 0.5
        y_max <- max(data_df$Expression, na.rm = TRUE) + 0.5
        
        p <- ggplot(data_df, aes(x = Pseudotime, y = Expression)) +
          geom_point(alpha = 0.1, size = 0.5, color = "grey50") +
          geom_smooth(method = "loess", span = 0.3, color = "darkblue", se = FALSE) +
          # Highlight Segment 1
          geom_rect(aes(xmin = bounds$seg1[1] - window, xmax = bounds$seg1[2] + window, 
                        ymin = y_min, ymax = y_max), fill = "red", alpha = 0.01, inherit.aes = FALSE) +
          # Highlight Segment 2
          geom_rect(aes(xmin = bounds$seg2[1] - window, xmax = bounds$seg2[2] + window, 
                        ymin = y_min, ymax = y_max), fill = "blue", alpha = 0.01, inherit.aes = FALSE) +
          theme_minimal() +
          labs(title = paste(gene, "Expression Along Gut -", assoc_name),
               subtitle = score_text)
        
        # Save to interactive list for HTML Export (Base64 static for extreme speed and perfect layout)
        tmp_img <- tempfile(fileext = ".png")
        suppressWarnings(ggsave(tmp_img, plot = p, width = 8, height = 4, dpi = 150))
        img_uri <- base64enc::dataURI(file = tmp_img, mime = "image/png")
        
        div_wrapper <- htmltools::tags$div(
          style = "margin-bottom: 30px; padding: 15px; border-left: 5px solid #007bff; background: #f8f9fa;",
          htmltools::tags$h4(paste("Gene:", gene), style = "margin-top: 0; color: #007bff;"),
          htmltools::tags$p(score_text, style = "font-weight: bold; color: #444; margin-bottom: 15px;"),
          htmltools::tags$img(src = img_uri, style = "width: 100%; max-width: 800px; display: block;")
        )
        all_plots_list[[paste0(assoc_name, "_1_Gene_", gene)]] <- div_wrapper
      }
    } else {
      cat("No significant co-expressed markers found for this association.\n")
    }
  } else {
    cat(paste("Not enough cells for DGE in", assoc_name, "\n"))
  }
}
```

## Gene Ontology Enrichment
```{r go_enrichment, fig.width=8, fig.height=6}
# Load genes previously found in the Spatial Topological Association
spatial_html <- readLines("/home/mattia/Thesis/111125ST_def/Random data/Topological_Associations_Report.html")
spatial_genes <- unique(gsub(".*Gene: (.*?)</h4.*", "\\1", grep("Gene: ", spatial_html, value = TRUE)))
spatial_genes_upper <- toupper(spatial_genes)
cat(paste("Loaded", length(spatial_genes_upper), "genes from the spatial topological analysis to exclude from GO enrichment.\n"))

for (region_name in names(dge_results)) {
  markers <- dge_results[[region_name]]
  
  # Remove the genes that were also in the spatial topological analysis to get SC-specific ones (or baseline)
  up_genes <- rownames(markers %>% filter(p_val_adj < 0.05))
  
  # EXCLUDE the topologically associated genes found in the spatial dataset
  novel_genes <- setdiff(up_genes, spatial_genes_upper)
  
  if (length(novel_genes) > 0) {
    cat(paste("\nRunning GO enrichment for", region_name, "with", length(novel_genes), "novel genes (excluded", length(up_genes) - length(novel_genes), "known spatial genes)...\n"))
    
    gene_df <- bitr(novel_genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Dm.eg.db)
    
    if (nrow(gene_df) > 0) {
      go_res <- enrichGO(gene = gene_df$ENTREZID,
                         OrgDb = org.Dm.eg.db,
                         ont = "BP",
                         pAdjustMethod = "BH",
                         pvalueCutoff = 0.05,
                         qvalueCutoff = 0.2,
                         readable = TRUE)
      
      if (!is.null(go_res) && nrow(go_res) > 0) {
        p <- dotplot(go_res, showCategory = 10, title = paste("GO BP Enrichment -", region_name))
        print(p)
      } else {
        cat("No significant GO terms found.\n")
      }
    }
  }
}
```

## Visualize Contact Areas in 3D
```{r plot_contacts}
# Create a separate 3D plot for each unique association
plots_list <- list()

for (c in unique(contact_pairs_df$cluster)) {
  assoc_name <- paste0("Association_", c)
  
  # Start with a grey mesh
  vertex_colors <- rep("rgb(220, 220, 220)", nrow(mesh_data$v_matrix))
  
  c_pairs <- contact_pairs_df %>% filter(cluster == c)
  involved_idx <- unique(c(c_pairs$idx1, c_pairs$idx2))
  
  # Find vertices that map to these centerline points
  v_involved <- mesh_data$vertex_centerline_idx %in% involved_idx
  
  # Highlight the two touching segments in orange
  vertex_colors[v_involved] <- "rgb(255, 140, 0)"
  
  fig <- plot_ly(
    x = mesh_data$v_matrix[, 1],
    y = mesh_data$v_matrix[, 2],
    z = mesh_data$v_matrix[, 3],
    i = mesh_data$f_matrix[, 1],
    j = mesh_data$f_matrix[, 2],
    k = mesh_data$f_matrix[, 3],
    type = "mesh3d",
    vertexcolor = vertex_colors,
    name = assoc_name
  )
  
  fig <- fig %>% layout(
    scene = list(aspectmode = "data")
  )
  
  plots_list[[assoc_name]] <- fig
  
  fig_wrapper <- htmltools::tags$div(
    style = "margin-top: 60px; margin-bottom: 30px; padding-top: 20px; border-top: 2px solid #ccc;",
    htmltools::tags$h2(paste("Topological Contact:", assoc_name), style = "color: #333;"),
    fig
  )
  all_plots_list[[paste0(assoc_name, "_0_3DMesh")]] <- fig_wrapper
}

# Sort all plots so 3D mesh comes first for each association, followed by its linear plots
sorted_plots <- all_plots_list[sort(names(all_plots_list))]

# Generate combined HTML report for ALL plots (Linear + 3D) without knitting
html_path <- "../Random data/Topological_Associations_SC_Report.html"

full_html <- htmltools::tags$div(
  style = "font-family: Arial, sans-serif; max-width: 1000px; margin: 0 auto; padding: 20px;",
  htmltools::tags$h1("Single-Cell Topological Associations Report", style = "text-align: center; color: #2c3e50; border-bottom: 2px solid #2c3e50; padding-bottom: 10px;"),
  htmltools::tagList(sorted_plots)
)

htmltools::save_html(full_html, file = html_path)
cat(paste("\n\n>>> Successfully saved fast, perfectly formatted HTML report to:", html_path, "\n"))
```
