---
title: "Unbiased Genetic Topological Association Discovery (ST)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(Seurat)
library(dplyr)
library(ggplot2)
library(pheatmap)
library(viridis)
library(Matrix)
```

## 1. Introduction

Instead of relying on physical 3D mesh distances to find topological contacts, this script discovers them in a **purely unbiased, genetics-driven way**. We tile the gut (10–90% pseudotime) with overlapping 5% sliding windows (step = 2.5%), compute a pseudo-bulk expression profile per window, and then calculate pairwise Spearman correlations between all windows that are at least 10% apart. Unexpectedly high correlations between distant windows indicate genetically similar regions that may be in physical contact.

## 2. Load ST Data

```{r load_data}
st_path <- "../Random data/merged_guts.rds"
st_obj <- readRDS(st_path)
st_obj <- subset(st_obj, cells = grep("^Gut3_", Cells(st_obj), value = TRUE))

valid_spots <- !is.na(st_obj$Spatial_Pseudotime)
st_obj <- subset(st_obj, cells = Cells(st_obj)[valid_spots])

if ("SCT" %in% Assays(st_obj)) {
  DefaultAssay(st_obj) <- "SCT"
} else if ("Segmented" %in% Assays(st_obj)) {
  DefaultAssay(st_obj) <- "Segmented"
}

cell_pt <- st_obj$Spatial_Pseudotime
cat("Loaded", ncol(st_obj), "spots with valid pseudotime.\n")
```

## 3. Define Sliding Windows and Build Pseudo-Bulk Matrix

```{r build_pseudobulk}
# Window parameters
win_size <- 3      # 3% pseudotime width (finer resolution)
step_size <- 1     # 1% step (dense tiling)
pt_min <- 0
pt_max <- 100

# Generate window starts
win_starts <- seq(pt_min, pt_max - win_size, by = step_size)
win_labels <- paste0(win_starts, "-", win_starts + win_size, "%")

cat("Number of sliding windows:", length(win_starts), "\n")

# Select variable genes for the comparison (top 2000 HVGs)
st_obj <- FindVariableFeatures(st_obj, nfeatures = 2000)
hvg <- VariableFeatures(st_obj)

# Pre-extract the expression matrix once (huge optimization vs. repeated subsetting)
expr_mat <- GetAssayData(st_obj, layer = "data")[hvg, ]

# Build pseudo-bulk matrix: genes x windows (mean expression per window)
pseudobulk <- matrix(0, nrow = length(hvg), ncol = length(win_starts))
rownames(pseudobulk) <- hvg
colnames(pseudobulk) <- win_labels

spot_counts <- integer(length(win_starts))

for (i in seq_along(win_starts)) {
  w_start <- win_starts[i]
  w_end   <- w_start + win_size
  in_window <- which(cell_pt >= w_start & cell_pt < w_end)
  spot_counts[i] <- length(in_window)
  
  if (length(in_window) >= 3) {
    pseudobulk[, i] <- Matrix::rowMeans(expr_mat[, in_window, drop = FALSE])
  }
}

# Remove windows with too few spots
valid_windows <- spot_counts >= 3
cat("Valid windows (>= 5 spots):", sum(valid_windows), "/", length(win_starts), "\n")

pseudobulk <- pseudobulk[, valid_windows]
win_starts_valid <- win_starts[valid_windows]
win_labels_valid <- win_labels[valid_windows]
```

## 4. Compute Pairwise Spearman Correlation (Distant Windows Only)

```{r compute_correlations}
n_windows <- ncol(pseudobulk)
min_dist <- 10 # Minimum pseudotime distance between window midpoints

# Full correlation matrix (Spearman) — vectorized, extremely fast
cor_mat <- cor(pseudobulk, method = "spearman")
rownames(cor_mat) <- win_labels_valid
colnames(cor_mat) <- win_labels_valid

# Mask out pairs that are too close (< 10% midpoint distance)
midpoints <- win_starts_valid + win_size / 2
midpoint_dist <- as.matrix(dist(midpoints))

# Set close pairs to NA so they don't distract in the heatmap
cor_mat_distant <- cor_mat
cor_mat_distant[midpoint_dist < min_dist] <- NA
```

## 5. Full Correlation Heatmap

This heatmap shows the Spearman correlation between **all** windows. The diagonal and near-diagonal (neighboring windows) are expected to be high. The interesting signal is off-diagonal hotspots between distant windows.

```{r plot_full_heatmap, fig.width=10, fig.height=8}
# For the full heatmap, keep all values but annotate the diagonal proximity
pheatmap(cor_mat, 
         cluster_rows = FALSE, cluster_cols = FALSE,
         color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
         breaks = seq(-1, 1, length.out = 101),
         main = "Pairwise Spearman Correlation (All Sliding Windows)",
         fontsize_row = 7, fontsize_col = 7,
         border_color = NA)
```

## 6. Distant-Only Correlation Heatmap

Now we mask out all pairs within 10% pseudotime of each other (the trivially correlated neighbors), revealing only the long-range genetic similarities.

```{r plot_distant_heatmap, fig.width=10, fig.height=8}
pheatmap(cor_mat_distant, 
         cluster_rows = FALSE, cluster_cols = FALSE,
         color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
         breaks = seq(-1, 1, length.out = 101),
         na_col = "grey95",
         main = "Distant Windows Only (>= 10% Apart)",
         fontsize_row = 7, fontsize_col = 7,
         border_color = NA)
```

## 7. Identify Top Genetic Associations

```{r top_associations}
# Extract the upper triangle of distant correlations
cor_distant_upper <- cor_mat_distant
cor_distant_upper[lower.tri(cor_distant_upper, diag = TRUE)] <- NA

# Convert to long format
assoc_df <- data.frame(
  Window1 = rep(win_labels_valid, each = n_windows),
  Window2 = rep(win_labels_valid, times = n_windows),
  Midpoint1 = rep(midpoints, each = n_windows),
  Midpoint2 = rep(midpoints, times = n_windows),
  Correlation = as.vector(cor_distant_upper)
) %>% 
  filter(!is.na(Correlation)) %>%
  mutate(Distance = abs(Midpoint2 - Midpoint1)) %>%
  arrange(desc(Correlation))

cat("\n=== Top 15 Genetically Associated Distant Windows ===\n")
print(head(assoc_df, 15))
```

## 8. Scatter Plot of Correlation vs. Distance

This plot reveals whether certain long-range distances show unexpectedly high correlations (above the baseline trend), which would indicate genetically-driven topological associations.

```{r scatter_cor_vs_dist, fig.width=8, fig.height=5}
ggplot(assoc_df, aes(x = Distance, y = Correlation)) +
  geom_point(alpha = 0.4, size = 2, color = "steelblue") +
  geom_smooth(method = "loess", span = 0.5, color = "firebrick", se = TRUE) +
  geom_hline(yintercept = mean(assoc_df$Correlation, na.rm = TRUE), 
             linetype = "dashed", color = "grey40") +
  theme_minimal() +
  labs(title = "Spearman Correlation vs. Pseudotime Distance",
       subtitle = "Points above the trend suggest genetically-driven topological contact",
       x = "Pseudotime Distance Between Windows (%)",
       y = "Spearman Correlation")
```

## 9. Highlight Significant Outlier Pairs

We flag pairs whose correlation is more than 2 standard deviations above the mean for their distance bin. These are the candidate genetic topological associations.

```{r find_outliers, fig.width=10, fig.height=8}
# Bin by distance (5% bins)
assoc_df$dist_bin <- cut(assoc_df$Distance, breaks = seq(10, max(assoc_df$Distance) + 5, by = 5))

# Calculate z-score within each distance bin
assoc_df <- assoc_df %>%
  group_by(dist_bin) %>%
  mutate(
    bin_mean = mean(Correlation, na.rm = TRUE),
    bin_sd = sd(Correlation, na.rm = TRUE),
    z_score = (Correlation - bin_mean) / bin_sd
  ) %>%
  ungroup()

outliers <- assoc_df %>% filter(z_score > 2) %>% arrange(desc(z_score))

cat("\n=== Significant Genetic Associations (z > 2 within distance bin) ===\n")
if(nrow(outliers) > 0) {
  print(outliers %>% select(Window1, Window2, Distance, Correlation, z_score))
} else {
  cat("No significant outlier pairs found at z > 2 threshold.\n")
  cat("Showing top pairs with z > 1.5:\n")
  outliers <- assoc_df %>% filter(z_score > 1.5) %>% arrange(desc(z_score))
  print(outliers %>% select(Window1, Window2, Distance, Correlation, z_score))
}

# Re-plot the heatmap highlighting outlier pairs
if(nrow(outliers) > 0) {
  highlight_mat <- matrix(NA, nrow = n_windows, ncol = n_windows)
  rownames(highlight_mat) <- win_labels_valid
  colnames(highlight_mat) <- win_labels_valid
  
  for (r in 1:nrow(outliers)) {
    w1 <- outliers$Window1[r]
    w2 <- outliers$Window2[r]
    highlight_mat[w1, w2] <- outliers$Correlation[r]
    highlight_mat[w2, w1] <- outliers$Correlation[r]
  }
  
  pheatmap(highlight_mat, 
           cluster_rows = FALSE, cluster_cols = FALSE,
           color = colorRampPalette(c("white", "darkorange", "firebrick3"))(50),
           na_col = "grey95",
           main = "Significant Genetic Associations (Outlier Pairs)",
           fontsize_row = 7, fontsize_col = 7,
           border_color = NA)
}
```

## 10. Cluster Genetic Associations and Map to 3D Mesh

To visualize these genetic hotspots on the 3D model, we cluster the outlier pairs (to group adjacent sliding windows into continuous regions) and map them to the 3D mesh.

```{r load_mesh_and_cluster, rgl=TRUE}
library(dbscan)
library(Rvcg)
library(rgl)

# 1. Cluster Outliers
if(nrow(outliers) == 0) {
  outliers <- assoc_df %>% filter(z_score > 1.5)
}

if(nrow(outliers) > 0) {
  # Cluster based on their midpoints
  eps_val <- 4  # 4% pseudotime clustering window
  clusters <- dbscan::dbscan(outliers[, c("Midpoint1", "Midpoint2")], eps = eps_val, minPts = 1)
  outliers$cluster <- clusters$cluster
  
  num_clusters <- length(unique(outliers$cluster))
  cat(paste("Grouped into", num_clusters, "distinct genetic association regions.\n"))
  
  contact_regions_list <- list()
  for (c in unique(outliers$cluster)) {
    c_pairs <- outliers %>% filter(cluster == c)
    # Windows are win_size wide, centered on Midpoint
    seg1_min <- min(c_pairs$Midpoint1) - (win_size/2)
    seg1_max <- max(c_pairs$Midpoint1) + (win_size/2)
    seg2_min <- min(c_pairs$Midpoint2) - (win_size/2)
    seg2_max <- max(c_pairs$Midpoint2) + (win_size/2)
    
    contact_regions_list[[as.character(c)]] <- list(
      seg1 = c(seg1_min, seg1_max),
      seg2 = c(seg2_min, seg2_max)
    )
  }
  
  # 2. Load 3D Mesh
  blender_dir <- "/home/mattia/Thesis/3D model"
  swc_dir <- "/mnt/d/BlackieGasparEtAl2024-sourceData/BlackieGasparEtAl2024-sourceData/OregonR/Centrelines"
  
  gut_mesh <- vcgImport(file.path(blender_dir, "gut.obj"))
  
  swc_file <- file.path(swc_dir, "M4T.swc")
  centreline <- read.table(swc_file, comment.char = "#")
  colnames(centreline) <- c("id", "type", "x", "y", "z", "radius", "parent")
  
  c_pts <- centreline[, c("x", "y", "z")]
  gut_pts <- t(gut_mesh$vb[1:3, ])
  
  c_center <- colMeans(c_pts)
  gut_center <- colMeans(gut_pts)
  
  c_aligned <- sweep(c_pts, 2, c_center, "-")
  c_aligned <- c_aligned * 1.20 # Scale
  c_aligned <- sweep(c_aligned, 2, gut_center, "+")
  
  centreline$x <- c_aligned[, "x"]
  centreline$y <- c_aligned[, "y"]
  centreline$z <- c_aligned[, "z"]
  
  pts <- as.matrix(centreline[, c("x", "y", "z")])
  dists <- sqrt(rowSums((pts[-1,] - pts[-nrow(pts),])^2))
  centreline$cum_dist <- c(0, cumsum(dists))
  centreline$pct_length <- centreline$cum_dist / max(centreline$cum_dist)
  
  pt_nodes <- (centreline$pct_length - 0.05) / 0.90 * 100
  
  cat("Mapping mesh vertices...\n")
  nearest_node_idx <- integer(nrow(gut_pts))
  for(i in 1:nrow(gut_pts)) {
    dsq <- (pts[,1] - gut_pts[i,1])^2 + 
           (pts[,2] - gut_pts[i,2])^2 + 
           (pts[,3] - gut_pts[i,3])^2
    nearest_node_idx[i] <- which.min(dsq)
  }
  
  # For each region, find matching nodes
  for (c in names(contact_regions_list)) {
    bounds <- contact_regions_list[[c]]
    contact_regions_list[[c]]$nodes1 <- which(pt_nodes >= bounds$seg1[1] & pt_nodes <= bounds$seg1[2])
    contact_regions_list[[c]]$nodes2 <- which(pt_nodes >= bounds$seg2[1] & pt_nodes <= bounds$seg2[2])
  }
  
  # 3. Plot Individual Meshes per Association
  for (c in names(contact_regions_list)) {
    bounds <- contact_regions_list[[c]]
    
    v_colors <- rep("grey80", nrow(gut_pts))
    v_colors[nearest_node_idx %in% bounds$nodes1] <- "firebrick1"
    v_colors[nearest_node_idx %in% bounds$nodes2] <- "dodgerblue1"
    
    gut_mesh_assoc <- gut_mesh
    gut_mesh_assoc$material$color <- v_colors
    
    open3d()
    shade3d(gut_mesh_assoc, alpha = 0.9)
    lines3d(centreline$x, centreline$y, centreline$z, color = "black", lwd = 2, alpha = 0.5)
    aspect3d("iso")
    view3d(theta = 45, phi = 30, zoom = 0.8)
    text3d(x = min(gut_pts[,1]), y = max(gut_pts[,2]), z = max(gut_pts[,3]), 
           texts = paste("Genetic Association:", c), col = "black", cex = 2)
           
    # Explicitly print the widget so it appears in the RStudio Viewer when run inside a loop!
    print(rglwidget())
  }
}
```

## 11. Differential Expression of Genetic Associations

```{r dge_and_plot, fig.width=8, fig.height=4}
if(nrow(outliers) > 0) {
  window <- 2 
  for (c in names(contact_regions_list)) {
    assoc_name <- paste0("GenCluster_", c)
    cat(paste("\n\n###", assoc_name, "\n"))
    
    bounds <- contact_regions_list[[c]]
    
    # --- DIFFERENTIAL GENE EXPRESSION ---
    
    is_in_seg1 <- !is.na(cell_pt) & (cell_pt >= (bounds$seg1[1] - window) & cell_pt <= (bounds$seg1[2] + window))
    is_in_seg2 <- !is.na(cell_pt) & (cell_pt >= (bounds$seg2[1] - window) & cell_pt <= (bounds$seg2[2] + window))
    
    cells_in_seg1 <- colnames(st_obj)[is_in_seg1]
    cells_in_seg2 <- colnames(st_obj)[is_in_seg2]
    cells_in_contact <- unique(c(cells_in_seg1, cells_in_seg2))
    cells_out <- colnames(st_obj)[!colnames(st_obj) %in% cells_in_contact]
    
    if (length(cells_in_contact) > 10 && length(cells_out) > 10) {
      ident_vec <- rep("Rest", ncol(st_obj))
      names(ident_vec) <- colnames(st_obj)
      ident_vec[cells_in_contact] <- "Contact"
      Idents(st_obj) <- ident_vec
      
      markers <- FindMarkers(st_obj, ident.1 = "Contact", ident.2 = "Rest", 
                             min.pct = 0.20, logfc.threshold = 0.40, return.thresh = 0.05, only.pos = TRUE)
      
      if(nrow(markers) > 0) {
        markers <- markers %>% filter((pct.1 - pct.2) > 0.15)
      }
      
      if(nrow(markers) > 0) {
        expr_data <- GetAssayData(st_obj, layer = "data")[rownames(markers), , drop=FALSE]
        pct_seg1 <- rowMeans(expr_data[, cells_in_seg1, drop=FALSE] > 0)
        pct_seg2 <- rowMeans(expr_data[, cells_in_seg2, drop=FALSE] > 0)
        
        valid_genes <- rownames(markers)[pct_seg1 >= 0.20 & pct_seg2 >= 0.20]
        markers <- markers[valid_genes, ]
        
        top_markers <- markers %>% filter(p_val_adj < 0.05) %>% head(3)
        
        if(nrow(top_markers) > 0) {
          for (gene in rownames(top_markers)) {
            data_df <- FetchData(st_obj, vars = c(gene, "Spatial_Pseudotime"))
            colnames(data_df) <- c("Expression", "Pseudotime")
            data_df <- data_df[!is.na(data_df$Pseudotime), ]
            
            y_min <- min(data_df$Expression, na.rm = TRUE) - 0.5
            y_max <- max(data_df$Expression, na.rm = TRUE) + 0.5
            
            p <- ggplot(data_df, aes(x = Pseudotime, y = Expression)) +
              geom_point(alpha = 0.3, size = 1, color = "grey50") +
              geom_smooth(method = "loess", span = 0.3, color = "darkblue", se = FALSE) +
              geom_rect(aes(xmin = bounds$seg1[1] - window, xmax = bounds$seg1[2] + window, 
                            ymin = y_min, ymax = y_max), fill = "firebrick1", alpha = 0.02, inherit.aes = FALSE) +
              geom_rect(aes(xmin = bounds$seg2[1] - window, xmax = bounds$seg2[2] + window, 
                            ymin = y_min, ymax = y_max), fill = "dodgerblue1", alpha = 0.02, inherit.aes = FALSE) +
              theme_minimal() +
              labs(title = paste(gene, "Expression -", assoc_name),
                   subtitle = paste("Log2FC:", round(top_markers[gene, "avg_log2FC"], 2)))
            
            print(p)
          }
        } else {
          cat("No significant co-expressed markers found.\n")
        }
      } else {
        cat("No significant markers passed spatial thresholds.\n")
      }
    } else {
      cat("Not enough cells for DGE.\n")
    }
  }
}
```
