---
title: "3D Spatial Gene Expression Projection (03Slide4)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(Seurat)
library(plotly)
library(dplyr)
library(htmlwidgets)
```

## Introduction
This script maps Seurat spatial pseudotime data onto the 3D midgut mesh. It extracts the `Spatial_Pseudotime` from the provided Seurat object (for 03Slide4 dataset) and maps the values along the anterior-posterior axis of the 3D `.obj` mesh.

```{r define_parameters}
# Parameters
seurat_obj_path <- "/home/mattia/Thesis/03Slide4/Random data/merged_guts.rds"
obj_file_path <- "/home/mattia/Thesis/3D model/flygut_midgut_3D.obj"
output_dir <- "/home/mattia/Thesis/03Slide4/figures/3D_mesh_plots"

# The gene to visualize
gene_name <- "iotaTry"
```

## Load Data
```{r load_data}
cat("Loading Seurat Object...\n")
seurat_obj <- readRDS(seurat_obj_path)

if (!gene_name %in% rownames(seurat_obj[["Segmented"]])) {
  stop(paste("Gene", gene_name, "not found in the 'Segmented' assay of the Seurat object."))
}

cat("Extracting expression and pseudotime data...\n")
data_df <- FetchData(seurat_obj, vars = c(gene_name, "Spatial_Pseudotime"))
colnames(data_df) <- c("Expression", "Pseudotime")

# Remove NA pseudotime cells
data_df <- data_df[!is.na(data_df$Pseudotime), ]
```

## Parse 3D Mesh
We read the `.obj` file directly into R to extract the vertices and faces.
```{r parse_mesh}
cat("Parsing OBJ file...\n")
obj_lines <- readLines(obj_file_path)

v_lines <- obj_lines[startsWith(obj_lines, "v ")]
f_lines <- obj_lines[startsWith(obj_lines, "f ")]

# Extract vertices
v_matrix <- do.call(rbind, lapply(strsplit(v_lines, " "), function(x) {
  as.numeric(x[x != "v" & x != ""])
}))

# Extract faces (OBJ is 1-indexed, plotly needs 0-indexed)
f_matrix <- do.call(rbind, lapply(strsplit(f_lines, " "), function(x) {
  v_indices <- sapply(x[x != "f" & x != ""], function(y) as.numeric(strsplit(y, "/")[[1]][1]))
  v_indices - 1 # convert to 0-indexed
}))

num_vertices <- nrow(v_matrix)
```

## Map Pseudotime to Mesh Vertices
We map the 0-100 pseudotime scale onto the midgut portion of the mesh.
The first 260 rings (Malpighian/posterior) and the last 50 rings (anterior non-midgut) are excluded.
```{r map_pseudotime}
cat("Mapping vertices to pseudotime...\n")
# 1000 rings * 60 vertices = 60000 vertices + 2 cap centers = 60002
# Rings 0-259 are Malpighian (exclude). Rings 950-999 are Anterior non-midgut (exclude).
# Rings 260-949 are Midgut.
# Ring 949 = Anterior (Pseudotime 0), Ring 260 = Posterior (Pseudotime 100)

vertex_ring_idx <- floor((0:(num_vertices - 1)) / 60)

# Handle cap centers (last 2 vertices)
vertex_ring_idx[vertex_ring_idx > 999] <- c(0, 999)

# Map ring to pseudotime
vertex_pseudotime <- rep(NA, num_vertices)
midgut_mask <- vertex_ring_idx >= 260 & vertex_ring_idx <= 949

# pt = (949 - ring_idx) / (949 - 260) * 100
vertex_pseudotime[midgut_mask] <- ((949 - vertex_ring_idx[midgut_mask]) / (949 - 260)) * 100
```

## Compute Expression
We compute both a local binned average (raw) and a LOESS smoothed expression profile along the pseudotime axis.
```{r compute_expression}
# Compute Raw (Binned) Expression
data_df$bin <- cut(data_df$Pseudotime, breaks = seq(0, 100, length.out = 101), include.lowest = TRUE, labels = FALSE)
bin_means <- data_df %>% group_by(bin) %>% summarize(mean_expr = mean(Expression), .groups = 'drop')

vertex_raw_expr <- rep(NA, num_vertices)
vertex_bin <- cut(vertex_pseudotime[midgut_mask], breaks = seq(0, 100, length.out = 101), include.lowest = TRUE, labels = FALSE)

matched_expr <- bin_means$mean_expr[match(vertex_bin, bin_means$bin)]
matched_expr[is.na(matched_expr)] <- 0
vertex_raw_expr[midgut_mask] <- matched_expr

# Compute LOESS Smoothed Expression
loess_model <- loess(Expression ~ Pseudotime, data = data_df, span = 0.3)

vertex_smooth_expr <- rep(NA, num_vertices)
predicted_expr <- predict(loess_model, newdata = data.frame(Pseudotime = vertex_pseudotime[midgut_mask]))

# LOESS can sometimes predict negative values for 0-bounded data, clip at 0
predicted_expr[predicted_expr < 0] <- 0
vertex_smooth_expr[midgut_mask] <- predicted_expr
```

## Plotting Function
```{r define_plot_function}
generate_mesh_plot <- function(expr_values, title, filename) {
  pal <- colorRamp(c("#440154", "#3b528b", "#21908c", "#5dc863", "#fde725"))
  
  min_expr <- min(expr_values, na.rm = TRUE)
  max_expr <- max(expr_values, na.rm = TRUE)
  
  if (max_expr == min_expr) {
    norm_expr <- rep(0, length(expr_values))
  } else {
    norm_expr <- (expr_values - min_expr) / (max_expr - min_expr)
  }
  
  vertex_colors <- rep("rgb(200, 200, 200)", num_vertices) # Default grey for non-midgut
  
  valid_idx <- !is.na(norm_expr)
  colors_rgb <- pal(norm_expr[valid_idx])
  colors_rgb[is.na(colors_rgb)] <- 0
  
  vertex_colors[valid_idx] <- apply(colors_rgb, 1, function(x) paste0("rgb(", x[1], ",", x[2], ",", x[3], ")"))
  
  fig <- plot_ly(
    x = v_matrix[, 1],
    y = v_matrix[, 2],
    z = v_matrix[, 3],
    i = f_matrix[, 1],
    j = f_matrix[, 2],
    k = f_matrix[, 3],
    type = "mesh3d",
    vertexcolor = vertex_colors,
    name = title,
    hoverinfo = "text",
    text = ifelse(valid_idx, paste("Expr:", round(expr_values, 2), "<br>Pseudotime:", round(vertex_pseudotime, 1)), "Non-Midgut Section")
  )
  
  fig <- fig %>% layout(
    title = title,
    scene = list(
      aspectmode = "data",
      xaxis = list(title = "X"),
      yaxis = list(title = "Y"),
      zaxis = list(title = "Z")
    )
  )
  
  out_file <- file.path(output_dir, filename)
  saveWidget(fig, file = out_file, selfcontained = FALSE)
  cat(paste("Saved plot to:", out_file, "\n"))
  
  return(fig)
}
```

## Generate Plots
### Raw Expression
```{r plot_raw}
fig_raw <- generate_mesh_plot(vertex_raw_expr, paste(gene_name, "- Raw Expression (03Slide4)"), paste0(gene_name, "_raw_3D_03Slide4.html"))
fig_raw
```

### Smoothed Expression
```{r plot_smoothed}
fig_smoothed <- generate_mesh_plot(vertex_smooth_expr, paste(gene_name, "- Smoothed Expression (03Slide4)"), paste0(gene_name, "_smoothed_3D_03Slide4.html"))
fig_smoothed
```
