---
title: "Progenitor Bin Analysis (ISC and EB)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(Seurat)
library(dplyr)
library(ggplot2)
library(viridis)
library(Matrix)
library(Matrix.utils)
library(zoo)
library(patchwork)
library(shiny)
library(plotly)

# Set path to 03Slide4 data
data_path <- "../Random data/VisiumHD_Drosophila_Resolution_List.rds"
```

## 1. Load Data
We load the Visium HD segmented data from the saved resolution list.

```{r load_data}
# --- 1. Load Data (Raw Visium HD Strategy) ---
data_dir <- "/home/mattia/Thesis/03Slide4/binned_data"

# Clean up Windows :Zone.Identifier metadata files
zone_files <- list.files(path = data_dir, pattern = "Zone.Identifier$", recursive = TRUE, full.names = TRUE)
if (length(zone_files) > 0) {
  message("Cleaning up metadata files...")
  unlink(zone_files)
}

# Fix Visium HD directory structure for Seurat's Load10X_Spatial
if (!dir.exists(file.path(data_dir, "binned_outputs"))) {
  message("Restoring SpaceRanger Visium HD directory structure...")
  dir.create(file.path(data_dir, "binned_outputs"), showWarnings = FALSE)
  for (res in c("square_002um", "square_008um", "square_016um")) {
    old_p <- file.path(data_dir, res)
    new_p <- file.path(data_dir, "binned_outputs", res)
    if (dir.exists(old_p)) file.rename(old_p, new_p)
  }
}

# Load the data
# Slice name must match available H&E image if present
seurat_obj <- Load10X_Spatial(data_dir, slice = "HE_Stain_Slice")

# Set Default Assay to 16um Bins
DefaultAssay(seurat_obj) <- "Spatial.016um"

# Filter out bins with zero counts
seurat_obj <- subset(seurat_obj, subset = nCount_Spatial.016um > 0)
message("Initial Object Count (Loaded & Filtered): ", ncol(seurat_obj))

# Normalize with SCTransform (Standard for Visium HD)
message("Running SCTransform on 16um bins...")
seurat_obj <- SCTransform(seurat_obj, assay = "Spatial.016um", verbose = FALSE)
message("Object Count after SCTransform: ", ncol(seurat_obj))

# --- Debugging Information ---
message("Assays available: ", paste(Assays(seurat_obj), collapse = ", "))
message("Default Assay: ", DefaultAssay(seurat_obj))

# 0. Handle Seurat V5 multi-layer objects
if (inherits(seurat_obj[[DefaultAssay(seurat_obj)]], "Assay5")) {
  message("Seurat V5 assay detected. Joining layers...")
  seurat_obj <- JoinLayers(seurat_obj, assay = DefaultAssay(seurat_obj))
}

# 1. Check for duplicate cell names (barcodes)
if (any(duplicated(Cells(seurat_obj)))) {
  message("Duplicate cell barcodes found! This is likely causing the row.names error.")
  # We subset to unique barcodes to be absolutely sure
  seurat_obj <- seurat_obj[, !duplicated(Cells(seurat_obj))]
  message("Reduced to ", ncol(seurat_obj), " unique cell barcodes.")
}

# 2. Check for duplicate gene names
if (any(duplicated(rownames(seurat_obj)))) {
  message("Duplicate gene names detected. Unique-ifying names...")
  new_names <- make.unique(rownames(seurat_obj))
  if (inherits(seurat_obj[[DefaultAssay(seurat_obj)]], "Assay5")) {
    rownames(seurat_obj[[DefaultAssay(seurat_obj)]]) <- new_names
  } else {
    rownames(seurat_obj[[DefaultAssay(seurat_obj)]]@counts) <- new_names
    rownames(seurat_obj[[DefaultAssay(seurat_obj)]]@data) <- new_names
  }
}

# Ensure data is normalized
if (!"data" %in% Layers(seurat_obj, assay = DefaultAssay(seurat_obj))) {
  message("Normalizing data...")
  seurat_obj <- NormalizeData(seurat_obj, verbose = FALSE)
}

# Final Check: Are cell names unique? 
```

## 2. Define Refined Progenitor Markers
To address potential bias towards large, transcriptionally dominant Enterocytes (ECs), we use an expanded marker set including both **Positive (Progenitor)** and **Negative (EC/Copper Cell)** signatures.

```{r define_markers}
# 1. Positive Markers (ISC and EB)
# Standard: esg, Dl (ISC), Su(H) (EB), klu (EB), sanpodo (ISC), Nrt
# Custom: GFP-construct, Cas9-construct
pos_markers_raw <- c("esg", "Dl", "Su(H)", "klu", "sanpodo", "Nrt", "GFP-construct", "Cas9-construct")

# 2. Negative Markers (Enterocytes and Copper Cells)
# ECs: Myo61F, LManVI, alpha-Try, beta-Try
# Copper Cells (mEC): MtnA, Npc2e
neg_markers_raw <- c("Myo61F", "LManVI", "alpha-Try", "beta-Try", "MtnA", "Npc2e")

# Helper to find present markers (case-insensitive)
find_present <- function(m_list, obj) {
  found <- m_list[m_list %in% rownames(obj)]
  missing <- m_list[!m_list %in% found]
  for (m in missing) {
    match <- grep(paste0("^", m, "$"), rownames(obj), ignore.case = TRUE, value = TRUE)
    if (length(match) > 0) found <- c(found, match[1])
  }
  return(unique(found))
}

pos_markers <- find_present(pos_markers_raw, seurat_obj)
neg_markers <- find_present(neg_markers_raw, seurat_obj)

message("Positive Markers used: ", paste(pos_markers, collapse = ", "))
message("Negative Markers (EC filter) used: ", paste(neg_markers, collapse = ", "))
```

## 3. Calculate Refined Progenitor Score
We calculate a **Contrast Score** by subtracting a fraction of the EC signature from the Progenitor signature. This ensures that bins with high EC contamination are penalized, even if they have some progenitor signal.

```{r calculate_score}
# 1. Calculate Average Positive Expression
pos_mat <- GetAssayData(seurat_obj, assay = "SCT", layer = "data")[pos_markers, , drop = FALSE]
pos_score <- colMeans(as.matrix(pos_mat))

# 2. Calculate Average Negative Expression (EC bias)
neg_mat <- GetAssayData(seurat_obj, assay = "SCT", layer = "data")[neg_markers, , drop = FALSE]
neg_score <- colMeans(as.matrix(neg_mat))

# 3. Final Weighted Contrast Score
# We subtract 50% of the EC signature to clean up the identification
seurat_obj$Progenitor_Score_Raw <- pos_score
seurat_obj$EC_Penalty <- neg_score
seurat_obj$Progenitor_Score <- pos_score - (0.5 * neg_score)

# Visualize the scores
# Ensure score doesn't go below 0 for standard range
seurat_obj$Progenitor_Score[seurat_obj$Progenitor_Score < 0] <- 0

hist(seurat_obj$Progenitor_Score, breaks = 50, main = "Weighted Progenitor Score Distribution", xlab = "Score (Positive - 0.5*Negative)")
message("Object Count after Progenitor Scoring: ", ncol(seurat_obj))
```

## 4. Identify Top 10% Bins
We rank the bins by the Progenitor Score and select the top 10%. We use rank with ties handled correctly to ensure we have two groups even in zero-inflated data.

```{r identify_bins}
# Using rank to avoid issues with many identical scores (like 0)
# We handle ties by assigning them the same rank, but we want a clear split
scores <- seurat_obj$Progenitor_Score
n_cells <- length(scores)
top_n <- ceiling(n_cells * 0.10)

# Identify the score threshold for the top 10%
# If many cells have the same score at the boundary, we might get more than 10%
# But we must ensure 'Others' exists.
threshold <- quantile(scores, 0.90)

if (threshold == 0 && sum(scores > 0) > 0) {
  # If 90% or more cells are 0, take all cells with score > 0 as "High"
  # or a subset of them if they are too many. 
  # Here we'll take everyone > 0 if threshold is 0.
  message("90th percentile is 0. Selecting all bins with score > 0 as Progenitor-High.")
  seurat_obj$Progenitor_Status <- ifelse(scores > 0, "Progenitor-High", "Others")
} else if (all(scores == 0)) {
  stop("All Progenitor Scores are 0. No bins identified.")
} else {
  # Standard case
  seurat_obj$Progenitor_Status <- ifelse(scores >= threshold, "Progenitor-High (Top 10%)", "Others (90%)")
}

# Ensure we have at least two levels
seurat_obj$Progenitor_Status <- factor(seurat_obj$Progenitor_Status)
message("Cell count per identity:")
print(table(seurat_obj$Progenitor_Status))

# Check if we have two groups
if (length(unique(seurat_obj$Progenitor_Status)) < 2) {
  stop("Could not identify two groups (Progenitor-High and Others). Try a different threshold or markers.")
}
message("Object Count after Status Assignment: ", ncol(seurat_obj))
```

## 5. Differential Gene Expression (DGE) Analysis
We compare the identified progenitor-high bins against the rest of the cells.

```{r dge_analysis}
# Set identity to Progenitor_Status
Idents(seurat_obj) <- "Progenitor_Status"

# Get the names of the two identities
idents <- levels(seurat_obj$Progenitor_Status)
high_ident <- idents[grep("High", idents)]
low_ident <- idents[grep("Other", idents)]

# Perform DGE
progenitor_markers <- FindMarkers(seurat_obj, 
                                 ident.1 = high_ident, 
                                 ident.2 = low_ident, 
                                 only.pos = TRUE, 
                                 min.pct = 0.1, 
                                 logfc.threshold = 0.25)

# Display top 20 DEGs
head(progenitor_markers, 50)

# Save result to Excel folder
output_csv <- "../Excels/Progenitor_Top10_DEGs.csv"
write.csv(progenitor_markers, output_csv)
message("DGE results saved to: ", output_csv)
message("Object Count after DGE: ", ncol(seurat_obj))
```

## 6. Visualization
We visualize the spatial distribution of the score and the identified bins.

```{r visualization, fig.width=12, fig.height=10}
# --- Standard Spatial Visualization ---
message("Generating spatial plots...")

# Diagnostic Plots: Progenitor Score, EC Penalty, and Final Filtered Score
p1 <- SpatialFeaturePlot(seurat_obj, features = "Progenitor_Score_Raw") + ggtitle("Raw Progenitor Marker Avg")
p2 <- SpatialFeaturePlot(seurat_obj, features = "EC_Penalty") + ggtitle("EC Signature (Penalty)")
p3 <- SpatialFeaturePlot(seurat_obj, features = "Progenitor_Score") + ggtitle("Final Contrast Score")
p4 <- SpatialDimPlot(seurat_obj, group.by = "Progenitor_Status") + ggtitle("Identified Progenitor Bins")

print((p1 + p2) / (p3 + p4))
message("Object Count after Visualization: ", ncol(seurat_obj))


# Individual marker plots using Seurat native function
if (length(pos_markers) > 0) {
  p_markers <- SpatialFeaturePlot(seurat_obj, features = pos_markers, ncol = 3, combine = FALSE)
  if (length(p_markers) > 0) print(wrap_plots(p_markers, ncol = 3))
}
```


## 7. Sample Subsetting (Interactive Lasso Selection)
Before performing the spatial similarity analysis, we split the sample into WT and Mutant areas. This uses an interactive Shiny lasso tool.

```{r interactive_subsetting}
# --- File Paths for Selections ---
selection_wt_file <- "../Random data/Slide4_Progenitor_WT_Selection.csv"
selection_mut_file <- "../Random data/Slide4_Progenitor_Mutant_Selection.csv"

# Barcode Matching Helper (Precise for Visium HD Bins)
clean_barcodes_local <- function(selection, all_bins) {
  if (is.null(selection)) return(character(0))
  # Use direct character intersection to avoid numeric conversion issues
  selection <- as.character(selection)
  valid <- intersect(selection, all_bins)
  
  if (length(valid) == 0 && length(selection) > 0) {
      # Fallback: try prefix-based numeric matching if direct fails
      message("Direct barcode match failed. Attempting numeric extraction...")
      sel_nums <- gsub("[^0-9]", "", selection)
      all_nums <- gsub("[^0-9]", "", all_bins)
      match_idx <- match(sel_nums, all_nums)
      valid <- all_bins[match_idx[!is.na(match_idx)]]
  }
  return(valid)
}

if (file.exists(selection_wt_file) && file.exists(selection_mut_file)) {
    message("Loading existing WT/Mutant bin selections from CSV...")
    wt_bins <- as.character(read.csv(selection_wt_file)$x)
    mut_bins <- as.character(read.csv(selection_mut_file)$x)
} else {
    message("Launching Shiny app for WT and Mutant bin selection...")
    coords <- GetTissueCoordinates(seurat_obj)
    # Ensure coords and barcodes match
    common_barcodes <- intersect(rownames(coords), Cells(seurat_obj))
    coords_df <- data.frame(x = coords[common_barcodes, 1], 
                            y = coords[common_barcodes, 2], 
                            barcode = common_barcodes)
    
    # Debug info
    message("Plotting ", nrow(coords_df), " bins.")
    message("Barcode Format: ", head(common_barcodes, 1))

    ui <- fluidPage(
        titlePanel("Progenitor Bin Splitting (03Slide4)"),
        sidebarLayout(
            sidebarPanel(
                h4("Instructions"),
                p("1. Use the Lasso tool on the plot to select the WT area."),
                p("2. Click 'Save WT Bins'."),
                p("3. Select the Mutant area and click 'Save Mutant Bins'."),
                hr(),
                actionButton("save_wt", "Save WT Bins"),
                verbatimTextOutput("st_wt"),
                hr(),
                actionButton("save_mut", "Save Mutant Bins"),
                verbatimTextOutput("st_mut"),
                hr(), 
                actionButton("done", "DONE (Continue)", class = "btn-success")
            ),
            mainPanel(plotlyOutput("plot", height = "700px"))
        )
    )
    
    server <- function(input, output, session) {
        v <- reactiveValues(wt = NULL, mut = NULL)
        
        output$plot <- renderPlotly({
            plot_ly(coords_df, x=~x, y=~y, key=~barcode, type="scatter", mode="markers",
                    marker=list(size=3, color='gray', opacity=0.5), source="src") %>%
                layout(dragmode="lasso", yaxis=list(autorange="reversed"), title = "Select WT and Mutant Regions")
        })
        
        observeEvent(input$save_wt, {
            d <- event_data("plotly_selected", source="src")
            if(!is.null(d)) {
                v$wt <- unique(as.character(d$key))
                message("Saved ", length(v$wt), " IDs to WT.")
            }
        })
        
        observeEvent(input$save_mut, {
            d <- event_data("plotly_selected", source="src")
            if(!is.null(d)) {
                v$mut <- unique(as.character(d$key))
                message("Saved ", length(v$mut), " IDs to Mutant.")
            }
        })
        
        output$st_wt <- renderText({ 
            if(is.null(v$wt)) "WT: No bins saved" 
            else paste("WT:", length(v$wt), "bins saved") 
        })
        
        output$st_mut <- renderText({ 
            if(is.null(v$mut)) "Mutant: No bins saved" 
            else paste("Mutant:", length(v$mut), "bins saved") 
        })
        
        observeEvent(input$done, { stopApp(list(wt=v$wt, mut=v$mut)) })
    }
    
    sel <- runApp(shinyApp(ui, server))
    all_bins <- Cells(seurat_obj)
    wt_bins <- clean_barcodes_local(sel$wt, all_bins)
    mut_bins <- clean_barcodes_local(sel$mut, all_bins)
    
    # Save the cleaned, unique barcodes
    write.csv(data.frame(x = unique(wt_bins)), selection_wt_file, row.names = FALSE)
    write.csv(data.frame(x = unique(mut_bins)), selection_mut_file, row.names = FALSE)
    message("Selections saved. WT:", length(wt_bins), " Mutant:", length(mut_bins))
}

# Apply to Seurat object
seurat_obj$Sample_Group <- "Background"
if(length(wt_bins) > 0) seurat_obj$Sample_Group[Cells(seurat_obj) %in% wt_bins] <- "WT"
if(length(mut_bins) > 0) seurat_obj$Sample_Group[Cells(seurat_obj) %in% mut_bins] <- "Mutant"

message("Final group assignment counts:")
print(table(seurat_obj$Sample_Group))

# Visualize the splitting results
if ("Sample_Group" %in% colnames(seurat_obj@meta.data)) {
  print(SpatialDimPlot(seurat_obj, group.by = "Sample_Group") + ggtitle("WT/Mutant Selection"))
}

message("Total Object Count (Bins): ", ncol(seurat_obj))
```

## 8. Progenitor Spatial Similarity (Splitted WT vs Mutant)
This section implements a refined "triangle plot" for both WT and Mutant progenitor populations.

```{r similarity_plot_refined_split, fig.width=15, fig.height=8}
# 1. Refined Calculation Function
calculate_triangle_matrix_refined <- function(obj, n_bins = 100, smoothing_window = 3) {
    if (ncol(obj) < 10) return(NULL)
    
    var_genes <- VariableFeatures(obj)
    if (length(var_genes) < 100) {
      obj <- FindVariableFeatures(obj, nfeatures = 2000, verbose = FALSE)
      var_genes <- VariableFeatures(obj)
    }
    
    obj$Bin <- factor(cut(obj$Spatial_Pseudotime, breaks = n_bins, labels = FALSE), levels = 1:n_bins)
    
    # Use SCT assay for high-quality spatial correlation
    exp_mat <- GetAssayData(obj, assay = "SCT", layer = "data")[var_genes, ]
    bin_means <- t(apply(exp_mat, 1, function(x) tapply(x, obj$Bin, mean, na.rm = TRUE)))
    
    # A. Interpolate missing bins
    bin_means <- t(apply(bin_means, 1, function(x) {
        if (all(is.na(x))) return(x)
        na.approx(x, na.rm = FALSE, rule = 2)
    }))
    
    # C. Noise Filtering (Moving Average Smoothing)
    bin_smoothed <- t(apply(bin_means, 1, function(x) {
        if(all(x == 0)) return(x)
        res <- as.numeric(stats::filter(x, rep(1/smoothing_window, smoothing_window), sides = 2))
        res[is.na(res)] <- mean(x, na.rm=T)
        return(res)
    }))
    
    cor_mat <- cor(bin_smoothed, method = "pearson", use = "pairwise.complete.obs")
    diag(cor_mat) <- NA
    global_mean <- mean(cor_mat, na.rm = TRUE)
    cor_mat_adj <- cor_mat - global_mean
    diag(cor_mat_adj) <- 1
    
    return(cor_mat_adj)
}

# 2. Plotting Function
plot_triangle_local <- function(sim_mat, title) {
    if (is.null(sim_mat)) return(NULL)
    n <- nrow(sim_mat)
    tri_data <- expand.grid(i = 1:n, j = 1:n) %>% filter(j >= i) %>% rowwise() %>%
        mutate(Similarity = sim_mat[i, j]) %>% filter(!is.na(Similarity))
    
    poly_list <- lapply(1:nrow(tri_data), function(idx) {
        row <- tri_data[idx, ]
        cx <- (row$i + row$j) / 2; cy <- (row$j - row$i) / 2
        data.frame(x = c(cx, cx + 0.5, cx, cx - 0.5), y = c(cy + 0.5, cy, cy - 0.5, cy), Similarity = row$Similarity, ID = idx)
    })
    poly_df <- do.call(rbind, poly_list)
    
    ggplot(poly_df, aes(x = x, y = y, fill = Similarity, group = ID)) +
        geom_polygon() +
        scale_fill_gradient2(high = "#B2182B", mid = "#F7F7F7", low = "#2166AC", midpoint = 0, name = "Similarity", limits = c(-0.4, 0.4), oob = scales::squish) +
        theme_minimal() + coord_fixed(ratio = 1) + labs(title = title, x = "AP Axis", y = "Dist") +
        theme(panel.grid = element_blank(), axis.text.y = element_blank())
}

# 3. Analyze each group
# Ensure Spatial_Pseudotime exists
if (!"Spatial_Pseudotime" %in% colnames(seurat_obj@meta.data)) {
  coords <- GetTissueCoordinates(seurat_obj)
  seurat_obj$Spatial_Pseudotime <- 100 * (coords[, 1] - min(coords[, 1])) / (max(coords[, 1]) - min(coords[, 1]))
}

# Diagnostic: Check Progenitor-High counts in each group
is_high_global <- grepl("High", seurat_obj$Progenitor_Status)
group_counts <- table(seurat_obj$Sample_Group, is_high_global)
message("Progenitor Distribution across selected areas (Global High = TRUE):")
print(group_counts)

# Helper for robust subsetting
get_subset_prog <- function(obj, group_name) {
    # 1. Try Global Top 10%
    cells_global <- Cells(obj)[is_high_global & obj$Sample_Group == group_name]
    
    if (length(cells_global) >= 10) {
        message(paste("Using", length(cells_global), "global progenitor-high bins for", group_name))
        return(subset(obj, cells = cells_global))
    }
    
    # 2. Fallback: Top 10% WITHIN the group if global is empty/too small
    message(paste("Warning: Global progenitor-high bins not found in", group_name, ". Falling back to local top 10%..."))
    grp_cells <- Cells(obj)[obj$Sample_Group == group_name]
    if (length(grp_cells) < 10) return(NULL)
    
    sub_grp <- subset(obj, cells = grp_cells)
    threshold_local <- quantile(sub_grp$Progenitor_Score, 0.90)
    cells_local <- Cells(sub_grp)[sub_grp$Progenitor_Score >= threshold_local]
    
    if (length(cells_local) >= 10) {
        message(paste("Using", length(cells_local), "local progenitor-high bins for", group_name))
        return(subset(sub_grp, cells = cells_local))
    }
    return(NULL)
}

# Analyze WT
wt_prog <- get_subset_prog(seurat_obj, "WT")
p_wt <- if(!is.null(wt_prog)) plot_triangle_local(calculate_triangle_matrix_refined(wt_prog), "WT Progenitors (Slide 4)") else NULL

# Analyze Mutant
mut_prog <- get_subset_prog(seurat_obj, "Mutant")
p_mut <- if(!is.null(mut_prog)) plot_triangle_local(calculate_triangle_matrix_refined(mut_prog), "Mutant Progenitors (Slide 4)") else NULL

# Display side-by-side
if (!is.null(p_wt) && !is.null(p_mut)) {
    p_final <- (p_wt | p_mut) + plot_layout(guides = "collect")
    print(p_final)
    ggsave("../figures/Progenitor_Triangle_Comparison_WT_Mutant.png", p_final, width = 15, height = 7)
}
# ==============================================================================
# --- PART II: SEGMENTED CELL ANALYSIS (COMPARISON) ---
# ==============================================================================

# Path to RDS list (should be consistent with previous runs)
rds_path <- "../Random data/VisiumHD_Drosophila_Resolution_List.rds"
if (!file.exists(rds_path)) stop("RDS Resolution List not found at: ", rds_path)

vhd_list <- readRDS(rds_path)
seurat_seg <- vhd_list$seg
message("Loaded Segmented Assay: ", ncol(seurat_seg), " cells.")

# Standard Log-Normalization for segmented data
seurat_seg <- NormalizeData(seurat_seg, verbose = FALSE)

# Deduplicate row.names if any (safeguard)
if (any(duplicated(Cells(seurat_seg)))) {
  seurat_seg <- seurat_seg[, !duplicated(Cells(seurat_seg))]
  message("Deduplicated barcodes: ", ncol(seurat_seg), " unique cells.")
}

# Run Dimensional Reduction for UMAP plots (user-requested)
message("Running PCA and UMAP for segmented data...")
seurat_seg <- FindVariableFeatures(seurat_seg, selection.method = "vst", nfeatures = 2000, verbose = FALSE)
seurat_seg <- ScaleData(seurat_seg, verbose = FALSE)
seurat_seg <- RunPCA(seurat_seg, features = VariableFeatures(seurat_seg), verbose = FALSE)
seurat_seg <- RunUMAP(seurat_seg, dims = 1:20, verbose = FALSE)
```

## 10. Basic Progenitor Scoring (Segmented)
For the segmented assay, we use the original 5 markers and a standard sum-based score with a top 5% threshold.

```{r score_seg}
# Original marker set
seg_markers_base <- c("esg", "Dl", "Su(H)", "GFP-construct", "Cas9-construct")
seg_markers <- seg_markers_base[seg_markers_base %in% rownames(seurat_seg)]

# Basic Sum Scoring
seg_scores <- colSums(as.matrix(GetAssayData(seurat_seg, layer = "data")[seg_markers, ]))
seurat_seg$Progenitor_Score <- seg_scores

# Identify Top 5%
thresh_seg <- quantile(seg_scores, 0.95)
seurat_seg$Progenitor_Status <- ifelse(seg_scores >= thresh_seg, "Progenitor-High (Top 5%)", "Others (95%)")
seurat_seg$Progenitor_Status <- factor(seurat_seg$Progenitor_Status)

print(table(seurat_seg$Progenitor_Status))

# Visualization (Using UMAP as requested)
p_seg1 <- FeaturePlot(seurat_seg, features = "Progenitor_Score") + ggtitle("Segmented: Progenitor Score (UMAP)")
p_seg2 <- DimPlot(seurat_seg, group.by = "Progenitor_Status") + ggtitle("Segmented: Top 5% Bins (UMAP)")
print(p_seg1 / p_seg2)
```

## 11. Interactive WT/Mutant Selection (Segmented)
We provide a dedicated lasso selection for the segmented data to ensure spatial accuracy for these specific coordinates.

```{r lasso_seg, eval=FALSE}
library(shiny)
library(plotly)

# Prepare coordinates from centroids (3D UMAP Logic)
if (all(c("x_centroid", "y_centroid") %in% colnames(seurat_seg@meta.data))) {
    coords_seg_raw <- seurat_seg@meta.data[, c("x_centroid", "y_centroid")]
    colnames(coords_seg_raw) <- c("x", "y")
} else {
    coords_seg_raw <- GetTissueCoordinates(seurat_seg)
    colnames(coords_seg_raw) <- c("x", "y")
}

coords_df_seg <- data.frame(
    x = coords_seg_raw$x,
    y = coords_seg_raw$y,
    barcode = rownames(coords_seg_raw)
)

ui <- fluidPage(
    titlePanel("Slide 4: Segmented Progenitor Selection"),
    sidebarLayout(
        sidebarPanel(
            actionButton("save_wt", "Save WT Selection"),
            actionButton("save_mut", "Save Mutant Selection"),
            helpText("Lasso area on plot and click Save."),
            textOutput("status")
        ),
        mainPanel(plotlyOutput("plot", height = "800px"))
    )
)

server <- function(input, output, session) {
    output$plot <- renderPlotly({
        plot_ly(coords_df_seg, x=~x, y=~y, key=~barcode, type="scatter", mode="markers",
                marker = list(size = 3, opacity = 0.5, color = 'gray'), source = "subset_source") %>%
        layout(dragmode = "lasso", xaxis = list(title = "X"), yaxis = list(title = "Y", autorange = "reversed"))
    })
    
    # Helper to clean/format barcodes (from 3D UMAP Logic)
    clean_barcodes_local <- function(selection, all_cells) {
        if (is.null(selection)) return(NULL)
        normalized <- tryCatch({ sprintf("cellid_%09d-1", as.numeric(selection))
        }, warning = function(w) selection, error = function(e) selection)
        valid_fmt <- intersect(normalized, all_cells)
        valid_raw <- intersect(selection, all_cells)
        if (length(valid_fmt) > length(valid_raw)) return(valid_fmt) else return(valid_raw)
    }

    observeEvent(input$save_wt, {
        d <- event_data("plotly_selected", source = "subset_source")
        if(!is.null(d)) {
            selected <- clean_barcodes_local(d$key, Cells(seurat_seg))
            write.csv(data.frame(barcode=selected), "../Random data/Slide4_Segmented_WT_Selection.csv", row.names=F)
            output$status <- renderText(paste("Saved", length(selected), "WT cells."))
        }
    })
    observeEvent(input$save_mut, {
        d <- event_data("plotly_selected", source = "subset_source")
        if(!is.null(d)) {
            selected <- clean_barcodes_local(d$key, Cells(seurat_seg))
            write.csv(data.frame(barcode=selected), "../Random data/Slide4_Segmented_Mutant_Selection.csv", row.names=F)
            output$status <- renderText(paste("Saved", length(selected), "Mutant cells."))
        }
    })
}
shinyApp(ui, server)
```

## 12. Segmented Similarity Triangle Plots (Progenitors vs Others)
```{r similarity_seg, fig.width=15, fig.height=12}
# Load Segmented Selections
wt_seg_ids <- if(file.exists("../Random data/Slide4_Segmented_WT_Selection.csv")) read.csv("../Random data/Slide4_Segmented_WT_Selection.csv")$barcode else NULL
mut_seg_ids <- if(file.exists("../Random data/Slide4_Segmented_Mutant_Selection.csv")) read.csv("../Random data/Slide4_Segmented_Mutant_Selection.csv")$barcode else NULL

if(!is.null(wt_seg_ids) && !is.null(mut_seg_ids)){
    # Setup Groups
    seurat_seg$Sample_Group <- "Background"
    seurat_seg$Sample_Group[Cells(seurat_seg) %in% wt_seg_ids] <- "WT"
    seurat_seg$Sample_Group[Cells(seurat_seg) %in% mut_seg_ids] <- "Mutant"
    
    # Calculate Spatial Pseudotime for Segmented
    coords_seg <- GetTissueCoordinates(seurat_seg)
    x_col_seg <- grep("x|centroid", colnames(coords_seg), ignore.case = TRUE, value = TRUE)[1]
    seurat_seg$Spatial_Pseudotime <- 100 * (coords_seg[[x_col_seg]] - min(coords_seg[[x_col_seg]])) / (max(coords_seg[[x_col_seg]]) - min(coords_seg[[x_col_seg]]))
    
    # Function to plot triangle helper (assumes ggplot2 in env)
    plot_tri_helper <- function(sim_mat, title) {
        if(is.null(sim_mat)) return(NULL)
        n <- nrow(sim_mat)
        tri_data <- expand.grid(i = 1:n, j = 1:n) %>% filter(j >= i) %>% rowwise() %>% mutate(Similarity = sim_mat[i, j])
        poly_list <- lapply(1:nrow(tri_data), function(idx) {
            row <- tri_data[idx, ]; cx <- (row$i + row$j) / 2; cy <- (row$j - row$i) / 2
            data.frame(x = c(cx, cx + 0.5, cx, cx - 0.5), y = c(cy + 0.5, cy, cy - 0.5, cy), Similarity = row$Similarity, ID = idx)
        })
        ggplot(do.call(rbind, poly_list), aes(x = x, y = y, fill = Similarity, group = ID)) +
            geom_polygon() + scale_fill_gradient2(high = "#B2182B", mid = "#F7F7F7", low = "#2166AC", midpoint = 0, name = "Rel. Sim", limits = c(-0.4, 0.4), oob = scales::squish) +
            theme_minimal() + coord_fixed(ratio = 1) + labs(title = title) + theme(panel.grid = element_blank(), axis.text.y = element_blank())
    }

    # Redefining refined triangle matrix function for segmented assay (standalone logic)
    calculate_triangle_matrix_refined <- function(obj, n_bins = 100, smoothing_window = 3) {
        if (is.null(obj) || ncol(obj) < 10) return(NULL)
        var_genes <- VariableFeatures(obj)
        if (length(var_genes) < 100) { obj <- FindVariableFeatures(obj, nfeatures = 2000, verbose = FALSE); var_genes <- VariableFeatures(obj) }
        
        obj$Bin <- factor(cut(obj$Spatial_Pseudotime, breaks = n_bins, labels = FALSE), levels = 1:n_bins)
        exp_mat <- GetAssayData(obj, layer = "data")[var_genes, ]
        bin_means <- t(apply(exp_mat, 1, function(x) tapply(x, obj$Bin, mean, na.rm = TRUE)))
        bin_means <- t(apply(bin_means, 1, function(x) zoo::na.approx(x, na.rm = FALSE, rule = 2)))
        bin_smoothed <- t(apply(bin_means, 1, function(x) {
            if(all(x == 0)) return(x)
            res <- as.numeric(stats::filter(x, rep(1/smoothing_window, smoothing_window), sides = 2)); res[is.na(res)] <- mean(x, na.rm=T); return(res)
        }))
        cor_mat <- cor(bin_smoothed, method = "pearson", use = "pairwise.complete.obs")
        diag(cor_mat) <- NA; global_mean <- mean(cor_mat, na.rm = TRUE)
        cor_mat_adj <- cor_mat - global_mean; diag(cor_mat_adj) <- 1; return(cor_mat_adj)
    }

    # Subset Logic
    is_prog <- seurat_seg$Progenitor_Status == "Progenitor-High (Top 5%)"
    is_others <- seurat_seg$Progenitor_Status == "Others (95%)"
    
    # Progenitor Subsets
    prog_wt <- subset(seurat_seg, cells = Cells(seurat_seg)[is_prog & seurat_seg$Sample_Group == "WT"])
    prog_mut <- subset(seurat_seg, cells = Cells(seurat_seg)[is_prog & seurat_seg$Sample_Group == "Mutant"])
    # Others Subsets
    others_wt <- subset(seurat_seg, cells = Cells(seurat_seg)[is_others & seurat_seg$Sample_Group == "WT"])
    others_mut <- subset(seurat_seg, cells = Cells(seurat_seg)[is_others & seurat_seg$Sample_Group == "Mutant"])

    # Generate Plots
    p1 <- plot_tri_helper(calculate_triangle_matrix_refined(prog_wt), "Progenitor WT")
    p2 <- plot_tri_helper(calculate_triangle_matrix_refined(prog_mut), "Progenitor Mutant")
    p3 <- plot_tri_helper(calculate_triangle_matrix_refined(others_wt), "Others WT (Non-Prog)")
    p4 <- plot_tri_helper(calculate_triangle_matrix_refined(others_mut), "Others Mutant (Non-Prog)")
    
    # 4-Panel Figure
    p_final <- (p1 + p2) / (p3 + p4)
    print(p_final)
    ggsave("../figures/Progenitor_vs_Others_Similarity_Segmented.png", p_final, width = 15, height = 12)
    message("Segmented Triangle Comparison saved to figures/.")
}
```

## 13. Segmented Identity DGE (Condition-Specific)
```{r dge_seg}
# Prepare Subsets by Selection Group
wt_cells_obj <- subset(seurat_seg, subset = Sample_Group == "WT")
mut_cells_obj <- subset(seurat_seg, subset = Sample_Group == "Mutant")

# A. WT Progenitor Identity (Progenitor vs Others in WT)
Idents(wt_cells_obj) <- wt_cells_obj$Progenitor_Status
message("Running DGE: WT Progenitor vs Others...")
wt_id_markers <- FindMarkers(wt_cells_obj, ident.1 = "Progenitor-High (Top 5%)", ident.2 = "Others (95%)", min.pct = 0.2, logfc.threshold = 0.25)
wt_id_markers$gene <- rownames(wt_id_markers)

# B. Mutant Progenitor Identity (Progenitor vs Others in Mutant)
Idents(mut_cells_obj) <- mut_cells_obj$Progenitor_Status
message("Running DGE: Mutant Progenitor vs Others...")
mut_id_markers <- FindMarkers(mut_cells_obj, ident.1 = "Progenitor-High (Top 5%)", ident.2 = "Others (95%)", min.pct = 0.2, logfc.threshold = 0.25)
mut_id_markers$gene <- rownames(mut_id_markers)

# Display Results
print("Top 20 WT Progenitor Identity Markers:")
print(head(wt_id_markers, 20))

print("Top 20 Mutant Progenitor Identity Markers:")
print(head(mut_id_markers, 20))

# Save
write.csv(wt_id_markers, "../Excels/Slide4_Segmented_WT_Progenitor_Markers.csv")
write.csv(mut_id_markers, "../Excels/Slide4_Segmented_Mutant_Progenitor_Markers.csv")
```

## 14. Progenitor Identity Shift Comparison
```{r identity_comparison}
# Filter for significant enriched markers (p_adj < 0.05, positive logFC)
wt_sig_markers <- wt_id_markers$gene[wt_id_markers$p_val_adj < 0.05 & wt_id_markers$avg_log2FC > 0]
mut_sig_markers <- mut_id_markers$gene[mut_id_markers$p_val_adj < 0.05 & mut_id_markers$avg_log2FC > 0]

# 1. Conserved Progenitor Markers (Define identity in both states)
conserved_markers <- intersect(wt_sig_markers, mut_sig_markers)

# 2. Lost Progenitor Markers (Reduced enrichment in mutant)
lost_markers <- setdiff(wt_sig_markers, mut_sig_markers)

# 3. Gained (Ectopic) Progenitor Markers (Newly enriched in mutant)
gained_markers <- setdiff(mut_sig_markers, wt_sig_markers)

message("\n--- Conserved Progenitor Markers ---")
print(conserved_markers)

message("\n--- Lost Progenitor Identity Genes (WT specific) ---")
print(lost_markers)

message("\n--- Gained (Ectopic) Progenitor Markers (Mutant specific) ---")
print(gained_markers)

# Export Comparison
write.csv(data.frame(gene = conserved_markers), "../Excels/Slide4_Conserved_Progenitor_Markers.csv")
write.csv(data.frame(gene = lost_markers), "../Excels/Slide4_Lost_Progenitor_Markers.csv")
write.csv(data.frame(gene = gained_markers), "../Excels/Slide4_Gained_Progenitor_Markers.csv")
```
