---
title: "Gut 3 Spatial Analysis"
output: html_document
date: "2026-02-13"
---

```{r setup, include=FALSE}
library(Seurat)
library(dplyr)
library(ggplot2)
library(cowplot)
library(viridis)
library(clusterProfiler)
library(org.Dm.eg.db)
library(pheatmap)
library(RColorBrewer)

# Set base theme
theme_set(theme_bw())
```

# 1. Load Data
We load the processed Gut 3 object from the 3D UMAP analysis. This object already contains:
- `Spatial_Pseudotime`: Position along the gut axis (0-100)
- `Predicted_CellType`: Cell type labels from label transfer

```{r load_data}
input_file <- "../Random data/gut3_processed_for_analysis.rds"

if (!file.exists(input_file)) {
  stop("Input file not found! Please run '3D_UMAP_Analysis.Rmd' first to generate 'gut3_processed_for_analysis.rds'.")
}

seurat_obj <- readRDS(input_file)

# Basic QC Check
print("Loaded Gut 3 Object:")
print(seurat_obj)
print(head(seurat_obj@meta.data[, c("Predicted_CellType", "Spatial_Pseudotime")]))
```

# 2. Identify Cell-Type Specific Markers
We identify markers for each cell type present in Gut 3.

```{r find_markers}
Idents(seurat_obj) <- seurat_obj$Predicted_CellType

# Find markers for every cluster compared to all remaining cells
# Using standard parameters for robustness
markers_all <- FindAllMarkers(
  seurat_obj,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25,
  verbose = FALSE
)

# Filter for significance
markers_sig <- markers_all %>%
  filter(p_val_adj < 0.05) %>%
  group_by(cluster) %>%
  top_n(n = 50, wt = avg_log2FC) # Keep top 50 for plotting candidates

# Save for reference
write.csv(markers_sig, "../Excels/gut3_celltype_markers.csv")
print(table(markers_sig$cluster))
```

# 3. Trajectory line Plots (Expression vs Length)
We plot specific marker genes for each cell type as smooth lines along the gut length.

```{r trajectory_plots, fig.width=10, fig.height=6}
# Function to plot expression trends
plot_trajectory_trend <- function(obj, cell_type, genes, y_limit = NULL) {
  
  # Subset to specific cell type
  sub_obj <- subset(obj, idents = cell_type)
  
  # Check if genes exist
  valid_genes <- genes[genes %in% rownames(sub_obj)]
  if(length(valid_genes) == 0) return(NULL)
  
  # Fetch data
  expr_data <- FetchData(sub_obj, vars = c(valid_genes, "Spatial_Pseudotime"))
  
  # Pivot to long format for ggplot
  expr_long <- tidyr::pivot_longer(
    expr_data, 
    cols = all_of(valid_genes), 
    names_to = "Gene", 
    values_to = "Expression"
  )
  
  # Max-Scaling normalization per gene (0 to 1) 
  # Keeps 0 at 0 (no expression) and makes max expression 1
  expr_long <- expr_long %>%
    group_by(Gene) %>%
    mutate(Scaled_Expr = Expression / max(Expression, na.rm = TRUE)) %>%
    ungroup()

  # Plot
  p <- ggplot(expr_long, aes(x = Spatial_Pseudotime, y = Scaled_Expr, color = Gene)) +
    # Use GAM smoothing for clean trajectory lines
    geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), se = FALSE, linewidth = 1.2) +
    
    # Enforce X-axis limits (0-100)
    coord_cartesian(xlim = c(0, 100), ylim = c(0, 1.05), expand = FALSE) +
    
    labs(
      title = paste("Trajectory: ", cell_type),
      x = "Gut Length (Pseudotime 0-100)",
      y = "Relative Expression (Scaled 0-1)"
    ) +
    theme_minimal() +
    theme(legend.position = "right")
  
  return(p)
}

# Generate plots for each cell type
unique_types <- unique(seurat_obj$Predicted_CellType)
plot_list <- list()

for (ctype in unique_types) {
  # Pick top 5 markers for this cell type to plot
  top_genes <- markers_sig %>% 
    filter(cluster == ctype) %>% 
    top_n(5, wt = avg_log2FC) %>% 
    pull(gene)
    
  if (length(top_genes) > 0) {
    p <- plot_trajectory_trend(seurat_obj, ctype, top_genes)
    print(p)
    ggsave(paste0("../figures/gut3_trajectory_", gsub("/", "_", ctype), ".png"), plot = p, width = 8, height = 5)
  }
}
```

# 3.1. Collapsed Trajectories (Progenitors & Enterocytes)
We group cells into broader categories to see general trends.

```{r collapsed_trajectories, fig.width=10, fig.height=6}
# 1. Create Collapsed Metadata
current_types <- as.character(seurat_obj$Predicted_CellType)
collapsed_types <- current_types

# Define groups
progenitors <- c("ISC", "EB")
enterocytes <- c("aEC", "mEC", "pEC", "dEC", "Enterocyte", "LFC", "Copper", "Iron")

# Replace
collapsed_types[collapsed_types %in% progenitors] <- "Progenitor"
collapsed_types[collapsed_types %in% enterocytes] <- "Enterocyte"
# Keep others as is (e.g. EE, etc) or group into "Other"
# collapsed_types[!collapsed_types %in% c("Progenitor", "Enterocyte")] <- "Other"

seurat_obj$Collapsed_Type <- collapsed_types

# 2. Find Markers for Collapsed Groups
Idents(seurat_obj) <- seurat_obj$Collapsed_Type
markers_collapsed <- FindAllMarkers(
  seurat_obj,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25,
  verbose = FALSE
)

# 3. Plot for Progenitor and Enterocyte
target_groups <- c("Progenitor", "Enterocyte")

for (group in target_groups) {
  if (group %in% unique(seurat_obj$Collapsed_Type)) {
    # Get top 5 markers for this BROAD group
    top_genes <- markers_collapsed %>% 
      filter(cluster == group) %>% 
      top_n(5, wt = avg_log2FC) %>% 
      pull(gene)
    
    if (length(top_genes) > 0) {
      p <- plot_trajectory_trend(seurat_obj, group, top_genes)
      print(p)
      ggsave(paste0("../figures/gut3_trajectory_COLLAPSED_", group, ".png"), plot = p, width = 8, height = 5)
    }
  }
}
```


# 3.2. Functional Trajectories (Signaling, Identity, Metabolism)
We plot trajectories for genes associated with specific biological functions.

```{r functional_trajectories, fig.width=10, fig.height=6}
# Helper: Fetch genes from GO Term
get_genes_from_go <- function(go_id, org_db = org.Dm.eg.db) {
  tryCatch({
    genes <- AnnotationDbi::select(org_db, keys = go_id, columns = "SYMBOL", keytype = "GOALL")
    return(unique(genes$SYMBOL))
  }, error = function(e) {
    message(paste("Error fetching GO:", go_id, "-", e$message))
    return(NULL)
  })
}

# Define Categories & GO Terms
functional_categories <- list(
  # Signaling
  "Notch_Signaling" = "GO:0007219",
  "Wnt_Signaling"   = "GO:0016055",
  "EGFR_Signaling"  = "GO:0007173",
  "JAK_STAT"        = "GO:0007259",
  
  # Identity
  "Transcription_Factors" = "GO:0003700", # DNA-binding transcription factor activity
  
  # Defense
  "Antimicrobial_Peptides" = "GO:0042742", # Defense response to bacterium
  
  # Metabolism
  "Lipid_Metabolism" = "GO:0006629",
  "Proteases"        = "GO:0004252"  # Serine-type endopeptidase activity
)

# Plotting Loop
for (cat_name in names(functional_categories)) {
  go_term <- functional_categories[[cat_name]]
  
  # 1. Get Genes
  go_genes <- get_genes_from_go(go_term)
  
  # 2. Filter: Must be in Seurat Object AND Variable/Expressed
  # Strategy: Intersect with Variable Features OR Top Markers to avoid noise
  valid_genes <- intersect(go_genes, rownames(seurat_obj))
  
  # Further filter: Keep only those with meaningful expression
  # Calculate average expression per gene across all cells
  # This is computationally expensive, so let's stick to VariableFeatures relative to dataset
  # OR: Use the markers_sig list we already generated!
  
  # Let's prioritize genes that are MARKERS for at least one cell type
  # This ensures they have spatial patterns
  relevant_genes <- intersect(valid_genes, unique(markers_sig$gene))
  
  # If too few, look at highly expressed ones
  if (length(relevant_genes) < 3) {
      # Fallback 1: Top variable features present in GO list
      top_var <- VariableFeatures(seurat_obj)[1:2000]
      relevant_genes <- intersect(valid_genes, top_var)
  }
  
  # Fallback 2: If still too few, just take highest expressed valid genes
  if (length(relevant_genes) < 3) {
      if(length(valid_genes) > 0) {
        relevant_genes <- valid_genes
        message("    Using all detected genes (not just variable ones) for this category.")
      }
  }
  
  # Pick Top 5-8 by total variance/expression to keep plot readable
  if (length(relevant_genes) > 0) {
    # Sort by total expression to pick "major" players
    counts_mat <- GetAssayData(seurat_obj, layer = "counts")
    # Backup to data layer if counts is missing (e.g. if only data slot populated after SCT/integration)
    if (is.null(counts_mat)) {
        counts_mat <- GetAssayData(seurat_obj, layer = "data")
    }
    
    total_expr <- rowSums(counts_mat[relevant_genes, , drop=FALSE])
    top_selection <- names(sort(total_expr, decreasing = TRUE))[1:min(8, length(relevant_genes))]
    
    message(paste("Plotting", cat_name, ":", paste(top_selection, collapse=", ")))
    
    # Plot Global Trend across all cells
    # We must set a temporary global identity so the subset function works
    obj_global <- seurat_obj
    obj_global$Global_ID <- "All Cells"
    Idents(obj_global) <- "Global_ID"
    
    p <- plot_trajectory_trend(obj_global, "All Cells", top_selection) 
      
    if (!is.null(p)) {
       p <- p + ggtitle(paste("Functional Trajectory:", cat_name, "(Global)"))
       print(p)
       ggsave(paste0("../figures/gut3_func_traj_", cat_name, ".png"), plot = p, width = 9, height = 6)
    }
  } else {
    message(paste("No relevant variable genes found for", cat_name))
  }
}
```

# 4. Binned Gene Ontology Analysis
We bin cells by gut length (size=2), find markers for each bin, and analyze GO enrichment.

## 4.1. Helper Functions
```{r binning_setup}
# 1. Create Bins
bin_size <- 2
seurat_obj$Bin_ID <- floor(seurat_obj$Spatial_Pseudotime / bin_size) * bin_size
seurat_obj$Bin_Label <- paste0("Pos_", sprintf("%03d", seurat_obj$Bin_ID))

# Ensure bins are ordered factors
bins <- sort(unique(seurat_obj$Bin_ID))
labels <- paste0("Pos_", sprintf("%03d", bins))
seurat_obj$Bin_Label <- factor(seurat_obj$Bin_Label, levels = labels)

message(paste("Created", length(bins), "spatial bins."))

# 2. GO Analysis Function
run_binned_go <- function(obj, subset_name) {
  message(paste("Running Binned GO for:", subset_name))
  
  # Set Indents to Bins
  Idents(obj) <- obj$Bin_Label
  
  # Find Markers for each BIN
  # We check "one-vs-rest" to find what is specific to that spatial position
  bin_markers <- FindAllMarkers(
    obj,
    only.pos = TRUE,
    min.pct = 0.1,      # Lower threshold to catch spatially restricted signals
    logfc.threshold = 0.25,
    verbose = FALSE
  )
  
  if (nrow(bin_markers) == 0) {
    message("No markers found for bins.")
    return(NULL)
  }
  
  # Run GO Enrichment
  go_results <- compareCluster(
    gene ~ cluster, 
    data = bin_markers, 
    fun = "enrichGO",
    OrgDb = org.Dm.eg.db,
    keyType = "SYMBOL",
    ont = "BP",  # Biological Process
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.2
  )
  
  # SIMPLIFY: Remove redundant terms to condense the list
  try({
      go_results <- clusterProfiler::simplify(go_results, cutoff = 0.7, by = "p.adjust", select_fun = min)
      message("Simplified GO terms.")
  }, silent = TRUE)
  
  return(go_results)
}
```

## 4.2. Run Analysis on Subsets
```{r run_go_subsets}
# A. All Cell Types
go_all <- run_binned_go(seurat_obj, "All Cells")

# B. Progenitors (ISC + EB)
progenitors <- c("ISC", "EB")
# Check if present
valid_prog <- progenitors[progenitors %in% unique(seurat_obj$Predicted_CellType)]
if(length(valid_prog) > 0) {
  # Better safe: subset by metadata
  obj_prog <- subset(seurat_obj, subset = Predicted_CellType %in% valid_prog)
  go_prog <- run_binned_go(obj_prog, "Progenitors")
} else {
  go_prog <- NULL
}

# C. Enterocytes (All EC types)
ec_types <- c("aEC", "mEC", "pEC", "dEC", "Enterocyte")
valid_ec <- ec_types[ec_types %in% unique(seurat_obj$Predicted_CellType)]
if(length(valid_ec) > 0) {
  obj_ec <- subset(seurat_obj, subset = Predicted_CellType %in% valid_ec)
  go_ec <- run_binned_go(obj_ec, "Enterocytes")
} else {
  go_ec <- NULL
}
```

# 5. GO Term Visualization
We create two specific plots:
1. **Specific Terms**: Terms found present specifically in one area.
2. **Bimodal Terms**: Terms that appear early and late, but not in the middle.

```{r go_visualization, fig.width=12, fig.height=10}
plot_go_heatmap <- function(go_res, title, filter_type = "specific", max_terms = 50) {
  if (is.null(go_res)) return(NULL)
  
  df <- as.data.frame(go_res)
  
  # Matrix: Terms x Bins (p.adjust or count)
  # We use -log10(p.adjust) for intensity
  df$intensity <- -log10(df$p.adjust)
  
  # Pivot
  mat_long <- df[, c("Description", "cluster", "intensity")]
  mat_wide <- tidyr::pivot_wider(mat_long, names_from = cluster, values_from = intensity, values_fill = 0)
  
  # Convert to matrix
  mat <- as.matrix(mat_wide[, -1])
  rownames(mat) <- mat_wide$Description
  
  # --- FILTERING LOGIC ---
  n_bins <- ncol(mat)
  
  # Define indices (approximate thirds)
  early_idx  <- 1:floor(n_bins/3)
  mid_idx    <- (floor(n_bins/3)+1):floor(2*n_bins/3)
  late_idx   <- (floor(2*n_bins/3)+1):n_bins
  
  keep_rows <- c()
  
  for (i in 1:nrow(mat)) {
    row_vals <- mat[i, ]
    vals_early <- mean(row_vals[early_idx] > 1.3)
    vals_mid   <- mean(row_vals[mid_idx] > 1.3)
    vals_late  <- mean(row_vals[late_idx] > 1.3)
    
    if (filter_type == "specific") {
      if (max(row_vals) > 0 && (max(row_vals) / sum(row_vals)) > 0.4) {
        keep_rows <- c(keep_rows, i)
      }
    } else if (filter_type == "bimodal") {
      if (vals_early > 0 && vals_late > 0 && vals_mid == 0) {
        keep_rows <- c(keep_rows, i)
      }
    }
  }
  
  if (length(keep_rows) < 2) {
    message(paste("Not enough terms found for filter:", filter_type))
    return(NULL)
  }
  
  mat_filtered <- mat[keep_rows, , drop = FALSE]
  
  # --- EVEN SPATIAL DISTRIBUTION: Pick Top N per Region ---
  # Find peak bin index for each row
  peak_bin <- apply(mat_filtered, 1, which.max)
  
  # Assign region to each row based on peak
  row_regions <- ifelse(peak_bin %in% early_idx, "Anterior",
                 ifelse(peak_bin %in% mid_idx,   "Mid", "Posterior"))
  
  # Number of terms per region
  n_per_reg <- ceiling(max_terms / 3)
  
  distributed_rows <- c()
  for (reg in c("Anterior", "Mid", "Posterior")) {
    reg_rows <- rownames(mat_filtered)[row_regions == reg]
    if (length(reg_rows) > 0) {
      # Sort region-specific rows by their max significance
      reg_mat <- mat_filtered[reg_rows, , drop=FALSE]
      reg_sig <- apply(reg_mat, 1, max)
      # Keep top N
      keep_reg <- names(sort(reg_sig, decreasing = TRUE))[1:min(n_per_reg, length(reg_rows))]
      distributed_rows <- c(distributed_rows, keep_reg)
    }
  }
  
  mat_final <- mat_filtered[distributed_rows, , drop=FALSE]
  # Re-order final matrix by peak position for smooth visual flow
  peak_final <- apply(mat_final, 1, which.max)
  mat_final <- mat_final[order(peak_final), , drop=FALSE]
  
  # Heatmap
  pheatmap(
    mat_final,
    cluster_cols = FALSE, 
    cluster_rows = FALSE, 
    main = paste(title, "-", filter_type, "(Even Dist)"),
    show_colnames = TRUE,
    fontsize_row = 8,
    color = colorRampPalette(c("white", "red"))(50)
  )
}
```

# 5.1 Plot Heatmaps (Even Distribution)
```{r plot_heatmaps, fig.width=12, fig.height=10}
# 1. All Cells
plot_go_heatmap(go_all, "All Cells - Specific", "specific")
plot_go_heatmap(go_all, "All Cells - Bimodal", "bimodal")

# 2. Progenitors
if(!is.null(go_prog)) {
  plot_go_heatmap(go_prog, "Progenitors - Specific", "specific")
  plot_go_heatmap(go_prog, "Progenitors - Bimodal", "bimodal")
}

# 3. Enterocytes
if(!is.null(go_ec)) {
  plot_go_heatmap(go_ec, "Enterocytes - Specific", "specific")
  plot_go_heatmap(go_ec, "Enterocytes - Bimodal", "bimodal")
}
```

# 5.2 Functional Grouped Heatmap (All Cells)
Focus on Signaling, Metabolism, Identity, and others.

```{r go_functional_grouped, fig.width=12, fig.height=12}
plot_functional_grouped_heatmap <- function(go_res, title, max_per_cat = 15) {
  if (is.null(go_res)) return(NULL)
  
  df <- as.data.frame(go_res)
  df$intensity <- -log10(df$p.adjust)
  
  # --- Helper: Categorize Terms ---
  categorize_term <- function(desc) {
    desc_lc <- tolower(desc)
    
    # signaling pathways (grouped together as requested)
    signaling_keywords <- c("signal", "notch", "wnt", "egfr", "jak", "stat", "tgf", "hedgehog", "hippo", "pathway", "receptor", "ligand", "transduction")
    if (any(sapply(signaling_keywords, grepl, x = desc_lc))) return("Signaling")
    
    # Identity / Differentiation
    identity_keywords <- c("differentiation", "fate", "identity", "specification", "development", "transcription factor", "morphogenesis")
    if (any(sapply(identity_keywords, grepl, x = desc_lc))) return("Identity/Development")
    
    # Metabolism
    metabolism_keywords <- c("metabolic", "metabolism", "biosynthetic", "catabolic", "lipid", "protease", "peptidase", "carbohydrate", "energy")
    if (any(sapply(metabolism_keywords, grepl, x = desc_lc))) return("Metabolism")
    
    # Defense/Stress
    defense_keywords <- c("defense", "immune", "stress", "stimulus", "bacterium", "antimicrobial", "damage", "wounding")
    if (any(sapply(defense_keywords, grepl, x = desc_lc))) return("Stress/Defense")
    
    return("Other Processes")
  }
  
  df$Category <- sapply(df$Description, categorize_term)
  
  # --- EVEN SELECTION per category and region ---
  # Pivot to matrix first to find peaks
  mat_long <- df[, c("Description", "cluster", "intensity", "Category")]
  mat_wide <- tidyr::pivot_wider(mat_long, names_from = cluster, values_from = intensity, values_fill = 0)
  
  # Determine peak region for each term
  bin_names <- colnames(mat_wide)[3:ncol(mat_wide)]
  n_bins <- length(bin_names)
  early_idx <- 1:floor(n_bins/3)
  mid_idx   <- (floor(n_bins/3)+1):floor(2*n_bins/3)
  late_idx  <- (floor(2*n_bins/3)+1):n_bins
  
  # Extract values matrix
  val_mat <- as.matrix(mat_wide[, bin_names])
  peak_bin <- apply(val_mat, 1, which.max)
  mat_wide$Region <- ifelse(peak_bin %in% early_idx, "Anterior",
                     ifelse(peak_bin %in% mid_idx,   "Mid", "Posterior"))
  
  # Add Max Intensity as column to avoid dimension mismatch in grouped top_n
  mat_wide$MaxIntensity <- apply(val_mat, 1, max)
  
  # Select Top N per Category to ensure coverage
  selected_terms <- mat_wide %>%
    group_by(Category) %>%
    # Within each category, we still want a bit of spatial distribution if possible, 
    # but top significance takes priority for functional groups
    top_n(max_per_cat, wt = MaxIntensity) %>%
    ungroup() %>%
    pull(Description)
    
  # Final Matrix
  final_df <- mat_wide[mat_wide$Description %in% selected_terms, ]
  mat_final <- as.matrix(final_df[, bin_names])
  rownames(mat_final) <- final_df$Description
  
  # Annotation for rows
  annotation_row <- data.frame(Category = factor(final_df$Category))
  rownames(annotation_row) <- final_df$Description
  
  # Ordering: Group by Category first, then by Peak Bin
  df_order <- data.frame(
    Description = final_df$Description,
    Category = final_df$Category,
    Peak = apply(mat_final, 1, which.max)
  )
  # Order by category (Grouped) and then by spatial peak
  df_order <- df_order[order(df_order$Category, df_order$Peak), ]
  
  mat_final <- mat_final[df_order$Description, ]
  annotation_row <- annotation_row[df_order$Description, , drop=FALSE]
  
  # Plot
  pheatmap(
    mat_final,
    cluster_cols = FALSE,
    cluster_rows = FALSE,
    annotation_row = annotation_row,
    main = paste(title, "- Functional Groups"),
    show_colnames = TRUE,
    fontsize_row = 7,
    color = colorRampPalette(c("white", "blue", "red"))(100)
  )
}

# Run for All Cells
plot_functional_grouped_heatmap(go_all, "All Cells")
```
