---
title: "Hierarchical Clustering of Regions"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(Seurat)
library(dplyr)
library(ggplot2)
library(reshape2)
library(zoo)

# Clustering and Gap Stat packages
library(cluster)
library(factoextra)
library(dendextend)
library(rioja)
```

## 1. Load Data
```{r load_data}
if (exists("merged_guts")) {
    merged_guts_local <- merged_guts
} else {
    path_def <- "../../111125ST_def/Random data/merged_guts.rds"
    path_norm <- "../Random data/merged_guts.rds"
    if (file.exists(path_norm)) {
        cat("Loading merged_guts from", path_norm, "\n")
        merged_guts_local <- readRDS(path_norm)
    } else if (file.exists(path_def)) {
        cat("Loading merged_guts from", path_def, "\n")
        merged_guts_local <- readRDS(path_def)
    } else {
        stop("merged_guts not found in environment, and merged_guts.rds not found in Random data directory. Please load it manually.")
    }
}

# Filter out ambiguous cells, similarly to previous analysis
is_ambiguous <- grepl("ambiguous", merged_guts_local$Predicted_CellType, ignore.case = TRUE)
clean_cells <- Cells(merged_guts_local)[!is_ambiguous]
sub_guts <- subset(merged_guts_local, cells = clean_cells)

cat("✓ Data loaded and cleaned.\n")
```

## 2. Binning Strategy & Missing Data Handling
We apply the 2.5% binning strategy (40 bins) to the spatial pseudotime.

```{r binning}
# Define 40 bins (2.5% resolution)
n_bins <- 107

# --- Customizable Option ---
# Set start_bin > 1 to remove early anterior/uninformative bins.
# Example: 3 will remove bins 1 and 2.
start_bin <- 7
# -------------------------

# Identify variable features
var_genes <- VariableFeatures(sub_guts)
if(length(var_genes) == 0) {
    sub_guts <- FindVariableFeatures(sub_guts, nfeatures = 2000, verbose = FALSE)
    var_genes <- VariableFeatures(sub_guts)
}

# Bin spatial pseudotime
sub_guts$Bin <- factor(cut(sub_guts$Spatial_Pseudotime, breaks = n_bins, labels = FALSE), levels = 1:n_bins)

# Calculate mean expression per bin
exp_mat <- GetAssayData(sub_guts, layer = "data")[var_genes, ]
bin_means <- t(apply(exp_mat, 1, function(x) tapply(x, sub_guts$Bin, mean, na.rm = TRUE)))

# Handle missing data / tissue breaks
# Transpose so rows are bins, columns are genes
bin_mat <- t(bin_means)

# Smoothing over NAs (tissue breaks)
cat("Assessing tissue breaks and imputing missing bins...\n")
library(zoo)

# Z-scale per gene (column) and smooth
bin_scaled <- apply(bin_mat, 2, function(x) {
    if(all(is.na(x)) || sd(x, na.rm=T) == 0) return(rep(0, length(x)))
    # Impute NAs with interpolation to cover tissue breaks before scaling
    x_imputed <- na.approx(x, na.rm = FALSE) 
    # If NAs at boundaries, carry forward/backward
    x_imputed <- na.locf(na.locf(x_imputed, na.rm=FALSE), fromLast=TRUE, na.rm=FALSE)
    
    # Scale
    scaled <- (x_imputed - mean(x_imputed, na.rm=T)) / sd(x_imputed, na.rm=T)
    return(scaled)
})

# Any remaining NAs (e.g. all 0 genes) can be zeroed out
bin_scaled[is.na(bin_scaled)] <- 0

# Subset bins if start_bin > 1
if (start_bin > 1) {
    cat("Removing bins before bin:", start_bin, "\n")
    bin_scaled <- bin_scaled[start_bin:nrow(bin_scaled), , drop = FALSE]
}

cat("✓ Applied 2.5% binning and handled gaps via interpolation.\n")
```

## 3. Gap Statistic for Optimal Spatial Clusters
We use `clusGap` to determine the optimal number of clusters, evaluating up to K=10.
We wrap `rioja::chclust` inside a custom function so `clusGap` can sequentially evaluate K.

```{r gap_stat, fig.width=8, fig.height=5}
# Distance matrix (Euclidean on scaled profiles)
dist_mat <- dist(bin_scaled, method = "euclidean")

# Custom clustering function wrapper for clusGap
# Since CONISS is hierarchical, we cut the tree into k clusters
coniss_clust_wrapper <- function(x, k) {
    d <- dist(x, method = "euclidean")
    ch <- chclust(d, method = "coniss")
    cl <- cutree(ch, k = k)
    return(list(cluster = cl))
}

cat("Calculating Gap Statistic using CONISS constrained clustering (may take a few moments)...\n")
set.seed(42)
gap_stat <- clusGap(bin_scaled, FUN = coniss_clust_wrapper, K.max = 10, B = 50)

# Optimal cluster determination using the "firstMax" or "Tibs2001SEmax" method
# Commonly used is the standard Tibshirani approach
optimal_k <- maxSE(gap_stat$Tab[, "gap"], gap_stat$Tab[, "SE.sim"], method = "Tibs2001SEmax")
cat("Optimal number of clusters calculated:", optimal_k, "\n")

# Plot Gap Statistic
p_gap <- fviz_gap_stat(gap_stat, linecolor = "black", maxSE = list(method = "Tibs2001SEmax", color = "deeppink")) +
    theme_minimal() +
    labs(title = "Gap Statistic: Optimal Spatial Clusters",
         x = "Number of clusters", y = "Gap") +
    theme(plot.title = element_text(face = "bold", size = 14))

# Save
if (!dir.exists("../figures/3D_glo")) dir.create("../figures/3D_glo", recursive=TRUE, showWarnings=FALSE)
ggsave("../figures/3D_glo/Gap_Statistic_Clusters.png", p_gap, width = 6, height = 5)

print(p_gap)
```

## 4. Spatially Constrained Hierarchical Clustering Profile
Using the optimal K found above (or manually defining it), we run CONISS to trace strictly contiguous clustering blocks that adhere to midgut morphology constraints.

```{r constrained_clustering, fig.width=10, fig.height=7}
# Perform hierarchical constrained clustering using CONISS
spatial_hc <- chclust(dist_mat, method = "coniss")

# Hard-wipe all labels at the root object level so ggplot doesn't forcefully draw them
spatial_hc$labels <- rep("", length(spatial_hc$order))

# --- Customizable Option: Override optimal K ---
# Set manual_k to a number (e.g., 5 or 5) to override the mathematical estimation if the Gap statistic is conservative.
manual_k <- 5 
# -------------------------

# Use optimal K from Gap Stat or manual override
k_to_use <- ifelse(!is.null(manual_k), manual_k, ifelse(optimal_k > 1, optimal_k, 5)) 

# Create dendrogram
dend <- as.dendrogram(spatial_hc)

# Assign colors to clusters
dend <- color_branches(dend, k = k_to_use)

# Extract cluster assignments for heatmap visualization or backgrounds
cluster_labels <- cutree(spatial_hc, k = k_to_use)

# Safely extract the cut height as a numeric value
cut_height <- as.numeric(names(dendextend::heights_per_k.dendrogram(dend)[as.character(k_to_use)]))

p_dend <- ggplot(as.ggdend(dend)) +
    theme_minimal() +
    labs(title = "Spatially Constrained Hierarchical Clustering",
         subtitle = paste("Resolution: 40 Bins (2.5%) | Optimal K =", k_to_use),
         y = "Height (Dissimilarity)",
         x = "Anatomical Bin Position (Anterior -> Posterior)") +
    geom_hline(yintercept = cut_height, linetype = "dashed", color = "deeppink", size = 1) +
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank()
    )

# Save
ggsave("../figures/3D_glo/Spatially_Constrained_Dendrogram.png", p_dend, width = 10, height = 7)

print(p_dend)
```

## 5. Overlaying Regional Backgrounds (Optional Aesthetics)
To closely match highly polished plots (e.g., adding colored background slices for the regions), we can extract the breakpoints of the clusters and apply rectangles in `ggplot`.

```{r dendro_backgrounds, fig.width=10, fig.height=7}
# Find boundaries of clusters to draw backing rectangles
cluster_rle <- rle(cluster_labels)
ends <- cumsum(cluster_rle$lengths)
starts <- c(1, ends[-length(ends)] + 1)

# Subtle background colors matching the branches
bg_colors <- scales::hue_pal()(k_to_use)

rects_df <- data.frame(
    xmin = starts - 0.5,
    xmax = ends + 0.5,
    ymin = -Inf,
    ymax = max(spatial_hc$height) * 1.05,
    fill = factor(cluster_rle$values)
)

# Dutta Regions Math
region_shift <- 4 # Customizable shift for region label plotting
dutta_breaks <- c(0, 15, 39, 53, 85, 100) * (n_bins / 100)
display_breaks <- dutta_breaks - (start_bin - 1) + 0.5 + region_shift
display_mids <- (display_breaks[-1] + display_breaks[-length(display_breaks)]) / 2
region_labs <- c("R1", "R2", "R3", "R4", "R5")

valid_regions <- which(display_mids > 0.5 & display_mids < nrow(bin_scaled) + 0.5)
valid_lines <- display_breaks[display_breaks > 0.5 & display_breaks < nrow(bin_scaled) + 0.5]

# Spradling Subregions Mapping
sp_breaks <- c(0, 15, 28, 39, 44, 48, 53, 70, 85, 92, 100) * (n_bins / 100)
sp_breaks_adj <- sp_breaks - (start_bin - 1) + 0.5 + region_shift
sp_mids <- (sp_breaks_adj[-1] + sp_breaks_adj[-length(sp_breaks_adj)]) / 2
sp_labs <- c("A1", "A2", "A3", "mEC (R3)", "Cu (R3)", "LFC (R3)", "P1", "P2", "P3", "P4")
valid_sp <- which(sp_mids > 0.5 & sp_mids < nrow(bin_scaled) + 0.5)

# Add as.ggdend(dend) natively rather than empty ggplot to avoid missing trees!
p_dend_bg <- ggplot(as.ggdend(dend)) +
    geom_rect(data = rects_df, aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=fill), alpha = 0.2, inherit.aes = FALSE) +
    scale_fill_manual(values = bg_colors, guide = "none") +
    geom_vline(xintercept = valid_lines, color = "black", linetype = "dotted", size = 0.8) +
    annotate("text", x = display_mids[valid_regions], y = max(spatial_hc$height) * 1.02, 
             label = region_labs[valid_regions], fontface = "bold", size = 4) +
    geom_hline(yintercept = cut_height, linetype = "dashed", color = "deeppink", size = 1) +
    theme_minimal() +
    labs(title = "Constrained Hierarchical Clustering with Anatomical Domains",
         subtitle = paste("CONISS Clustering | 40 Bins (2.5%) | K =", k_to_use),
         y = "Height", x = "Spradling Subregions & Major Celltypes") +
    scale_x_continuous(breaks = sp_mids[valid_sp], labels = sp_labs[valid_sp]) +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10, face = "bold.italic", color = "darkblue"),
        axis.ticks.x = element_line(color="black"),
        panel.grid = element_blank()
    )

ggsave("../figures/3D_glo/Spatially_Constrained_Dendro_Backgrounds.png", p_dend_bg, width = 10, height = 7)
print(p_dend_bg)

# Standard R plot:
png("../figures/3D_glo/Constrained_Dendrogram_BaseR.png", width=2500, height=1800, res=300)
par(mar = c(5, 4, 3, 1)) # Shifted bottom margin lower for Spradling labels
plot(dend, main = "Constrained Hierarchical Clustering (CONISS)", ylab = "Height", xlab="", leaflab = "none", xaxt = "n")
rect(starts - 0.5, -100, ends + 0.5, 100, col = adjustcolor(bg_colors, alpha.f = 0.2), border = NA)
lines(x = c(0, nrow(bin_scaled) + 1), y = rep(cut_height, 2), col="deeppink", lty=2, lwd=2)
abline(v = valid_lines, col = "black", lty = 3, lwd = 2)
mtext(region_labs[valid_regions], side = 3, at = display_mids[valid_regions], line = 0.5, font = 2)
axis(1, at = sp_mids[valid_sp], labels = sp_labs[valid_sp], las = 2, cex.axis = 0.75, font = 3, col.axis="darkblue")
dev.off()

cat("✓ Clustering plots constructed and saved.\n")
```

## 6. Domain Signature Gene Identification
```{r domain_signatures}
# Map cluster labels (k_to_use clusters) back to bins
bin_clusters <- data.frame(Bin = as.character(start_bin:n_bins), Cluster = paste0("R", cluster_labels))

# Append to metadata safely
md <- sub_guts@meta.data
md$Bin_Char <- as.character(md$Bin)
md <- left_join(md, bin_clusters, by = c("Bin_Char" = "Bin"))
sub_guts$Spatial_Domain <- md$Cluster

# Filter out NA (cells in bins < start_bin)
sub_domains <- subset(sub_guts, subset = !is.na(Spatial_Domain))
Idents(sub_domains) <- "Spatial_Domain"

cat("Running FindAllMarkers to identify domain signatures...\n")
markers <- FindAllMarkers(sub_domains, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
top20 <- markers %>% group_by(cluster) %>% top_n(n = 20, wt = avg_log2FC)

cat("✓ Identified top 20 signature genes for each spatial domain.\n")

# Dynamically print the top 20 table for immediate viewing (not saved)
library(knitr)
cat("\n--- DOMAIN SIGNATURE GENES (TOP 20) ---\n")
print(kable(
    top20 %>% dplyr::select(cluster, gene, avg_log2FC, p_val_adj) %>% arrange(cluster, desc(avg_log2FC)), 
    format = "pipe"
))
```

## 7. Regional Gene Module Trajectories
```{r gene_modules, fig.width=10, fig.height=6}
module_list <- split(top20$gene, top20$cluster)
sub_domains <- AddModuleScore(sub_domains, features = module_list, name = "Domain_Mod_")

mod_cols <- paste0("Domain_Mod_", 1:length(module_list))
domain_names <- names(module_list)

plot_data <- sub_domains@meta.data %>% 
    dplyr::select(Spatial_Pseudotime, all_of(mod_cols)) %>%
    melt(id.vars="Spatial_Pseudotime", variable.name="ModuleRaw", value.name="Score")

plot_data$Domain <- factor(domain_names[match(plot_data$ModuleRaw, mod_cols)], levels = paste0("R", unique(cluster_labels)))

# Calculate boundaries for the vertical dotted lines dynamically across chosen cluster limits
# Crucially, we calculate this from the single-cell annotations, NOT the melted multi-module tracking data!
domain_bounds <- sub_domains@meta.data %>% 
    group_by(Spatial_Domain) %>% 
    summarize(min_pt = min(Spatial_Pseudotime), max_pt = max(Spatial_Pseudotime)) %>%
    arrange(min_pt)
    
v_lines <- (domain_bounds$max_pt[-nrow(domain_bounds)] + domain_bounds$min_pt[-1]) / 2
x_mids <- c(0, v_lines, 100)
x_mids <- (x_mids[-1] + x_mids[-length(x_mids)]) / 2

if (!require("ggsci", character.only = TRUE)) install.packages("ggsci", ask = FALSE)
if (!require("ggpubr", character.only = TRUE)) install.packages("ggpubr", ask = FALSE)

# Professional publication-grade aesthetic override!
min_y <- min(plot_data$Score, na.rm=TRUE)

p_modules <- ggplot(plot_data, aes(x=Spatial_Pseudotime, y=Score, color=Domain)) +
    # Add subtle shaded domain backgrounds alternating sequentially
    annotate("rect", xmin = c(0, v_lines), xmax = c(v_lines, 100), 
             ymin = -Inf, ymax = Inf, 
             fill = rep(c("grey95", "white"), length.out = k_to_use), alpha = 1) +
    # Clean dashed separation lines
    geom_vline(xintercept = v_lines, linetype="dashed", color="grey60", size=0.5) +
    # Draw thicker, premium smoothed lines over the background
    geom_smooth(method = "loess", span=0.3, se=FALSE, size=1.8, alpha=0.9, lineend="round") +
    # Map domain labels deeply to the bottom
    annotate("text", x = x_mids, y = -Inf, 
             label = domain_bounds$Spatial_Domain, vjust = -0.5, size = 5.5, fontface = "bold", color="grey20") +
    # Nature Publishing Group (NPG) premium scientific palette
    scale_color_npg(name = "Related\ndomain:") +
    # Base ggpubr layout (widely used in journals)
    theme_pubr(base_size = 14, border = TRUE) +
    labs(x="Spatial Position (0-100%)", y="Module Score (Loess Fit)") +
    theme(
        axis.title = element_text(face="bold"),
        legend.title = element_text(face="bold"),
        legend.position = "right",
        # Add bottom margin padding so the annotation text doesn't clip below axis
        plot.margin = margin(t=15, r=20, b=25, l=15)
    ) +
    scale_x_continuous(breaks = seq(0, 100, by=10), expand = c(0, 0)) +
    scale_y_continuous(expand = expansion(mult = c(0.15, 0.05))) +
    coord_cartesian(clip = "off")

ggsave("../figures/3D_glo/Regional_Gene_Modules.png", p_modules, width = 9, height = 5)
print(p_modules)
```

## 8. Metabolic GO Spatial Enrichment Profile
```{r go_enrichment, fig.width=8, fig.height=10}
if (!require("clusterProfiler", character.only = TRUE)) stop("Please install clusterProfiler")
if (!require("org.Dm.eg.db", character.only = TRUE)) stop("Please install org.Dm.eg.db")
library(stringr)

# 20 genes is mathematically too small an overlap pool to consistently trigger GO pathways across all districts.
# We slice deeper to the Top 150 specifically for the programmatic enrichment step so the ontology matching functions cleanly.
top_go_genes <- markers %>% group_by(cluster) %>% top_n(n = 150, wt = avg_log2FC)
module_list_go <- split(top_go_genes$gene, top_go_genes$cluster)

cat("Running compareCluster for GO enrichment...\n")
go_res <- compareCluster(geneCluster = module_list_go, fun = "enrichGO", 
                         OrgDb = org.Dm.eg.db, keyType = "SYMBOL", 
                         ont = "BP", pAdjustMethod = "BH", pvalueCutoff = 0.5, qvalueCutoff = 0.8)

if (!is.null(go_res) && nrow(go_res@compareClusterResult) > 0) {
    # Broaden the metabolic terms to include specific Drosophila gut digestive/transport pathways!
    df_go <- go_res@compareClusterResult %>% 
        filter(grepl("metabol|catabol|biosynthe|digest|transport|lipid|glyco|carbo|amino|acid|water|ion|peptide|protein|hormon|excret", Description, ignore.case=TRUE))
    
    go_res@compareClusterResult <- df_go
    
    # Simplify removes redundancy mathematically, we wrap it in a tryCatch in case semantic similarity drops too much
    go_sim <- tryCatch({
        simplify(go_res, cutoff = 0.7, by = "p.adjust", select_fun = min)
    }, error = function(e) {
        cat("Note: Semantic mapping simplification failed, falling back to manual trimming.\n")
        return(go_res)
    })
    
    df_sim <- go_sim@compareClusterResult
    
    # Identify the exact Names of the top 6 driving terms per cluster
    top_terms <- df_sim %>%
        group_by(Cluster) %>%
        arrange(p.adjust) %>%
        slice_head(n = 6) %>%
        pull(Description) %>%
        unique()
        
    # Map those top target terms comprehensively across ALL clusters for the dotplot matrix
    top_go <- df_sim %>% filter(Description %in% top_terms)
        
    top_go$GeneRatio_num <- sapply(top_go$GeneRatio, function(x) {
        p <- as.numeric(strsplit(x, "/")[[1]])
        p[1] / p[2]
    })
    
    top_go$Description <- str_wrap(top_go$Description, width = 45)
    
    top_go$Cluster <- factor(top_go$Cluster, levels = paste0("R", unique(cluster_labels)))
    
    # Order Y-axis sequentially spanning structural location mapping and strictness
    term_order <- top_go %>% 
        group_by(Description) %>% 
        summarize(min_p = min(p.adjust), PrimaryCluster = min(as.numeric(Cluster))) %>%
        arrange(PrimaryCluster, min_p) %>%
        pull(Description)
        
    top_go$Description <- factor(top_go$Description, levels = rev(term_order))
    
    p_dot <- ggplot(top_go, aes(x = Cluster, y = Description, size = GeneRatio_num, color = p.adjust)) +
        geom_point() +
        scale_color_viridis_c(option = "magma", direction = -1, name = "P_adj") +
        theme_minimal() +
        labs(size = "Gene ratio", title = "Highly Condensed Metabolic GO Terms", y = "", x = "") +
        theme(axis.text.x = element_text(size=12, face="bold"),
              axis.text.y = element_text(size=10),
              panel.grid.major.x = element_line(linetype="dotted", color="grey80"),
              panel.border = element_rect(color="black", fill=NA))
              
    ggsave("../figures/3D_glo/Metabolic_GO_Enrichment.png", p_dot, width = 8, height = 10)
    print(p_dot)
} else {
    cat("Warning: No significant GO enrichment results found to plot!\n")
}
```
---
End of script.
