---
title: "DOT Deconvolution (Exact Match)"
output: html_document
date: "2026-02-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Seurat)
library(SeuratDisk)
library(ggplot2)
library(dplyr)
library(Matrix)
library(spacexr)
library(SummarizedExperiment)
library(SpatialExperiment)
# SaezLab DOT package is loaded as DOTr
library(DOTr) 
library(org.Dm.eg.db) # from St111125
library(AnnotationDbi) # from St111125
library(Matrix.utils) # from Processing.Rmd logic
```

# 1. Prepare Spatial Data (Replicating `St111125.Rmd`)

We replicate the exact steps from `St111125.Rmd` to generate the `seurat_obj` in its final state.

```{r load_spatial}
# Path to spatial data
data_dir <- "/home/mattia/Thesis/111125ST_def/binned_data"

print("Loading 10X Spatial data...")
spe_object <- Load10X_Spatial(data_dir, slice = "HE_Stain_Slice")

# 1. Set default assay
DefaultAssay(spe_object) <- "Spatial.016um"

# 2. Filter out zero count spots
spe_object <- subset(spe_object, subset = nCount_Spatial.016um > 0)

# 3. SCTransform
print("Running SCTransform...")
spe_object <- SCTransform(spe_object, assay = "Spatial.016um", verbose = FALSE)

# 4. Run PCA
print("Running PCA...")
seurat_obj <- RunPCA(spe_object, assay = "SCT", features = VariableFeatures(spe_object))

# 5. Run UMAP
print("Running UMAP...")
seurat_obj <- RunUMAP(seurat_obj, dims = 1:30, reduction = "pca")

# 6. Find Neighbors & Clusters
print("Clustering...")
seurat_obj <- FindNeighbors(seurat_obj, dims = 1:30)
seurat_obj <- FindClusters(seurat_obj, resolution = 0.4)

# 7. Add Mitochondrial Percentage
print("Calculating Percentages...")
seurat_obj[["percent.mt"]] <- PercentageFeatureSet(seurat_obj, pattern = "^mt:")

# 8. Add Mean Expression
normalized_data <- GetAssayData(seurat_obj, assay = "SCT", layer = "data")
mean_expression_per_pixel <- colMeans(normalized_data)
seurat_obj$mean_expression <- mean_expression_per_pixel

# 9. Add Ribosomal Percentage (Final step of St111125.Rmd)
seurat_obj[["percent.ribo"]] <- PercentageFeatureSet(seurat_obj, pattern = "^Rp[LS]")

print("Spatial Object Prepared.")
```

# 2. Load Single Cell Reference

```{r load_sc}
sc_path <- "/home/mattia/Thesis/Single-cell/seurat_ctrl_with_metadata.rds"
print("Loading Single-Cell Reference...")
if (!file.exists(sc_path)) stop("SC file not found.")
seurat_ref <- readRDS(sc_path)

# Ensure celltype column
if (!"celltype" %in% colnames(seurat_ref@meta.data)) {
    print("Warning: 'celltype' column not found. Please check metadata.")
    print(head(seurat_ref@meta.data))
} else {
    Idents(seurat_ref) <- "celltype"
}
```

# 3. Exact Gene Matching (Replicating `Processing.Rmd`)

We use the *exact* logic provided in `Processing.Rmd` to clean, intersect, aggregate, and match genes between Spatial and Single-Cell data.

```{r gene_matching}
print("Starting Exact Gene Matching...")

# 1. Extract Raw Counts
# Note: St111125 uses "Spatial.016um" (raw) or "SCT" (normalized). Processing.Rmd used "Spatial.016um" count slot equivalent.
# We will use the 'counts' slot from the default assay of seurat_obj (which started as Spatial.016um)
# BUT seurat_obj is now the SCT transformed object. The original counts are in "Spatial.016um".
spatial_counts <- GetAssayData(seurat_obj, assay = "Spatial.016um", layer = "counts")
sc_counts <- GetAssayData(seurat_ref, assay = "RNA", layer = "counts")

## 1. Extract Gene Names
spatial_genes_raw <- rownames(spatial_counts)
sc_genes_raw <- rownames(sc_counts)

## 2. Aggressive Cleaning and Harmonization
# a. Remove anything after a period
# b. Convert to uppercase
spatial_genes_clean <- toupper(gsub("\\..*$", "", spatial_genes_raw))
sc_genes_clean <- toupper(gsub("\\..*$", "", sc_genes_raw))

# c. Assign back
rownames(spatial_counts) <- spatial_genes_clean
rownames(sc_counts) <- sc_genes_clean

## 3. Determine Valid Intersection
common_genes <- intersect(rownames(spatial_counts), rownames(sc_counts))

## 4. Check for Duplicates (Crucial Safety Check)
cat("Duplicates in Spatial:", sum(duplicated(rownames(spatial_counts))), "\n")
cat("Duplicates in SC:", sum(duplicated(rownames(sc_counts))), "\n")

## 5. Filter the Matrices
spatial_counts_filtered <- spatial_counts[common_genes, ]
sc_counts_filtered <- sc_counts[common_genes, ]

## 6. Aggregate (dMeltAndSum logic)
# Use Matrix.utils for robust aggregation
print("Aggregating duplicates...")
spatial_counts_agg <- Matrix.utils::aggregate.Matrix(
    spatial_counts, 
    group = rownames(spatial_counts), 
    fun = "sum"
)

sc_counts_agg <- Matrix.utils::aggregate.Matrix(
    sc_counts, 
    group = rownames(sc_counts), 
    fun = "sum"
)

## 7. Final Intersection & Filtering
common_genes_final <- intersect(rownames(spatial_counts_agg), rownames(sc_counts_agg))

spatial_counts_final <- spatial_counts_agg[common_genes_final, ]
sc_counts_final <- sc_counts_agg[common_genes_final, ]

print(paste("Final Common Genes:", length(common_genes_final)))
```

# 4. Prepare for DOT

```{r prepare_dot}
# Spatial Coords
spatial_coords <- GetTissueCoordinates(seurat_obj)[colnames(spatial_counts_final), c("x", "y")]

# Input Matrices for DOT
# DOT can work with raw counts or normalized. Given we just aggregated raw counts,
# we should verify if DOT requires normalization. The paper suggests normalized/scaled data 
# might be better for feature transfer, but let's stick to the matrix we have match.
# Typically, we pass the matched expression matrices.

spatial_mat <- as.matrix(spatial_counts_final)
sc_mat <- as.matrix(sc_counts_final)

# SC Labels (No Subclustering)
sc_labels <- as.character(Idents(seurat_ref))
names(sc_labels) <- colnames(seurat_ref)
# Filter sc_labels to match the cells in sc_mat (which might have changed order or been filtered if any dropped)
# Note: aggregate.Matrix aggregates ROWS (genes), columns (cells) should remain, but let's be safe.
sc_labels <- sc_labels[colnames(sc_mat)]

print("Data ready for DOT.")
```

# 5. Run DOT

```{r run_dot}
print("Running DOT...")

tryCatch({
    # Correct workflow for saezlab/DOT (DOTr package)
    
    print("Pre-processing Reference (Single-Cell)...")
    # sc_mat: genes x cells, sc_labels: character vector of labels
    prep_sc <- DOTr::setup.ref(
        ref_data = sc_mat, 
        ref_annotations = sc_labels
    )
    
    print("Pre-processing Target (Spatial)...")
    # spatial_mat: genes x spots, spatial_coords: spots x 2
    prep_srt <- DOTr::setup.srt(
        srt_data = spatial_mat, 
        srt_coords = spatial_coords,
        verbose = TRUE
    )
    
    print("Creating DOT Object...")
    dot_obj <- DOTr::create.DOT(ref = prep_sc,
        srt = prep_srt
        
    )
    
    print("Running DOT Algorithm (Frank-Wolfe)...")
    dot_results <- DOTr::run.DOT.highresolution(
        object = dot_obj
    )
    
    # Define output path
    output_dir <- "/home/mattia/Thesis/111125ST_def/Random data"
    
    saveRDS(dot_results, file.path(output_dir, "dot_results_exact_match.rds"))
    print(paste("DOT Results saved to:", output_dir))
    
    # Also save the updated Seurat object with labels
    # We do the argmax in the next step, but let's save the progress
    saveRDS(seurat_obj, file.path(output_dir, "seurat_obj_dot.rds"))
    
    print("DOT Complete.")
}, error = function(e) {
    print("DOT Failed:")
    print(e)
})
```

# 6. Visualization: Most Likely Cell Type

```{r plot_max}
if (exists("dot_results")) {
    print("Visualizing Results...")
    
    # Debug: Check slot names if any error occurs
    # print(slotNames(dot_results))
    
    # DOT results are S4 objects; use @ to access slots
    weights <- dot_results@weights 
    
    # 1. Find Max Cell Type per Spot
    # argmax for each row (spot)
    max_col_index <- max.col(weights, ties.method = "first")
    max_cell_type <- colnames(weights)[max_col_index]
    
    # 2. Add to Metadata
    seurat_obj <- AddMetaData(seurat_obj, metadata = max_cell_type, col.name = "Most_Likely_DOT")
    
    # 3. Plot (Tiled Version: No gaps between spots)
    p <- SpatialDimPlot(seurat_obj, 
                        group.by = "Most_Likely_DOT",
                        image.alpha = 0,        
                        pt.size.factor = 4.5,   # High value for perfect tiling/continuous palette
                        stroke = 0) +           
         ggtitle("Most Likely Cell Type (DOT Deconvolution)") +
         theme_minimal() + 
         theme(panel.grid = element_blank(),
               axis.text = element_blank(),
               axis.ticks = element_blank(),
               axis.title = element_blank(),
               legend.position = "right")
    print(p)
}
```

# 6.5 Interactive ROI Selection (Shiny Lasso)
Run this chunk to manually select a region of interest (ROI) using the lasso tool. This will filter the `seurat_obj` for all subsequent steps.

```{r shiny_lasso, eval=FALSE}
library(shiny)
library(plotly)

# 1. Prepare data for plotting
coords <- GetTissueCoordinates(seurat_obj)
# Detect coordinate columns (usually col 1 and 2, but let's be safe)
coord_cols <- colnames(coords)[1:2]
print(paste("Detected coordinate columns:", paste(coord_cols, collapse = ", ")))

plot_df <- as.data.frame(coords)
plot_df$CellType <- seurat_obj$Most_Likely_DOT
plot_df$Barcode <- rownames(plot_df)

# UI
ui <- fluidPage(
    titlePanel("Interactive Spatial Selector (Lasso)"),
    sidebarLayout(
        sidebarPanel(
            helpText("Use the Lasso tool on the plot to select spots to KEEP."),
            actionButton("done", "Apply Selection & Exit", class = "btn-success"),
            hr(),
            verbatimTextOutput("selection_status")
        ),
        mainPanel(
            plotlyOutput("spatial_plot", height = "700px")
        )
    )
)

# Server
server <- function(input, output, session) {
    # Store selected barcodes
    selected_barcodes <- reactiveVal(NULL)
    
    output$spatial_plot <- renderPlotly({
        # Dynamically map the detected columns to x and y
        p <- plot_ly(plot_df, 
                x = as.formula(paste0("~", coord_cols[1])), 
                y = as.formula(paste0("~", coord_cols[2])), 
                color = ~CellType,
                key = ~Barcode,
                type = 'scatter', 
                mode = 'markers',
                marker = list(size = 5)) %>%
            layout(dragmode = "lasso",
                   yaxis = list(autorange = "reversed"),
                   xaxis = list(scaleanchor = "y", scaleratio = 1))
        p
    })
    
    observeEvent(event_data("plotly_selected"), {
        d <- event_data("plotly_selected")
        if (!is.null(d)) {
            selected_barcodes(d$key)
        }
    })
    
    output$selection_status <- renderText({
        if (is.null(selected_barcodes())) {
            "No spots selected yet."
        } else {
            paste("Selected spots:", length(selected_barcodes()))
        }
    })
    
    observeEvent(input$done, {
        if (!is.null(selected_barcodes())) {
            # Update the global seurat_obj
            seurat_obj <<- subset(seurat_obj, cells = selected_barcodes())
            print(paste("Seurat object subsetted to", ncol(seurat_obj), "spots."))
        } else {
            print("No selection made. Seurat object remains unchanged.")
        }
        stopApp()
    })
}

# Run the app
runGadget(ui, server)
```

# 7. Post-Deconvolution Filtering (unk > 5%)

```{r filtering_comparison}
if (exists("dot_results")) {
    print("Filtering spots with >5% unk proportion...")
    
    # 1. Extract 'unk' proportion
    weights_mat <- dot_results@weights
    if ("unk" %in% colnames(weights_mat)) {
        unk_prop <- weights_mat[, "unk"]
        seurat_obj <- AddMetaData(seurat_obj, metadata = unk_prop, col.name = "unk_proportion")
        
        # 2. Before Plot
        p_before <- SpatialDimPlot(seurat_obj, group.by = "Most_Likely_DOT") + 
                    ggtitle("Before Filtering (All Spots)") + 
                    theme(legend.position = "right")
        
        # 3. Filter the object
        seurat_obj_filtered <- subset(seurat_obj, subset = unk_proportion <= 0.05)
        
        # 4. After Plot
        p_after <- SpatialDimPlot(seurat_obj_filtered, group.by = "Most_Likely_DOT") + 
                   ggtitle("After Filtering (unk <= 5%)") + 
                   theme(legend.position = "right")
        
        # 5. Display side-by-side (using cowplot or patchwork if available, else sequential)
        # We'll use wrap_plots or just print them if patchwork isn't confirmed.
        # But most Seurat environments have it.
        if (requireNamespace("patchwork", quietly = TRUE)) {
            library(patchwork)
            print(p_before + p_after)
        } else {
            print(p_before)
            print(p_after)
        }
        
        # 6. Save filtered object
        output_dir <- "/home/mattia/Thesis/111125ST_def/Random data"
        saveRDS(seurat_obj_filtered, file.path(output_dir, "seurat_obj_dot_filtered.rds"))
        print(paste("Filtered Seurat object saved to:", output_dir))
        
        # 7. CRITICAL: Overwrite main objects to ensure Section 8 uses filtered data
        # Syncing Seurat object, weights matrix, and spatial coordinates
        kept_barcodes <- colnames(seurat_obj_filtered)
        seurat_obj <<- seurat_obj_filtered
        dot_results@weights <<- dot_results@weights[kept_barcodes, ]
        if (exists("prep_srt") && "C" %in% names(prep_srt)) {
            prep_srt$C <<- prep_srt$C[kept_barcodes, ]
        }
        
    } else {
        print("Column 'unk' not found in weights results. Skipping filtering.")
    }
}
```

```{r filter_sync}
# 7.5 Synchronization & Output
# Ensures all downstream steps use ONLY the filtered spots.

if (exists("seurat_obj")) {
    kept_barcodes <- colnames(seurat_obj)
    print(paste("Finalizing filter: Keeping", length(kept_barcodes), "spots."))
    
    # Update Weights Matrix
    if (exists("dot_results")) {
        dot_results@weights <- dot_results@weights[kept_barcodes, , drop = FALSE]
    }
    
    # Update Preparation Object (for draw_maps)
    if (exists("prep_srt")) {
        prep_srt$C <- prep_srt$C[kept_barcodes, , drop = FALSE]
        prep_srt$X <- prep_srt$X[kept_barcodes, , drop = FALSE]
    }
    
    # Update original coordinates if used
    if (exists("spatial_coords")) {
        spatial_coords <- spatial_coords[kept_barcodes, , drop = FALSE]
    }
}
```

# 8. Advanced Visualizations (Paper Recommendations)

## 8.1 Spatial Abundance Maps (`draw_maps`)
The built-in `DOTr::draw_maps()` is optimized for the DOT object structure.

```{r draw_maps}
if (exists("dot_results") && exists("prep_srt")) {
    print("Generating DOTr::draw_maps (All Cell Types)...")
    
    # source data from synchronized objects
    weights_mat <- dot_results@weights
    coords_mat <- prep_srt$C
    
    # Final check: intersection of barcodes
    common_barcodes <- intersect(rownames(weights_mat), rownames(coords_mat))
    weights_mat <- weights_mat[common_barcodes, , drop = FALSE]
    coords_mat <- coords_mat[common_barcodes, , drop = FALSE]
    
    # Use all cell type columns instead of just the top 4
    all_cts <- colnames(weights_mat)
    
    tryCatch({
        # Using the signature provided by the user
        DOTr::draw_maps(
            spatial = coords_mat, 
            weights = weights_mat[, all_cts, drop = FALSE],
            normalize = TRUE,
            ncol = 4, # Increased ncol to accommodate more maps
            trans = "identity",
            size = 0.5,
            shape = 19,
            flip_y = TRUE,
            viridis_option = "magma",
            legend_title = "abundance",
            background = "gray90"
        )
    }, error = function(e) {
        print("draw_maps failed with list input. Checking coordinates directly...")
        print(e)
    })
}
```

## 8.2 Uncertainty Mapping (Entropy)
Identifies spots where deconvolution is "uncertain" (high entropy).

```{r uncertainty}
if (exists("dot_results")) {
    print("Calculating Deconvolution Uncertainty...")
    
    # Get weights matrix
    w <- dot_results@weights
    
    # Calculate Shannon Entropy: -sum(p * log(p))
    # Add a small epsilon to avoid log(0)
    calc_entropy <- function(p) {
        p <- p[p > 0]
        -sum(p * log(p))
    }
    
    entropy_scores <- apply(w, 1, calc_entropy)
    
    # Add to Seurat object
    seurat_obj <- AddMetaData(seurat_obj, metadata = entropy_scores, col.name = "DOT_Entropy")
    
    # Plot
    p <- SpatialFeaturePlot(seurat_obj, features = "DOT_Entropy") + 
        ggtitle("DOT Deconvolution Uncertainty (Entropy)") +
        scale_fill_viridis_c(option = "magma")
    print(p)
}
```

## 8.3 Spatial Correlation Analysis
Correlation between DOT scores and spatial QC features.

```{r spatial_cor}
if (exists("dot_results")) {
    print("Running Spatial Correlation Analysis...")
    
    # 1. Ensure QC metrics are present (Recalculate or check)
    if (!"mean_expression" %in% colnames(seurat_obj@meta.data)) {
        normalized_data <- GetAssayData(seurat_obj, assay = "SCT", layer = "data")
        seurat_obj$mean_expression <- colMeans(normalized_data)
    }
    
    if (!"percent.mt" %in% colnames(seurat_obj@meta.data)) {
        seurat_obj[["percent.mt"]] <- PercentageFeatureSet(seurat_obj, pattern = "^mt:")
    }
    
    if (!"percent.ribo" %in% colnames(seurat_obj@meta.data)) {
        seurat_obj[["percent.ribo"]] <- PercentageFeatureSet(seurat_obj, pattern = "^Rp[LS]")
    }

    # 2. Add DOT weights to metadata for correlation
    weights_mat <- dot_results@weights
    seurat_obj <- AddMetaData(seurat_obj, metadata = as.data.frame(weights_mat))
    
    # 3. Perform Correlation
    meta <- seurat_obj@meta.data
    dot_cols <- colnames(weights_mat)
    qc_cols <- c("mean_expression", "percent.mt", "percent.ribo")
    
    # Filter for existing QC cols
    qc_cols <- qc_cols[qc_cols %in% colnames(meta)]
    
    if (length(dot_cols) > 0 && length(qc_cols) > 0) {
        cor_mat <- cor(meta[, dot_cols], meta[, qc_cols], use = "complete.obs")
        
        library(corrplot)
        # cl.pos = 'n' removes the color legend labels (numbers on the right)
        corrplot(cor_mat, method = "circle", 
                 title = "Correlation: DOT vs Spatial QC", 
                 mar = c(0,0,1,0),
                 cl.pos = "n")
    }
}
```

# 9. Sub-clustering: Generating High-Resolution Reference Labels
Objective: Split each cell type into 3 subclusters to resolve finer spatial patterns.

```{r subclustering}
if (exists("seurat_ref")) {
    print("Starting sub-clustering of reference labels (k=3)...")
    
    # Identify the metadata column for cell types 
    ct_col <- "cell_type" 
    if (!ct_col %in% colnames(seurat_ref@meta.data)) {
        ct_col <- "celltype"
    }
    
    cell_types <- unique(seurat_ref[[ct_col, drop = TRUE]])
    new_subcluster_labels <- character(ncol(seurat_ref))
    names(new_subcluster_labels) <- colnames(seurat_ref)
    
    for (ct in cell_types) {
        print(paste("Sub-clustering:", ct))
        
        # Subset to specific cell type
        cells_of_type <- colnames(seurat_ref)[seurat_ref[[ct_col, drop = TRUE]] == ct]
        subset_obj <- subset(seurat_ref, cells = cells_of_type)
        
        # Pipelines for Variable Features and PCA
        try({
            # Ensure normalization for the subset (fixes ScaleData error)
            subset_obj <- NormalizeData(subset_obj, verbose = FALSE)
            subset_obj <- FindVariableFeatures(subset_obj, selection.method = "vst", nfeatures = 1000, verbose = FALSE)
            subset_obj <- ScaleData(subset_obj, verbose = FALSE)
            subset_obj <- RunPCA(subset_obj, npcs = 10, verbose = FALSE)
            
            # K-means clustering (exactly 3)
            pca_data <- Embeddings(subset_obj, "pca")
            if (nrow(pca_data) >= 3) {
                km_res <- kmeans(pca_data, centers = 3, nstart = 25)
                new_labels <- paste0(ct, "_", km_res$cluster)
            } else {
                new_labels <- paste0(ct, "_1")
            }
            new_subcluster_labels[colnames(subset_obj)] <- new_labels
        }, silent = FALSE)
    }
    
    # Add to main reference object
    seurat_ref$Subcluster_Labels <- new_subcluster_labels
    
    # Verify
    print(table(seurat_ref$Subcluster_Labels))
}
```

# 10. Subcluster Deconvolution: High-Resolution DOT
Re-running the DOT algorithm using the new high-resolution labels.

```{r dot_subclusters}
if (exists("seurat_ref") && "Subcluster_Labels" %in% colnames(seurat_ref@meta.data)) {
    print("Preparing data for Subcluster Deconvolution...")
    
    # 1. Get counts
    spatial_counts_sub <- GetAssayData(seurat_obj, assay = "Spatial.016um", layer = "counts")
    sc_counts_sub <- GetAssayData(seurat_ref, assay = "RNA", layer = "counts")
    
    # 2. Gene Cleaning
    spatial_genes_clean_sub <- toupper(gsub("\\..*$", "", rownames(spatial_counts_sub)))
    sc_genes_clean_sub <- toupper(gsub("\\..*$", "", rownames(sc_counts_sub)))
    
    rownames(spatial_counts_sub) <- spatial_genes_clean_sub
    rownames(sc_counts_sub) <- sc_genes_clean_sub
    
    # 3. Aggressive Aggregation (dMeltAndSum logic)
    print("Aggregating duplicates (Subclusters)...")
    spatial_counts_agg_sub <- Matrix.utils::aggregate.Matrix(
        spatial_counts_sub, 
        group = rownames(spatial_counts_sub), 
        fun = "sum"
    )
    
    sc_counts_agg_sub <- Matrix.utils::aggregate.Matrix(
        sc_counts_sub, 
        group = rownames(sc_counts_sub), 
        fun = "sum"
    )
    
    # 4. Final Intersection
    common_genes_sub <- intersect(rownames(spatial_counts_agg_sub), rownames(sc_counts_agg_sub))
    spatial_mat_sub <- as.matrix(spatial_counts_agg_sub[common_genes_sub, ])
    sc_mat_sub <- as.matrix(sc_counts_agg_sub[common_genes_sub, ])
    
    # 5. Prepare Labels
    sc_labels_sub <- seurat_ref$Subcluster_Labels
    names(sc_labels_sub) <- colnames(seurat_ref)
    sc_labels_sub <- sc_labels_sub[colnames(sc_mat_sub)]
    
    # 6. Prepare Coords
    spatial_coords_sub <- GetTissueCoordinates(seurat_obj)[colnames(spatial_mat_sub), c("x", "y")]
    
    # 2. DOTr Workflow
    print("Pre-processing High-Res Reference...")
    # Using setup.ref on the matched matrix
    prep_sc_sub <- DOTr::setup.ref(
        ref_data = sc_mat_sub, 
        ref_annotations = sc_labels_sub
    )
    
    print("Pre-processing High-Res Target...")
    # prep_srt_sub is derived from the sub-matched spatial_mat_sub
    prep_srt_sub <- DOTr::setup.srt(
        srt_data = spatial_mat_sub, 
        srt_coords = spatial_coords_sub,
        verbose = TRUE
    )
    
    print("Creating DOT Object (Subcl.)...")
    dot_obj_sub <- DOTr::create.DOT(
        ref = prep_sc_sub,
        srt = prep_srt_sub
    )
    
    print("Running High-Resolution DOT (Subclusters)...")
    # Frank-Wolfe optimization at subcluster resolution
    dot_results_sub <- DOTr::run.DOT.highresolution(
        object = dot_obj_sub
    )
    
    # Save Subcluster Results
    output_dir <- "/home/mattia/Thesis/111125ST_def/Random data"
    saveRDS(dot_results_sub, file.path(output_dir, "dot_results_subclusters.rds"))
    print("Subcluster deconvolution complete.")
}
```

# 11. Subcluster Filtering: ROI and Proportion-based QC
Objective: Filter the high-resolution results to remove noisy spots.

```{r subcluster_filtering}
if (exists("dot_results_sub")) {
    print("Applying filtering to Subcluster Deconvolution...")
    
    # 1. Identify Most Likely Subcluster
    weights_sub <- dot_results_sub@weights
    if (!is.matrix(weights_sub)) weights_sub <- as.matrix(weights_sub)
    
    # Argmax
    max_idx <- max.col(weights_sub, ties.method = "first")
    max_subcluster_vals <- colnames(weights_sub)[max_idx]
    names(max_subcluster_vals) <- rownames(weights_sub)
    
    # 2. Sync with ROI
    common_spots <- intersect(colnames(seurat_obj), rownames(weights_sub))
    print(paste("Common Spots (ROI & DOT):", length(common_spots)))
    
    if (length(common_spots) == 0) {
        stop("No common spots found.")
    }
    
    seurat_sub <- subset(seurat_obj, cells = common_spots)
    
    # Add labels (Categorical)
    # Ensure we only pick labels for the spots we have
    matched_labels <- max_subcluster_vals[colnames(seurat_sub)]
    seurat_sub$Subcluster_Label_Final <- as.character(matched_labels)
    
    # 3. unk Filtering (>5%)
    if ("unk" %in% colnames(weights_sub)) {
        unk_vec <- weights_sub[colnames(seurat_sub), "unk"]
        seurat_sub$unk_prop_sub <- unk_vec
        seurat_sub_filtered <- subset(seurat_sub, subset = unk_prop_sub < 0.05)
    } else {
        seurat_sub_filtered <- seurat_sub
    }
    
    # 4. Cleanup & Diagnostics
    print(paste("Final Spots for Subcluster Plotting:", ncol(seurat_sub_filtered)))
    
    if (ncol(seurat_sub_filtered) > 0) {
        # Force discrete factors and drop NAs
        final_vals <- seurat_sub_filtered$Subcluster_Label_Final
        # Remove NA/Empty
        valid_spots <- !is.na(final_vals) & final_vals != "" & final_vals != "unk"
        seurat_sub_filtered <- subset(seurat_sub_filtered, cells = colnames(seurat_sub_filtered)[valid_spots])
        
        # FINAL CATEGORICAL ENFORCEMENT
        seurat_sub_filtered$Subcluster_Label_Final <- factor(seurat_sub_filtered$Subcluster_Label_Final)
        
        print("Final Categories identified:")
        print(table(seurat_sub_filtered$Subcluster_Label_Final))
    }
    
    # Save objects
    output_dir <- "/home/mattia/Thesis/111125ST_def/Random data"
    saveRDS(seurat_sub_filtered, file.path(output_dir, "seurat_obj_subclusters_filtered.rds"))
    
    # Final weights sync for draw_maps
    kept_barcodes_sub <- colnames(seurat_sub_filtered)
    weights_sub_filtered <- weights_sub[kept_barcodes_sub, , drop = FALSE]
}
```

# 12. Subcluster Visualization (High-Resolution Atlas)
Objective: Generate the final continuous palette and uncertainty maps for subclusters.

```{r subcluster_plots}
if (exists("seurat_sub_filtered") && ncol(seurat_sub_filtered) > 0) {
    print("Generating Subcluster Atlas...")
    
    # Final Discrete Check
    if (is.factor(seurat_sub_filtered$Subcluster_Label_Final) && 
        length(levels(seurat_sub_filtered$Subcluster_Label_Final)) > 0) {
        
        # 1. Tiled Most Likely Subcluster
        p1 <- SpatialDimPlot(seurat_sub_filtered, 
                            group.by = "Subcluster_Label_Final",
                            image.alpha = 0,        
                            pt.size.factor = 4.5,   
                            stroke = 0) +           
             ggtitle("High-Resolution Atlas: Subclusters") +
             theme_void() + 
             theme(legend.position = "right")
        
        print(p1)
    } else {
        cat("Error: Cannot plot. Subcluster_Label_Final has no valid category levels.\n")
    }
    
    # 2. Detailed Abundance Maps (draw_maps)
    # Re-extract coordinates for filtered spots
    # CLEANING: Ensure only x and y columns are passed to avoid "discrete value" errors with barcodes
    coords_sub <- GetTissueCoordinates(seurat_sub_filtered)
    # Force numeric and keep only x/y
    coords_sub_clean <- as.matrix(coords_sub[, c(1, 2)])
    
    # ENSURE WEIGHTS ARE NUMERIC
    weights_numeric <- as.matrix(weights_sub_filtered)
    # Remove 'unk' from abundance maps if present (optional but recommended for cleaner maps)
    map_cols <- setdiff(colnames(weights_numeric), "unk")
    
    # BATCHING: Plot 9 subclusters per "window" (batch)
    n_cols <- length(map_cols)
    batch_size <- 9
    n_batches <- ceiling(n_cols / batch_size)
    
    print(paste("Plotting", n_cols, "subcluster maps in", n_batches, "batches of 9..."))
    
    for (i in 1:n_batches) {
        start_idx <- (i - 1) * batch_size + 1
        end_idx <- min(i * batch_size, n_cols)
        current_cols <- map_cols[start_idx:end_idx]
        
        print(paste("Batch", i, ": columns", start_idx, "to", end_idx))
        
        tryCatch({
            # Capture and print since it's inside a loop
            p_batch <- DOTr::draw_maps(
                spatial = coords_sub_clean, 
                weights = weights_numeric[, current_cols, drop = FALSE],
                normalize = TRUE,
                ncol = 3, # 3x3 grid for batch of 9
                flip_y = TRUE,
                viridis_option = "magma",
                background = "gray90",
                size = 0.5,
                shape = 19
            )
            print(p_batch)
        }, error = function(e) {
            print(paste("Batch", i, "failed:"))
            print(e)
        })
    }
}
```
