---
title: "EE Regional Correlation Analysis"
output: html_document
---

```{r setup, include=FALSE}
library(Seurat)
library(ggplot2)
library(dplyr)
library(Matrix)
library(Matrix.utils)
library(pheatmap)
library(patchwork)

# Paths
sc_data_path <- "/home/mattia/Thesis/111125ST_def/Random data/EE_with_spatial_scores.rds"
sc_ref_path <- "/home/mattia/Thesis/Single-cell/EE-single-cell/GSE132274_RAW(1)"
```

## 1. Load and Harmonize Gene Names

We must clean the gene names (remove version decimals, uppercase) and aggregate duplicates before proceeding to ensure markers are found.

```{r load_and_clean}
# 1. Load Spatial EE Data
if (!file.exists(sc_data_path)) stop("Spatial EE data not found.")
sc_data <- readRDS(sc_data_path)

# 2. Load Reference EE Data
counts <- Read10X(data.dir = sc_ref_path)

# --- ROBUST GENE CLEANING ---
clean_genes <- function(counts_matrix) {
  raw_names <- rownames(counts_matrix)
  # Remove version suffixes like .1, .2 and convert to uppercase
  clean_names <- toupper(gsub("\\..*$", "", raw_names))
  
  # Aggregate duplicates by summing counts
  if (any(duplicated(clean_names))) {
    message("Aggregating duplicated gene names...")
    counts_matrix <- Matrix.utils::aggregate.Matrix(
      counts_matrix, 
      group = clean_names, 
      fun = "sum"
    )
  } else {
    rownames(counts_matrix) <- clean_names
  }
  return(counts_matrix)
}

# Clean both
sc_ref_counts_clean <- clean_genes(counts)
sc_data_counts_clean <- clean_genes(GetAssayData(sc_data, assay = "SCT", layer = "counts"))

# Re-initialize Seurat objects with clean names
sc_ref <- CreateSeuratObject(counts = sc_ref_counts_clean)
sc_data_clean <- CreateSeuratObject(counts = sc_data_counts_clean)

# Transmit metadata back to sc_data_clean
sc_data_clean@meta.data <- sc_data@meta.data
sc_data <- sc_data_clean

# Normalize sc_data since it is now a fresh object with "RNA" assay
sc_data <- NormalizeData(sc_data)

message("Gene name harmonization and normalization complete.")
```

## 2. Process Reference and Map Clusters to Regions

```{r process_ref}
# 2. Reference Processing (t-SNE at Res 0.4)
sc_ref <- NormalizeData(sc_ref)
sc_ref <- FindVariableFeatures(sc_ref, nfeatures = 2000)
sc_ref <- ScaleData(sc_ref)
sc_ref <- RunPCA(sc_ref, verbose = FALSE)
sc_ref <- FindNeighbors(sc_ref, dims = 1:20)
sc_ref <- FindClusters(sc_ref, resolution = 0.4) # Requested resolution
sc_ref <- RunTSNE(sc_ref, dims = 1:20) # Use t-SNE

# --- FOOLPROOF MANUAL MAPPING ---
# Logic: Cluster IDs are strings. We map them directly to regions.
cluster_ids <- as.character(Idents(sc_ref))
regions_vec <- rep("Unknown", length(cluster_ids))

# According to the image mapping:
regions_vec[cluster_ids %in% c("2", "3", "8", "9", "10")] <- "Anterior"
regions_vec[cluster_ids %in% c("0", "1", "4")] <- "Middle"
regions_vec[cluster_ids %in% c("5", "6", "7")] <- "Posterior"

# Assign back to the object metadata
sc_ref[["Region"]] <- factor(regions_vec, levels = c("Anterior", "Middle", "Posterior", "Unknown"))

# PLOT: Verify results
p1 <- DimPlot(sc_ref, reduction = "tsne", label = TRUE) + NoLegend() + ggtitle("Reference t-SNE (Res 0.4)")
p2 <- DimPlot(sc_ref, group.by = "Region", reduction = "tsne", 
              cols = c("Anterior" = "blue", "Middle" = "green", "Posterior" = "red", "Unknown" = "grey80")) + 
      ggtitle("Manual Regional Assignment")

p1 + p2

# --- DIAGNOSTIC PRINT ---
message("Manual Cluster Assignments Summary:")
print(table(sc_ref$Region, Idents(sc_ref)))
```

## 3. Verify Spatial Data

Using the `region_prediction` (R1-R5) labels for the spatial datasets.

```{r verify_spatial}
if ("region_prediction" %in% colnames(sc_data@meta.data)) {
  message("Collapsing spatial data regions:")
  # Group R1+R2 and R4+R5
  sc_data$Spatial_Region <- case_when(
    sc_data$region_prediction %in% c("R1", "R2") ~ "R1-R2",
    sc_data$region_prediction == "R3" ~ "R3",
    sc_data$region_prediction %in% c("R4", "R5") ~ "R4-R5",
    TRUE ~ "Other"
  )
  
  sc_data$Spatial_Region <- factor(sc_data$Spatial_Region, levels = c("R1-R2", "R3", "R4-R5"))
  print(table(sc_data$Spatial_Region))
} else {
  stop("Column 'region_prediction' not found in spatial data.")
}
```

## 4. Region-wise Correlation Analysis

```{r correlation}
# 1. Calculate average expression per region for common genes
common_genes <- intersect(rownames(sc_data), rownames(sc_ref))
message("Number of common genes for correlation: ", length(common_genes))

# We use the factor levels we established earlier
avg_spatial <- AverageExpression(sc_data, features = common_genes, group.by = "Spatial_Region", assay = "RNA")$RNA
avg_ref <- AverageExpression(sc_ref, features = common_genes, group.by = "Region", assay = "RNA")$RNA

# Ensure we only use the three biological regions (Anterior, Middle, Posterior) for the comparison
# even if some clusters were 'Unknown'
biological_regions <- c("Anterior", "Middle", "Posterior")
avg_ref_sub <- avg_ref[, intersect(colnames(avg_ref), biological_regions), drop=FALSE]

# 2. Compute Spearman Correlation Matrix
cor_matrix <- cor(as.matrix(avg_spatial), as.matrix(avg_ref_sub), method = "spearman")

# 3. Visualization
library(pheatmap)
pheatmap(cor_matrix, 
         cluster_rows = FALSE, 
         cluster_cols = FALSE,
         display_numbers = TRUE,
         main = "Region-wise Correlation (Collapsed Spatial vs Ref Regions)",
         color = colorRampPalette(c("blue", "white", "red"))(100),
         labels_row = paste0("Spatial: ", rownames(cor_matrix)),
         labels_col = paste0("Ref: ", colnames(cor_matrix)))
```
## 5. Marker Visualization on Reference t-SNE

Visualizing the 56 EE spatial markers in batches of 6. We also include TK (Anterior marker) and AstC (Posterior/Middle marker) for regional context.

```{r marker_visualization}
# 1. Get the list of 56 EE markers from the research
ee_markers_file <- "/home/mattia/Thesis/111125ST_def/Random data/celltype_specific_spatial_EE.csv"
ee_markers_df <- read.csv(ee_markers_file)
ee_markers <- ee_markers_df$gene

# 2. Filter for genes present in reference
ee_markers_present <- intersect(ee_markers, rownames(sc_ref))
message("Found ", length(ee_markers_present), " out of ", length(ee_markers), " markers in reference data.")

# 3. Add Key Regional Markers for Reference
key_ref_markers <- c("TK", "ASTC")
all_to_plot <- unique(c(key_ref_markers, ee_markers_present))

# 4. Batch Plotting (6 per window)
batch_size <- 6
num_batches <- ceiling(length(all_to_plot) / batch_size)

for (i in 1:num_batches) {
  start_idx <- (i - 1) * batch_size + 1
  end_idx <- min(i * batch_size, length(all_to_plot))
  current_batch <- all_to_plot[start_idx:end_idx]
  
  p <- FeaturePlot(sc_ref, 
                   features = current_batch, 
                   reduction = "tsne", 
                   cols = c("grey90", "red"),
                   ncol = 3) + 
       plot_annotation(title = paste0("EE Spatial Markers - Batch ", i))
  
  print(p)
}

# 5. Regional Enrichment Logic
# We can determine if a gene is in an 'Anterior' or 'Posterior' cluster 
# by looking at overlap with TK (Anterior) vs ASTC (Posterior/Middle).
avg_expr_by_cluster <- AverageExpression(sc_ref, features = all_to_plot, group.by = "Region", assay = "RNA")$RNA

message("\n--- Regional Enrichment Summary for Top Markers ---")
# Print a simple logic check for the top 10 markers
top_10_markers <- head(ee_markers_present, 10)
for (m in top_10_markers) {
  exprs <- avg_expr_by_cluster[m, ]
  peak_region <- names(which.max(exprs))
  message(m, " peaks in: ", peak_region)
}
```
