---
title: "LFC Sub-Cluster Analysis & Similarity Mapping"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(Seurat)
library(dplyr)
library(ggplot2)
library(plotly)
library(htmlwidgets)
library(pheatmap)
```

## 1. Load Data
```{r load_data}
path_sc <- "../Random data/sc_with_spatial_scores.rds"
if(!file.exists(path_sc)) stop("Cannot locate single-cell object at ", path_sc)

cat("Loading Single Cell Object...\n")
sc_obj <- readRDS(path_sc)

# Verify spatial data is present
if (!"Smoothed_Pseudotime" %in% colnames(sc_obj@meta.data)) {
  warning("Smoothed_Pseudotime not found in metadata. Defaulting to Transferred_Pseudotime if available.")
  if ("Transferred_Pseudotime" %in% colnames(sc_obj@meta.data)) {
    sc_obj$Smoothed_Pseudotime <- sc_obj$Transferred_Pseudotime
  } else {
    stop("No spatial pseudotime found in the Seurat object.")
  }
}

# Subset for LFCs purely based on prior cell classifications
# Using the user-confirmed 'celltype' metadata column
lfc_cells <- subset(sc_obj, subset = celltype == "LFC")
cat("Isolated", ncol(lfc_cells), "LFC cells out of", ncol(sc_obj), "total cells.\n")
```

## 2. Re-Clustering & 3D UMAP Generation
We re-process the LFC subset uniquely and generate a 3-dimensional UMAP to visualize the structural divide between the distinct populations.

```{r clustering_3d}
# Instead of re-running PCA and UMAP, we retain the global transcription space
# inherited from the full single-cell dataset. This ensures clustering uses the global context.

# Cluster directly on the global PCA space. 
# Increase k.param slightly and lower resolution to limit to ~3 clusters
lfc_cells <- FindNeighbors(lfc_cells, reduction = "pca", dims = 1:20, k.param = 15)

# Moderate resolution to capture the main splits (approx 3 clusters)
lfc_cells <- FindClusters(lfc_cells, resolution = 0.5)

# Extract rendering coordinates and inject Space as the 3rd Dimension
umap_coords <- as.data.frame(Embeddings(lfc_cells, "umap"))
umap_coords$Space <- lfc_cells$Smoothed_Pseudotime
umap_coords$Cluster <- as.character(lfc_cells$seurat_clusters)

# Define global output directory constraints
dir.create("../figures/LFC_Analysis", recursive=TRUE, showWarnings=FALSE)

cat("✓ Re-clustering complete. Defined", length(unique(lfc_cells$seurat_clusters)), "clusters natively.\n")

# 3D Plot for Cluster Imputation (2 Genetic + 1 Spatial Phase)
p_3d_cluster <- plot_ly(data = umap_coords, 
                        x = ~umap_1, y = ~umap_2, z = ~Space, 
                        color = ~Cluster, colors = c("#E64B35", "#4DBBD5", "#00A087", "#3C5488", "#F39B7F", "#8491B4", "#91D1C2"),
                        type = "scatter3d", mode = "markers",
                        marker = list(size = 3, opacity = 0.8)) %>%
    layout(title = "LFC Sub-Clusters (X/Y: Genetic, Z: Space)",
           scene = list(xaxis = list(title = "UMAP_1 (Genetic)"),
                        yaxis = list(title = "UMAP_2 (Genetic)"),
                        zaxis = list(title = "Smoothed Pseudotime (Space)")))

htmlwidgets::saveWidget(p_3d_cluster, file = "../figures/LFC_Analysis/3D_UMAP_LFC_Clusters.html")
p_3d_cluster
```

```{r analysis}
## 3. 3D UMAP Feature Projections
# Visualizing top driving phenotypic parameters (such as structural layout spatiality) directly across the 3D space.

p_3d_feature <- plot_ly(data = umap_coords, 
                        x = ~umap_1, y = ~umap_2, z = ~Space, 
                        color = ~Space, colors = colorRamp(c("navy", "yellow", "firebrick3")),
                        type = "scatter3d", mode = "markers",
                        marker = list(size = 3, opacity = 0.8)) %>%
    layout(title = "LFC Spatial Topography (X/Y: Genetic, Z: Space)",
           scene = list(xaxis = list(title = "UMAP_1 (Genetic)"),
                        yaxis = list(title = "UMAP_2 (Genetic)"),
                        zaxis = list(title = "Smoothed Pseudotime (Space)")))

htmlwidgets::saveWidget(p_3d_feature, file = "../figures/LFC_Analysis/3D_Hybrid_LFC_Pseudotime.html")
p_3d_feature
```

## 4. Differential Gene Expression (DGE) 
Identifying the unique transcription boundaries pushing the limits of the detected pools.

```{r dge_analysis}
Idents(lfc_cells) <- "seurat_clusters"
cat("Running FindAllMarkers to isolate distinct expression limits...\n")

lfc_markers <- FindAllMarkers(lfc_cells, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
top_lfc <- lfc_markers %>% group_by(cluster) %>% slice_max(n = 15, order_by = avg_log2FC)

write.csv(lfc_markers, "../figures/LFC_Analysis/LFC_Cluster_DGE.csv", row.names=FALSE)

cat("\n--- TOP DIFFERENTIAL MARKERS ---\n")
library(knitr)
# Safely handle potential empty marker results
if(nrow(top_lfc) > 0) {
    print(kable(top_lfc %>% dplyr::select(cluster, gene, avg_log2FC, p_val_adj), format="pipe"))
} else {
    cat("No significant markers found for the current clusters.\n")
}
```

## 5. Global Celltype Similarity Matrix
Computing average expression profiles seamlessly across all functional cell classes inside the original dataset to trace where the specific isolated LFC clusters map.

```{r similarity_matrix, fig.width=10, fig.height=8}
# Re-integrate the newly modeled LFC identities into the master dataset structurally
sc_obj$LFC_Split_Identity <- as.character(sc_obj$celltype)

# Safely swap purely the LFC cells for their new structurally defined IDs
new_ids <- paste0("LFC_Subcluster_", lfc_cells$seurat_clusters)
sc_obj$LFC_Split_Identity[Cells(lfc_cells)] <- new_ids

Idents(sc_obj) <- "LFC_Split_Identity"

cat("Computing holistic AverageExpression arrays across all structural lineages...\n")
# Get current default assay to be safe
def_assay <- DefaultAssay(sc_obj)
avg_exp <- AverageExpression(sc_obj, assays = def_assay)

# Extract the matrix from the list (Seurat V4/V5 returns a list of matrices)
if (is.list(avg_exp)) {
    exp_matrix <- as.matrix(avg_exp[[1]])
} else {
    exp_matrix <- as.matrix(avg_exp)
}

# Ensure there are no NA/Inf values which break correlation
exp_matrix[is.na(exp_matrix)] <- 0

cat("Calculating Pearson correlation matrix on assay:", names(avg_exp)[1], "\n")
sim_matrix <- cor(exp_matrix, method = "pearson")

# Trim generic uninformative or ambiguous arrays before plotting
valid_cols <- !grepl("ambiguous|unknown", colnames(sim_matrix), ignore.case=TRUE)
clean_matrix <- sim_matrix[valid_cols, valid_cols]

png("../figures/LFC_Analysis/LFC_Similarity_Heatmap.png", width=3200, height=2600, res=300)
pheatmap(clean_matrix, 
         color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
         display_numbers = FALSE,
         fontsize_row = 12, fontsize_col = 12,
         main = "Global Transcriptomic Similarity Matrix (Pearson)")
dev.off()

# Print inline for the RMarkdown
pheatmap(clean_matrix, 
         color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
         display_numbers = FALSE,
         fontsize_row = 12, fontsize_col = 12,
         main = "Global Transcriptomic Similarity Matrix (Pearson)")

cat("✓ Execution completed. Matrices stored securely in figures directory.\n")
```

## 5.5 Global 3D Visualization
Visualizing the full single-cell dataset in 3D (UMAP1, UMAP2, and Smoothed_Pseudotime) to see how the local subclusters or specific cell types map globally. 

```{r global_3d_viz}
# We use the global UMAP and Spatial coordinates present in the main sc_obj.
global_coords <- data.frame(
  UMAP1 = Embeddings(sc_obj, "umap")[, 1],
  UMAP2 = Embeddings(sc_obj, "umap")[, 2],
  Space = sc_obj$Smoothed_Pseudotime,
  CellType = as.character(sc_obj$celltype),
  SplitIdentity = sc_obj$LFC_Split_Identity
)

# 1. Global Plot with LFC Subclusters
p_global_split <- plot_ly(data = global_coords, 
                          x = ~UMAP1, y = ~UMAP2, z = ~Space, 
                          color = ~SplitIdentity, 
                          type = "scatter3d", mode = "markers",
                          marker = list(size = 2, opacity = 0.6)) %>%
  layout(title = "Global 3D Space (All Cells + LFC Subclusters)",
         scene = list(xaxis = list(title = "UMAP_1"),
                      yaxis = list(title = "UMAP_2"),
                      zaxis = list(title = "Smoothed Pseudotime")))

htmlwidgets::saveWidget(p_global_split, file = "../figures/LFC_Analysis/3D_Global_LFC_Split.html")
p_global_split

# 2. Custom Plot: All Cells Labeled + LFC Subcluster 0 Re-assigned to 'Iron cells'
# The user wants all cells labeled natively, but the specific LFC subcluster 0 should be renamed to "Iron cells",
# while the rest of the LFC subclusters remain labeled simply as "LFC".
global_coords$Custom_Label <- as.character(global_coords$CellType)

# Find the specific cell barcodes that belong to LFC Subcluster 0
lfc_subcluster_0_cells <- Cells(lfc_cells)[lfc_cells$seurat_clusters == 0]

# Override their label
# We use match() to ensure safe assignment even if ordering differs
global_coords$Custom_Label[rownames(global_coords) %in% lfc_subcluster_0_cells] <- "Iron cells"

p_global_custom <- plot_ly(data = global_coords, 
                          x = ~UMAP1, y = ~UMAP2, z = ~Space, 
                          color = ~Custom_Label, 
                          type = "scatter3d", mode = "markers",
                          marker = list(size = 2, opacity = 0.6)) %>%
  layout(title = "Global 3D Space (LFC Subcluster 0 -> Iron cells)",
         scene = list(xaxis = list(title = "UMAP_1"),
                      yaxis = list(title = "UMAP_2"),
                      zaxis = list(title = "Smoothed Pseudotime")))

htmlwidgets::saveWidget(p_global_custom, file = "../figures/LFC_Analysis/3D_Global_LFC_0_is_Iron.html")
p_global_custom
```

## 6. Gene Ontology (GO)
We run GO Biological Process enrichment on the specific markers of each subcluster.
```{r go_analysis, fig.width=10, fig.height=8}
library(clusterProfiler)
library(org.Dm.eg.db)

# Run GO Enrichment using significant markers only
lfc_markers_sig <- lfc_markers %>% filter(p_val_adj < 0.05)

if (nrow(lfc_markers_sig) > 0) {
  go_results <- compareCluster(
    gene ~ cluster, 
    data = lfc_markers_sig, 
    fun = "enrichGO",
    OrgDb = org.Dm.eg.db,
    keyType = "SYMBOL",
    ont = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.2
  )

  if (!is.null(go_results) && nrow(as.data.frame(go_results)) > 0) {
    # Simplify to reduce redundant terms
    tryCatch({
      go_results_simp <- clusterProfiler::simplify(go_results, cutoff = 0.7, by = "p.adjust", select_fun = min)
      
      p_go <- dotplot(go_results_simp, showCategory = 10) + 
        ggtitle("GO BP Enrichment: LFC Subclusters")
      print(p_go)
    }, error = function(e) {
      message("Could not simplify GO terms. Plotting original results.")
      p_go <- dotplot(go_results, showCategory = 10) + 
        ggtitle("GO BP Enrichment: LFC Subclusters")
      print(p_go)
    })
  } else {
    message("No significant GO terms found for the subclusters.")
  }
} else {
  message("Not enough significant markers for GO analysis.")
}
```

## 7. Global DGE & GO: Iron cells vs All Other Cells
In this section, we compare the newly defined "Iron cells" (LFC Subcluster 0) against all other cells in the dataset to define their global transcriptomic identity.

```{r global_iron_dge}
# Ensure the Custom_Label we created for plotting is added to the Seurat object
sc_obj$Custom_Label <- global_coords$Custom_Label
Idents(sc_obj) <- "Custom_Label"

cat("Running Global DGE: 'Iron cells' vs All Other Cells...\n")
global_iron_markers <- FindMarkers(sc_obj, ident.1 = "Iron cells", only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)

# Add gene names as a column for easier manipulation
global_iron_markers$gene <- rownames(global_iron_markers)

# Filter for significance
global_iron_sig <- global_iron_markers %>% filter(p_val_adj < 0.05)

write.csv(global_iron_sig, "../figures/LFC_Analysis/Global_Iron_Cells_vs_All_Markers.csv", row.names=FALSE)

cat("\n--- TOP GLOBAL MARKERS FOR 'IRON CELLS' ---\n")
if(nrow(global_iron_sig) > 0) {
    top_global_iron <- global_iron_sig %>% top_n(n = 15, wt = avg_log2FC)
    print(kable(top_global_iron %>% dplyr::select(gene, avg_log2FC, p_val_adj), format="pipe"))
} else {
    cat("No globally significant markers found for 'Iron cells' vs all other cells.\n")
}
```

```{r global_iron_go, fig.width=10, fig.height=8}
# Run standard enrichGO on the global Iron cell markers
if (nrow(global_iron_sig) > 0) {
  cat("Running Global GO Enrichment for 'Iron cells'...\n")
  global_iron_go <- enrichGO(
    gene = global_iron_sig$gene,
    OrgDb = org.Dm.eg.db,
    keyType = "SYMBOL",
    ont = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.2
  )
  
  if (!is.null(global_iron_go) && nrow(as.data.frame(global_iron_go)) > 0) {
    tryCatch({
      global_iron_go_simp <- clusterProfiler::simplify(global_iron_go, cutoff = 0.7, by = "p.adjust", select_fun = min)
      
      p_global_iron_go <- dotplot(global_iron_go_simp, showCategory = 15) + 
        ggtitle("Global GO BP Enrichment: Iron cells vs All Other Cells")
      print(p_global_iron_go)
    }, error = function(e) {
      p_global_iron_go <- dotplot(global_iron_go, showCategory = 15) + 
        ggtitle("Global GO BP Enrichment: Iron cells vs All Other Cells")
      print(p_global_iron_go)
    })
  } else {
    message("No significant global GO terms found for 'Iron cells'.")
  }
} else {
  message("Not enough significant global markers for GO analysis.")
}
```
