---
title: "Single-Cell Spatial Deconvolution (sc_spatial_dec)"
output: html_document
date: "2026-02-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Seurat)
library(spacexr)
library(ggplot2)
library(dplyr)
library(Matrix)
library(Matrix.utils)
library(SummarizedExperiment)
library(SpatialExperiment)
library(plotly)
library(writexl)
```

## 1. Data Loading

We load the Single-Cell reference and the Merged Spatial dataset produced by `3D_UMAP_Analysis.Rmd`.

```{r load_data}
# 1. Load Single-Cell Reference
sc_path <- "/home/mattia/Thesis/Single-cell/seurat_ctrl_with_metadata.rds"
if (!file.exists(sc_path)) stop("Single-cell reference not found.")
seurat_sc <- readRDS(sc_path)

# 2. Load Merged Spatial Data
spatial_path <- "/home/mattia/Thesis/111125ST_def/Random data/merged_guts_spatial.rds"
if (!file.exists(spatial_path)) stop("Merged spatial data not found. Please run 3D_UMAP_Analysis.Rmd first.")
seurat_spatial <- readRDS(spatial_path)

message("Datasets loaded successfully.")
```

## 2. Gene Name Harmonization

Single-cell and Spatial data often have different gene naming conventions. We clean and match them.

```{r gene_matching}
# Function to clean and aggregate matrices
harmonize_matrix <- function(mat) {
  # Clean names: Uppercase, remove periods, remove ALL whitespace
  genes <- gsub("[[:space:]]", "", toupper(gsub("\\..*$", "", rownames(mat))))
  rownames(mat) <- genes
  
  # Aggregate duplicates
  mat_agg <- aggregate.Matrix(mat, group = rownames(mat), fun = "sum")
  return(mat_agg)
}

# Process both
sc_counts <- harmonize_matrix(GetAssayData(seurat_sc, assay = "RNA", layer = "counts"))
spatial_counts <- harmonize_matrix(GetAssayData(seurat_spatial, assay = "Segmented", layer = "counts"))

# Intersect
common_genes <- intersect(rownames(sc_counts), rownames(spatial_counts))
sc_counts <- sc_counts[common_genes, ]
spatial_counts <- spatial_counts[common_genes, ]

message(paste("Harmonization complete. Common genes:", length(common_genes)))

# Create filtered objects with matched counts
seurat_sc <- CreateSeuratObject(counts = sc_counts, meta.data = seurat_sc@meta.data)
seurat_spatial <- CreateSeuratObject(counts = spatial_counts, meta.data = seurat_spatial@meta.data)

# Normalize
seurat_sc <- SCTransform(seurat_sc, verbose = FALSE)
seurat_spatial <- SCTransform(seurat_spatial, verbose = FALSE)
```

## 3. Approach A: Cell-by-cell Label Transfer (Spatial -> SC)

We treat the Spatial dataset as the **Reference** because it contains the spatial scores (Pseudotime, PC coordinates) we want to transfer.

```{r label_transfer}
# 1. Run PCA on Spatial to use as reference
seurat_spatial <- RunPCA(seurat_spatial, verbose = FALSE)

# 2. Find Transfer Anchors
message("--- Diagnostics: Spatial Reference ---")
print(colnames(seurat_spatial@meta.data))
print(summary(seurat_spatial$Spatial_Pseudotime))

# Robustness check: Remove cells with NA pseudotime if any exist
n_init <- ncol(seurat_spatial)
seurat_spatial <- subset(seurat_spatial, subset = !is.na(Spatial_Pseudotime))
if (ncol(seurat_spatial) < n_init) {
  message(paste("Removed", n_init - ncol(seurat_spatial), "cells with NA pseudotime."))
}

# Ensure PCA exists
if (!"pca" %in% names(seurat_spatial@reductions)) {
  message("Running PCA on Reference...")
  seurat_spatial <- RunPCA(seurat_spatial, verbose = FALSE)
}

# 2. Find Anchors
message("Finding Transfer Anchors...")

# Ensure Query has PCA and UMAP for visualization
if (!"pca" %in% names(seurat_sc@reductions)) {
  message("Running PCA on Query (Single-Cell)...")
  seurat_sc <- RunPCA(seurat_sc, verbose = FALSE)
}
if (!"umap" %in% names(seurat_sc@reductions)) {
  message("Running UMAP on Query (Single-Cell)...")
  seurat_sc <- RunUMAP(seurat_sc, dims = 1:30, verbose = FALSE)
}

transfer_anchors <- FindTransferAnchors(
  reference = seurat_spatial,
  query = seurat_sc,
  normalization.method = "SCT",
  reference.assay = "SCT",
  query.assay = "SCT",
  dims = 1:30,
  reference.reduction = "pca" 
)

# 4. Transfer Data
# Transfer Pseudotime
message("Transferring Pseudotime...")
pt_predictions <- TransferData(
  anchorset = transfer_anchors,
  refdata = "Spatial_Pseudotime",
  reference = seurat_spatial,
  weight.reduction = seurat_sc[["pca"]],
  dims = 1:30
)

# For numeric transfer, Seurat stores the result in predicted.id
# We also capture the prediction score (confidence)
seurat_sc$Transferred_Pseudotime <- as.numeric(as.character(pt_predictions$predicted.id))
seurat_sc$Transfer_Confidence <- pt_predictions$prediction.score.max

# 5. Noise Reduction: KNN Smoothing
# If a cell is surrounded by neighbors with very different pseudotimes, 
# it might be an anchor artifact. We smooth the scores using the SC UMAP/PCA neighborhood.
message("Refining scores with KNN smoothing...")

# Build or retrieve nearest neighbor graph
if (!"SCT_nn" %in% names(seurat_sc@graphs)) {
  message("Building nearest neighbor graph...")
  seurat_sc <- FindNeighbors(seurat_sc, reduction = "pca", dims = 1:30, graph.name = "SCT_nn")
}

neighbors <- seurat_sc@graphs$SCT_nn
# Simple smoothing: average of self + neighbors
pt_vec <- seurat_sc$Transferred_Pseudotime
# Convert graph to adjacency matrix
adj <- as.matrix(neighbors)
# Weight the neighbors (row-normalize)
adj_norm <- adj / rowSums(adj)
seurat_sc$Smoothed_Pseudotime <- as.numeric(adj_norm %*% pt_vec)

# 6. Visualization: UMAP Projections
p1 <- DimPlot(seurat_sc, reduction = "umap", group.by = "celltype", label = TRUE) + 
  NoLegend() + ggtitle("Cell Types (UMAP)")

p2 <- FeaturePlot(seurat_sc, features = "Transferred_Pseudotime") + 
  scale_color_viridis_c(option = "magma") + 
  ggtitle("Inferred spatial position")

p3 <- FeaturePlot(seurat_sc, features = "Smoothed_Pseudotime") + 
  scale_color_viridis_c(option = "magma") + 
  ggtitle("Pseudotime (Smoothed)")

p4 <- FeaturePlot(seurat_sc, features = "Transfer_Confidence") + 
  scale_color_viridis_c(option = "mako") + 
  ggtitle("Transfer Confidence")

library(patchwork)
(p1 + p2) / (p3 + p4)

# 7. Visualization: PCA Projections (Reversed X-Axis)
p5 <- DimPlot(seurat_sc, reduction = "pca", group.by = "celltype", label = TRUE) + 
  NoLegend() + ggtitle("Cell Types (PCA)") + scale_x_reverse()

p6 <- FeaturePlot(seurat_sc, features = "Smoothed_Pseudotime", reduction = "pca") + 
  scale_color_viridis_c(option = "magma") + 
  ggtitle("Smoothed Pseudotime (PCA)") + scale_x_reverse()

print(p5 + p6)

# 8. 3D Visualization: UMAP + Smoothed Pseudotime
# Prepare simplified plotting dataframe
plot_3d_df <- data.frame(
  UMAP1 = Embeddings(seurat_sc, "umap")[,1],
  UMAP2 = Embeddings(seurat_sc, "umap")[,2],
  Pseudotime = seurat_sc$Smoothed_Pseudotime,
  CellType = seurat_sc$celltype,
  Confidence = seurat_sc$Transfer_Confidence
)

plot_ly(plot_3d_df, x = ~UMAP1, y = ~UMAP2, z = ~Pseudotime, 
        color = ~CellType, colors = "Set3", type = "scatter3d", mode = "markers",
        marker = list(size = 2),
        text = ~paste("Type:", CellType, "<br>Conf:", round(Confidence, 2))) %>%
  layout(title = "3D Smoothed Spatial Map",
         scene = list(xaxis = list(title = 'UMAP 1'),
                      yaxis = list(title = 'UMAP 2'),
                      zaxis = list(title = 'Inferred spatial position')))
```


## 4. Approach B: RCTD Deconvolution (SC as Pixels)

We treat each Single Cell as a "pixel" and ask which Spatial "cluster" it belongs to.

```{r rctd_deconvolution}
# 1. Define Spatial Pseudotime Bins as "Clusters"
# We divide the 0-100 pseudotime scale into bins (e.g., bins of 5)
bin_width <- 5
seurat_spatial$Pseudotime_Bin <- floor(seurat_spatial$Spatial_Pseudotime / bin_width) * bin_width
seurat_spatial$Pseudotime_Bin_Label <- paste0("Bin_", seurat_spatial$Pseudotime_Bin)

# 2. Filter out bins with < 25 cells (RCTD Requirement)
bin_counts <- table(seurat_spatial$Pseudotime_Bin_Label)
valid_bins <- names(bin_counts)[bin_counts >= 25]
message(paste("Filtering bins. Keeping", length(valid_bins), "bins with >= 25 cells."))
seurat_spatial <- subset(seurat_spatial, subset = Pseudotime_Bin_Label %in% valid_bins)
seurat_spatial$Pseudotime_Bin_Label <- factor(seurat_spatial$Pseudotime_Bin_Label)

# 3. Create Reference Object (Binned Pseudotime)
spatial_counts_mat <- GetAssayData(seurat_spatial, assay = "SCT", layer = "counts")
spatial_labels <- seurat_spatial$Pseudotime_Bin_Label
names(spatial_labels) <- colnames(seurat_spatial)

# RCTD Reference setup (Using SummarizedExperiment for compatibility)
sc_ref_for_rctd <- SummarizedExperiment(
  assays = list(counts = spatial_counts_mat),
  colData = data.frame(cell_type = spatial_labels, row.names = names(spatial_labels))
)

# 3. Create Target Object (Single Cells as Pixels)
# Since RCTD expects coordinates, we give them "zero" coordinates or random ones
# as we don't have spatial positions for SC, but the algorithm needs the slot.
sc_counts_mat <- GetAssayData(seurat_sc, assay = "SCT", layer = "counts")
sc_coords <- data.frame(x = rep(0, ncol(seurat_sc)), y = rep(0, ncol(seurat_sc)), row.names = colnames(seurat_sc))

spatial_target_for_rctd <- SpatialExperiment(
  assays = list(counts = sc_counts_mat),
  spatialCoords = as.matrix(sc_coords)
)

# 4. Run RCTD (with Caching)
rctd_save_path <- "../Random data/myRCTD_spatial_bins_results.rds"

if (file.exists(rctd_save_path)) {
  message("Loading existing RCTD results from disk...")
  myRCTD <- readRDS(rctd_save_path)
} else {
  message("No existing results found. Computing RCTD deconvolution...")
  myRCTD <- createRctd(
    spatial_experiment = spatial_target_for_rctd,
    reference_experiment = sc_ref_for_rctd,
    cell_type_col = "cell_type"
  )
  myRCTD <- runRctd(myRCTD)
  
  message("Saving RCTD results for future use...")
  saveRDS(myRCTD, rctd_save_path)
}

# 5. Extract Results
# RCTD weights are Bins x Pixels (Cells). We transpose to get Cells x Bins for Seurat metadata.
weights_mat <- as.matrix(assay(myRCTD, "weights"))
weights <- as.data.frame(t(weights_mat))

# Diagnostic Check
message(paste("Weights dimensions:", nrow(weights), "x", ncol(weights)))
message(paste("Seurat cells:", ncol(seurat_sc)))
if (!all(rownames(weights) %in% colnames(seurat_sc))) {
  warning("Metadata rownames do not perfectly match Seurat colnames!")
}

# Add weights to Seurat SC object
seurat_sc <- AddMetaData(seurat_sc, metadata = weights)

# 6. Visualization: UMAP Level Comparison
# Plot weights for a specific pseudotime bin (e.g., Bin 0 or Bin 50)
available_bins <- colnames(weights)
target_bin <- available_bins[round(length(available_bins)/2)] # Pick center bin for demo

p1_u <- DimPlot(seurat_sc, reduction = "umap", group.by = "celltype", label = TRUE) + 
  NoLegend() + ggtitle("Cell Types (UMAP)")
p2_u <- FeaturePlot(seurat_sc, features = target_bin) + 
  scale_color_viridis_c(option = "magma") + 
  ggtitle(paste("RCTD:", target_bin, "(UMAP)"))
p3_u <- FeaturePlot(seurat_sc, features = "Transferred_Pseudotime") + 
  scale_color_viridis_c(option = "magma") + 
  ggtitle("Pseudotime (UMAP)")

library(patchwork)
print(p1_u + p3_u) # + p2_u

# 7. Visualization: PCA Level Comparison (Reversed X-Axis)
p1_p <- DimPlot(seurat_sc, reduction = "pca", group.by = "celltype", label = TRUE) + 
  NoLegend() + ggtitle("Cell Types (PCA)") + scale_x_reverse()
p2_p <- FeaturePlot(seurat_sc, features = target_bin, reduction = "pca") + 
  scale_color_viridis_c(option = "magma") + 
  ggtitle(paste("RCTD:", target_bin, "(PCA)")) + scale_x_reverse()
p3_p <- FeaturePlot(seurat_sc, features = "Transferred_Pseudotime", reduction = "pca") + 
  scale_color_viridis_c(option = "magma") + 
  ggtitle("Pseudotime (PCA)") + scale_x_reverse()

print(p1_p + p3_p) #+ p2_p

# 7. 3D Visualization: UMAP + Pseudotime (Colored by Cell Type)
plot_3d_rctd_df <- data.frame(
  UMAP1 = Embeddings(seurat_sc, "umap")[,1],
  UMAP2 = Embeddings(seurat_sc, "umap")[,2],
  Pseudotime = seurat_sc$Transferred_Pseudotime,
  CellType = seurat_sc$celltype
)

plot_ly(plot_3d_rctd_df, x = ~UMAP1, y = ~UMAP2, z = ~Pseudotime, 
        color = ~CellType, colors = "Set3", type = "scatter3d", mode = "markers",
        marker = list(size = 2)) %>%
  layout(title = "Approach B 3D: Cell Type Distribution",
         scene = list(xaxis = list(title = 'UMAP 1'),
                      yaxis = list(title = 'UMAP 2'),
                      zaxis = list(title = 'Transferred Pseudotime')))
```

## 6. Differential Gene Expression Analysis

We perform DGE analysis using pseudotime as a continuous covariate to identify genes that vary along the spatial axis within each cell type.

### Approach A: Smoothed Pseudotime DGE

```{r dge_approach_a}
library(presto)

# Function to perform correlation-based DGE for a cell type
run_pseudotime_dge <- function(seurat_obj, celltype_name, pseudotime_col = "Smoothed_Pseudotime") {
  # Subset to specific cell type
  cells_subset <- subset(seurat_obj, subset = celltype == celltype_name)
  
  if (ncol(cells_subset) < 50) {
    message(paste("Skipping", celltype_name, "- too few cells:", ncol(cells_subset)))
    return(NULL)
  }
  
  message(paste("Analyzing", celltype_name, "with", ncol(cells_subset), "cells..."))
  
  # Get expression matrix and pseudotime
  expr_mat <- GetAssayData(cells_subset, assay = "SCT", layer = "data")
  pt_values <- cells_subset@meta.data[[pseudotime_col]]
  
  # Calculate correlation for each gene
  correlations <- apply(expr_mat, 1, function(gene_expr) {
    if (sd(gene_expr) == 0) return(c(cor = 0, pval = 1))
    test <- cor.test(gene_expr, pt_values, method = "spearman")
    c(cor = test$estimate, pval = test$p.value)
  })
  
  # Create results dataframe
  results <- data.frame(
    gene = rownames(expr_mat),
    correlation = correlations[1, ],
    pvalue = correlations[2, ],
    celltype = celltype_name,
    n_cells = ncol(cells_subset)
  )
  
  # Adjust p-values
  results$padj <- p.adjust(results$pvalue, method = "BH")
  
  # Add mean expression
  results$mean_expr <- rowMeans(expr_mat)
  
  # Sort by absolute correlation
  results <- results[order(-abs(results$correlation)), ]
  
  return(results)
}

# Check if results already exist
dge_smoothed_path <- "../Random data/dge_smoothed_pseudotime.csv"

if (file.exists(dge_smoothed_path)) {
  message("Loading existing smoothed pseudotime DGE results...")
  dge_all_smoothed <- read.csv(dge_smoothed_path)
  
  # Reconstruct list format for downstream analysis
  celltypes <- unique(seurat_sc$celltype)
  dge_results_smoothed <- lapply(celltypes, function(ct) {
    subset(dge_all_smoothed, celltype == ct)
  })
  names(dge_results_smoothed) <- celltypes
  dge_results_smoothed <- dge_results_smoothed[sapply(dge_results_smoothed, nrow) > 0]
  
} else {
  message("Computing smoothed pseudotime DGE...")
  
  # Run for all cell types with Smoothed Pseudotime
  celltypes <- unique(seurat_sc$celltype)
  dge_results_smoothed <- lapply(celltypes, function(ct) {
    run_pseudotime_dge(seurat_sc, ct, "Smoothed_Pseudotime")
  })
  names(dge_results_smoothed) <- celltypes
  
  # Remove NULL results (cell types with too few cells)
  dge_results_smoothed <- dge_results_smoothed[!sapply(dge_results_smoothed, is.null)]
  
  # Combine all results
  dge_all_smoothed <- do.call(rbind, dge_results_smoothed)
  
  # Save results
  write.csv(dge_all_smoothed, dge_smoothed_path, row.names = FALSE)
  message("Saved smoothed pseudotime DGE results")
}
```

### Approach B: Raw Pseudotime DGE (RCTD-based)

```{r dge_approach_b}
# Check if results already exist
dge_raw_path <- "../Random data/dge_raw_pseudotime.csv"

if (file.exists(dge_raw_path)) {
  message("Loading existing raw pseudotime DGE results...")
  dge_all_raw <- read.csv(dge_raw_path)
  
  # Reconstruct list format for downstream analysis
  celltypes <- unique(seurat_sc$celltype)
  dge_results_raw <- lapply(celltypes, function(ct) {
    subset(dge_all_raw, celltype == ct)
  })
  names(dge_results_raw) <- celltypes
  dge_results_raw <- dge_results_raw[sapply(dge_results_raw, nrow) > 0]
  
} else {
  message("Computing raw pseudotime DGE...")
  
  # Run for all cell types with Raw Transferred Pseudotime
  celltypes <- unique(seurat_sc$celltype)
  dge_results_raw <- lapply(celltypes, function(ct) {
    run_pseudotime_dge(seurat_sc, ct, "Transferred_Pseudotime")
  })
  names(dge_results_raw) <- celltypes
  
  # Remove NULL results
  dge_results_raw <- dge_results_raw[!sapply(dge_results_raw, is.null)]
  
  # Combine all results
  dge_all_raw <- do.call(rbind, dge_results_raw)
  
  # Save results
  write.csv(dge_all_raw, dge_raw_path, row.names = FALSE)
  message("Saved raw pseudotime DGE results")
}
```

### Visualization: Top Pseudotime-Dependent Genes

```{r dge_visualization}
library(ggplot2)
library(dplyr)
library(viridis)

# Select a cell type for detailed visualization (e.g., EC - Enterocytes)
target_celltype <- "ISC"

if (target_celltype %in% names(dge_results_smoothed)) {
  ct_results <- dge_results_smoothed[[target_celltype]]
  
  # 1. Volcano Plot
  ct_results$significant <- ct_results$padj < 0.05 & abs(ct_results$correlation) > 0.3
  
  p_volcano <- ggplot(ct_results, aes(x = correlation, y = -log10(pvalue), color = significant)) +
    geom_point(alpha = 0.6, size = 1.5) +
    scale_color_manual(values = c("grey70", "red3")) +
    geom_vline(xintercept = c(-0.3, 0.3), linetype = "dashed", color = "grey40") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    labs(title = paste("Volcano Plot:", target_celltype),
         x = "Spearman Correlation with Pseudotime",
         y = "-log10(p-value)") +
    theme_minimal()
  
  print(p_volcano)
  
  # 2. Heatmap of Top Genes
  # Get top positively and negatively correlated genes
  top_pos <- ct_results %>% filter(correlation > 0, padj < 0.05) %>% head(20)
  top_neg <- ct_results %>% filter(correlation < 0, padj < 0.05) %>% head(20)
  top_genes <- c(top_pos$gene, top_neg$gene)
  
  if (length(top_genes) > 0) {
    # Subset cells
    cells_subset <- subset(seurat_sc, subset = celltype == target_celltype)
    
    # Order cells by pseudotime
    cells_ordered <- cells_subset@meta.data %>%
      arrange(Smoothed_Pseudotime) %>%
      rownames()
    
    # Get expression matrix
    expr_mat <- GetAssayData(cells_subset, assay = "SCT", layer = "data")[top_genes, cells_ordered]
    
    # Scale for visualization
    expr_scaled <- t(scale(t(as.matrix(expr_mat))))
    
    # Create annotation for pseudotime
    pt_anno <- cells_subset@meta.data[cells_ordered, "Smoothed_Pseudotime", drop = FALSE]
    
    library(pheatmap)
    pheatmap(expr_scaled,
             cluster_cols = FALSE,
             cluster_rows = TRUE,
             show_colnames = FALSE,
             annotation_col = pt_anno,
             color = colorRampPalette(c("blue", "white", "red"))(100),
             main = paste("Top Pseudotime Genes:", target_celltype),
             fontsize_row = 8)
  }
  
  # 3. Pseudotime-Ordered Heatmap (Genes ordered by peak expression position)
  # Get significant genes
  sig_genes <- ct_results %>% filter(padj < 0.05, abs(correlation) > 0.3)
  
  if (nrow(sig_genes) > 0) {
    # Limit to top genes if too many
    if (nrow(sig_genes) > 50) {
      sig_genes <- sig_genes %>% arrange(desc(abs(correlation))) %>% head(50)
    }
    
    cells_subset <- subset(seurat_sc, subset = celltype == target_celltype)
    
    # Get expression matrix for significant genes
    expr_mat <- GetAssayData(cells_subset, assay = "SCT", layer = "data")[sig_genes$gene, ]
    pt_values <- cells_subset$Smoothed_Pseudotime
    
    # Find peak expression position for each gene
    gene_peak_positions <- apply(expr_mat, 1, function(gene_expr) {
      # Bin pseudotime and find bin with highest mean expression
      pt_bins <- cut(pt_values, breaks = 20, labels = FALSE)
      bin_means <- tapply(gene_expr, pt_bins, mean, na.rm = TRUE)
      which.max(bin_means)
    })
    
    # Order genes by peak position
    gene_order <- names(sort(gene_peak_positions))
    
    # Create binned pseudotime for visualization
    pt_bins <- cut(pt_values, breaks = 50, labels = FALSE)
    bin_centers <- tapply(pt_values, pt_bins, mean)
    
    # Average expression per bin
    expr_binned <- t(apply(expr_mat[gene_order, ], 1, function(gene_expr) {
      tapply(gene_expr, pt_bins, mean, na.rm = TRUE)
    }))
    
    # Scale for visualization
    expr_scaled <- t(scale(t(expr_binned)))
    
    # Create annotation
    pt_anno <- data.frame(
      Pseudotime = bin_centers,
      row.names = colnames(expr_binned)
    )
    
    library(pheatmap)
    pheatmap(expr_scaled,
             cluster_cols = FALSE,
             cluster_rows = FALSE,  # Already ordered by peak
             show_colnames = FALSE,
             annotation_col = pt_anno,
             color = colorRampPalette(c("blue", "white", "red"))(100),
             main = paste("Pseudotime-Ordered Expression:", target_celltype),
             fontsize_row = 6,
             border_color = NA)
  }
  
  # 4. Expression Curves for Top Genes
  # Plot both increasing and decreasing genes
  if (nrow(top_pos) > 0) {
    # Select top 6 increasing genes
    genes_to_plot_pos <- head(top_pos$gene, 6)
    cells_subset <- subset(seurat_sc, subset = celltype == target_celltype)
    
    # Prepare data for plotting
    plot_data <- data.frame(
      Pseudotime = cells_subset$Smoothed_Pseudotime,
      cell = colnames(cells_subset)
    )
    
    for (gene in genes_to_plot_pos) {
      plot_data[[gene]] <- GetAssayData(cells_subset, assay = "SCT", layer = "data")[gene, ]
    }
    
    # Reshape for ggplot
    plot_data_long <- tidyr::pivot_longer(plot_data, 
                                          cols = all_of(genes_to_plot_pos),
                                          names_to = "Gene",
                                          values_to = "Expression")
    
    p_curves_pos <- ggplot(plot_data_long, aes(x = Pseudotime, y = Expression)) +
      geom_point(alpha = 0.3, size = 0.5, color = "grey60") +
      geom_smooth(method = "loess", color = "red3", se = TRUE) +
      facet_wrap(~Gene, scales = "free_y", ncol = 3) +
      labs(title = paste("Top Increasing Genes:", target_celltype),
           x = "Smoothed Pseudotime",
           y = "Normalized Expression") +
      theme_minimal()
    
    print(p_curves_pos)
  }
  
  # Plot top decreasing genes
  if (nrow(top_neg) > 0) {
    genes_to_plot_neg <- head(top_neg$gene, 6)
    cells_subset <- subset(seurat_sc, subset = celltype == target_celltype)
    
    plot_data_neg <- data.frame(
      Pseudotime = cells_subset$Smoothed_Pseudotime,
      cell = colnames(cells_subset)
    )
    
    for (gene in genes_to_plot_neg) {
      plot_data_neg[[gene]] <- GetAssayData(cells_subset, assay = "SCT", layer = "data")[gene, ]
    }
    
    plot_data_neg_long <- tidyr::pivot_longer(plot_data_neg, 
                                              cols = all_of(genes_to_plot_neg),
                                              names_to = "Gene",
                                              values_to = "Expression")
    
    p_curves_neg <- ggplot(plot_data_neg_long, aes(x = Pseudotime, y = Expression)) +
      geom_point(alpha = 0.3, size = 0.5, color = "grey60") +
      geom_smooth(method = "loess", color = "blue3", se = TRUE) +
      facet_wrap(~Gene, scales = "free_y", ncol = 3) +
      labs(title = paste("Top Decreasing Genes:", target_celltype),
           x = "Smoothed Pseudotime",
           y = "Normalized Expression") +
      theme_minimal()
    
    print(p_curves_neg)
  }
}
```

### Cell-Type-Specific Spatial Genes

```{r celltype_specific_genes}
# Identify genes that are both spatially regulated AND cell-type-specific
# This helps find regional specialization markers unique to each cell type

if (target_celltype %in% names(dge_results_smoothed)) {
  
  # 1. Calculate cell-type specificity for each gene
  # Use a simple approach: ratio of expression in target vs other cell types
  
  message(paste("Identifying cell-type-specific spatial genes for", target_celltype))
  
  cells_target <- subset(seurat_sc, subset = celltype == target_celltype)
  cells_other <- subset(seurat_sc, subset = celltype != target_celltype)
  
  # Get mean expression in each group
  expr_target <- rowMeans(GetAssayData(cells_target, assay = "SCT", layer = "data"))
  expr_other <- rowMeans(GetAssayData(cells_other, assay = "SCT", layer = "data"))
  
  # Calculate specificity score (log2 fold change)
  specificity_score <- log2((expr_target + 0.01) / (expr_other + 0.01))
  
  # Get DGE results for this cell type
  ct_results <- dge_results_smoothed[[target_celltype]]
  
  # Add specificity scores
  ct_results$specificity <- specificity_score[ct_results$gene]
  
  # Filter for genes that are:
  # 1. Spatially regulated (padj < 0.05, |cor| > 0.3)
  # 2. Cell-type-specific (specificity > 1, i.e., 2-fold enriched)
  specific_spatial_genes <- ct_results %>%
    filter(padj < 0.05, abs(correlation) > 0.3, specificity > 1) %>%
    arrange(desc(specificity))
  
  message(paste("Found", nrow(specific_spatial_genes), "cell-type-specific spatial genes"))
  
  if (nrow(specific_spatial_genes) > 0) {
    # 2. Volcano plot highlighting cell-type-specific genes
    ct_results$category <- "Other"
    ct_results$category[ct_results$padj < 0.05 & abs(ct_results$correlation) > 0.3] <- "Spatial"
    ct_results$category[ct_results$gene %in% specific_spatial_genes$gene] <- "Specific + Spatial"
    
    p_specific_volcano <- ggplot(ct_results, aes(x = correlation, y = -log10(pvalue), color = category)) +
      geom_point(alpha = 0.6, size = 1.5) +
      scale_color_manual(values = c("Other" = "grey70", 
                                     "Spatial" = "orange", 
                                     "Specific + Spatial" = "darkred")) +
      geom_vline(xintercept = c(-0.3, 0.3), linetype = "dashed", color = "grey40") +
      geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
      labs(title = paste("Cell-Type-Specific Spatial Genes:", target_celltype),
           x = "Expression Correlation with Position",
           y = "-log10(p-value)",
           color = "Gene Category") +
      theme_minimal()
    
    print(p_specific_volcano)
    
    # 3. Expression curves for ALL cell-type-specific spatial genes
    # Plot in batches of 6 genes per plot
    if (nrow(specific_spatial_genes) > 0) {
      n_genes <- nrow(specific_spatial_genes)
      genes_per_plot <- 6
      n_plots <- ceiling(n_genes / genes_per_plot)
      
      message(paste("Creating", n_plots, "plots for", n_genes, "cell-type-specific spatial genes"))
      
      for (i in 1:n_plots) {
        # Get genes for this batch
        start_idx <- (i - 1) * genes_per_plot + 1
        end_idx <- min(i * genes_per_plot, n_genes)
        genes_batch <- specific_spatial_genes$gene[start_idx:end_idx]
        
        # Prepare data
        plot_data_specific <- data.frame(
          Pseudotime = cells_target$Smoothed_Pseudotime,
          cell = colnames(cells_target)
        )
        
        for (gene in genes_batch) {
          plot_data_specific[[gene]] <- GetAssayData(cells_target, assay = "SCT", layer = "data")[gene, ]
        }
        
        # Reshape for ggplot
        plot_data_specific_long <- tidyr::pivot_longer(plot_data_specific, 
                                                       cols = all_of(genes_batch),
                                                       names_to = "Gene",
                                                       values_to = "Expression")
        
        # Create plot
        p_specific_curves <- ggplot(plot_data_specific_long, aes(x = Pseudotime, y = Expression)) +
          geom_point(alpha = 0.3, size = 0.5, color = "grey60") +
          geom_smooth(method = "loess", color = "darkred", se = TRUE) +
          facet_wrap(~Gene, scales = "free_y", ncol = 3) +
          labs(title = paste("Cell-Type-Specific Spatial Markers:", target_celltype, 
                            paste0("(", start_idx, "-", end_idx, " of ", n_genes, ")")),
               x = "Smoothed Pseudotime",
               y = "Normalized Expression",
               subtitle = paste("Genes enriched >2-fold in", target_celltype, "vs other cell types")) +
          theme_minimal()
        
        print(p_specific_curves)
      }
    }
    
    # 4. Save cell-type-specific results
    write.csv(specific_spatial_genes, 
              paste0("../Random data/celltype_specific_spatial_", target_celltype, ".csv"),
              row.names = FALSE)
    message(paste("Saved cell-type-specific spatial genes for", target_celltype))
  }
}

```

### Comparison: Smoothed vs Raw Results

```{r dge_comparison}
# Compare results between smoothed and raw for the same cell type
if (target_celltype %in% names(dge_results_smoothed) && 
    target_celltype %in% names(dge_results_raw)) {
  
  smoothed_res <- dge_results_smoothed[[target_celltype]]
  raw_res <- dge_results_raw[[target_celltype]]
  
  # Merge results
  comparison <- merge(smoothed_res[, c("gene", "correlation", "padj")],
                      raw_res[, c("gene", "correlation", "padj")],
                      by = "gene",
                      suffixes = c("_smoothed", "_raw"))
  
  # Plot correlation comparison
  p_compare <- ggplot(comparison, aes(x = correlation_raw, y = correlation_smoothed)) +
    geom_point(alpha = 0.4, size = 1) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
    labs(title = paste("Smoothed vs Raw Pseudotime Correlations:", target_celltype),
         x = "Correlation (Raw Pseudotime)",
         y = "Correlation (Smoothed Pseudotime)") +
    theme_minimal()
  
  print(p_compare)
  
  # Identify genes with large differences
  comparison$diff <- abs(comparison$correlation_smoothed - comparison$correlation_raw)
  top_diff <- comparison %>% arrange(desc(diff)) %>% head(10)
  
  message("Top genes with largest difference between smoothed and raw:")
  print(top_diff[, c("gene", "correlation_smoothed", "correlation_raw", "diff")])
}
```

## 6b. Collapsed Cell-Type DGE Analysis
This section performs DGE analysis on higher-level cell type categories to identify broader lineage-level spatial patterns.

```{r collapsed_celltype_dge}
library(ggplot2)
library(dplyr)
# 1. Create collapsed cell type labels
seurat_sc$celltype_collapsed <- seurat_sc$celltype
# Define mapping
ec_types <- c("aEC", "pEC", "mEC")
dc_types <- c("ISC", "EB", "dEC")
# Apply mapping
seurat_sc$celltype_collapsed[seurat_sc$celltype %in% ec_types] <- "EC"
seurat_sc$celltype_collapsed[seurat_sc$celltype %in% dc_types] <- "DC"
# 2. Filter to keep only EC and DC
seurat_collapsed <- subset(seurat_sc, subset = celltype_collapsed %in% c("EC", "DC"))
message(paste("Filtered to", ncol(seurat_collapsed), "cells"))
message(paste("EC cells:", sum(seurat_collapsed$celltype_collapsed == "EC")))
message(paste("DC cells:", sum(seurat_collapsed$celltype_collapsed == "DC")))
# 3. Run DGE analysis - with caching
dge_collapsed_path <- "../Random data/dge_collapsed_pseudotime.csv"
if (file.exists(dge_collapsed_path)) {
  message("Loading existing collapsed DGE results...")
  dge_all_collapsed <- read.csv(dge_collapsed_path)
  collapsed_types <- c("EC", "DC")
  dge_results_collapsed <- lapply(collapsed_types, function(ct) subset(dge_all_collapsed, celltype == ct))
  names(dge_results_collapsed) <- collapsed_types
  dge_results_collapsed <- dge_results_collapsed[sapply(dge_results_collapsed, nrow) > 0]
} else {
  message("Computing collapsed DGE...")
  
  # Modified DGE function for collapsed types
  run_pseudotime_dge_collapsed <- function(seurat_obj, celltype_name, pseudotime_col = "Smoothed_Pseudotime") {
    # Subset using celltype_collapsed instead of celltype
    cells_subset <- subset(seurat_obj, subset = celltype_collapsed == celltype_name)
    
    if (ncol(cells_subset) < 50) {
      message(paste("Skipping", celltype_name, "- too few cells:", ncol(cells_subset)))
      return(NULL)
    }
    
    message(paste("Analyzing", celltype_name, "with", ncol(cells_subset), "cells..."))
    
    expr_mat <- GetAssayData(cells_subset, assay = "SCT", layer = "data")
    pt_values <- cells_subset@meta.data[[pseudotime_col]]
    
    correlations <- apply(expr_mat, 1, function(gene_expr) {
      if (sd(gene_expr) == 0) return(c(cor = 0, pval = 1))
      test <- cor.test(gene_expr, pt_values, method = "spearman")
      c(cor = test$estimate, pval = test$p.value)
    })
    
    results <- data.frame(
      gene = rownames(expr_mat),
      correlation = correlations[1, ],
      pvalue = correlations[2, ],
      celltype = celltype_name,
      n_cells = ncol(cells_subset)
    )
    
    results$padj <- p.adjust(results$pvalue, method = "BH")
    results$mean_expr <- rowMeans(expr_mat)
    results <- results[order(-abs(results$correlation)), ]
    
    return(results)
  }
  
  collapsed_types <- c("EC", "DC")
  dge_results_collapsed <- lapply(collapsed_types, function(ct) {
    run_pseudotime_dge_collapsed(seurat_collapsed, ct, "Smoothed_Pseudotime")
  })
  names(dge_results_collapsed) <- collapsed_types
  dge_results_collapsed <- dge_results_collapsed[!sapply(dge_results_collapsed, is.null)]
  dge_all_collapsed <- do.call(rbind, dge_results_collapsed)
  write.csv(dge_all_collapsed, dge_collapsed_path, row.names = FALSE)
}
# 4. Visualizations for each collapsed type
for (ct in names(dge_results_collapsed)) {
  ct_results <- dge_results_collapsed[[ct]]
  ct_results$significant <- ct_results$padj < 0.05 & abs(ct_results$correlation) > 0.3
  
  # Volcano plot
  p_volcano <- ggplot(ct_results, aes(x = correlation, y = -log10(pvalue), color = significant)) +
    geom_point(alpha = 0.6, size = 1.5) +
    scale_color_manual(values = c("grey70", "red3")) +
    geom_vline(xintercept = c(-0.3, 0.3), linetype = "dashed") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
    labs(title = paste("Collapsed:", ct), subtitle = "EC=aEC+pEC+mEC, DC=ISC+EB+dEC") +
    theme_minimal()
  print(p_volcano)
  
  # Get top genes and create heatmap (same logic as before)
  sig_genes <- ct_results %>% filter(padj < 0.05, abs(correlation) > 0.3)
  if (nrow(sig_genes) > 50) sig_genes <- sig_genes %>% arrange(desc(abs(correlation))) %>% head(50)
  
  if (nrow(sig_genes) > 0) {
    cells_subset <- subset(seurat_collapsed, subset = celltype_collapsed == ct)
    expr_mat <- GetAssayData(cells_subset, assay = "SCT", layer = "data")[sig_genes$gene, ]
    pt_values <- cells_subset$Smoothed_Pseudotime
    
    gene_peak_positions <- apply(expr_mat, 1, function(x) {
      bins <- cut(pt_values, breaks = 20, labels = FALSE)
      which.max(tapply(x, bins, mean, na.rm = TRUE))
    })
    gene_order <- names(sort(gene_peak_positions))
    
    pt_bins <- cut(pt_values, breaks = 50, labels = FALSE)
    expr_binned <- t(apply(expr_mat[gene_order, ], 1, function(x) tapply(x, pt_bins, mean, na.rm = TRUE)))
    expr_scaled <- t(scale(t(expr_binned)))
    
    pheatmap(expr_scaled, cluster_cols = FALSE, cluster_rows = FALSE, show_colnames = FALSE,
             color = colorRampPalette(c("blue", "white", "red"))(100),
             main = paste(ct, "(Collapsed)"), fontsize_row = 6, border_color = NA)
  }
  
  # Expression curves for top increasing/decreasing
  top_pos <- ct_results %>% filter(correlation > 0, padj < 0.05) %>% head(6)
  top_neg <- ct_results %>% filter(correlation < 0, padj < 0.05) %>% head(6)
  
  for (direction in c("pos", "neg")) {
    genes_list <- if (direction == "pos") top_pos else top_neg
    if (nrow(genes_list) == 0) next
    
    cells_subset <- subset(seurat_collapsed, subset = celltype_collapsed == ct)
    plot_data <- data.frame(Pseudotime = cells_subset$Smoothed_Pseudotime)
    for (g in genes_list$gene) plot_data[[g]] <- GetAssayData(cells_subset, assay = "SCT", layer = "data")[g, ]
    
    plot_long <- tidyr::pivot_longer(plot_data, cols = -Pseudotime, names_to = "Gene", values_to = "Expression")
    color_val <- if (direction == "pos") "red3" else "blue3"
    title_text <- if (direction == "pos") "Increasing" else "Decreasing"
    
    p <- ggplot(plot_long, aes(x = Pseudotime, y = Expression)) +
      geom_point(alpha = 0.3, size = 0.5, color = "grey60") +
      geom_smooth(method = "loess", color = color_val, se = TRUE) +
      facet_wrap(~Gene, scales = "free_y", ncol = 3) +
      labs(title = paste(title_text, "Genes:", ct, "(Collapsed)")) +
      theme_minimal()
    print(p)
  }
}
```

### Cell-Type-Specific Spatial Genes (Collapsed Groups)
```{r collapsed_celltype_specific_genes}
# Identify genes that are both spatially regulated AND specific to EC or DC
for (target_ct in c("EC", "DC")) {
  if (target_ct %in% names(dge_results_collapsed)) {
    
    message(paste("Identifying cell-type-specific spatial genes for", target_ct))
    
    # Calculate specificity: EC vs DC or DC vs EC
    cells_target <- subset(seurat_collapsed, subset = celltype_collapsed == target_ct)
    cells_other <- subset(seurat_collapsed, subset = celltype_collapsed != target_ct)
    
    # Get mean expression
    expr_target <- rowMeans(GetAssayData(cells_target, assay = "SCT", layer = "data"))
    expr_other <- rowMeans(GetAssayData(cells_other, assay = "SCT", layer = "data"))
    
    # Specificity score
    specificity_score <- log2((expr_target + 0.01) / (expr_other + 0.01))
    
    ct_results <- dge_results_collapsed[[target_ct]]
    ct_results$specificity <- specificity_score[ct_results$gene]
    
    # Filter for specific + spatial genes
    specific_spatial_genes <- ct_results %>%
      filter(padj < 0.05, abs(correlation) > 0.3, specificity > 1) %>%
      arrange(desc(specificity))
    
    message(paste("Found", nrow(specific_spatial_genes), "genes for", target_ct))
    
    if (nrow(specific_spatial_genes) > 0) {
      # Volcano plot
      ct_results$category <- "Other"
      ct_results$category[ct_results$padj < 0.05 & abs(ct_results$correlation) > 0.3] <- "Spatial"
      ct_results$category[ct_results$gene %in% specific_spatial_genes$gene] <- "Specific + Spatial"
      
      p_volcano <- ggplot(ct_results, aes(x = correlation, y = -log10(pvalue), color = category)) +
        geom_point(alpha = 0.6, size = 1.5) +
        scale_color_manual(values = c("Other" = "grey70", "Spatial" = "orange", "Specific + Spatial" = "darkred")) +
        geom_vline(xintercept = c(-0.3, 0.3), linetype = "dashed") +
        geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
        labs(title = paste(target_ct, "Specific Spatial Genes (Collapsed)"),
             subtitle = paste(target_ct, "vs", ifelse(target_ct == "EC", "DC", "EC"))) +
        theme_minimal()
      print(p_volcano)
      
      # Expression curves - all genes in batches of 6
      n_genes <- nrow(specific_spatial_genes)
      for (i in 1:ceiling(n_genes / 6)) {
        start_idx <- (i - 1) * 6 + 1
        end_idx <- min(i * 6, n_genes)
        genes_batch <- specific_spatial_genes$gene[start_idx:end_idx]
        
        plot_data <- data.frame(Pseudotime = cells_target$Smoothed_Pseudotime)
        for (g in genes_batch) plot_data[[g]] <- GetAssayData(cells_target, assay = "SCT", layer = "data")[g, ]
        
        plot_long <- tidyr::pivot_longer(plot_data, cols = -Pseudotime, names_to = "Gene", values_to = "Expression")
        
        p <- ggplot(plot_long, aes(x = Pseudotime, y = Expression)) +
          geom_point(alpha = 0.3, size = 0.5, color = "grey60") +
          geom_smooth(method = "loess", color = "darkred", se = TRUE) +
          facet_wrap(~Gene, scales = "free_y", ncol = 3) +
          labs(title = paste(target_ct, "Specific Markers (", start_idx, "-", end_idx, "of", n_genes, ")")) +
          theme_minimal()
        print(p)
      }
      
      # Save results
      write.csv(specific_spatial_genes, 
                paste0("../Random data/celltype_specific_spatial_collapsed_", target_ct, ".csv"),
                row.names = FALSE)
    }
  }
}
```

## 7. Save Results

```{r save_results}
saveRDS(seurat_sc, "../Random data/sc_with_spatial_scores.rds")
message("Saved single-cell object with transferred spatial content.")

# 1. Set the identity to your cell type column
Idents(seurat_sc) <- "celltype"

# 2. Now subset using the identity you just set
subset_obj <- subset(seurat_sc, idents = "EE")

saveRDS(subset_obj, "../Random data/EE_with_spatial_scores.rds")
message("Saved single-cell object with transferred spatial content.")
```

## 8. Export for Research (Excel)

This section consolidates spatially cell-type specific genes for EE, ISC, EB, EC, and DC with specificity scores and direction of expression.

```{r research_export}
# 1. Define helper function for specificity
calculate_specificity <- function(seurat_obj, ct_name, is_collapsed = FALSE) {
  col_name <- if (is_collapsed) "celltype_collapsed" else "celltype"
  
  cells_target <- subset(seurat_obj, cells = colnames(seurat_obj)[seurat_obj[[col_name]] == ct_name])
  cells_other <- subset(seurat_obj, cells = colnames(seurat_obj)[seurat_obj[[col_name]] != ct_name])
  
  expr_target <- rowMeans(GetAssayData(cells_target, assay = "SCT", layer = "data"))
  expr_other <- rowMeans(GetAssayData(cells_other, assay = "SCT", layer = "data"))
  
  spec <- log2((expr_target + 0.01) / (expr_other + 0.01))
  return(spec)
}

# 2. Collect results for each requested type
target_types <- list(
  EE = list(obj = seurat_sc, results = dge_results_smoothed, collapsed = FALSE),
  ISC = list(obj = seurat_sc, results = dge_results_smoothed, collapsed = FALSE),
  EB = list(obj = seurat_sc, results = dge_results_smoothed, collapsed = FALSE),
  EC = list(obj = seurat_collapsed, results = dge_results_collapsed, collapsed = TRUE),
  DC = list(obj = seurat_collapsed, results = dge_results_collapsed, collapsed = TRUE)
)

all_research_genes <- list()

for (ct_name in names(target_types)) {
  info <- target_types[[ct_name]]
  
  if (ct_name %in% names(info$results)) {
    message(paste("Processing", ct_name, "for export..."))
    
    # Get DGE results
    ct_dge <- info$results[[ct_name]]
    
    # Calculate specificity
    ct_spec <- calculate_specificity(info$obj, ct_name, info$collapsed)
    ct_dge$specificity <- ct_spec[ct_dge$gene]
    
    # Filter for significant spatial genes
    # Criteria: padj < 0.05, |correlation| > 0.3, specificity > 1
    sig_spatial <- ct_dge %>% 
      dplyr::filter(padj < 0.05, abs(correlation) > 0.3, specificity > 1) %>%
      dplyr::arrange(desc(specificity)) %>%
      dplyr::mutate(
        Direction = ifelse(correlation > 0, "Increasing", "Decreasing"),
        Expression_Status = "Specific Marker"
      ) %>%
      dplyr::select(gene, celltype, correlation, Direction, pvalue, padj, mean_expr, specificity, Expression_Status)
    
    all_research_genes[[ct_name]] <- sig_spatial
  } else {
    message(paste("Warning: No DGE results found for", ct_name, "- skipping."))
  }
}

# 3. Consolidate and Save
if (length(all_research_genes) > 0) {
  final_df <- do.call(rbind, all_research_genes)
  
  # Export to Excel
  output_path <- "../Excels/sc_spatial_dec_markers.xlsx"
  write_xlsx(final_df, output_path)
  
  message(paste("Successfully exported spatial markers to", output_path))
  print(head(final_df))
} else {
  message("No significant spatial markers found for any requested cell type.")
}
```

## 9. Automated HTML Report Generation (Spatial Markers)

This chunk generates a standalone HTML report of spatial markers for the requested cell types (ISC, EE, EB, DC, EC).

```{r html_report_generation}
library(patchwork)
library(rmarkdown)

# 1. Configuration & Path Fix
report_types <- c("ISC", "EE", "EB", "DC", "EC")
project_root <- normalizePath("..")
output_dir <- file.path(project_root, "figures")
output_file <- file.path(output_dir, "spatial_markers_report.html")

# Automated Check
if (file.exists(output_file)) {
  message("--- INFO: HTML Report already exists at: ", output_file)
  message("--- To regenerate it, delete the file or set force_render <- TRUE")
  force_render <- FALSE 
} else {
  force_render <- TRUE
}

if (force_render) {
  message("Generating HTML report for:", paste(report_types, collapse = ", "))
  temp_rmd <- tempfile(fileext = ".Rmd")
  
  cat("---
title: 'Spatial Markers Report (Seurat Spatial Dec)'
date: '`r Sys.Date()`'
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
---

", file = temp_rmd)

  for (ct in report_types) {
    if (ct %in% names(target_types)) {
      data_entry <- target_types[[ct]]
      res <- data_entry$results[[ct]]
      
      # Calculate specificity for report filtering
      ct_spec <- calculate_specificity(data_entry$obj, ct, data_entry$collapsed)
      res$specificity <- ct_spec[res$gene]
      
      sig_markers <- res %>% 
        filter(padj < 0.05, abs(correlation) > 0.3, specificity > 1)
      sig_markers$gene <- gsub("[[:space:]]", "", as.character(sig_markers$gene))
      sig_markers <- sig_markers %>% arrange(desc(specificity))
      
      message(paste("---", ct, ": Found", nrow(sig_markers), "significant markers for report."))
      
      if (nrow(sig_markers) > 0) {
        cat(paste0("\n# ", ct, " {.tabset}\n\n"), file = temp_rmd, append = TRUE)
        n_markers <- nrow(sig_markers)
        batch_size <- 6
        n_batches <- ceiling(n_markers / batch_size)
        
        for (i in 1:n_batches) {
          start_idx <- (i - 1) * batch_size + 1
          end_idx <- min(i * batch_size, n_markers)
          markers_batch <- gsub("[[:space:]]", "", sig_markers$gene[start_idx:end_idx])
          
          cat(paste0("## Batch ", i, " (", start_idx, "-", end_idx, ")\n"), 
              file = temp_rmd, append = TRUE)
          chunk_header <- paste0("```{r plot_", ct, "_batch_", i, ", echo=FALSE, fig.width=12, fig.height=8, warning=FALSE, message=FALSE}\n")
          cat(chunk_header, file = temp_rmd, append = TRUE)
          cat(paste0("target_ct <- '", ct, "'\n"), file = temp_rmd, append = TRUE)
          cat(paste0("is_collapsed <- ", ifelse(data_entry$collapsed, "TRUE", "FALSE"), "\n"), file = temp_rmd, append = TRUE)
          
          cat("if (is_collapsed) {
            target_ids <- which(seurat_collapsed$celltype_collapsed == target_ct)
            cells_target <- if(length(target_ids) > 0) seurat_collapsed[, target_ids] else NULL
          } else {
            target_ids <- which(seurat_sc$celltype == target_ct)
            cells_target <- if(length(target_ids) > 0) seurat_sc[, target_ids] else NULL
          }\n", file = temp_rmd, append = TRUE)
          
          cat("if (!is.null(cells_target)) {
            plot_data <- data.frame(Pseudotime = cells_target$Smoothed_Pseudotime, row.names = colnames(cells_target))
            valid_markers <- c()\n", file = temp_rmd, append = TRUE)
          for (g in markers_batch) {
            cat(paste0("if ('", g, "' %in% rownames(GetAssayData(cells_target, assay = 'SCT'))) {
              plot_data[['", g, "']] <- GetAssayData(cells_target, assay = 'SCT', layer = 'data')['", g, "', ]
              valid_markers <- c(valid_markers, '", g, "')
            }\n"), file = temp_rmd, append = TRUE)
          }
          cat("if (length(valid_markers) > 0) {
            plot_list <- lapply(valid_markers, function(g) {
              ggplot(plot_data, aes(x = Pseudotime, y = .data[[g]])) +
                geom_point(alpha = 0.2, size = 0.5, color = 'grey70') +
                geom_smooth(method = 'loess', color = 'darkred', se = TRUE) +
                theme_minimal() + labs(title = g, y = 'Expr', x = 'Pseudotime')
            })
            patchwork::wrap_plots(plot_list, ncol = 3, nrow = 2)
          } else { print('No valid markers found.') }
        } else { print('No cells found.') }\n", file = temp_rmd, append = TRUE)
          cat("```\n\n", file = temp_rmd, append = TRUE)
        }
      }
    }
  }
  render(temp_rmd, output_file = output_file, quiet = FALSE, envir = globalenv())
}
```

## 10. Focused Granular EE Analysis (3D Spatial Mapping)

This section focuses specifically on the granular labels: **AstC-EE, EEP, classIII-EE, and Tk-EE**.

```{r granular_ee_3d_viz}
# 1. Temporarily Switch to Granular Identities
if ("celltype_granular" %in% colnames(seurat_sc@meta.data)) {
  message("Setting identities to celltype_granular for focused visualization...")
  temp_celltype <- seurat_sc$celltype
  seurat_sc$celltype <- seurat_sc$celltype_granular
  Idents(seurat_sc) <- "celltype"
}

target_granular_types <- c("AstC-EE", "EEP", "classIII-EE", "Tk-EE")
message(paste("Visualizing granular types:", paste(target_granular_types, collapse = ", ")))

# 3D Visualization
plot_3d_granular_df <- data.frame(
  UMAP1 = Embeddings(seurat_sc, "umap")[,1],
  UMAP2 = Embeddings(seurat_sc, "umap")[,2],
  Pseudotime = seurat_sc$Smoothed_Pseudotime,
  CellType = seurat_sc$celltype
)
plot_3d_granular_df_subset <- plot_3d_granular_df[plot_3d_granular_df$CellType %in% target_granular_types, ]

plot_ly(plot_3d_granular_df_subset, x = ~UMAP1, y = ~UMAP2, z = ~Pseudotime, 
        color = ~CellType, colors = "Set1", type = "scatter3d", mode = "markers",
        marker = list(size = 3, opacity = 0.8),
        text = ~paste("Type:", CellType, "<br>PT:", round(Pseudotime, 2))) %>%
  layout(title = "3D Granular EE & Progenitor Spatial Map",
         scene = list(xaxis = list(title = 'UMAP 1'),
                      yaxis = list(title = 'UMAP 2'),
                      zaxis = list(title = 'Inferred position')))

# Restore original identity column
if (exists("temp_celltype")) {
  seurat_sc$celltype <- temp_celltype
  Idents(seurat_sc) <- "celltype"
}
```
