---
title: "Spatial Centerline Projection & Tissue Painting"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load necessary libraries
library(Seurat)
library(png)
library(ggplot2)
library(dplyr)
library(mgcv)
library(viridis)
library(SpatialExperiment)
library(cowplot)
library(patchwork)
```

## 1. Parameters & Configuration

```{r config}
# Select the gene to plot
target_gene <- "CG17571
" # Change this gene as needed

# File paths
image_path <- "/home/mattia/Thesis/3D model/Immagine 2026-06-19 200559.png"
sc_path <- "/home/mattia/Thesis/03Slide4/Random data/sc_with_spatial_scores.rds"
spatial_path <- "/home/mattia/Thesis/03Slide4/Random data/merged_guts.rds"
geom_cache_path <- "/home/mattia/Thesis/03Slide4/Random data/tissue_geometry.rds"
```

## 2. Geometric Processing & Caching

We segment the midgut image, extract the smoothed centerline, and divide the entire tissue profile into 100 sections.

```{r geometry}
if (file.exists(geom_cache_path)) {
  cat("Loading cached tissue geometry...\n")
  geom <- readRDS(geom_cache_path)
  tissue_df <- geom$tissue
  centerline_smoothed <- geom$centerline
} else {
  cat("Processing tissue geometry from scratch...\n")
  
  # 1. Process Image
  img <- readPNG(image_path)
  if (length(dim(img)) == 3) {
    img_gray <- (img[,,1] + img[,,2] + img[,,3]) / 3
  } else {
    img_gray <- img
  }
  
  tissue_mask <- img_gray < 0.9
  coords <- which(tissue_mask, arr.ind = TRUE)
  tissue_df <- data.frame(x = coords[, "col"], y = coords[, "row"])
  
  centerline_raw <- tissue_df %>%
    group_by(x) %>%
    summarize(y = median(y), .groups = "drop")
  
  gam_model <- gam(y ~ s(x, bs = "cs", k = 20), data = centerline_raw)
  
  x_seq <- seq(min(centerline_raw$x), max(centerline_raw$x), length.out = 1000)
  centerline_smoothed <- data.frame(
    x = x_seq,
    y = predict(gam_model, newdata = data.frame(x = x_seq))
  )
  
  centerline_smoothed$dist <- 0
  for (i in 2:nrow(centerline_smoothed)) {
    dx <- centerline_smoothed$x[i] - centerline_smoothed$x[i-1]
    dy <- centerline_smoothed$y[i] - centerline_smoothed$y[i-1]
    centerline_smoothed$dist[i] <- centerline_smoothed$dist[i-1] + sqrt(dx^2 + dy^2)
  }
  
  max_dist <- max(centerline_smoothed$dist)
  centerline_smoothed$Section <- (centerline_smoothed$dist / max_dist) * 100
  
  # 2. Map Tissue Pixels
  cat("Mapping tissue pixels to centerline sections...\n")
  cl_x <- centerline_smoothed$x
  cl_y <- centerline_smoothed$y
  cl_sec <- centerline_smoothed$Section
  
  pixel_sections <- numeric(nrow(tissue_df))
  for (i in seq_len(nrow(tissue_df))) {
    dists <- (cl_x - tissue_df$x[i])^2 + (cl_y - tissue_df$y[i])^2
    pixel_sections[i] <- cl_sec[which.min(dists)]
  }
  tissue_df$Section <- pixel_sections
  
  # Cache geometry
  saveRDS(list(tissue = tissue_df, centerline = centerline_smoothed), geom_cache_path)
}

# Image dimension bounds for perfect alignment
img <- readPNG(image_path)
img_h <- nrow(img)
img_w <- ncol(img)

# Setup original image plot precisely scaled to coordinate grid
p_orig <- ggplot() +
  annotation_raster(img, xmin = 1, xmax = img_w, ymin = img_h, ymax = 1) +
  coord_fixed(xlim = c(1, img_w), ylim = c(img_h, 1)) +
  scale_y_reverse() +
  theme_void() +
  labs(title = "Original Histology")

# Plot spatial scores (0-100) mapped over the tissue
ggplot(tissue_df, aes(x = x, y = y, color = Section)) +
  geom_point(size = 0.5, alpha = 0.5) +
  geom_path(data = centerline_smoothed, aes(x = x, y = y), color = "red", linewidth = 1.5) +
  scale_color_viridis(option = "turbo") +
  coord_fixed(xlim = c(1, img_w), ylim = c(img_h, 1)) +
  scale_y_reverse() +
  theme_minimal() +
  labs(title = "Tissue Profile Colored by Section Score (0-100)", x = "Anterior -> Posterior", y = "Dorsal/Ventral")
```

## 3. Painting Tissue from Single-Cell Data

```{r paint_sc, fig.height=8, fig.width=10}
# Load Single Cell Object
if (!exists("sc_obj")) { sc_obj <- readRDS(sc_path) }

sc_expr <- FetchData(sc_obj, vars = target_gene)
sc_df <- data.frame(
  Pseudotime = sc_obj@meta.data$Smoothed_Pseudotime,
  Expression = sc_expr[[1]]
)

cat(paste("Fitting SC expression model for", target_gene, "...\n"))
expr_gam_sc <- gam(Expression ~ s(Pseudotime, bs = "cs"), data = sc_df)

section_bins <- data.frame(Section = seq(0, 100, length.out = 101))
section_bins$SC_Predicted_Expr <- predict(expr_gam_sc, newdata = data.frame(Pseudotime = section_bins$Section))
section_bins$SC_Predicted_Expr[section_bins$SC_Predicted_Expr < 0] <- 0
section_bins$SC_ZScore <- as.numeric(scale(section_bins$SC_Predicted_Expr))
if(any(is.na(section_bins$SC_ZScore))) section_bins$SC_ZScore[is.na(section_bins$SC_ZScore)] <- 0

# Merge and Plot
tissue_df$Rounded_Section <- round(tissue_df$Section)
tissue_painted_sc <- tissue_df %>%
  left_join(section_bins, by = c("Rounded_Section" = "Section"))

p_tissue_sc <- ggplot(tissue_painted_sc, aes(x = x, y = y, color = SC_ZScore)) +
  geom_point(size = 0.5, alpha = 0.8) +
  scale_color_viridis(option = "magma", name = "SC Expr (Z-Score)", limits = c(0, 2), oob = scales::squish) +
  coord_fixed(xlim = c(1, img_w), ylim = c(img_h, 1)) +
  scale_y_reverse() +
  theme_void() +
  labs(title = paste("Painted Tissue (Single-Cell):", target_gene))

# Perfectly align the plots using patchwork
(p_orig / p_tissue_sc) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

## 4. Painting Tissue from Spatial Data

```{r paint_spatial, fig.height=8, fig.width=10}
# Load Spatial Object
if (!exists("spatial_obj")) { spatial_obj <- readRDS(spatial_path) }

# Extract expression and spatial pseudotime
if (inherits(spatial_obj, "SpatialExperiment")) {
  library(SummarizedExperiment)
  if (target_gene %in% rownames(spatial_obj)) {
    if ("logcounts" %in% assayNames(spatial_obj)) {
      expr_vals <- as.numeric(assay(spatial_obj, "logcounts")[target_gene, ])
    } else {
      expr_vals <- as.numeric(assay(spatial_obj, "counts")[target_gene, ])
    }
  } else {
    expr_vals <- rep(0, ncol(spatial_obj))
  }
  sp_pseudotime <- colData(spatial_obj)[["Spatial_Pseudotime"]]
} else {
  expr_vals <- FetchData(spatial_obj, vars = target_gene)[[1]]
  sp_pseudotime <- spatial_obj@meta.data$Spatial_Pseudotime
}

spatial_df <- data.frame(
  Pseudotime = sp_pseudotime,
  Expression = expr_vals
)

cat(paste("Fitting Spatial expression model for", target_gene, "...\n"))
expr_gam_sp <- gam(Expression ~ s(Pseudotime, bs = "cs"), data = spatial_df)

section_bins$SP_Predicted_Expr <- predict(expr_gam_sp, newdata = data.frame(Pseudotime = section_bins$Section))
section_bins$SP_Predicted_Expr[section_bins$SP_Predicted_Expr < 0] <- 0
section_bins$SP_ZScore <- as.numeric(scale(section_bins$SP_Predicted_Expr))
if(any(is.na(section_bins$SP_ZScore))) section_bins$SP_ZScore[is.na(section_bins$SP_ZScore)] <- 0

# Merge and Plot
tissue_painted_sp <- tissue_df %>%
  left_join(section_bins, by = c("Rounded_Section" = "Section"))

p_tissue_sp <- ggplot(tissue_painted_sp, aes(x = x, y = y, color = SP_ZScore)) +
  geom_point(size = 0.5, alpha = 0.8) +
  scale_color_viridis(option = "magma", name = "Spatial Expr (Z-Score)", limits = c(0, 2), oob = scales::squish) +
  coord_fixed(xlim = c(1, img_w), ylim = c(img_h, 1)) +
  scale_y_reverse() +
  theme_void() +
  labs(title = paste("Painted Tissue (Spatial):", target_gene))

# Perfectly align the plots using patchwork
(p_orig / p_tissue_sp) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

## 5. Painting Tissue from a Gene List (SC and Spatial)

This section takes a list of genes, loops through them, and generates an aligned Single-Cell and Spatial plot for each gene. Plots are also saved to the figures directory.

```{r paint_gene_list, fig.height=12, fig.width=10}
# Load gene list from the TF reference file
tf_file <- "/mnt/c/Users/matti/Desktop/Master/Thesis/Reference genome/FlyBase_Fields_download.txt"
tf_data <- read.delim(tf_file, header = TRUE, sep = "\t", stringsAsFactors = FALSE, check.names=FALSE)

# The third column is SYMBOL. Proceed as previously: full capital, remove '-'
symbols <- tf_data[[3]]
gene_list <- toupper(gsub("-", "", symbols))

# Ensure objects are loaded
if (!exists("sc_obj")) { sc_obj <- readRDS(sc_path) }

# Remove ISC, EB, dEC, and daEC from single cell data before plotting
cells_to_remove <- c("ISC", "EB", "dEC", "daEC")
if (any(sc_obj$celltype %in% cells_to_remove)) {
  sc_obj_plot <- subset(sc_obj, subset = celltype %in% cells_to_remove == FALSE)
} else {
  sc_obj_plot <- sc_obj
}

if (!exists("spatial_obj")) { spatial_obj <- readRDS(spatial_path) }

tissue_df$Rounded_Section <- round(tissue_df$Section)

# Create output directory for these specific plots
dir.create("../figures/centerline_projections", showWarnings = FALSE, recursive = TRUE)

for (current_gene in gene_list) {
  cat(paste("\n==================================================\n"))
  cat(paste("Processing Gene:", current_gene, "\n"))
  
  # --- 5.1 Single-Cell Score ---
  if (current_gene %in% rownames(sc_obj_plot)) {
    sc_expr <- FetchData(sc_obj_plot, vars = current_gene)[[1]]
    sc_df <- data.frame(
      Pseudotime = sc_obj_plot@meta.data$Smoothed_Pseudotime,
      Expression = sc_expr
    )
    
    expr_gam_sc <- gam(Expression ~ s(Pseudotime, bs = "cs"), data = sc_df)
    
    section_bins <- data.frame(Section = seq(0, 100, length.out = 101))
    section_bins$SC_Predicted_Expr <- predict(expr_gam_sc, newdata = data.frame(Pseudotime = section_bins$Section))
    section_bins$SC_Predicted_Expr[section_bins$SC_Predicted_Expr < 0] <- 0
    section_bins$SC_ZScore <- as.numeric(scale(section_bins$SC_Predicted_Expr))
    if(any(is.na(section_bins$SC_ZScore))) section_bins$SC_ZScore[is.na(section_bins$SC_ZScore)] <- 0
  } else {
    cat(paste("Gene", current_gene, "not found in Single-Cell object. Skipping SC plot.\n"))
    section_bins <- data.frame(Section = seq(0, 100, length.out = 101), SC_ZScore = 0)
  }
  
  # --- 5.2 Spatial Score ---
  sp_found <- FALSE
  if (inherits(spatial_obj, "SpatialExperiment")) {
    library(SummarizedExperiment)
    if (current_gene %in% rownames(spatial_obj)) {
      if ("logcounts" %in% assayNames(spatial_obj)) {
        expr_vals <- as.numeric(assay(spatial_obj, "logcounts")[current_gene, ])
      } else {
        expr_vals <- as.numeric(assay(spatial_obj, "counts")[current_gene, ])
      }
      sp_found <- TRUE
    }
    sp_pseudotime <- colData(spatial_obj)[["Spatial_Pseudotime"]]
  } else {
    if (current_gene %in% rownames(spatial_obj)) {
      expr_vals <- FetchData(spatial_obj, vars = current_gene)[[1]]
      sp_found <- TRUE
    }
    sp_pseudotime <- spatial_obj@meta.data$Spatial_Pseudotime
  }
  
  if (sp_found) {
    spatial_df <- data.frame(Pseudotime = sp_pseudotime, Expression = expr_vals)
    expr_gam_sp <- gam(Expression ~ s(Pseudotime, bs = "cs"), data = spatial_df)
    
    section_bins$SP_Predicted_Expr <- predict(expr_gam_sp, newdata = data.frame(Pseudotime = section_bins$Section))
    section_bins$SP_Predicted_Expr[section_bins$SP_Predicted_Expr < 0] <- 0
    section_bins$SP_ZScore <- as.numeric(scale(section_bins$SP_Predicted_Expr))
    if(any(is.na(section_bins$SP_ZScore))) section_bins$SP_ZScore[is.na(section_bins$SP_ZScore)] <- 0
  } else {
    cat(paste("Gene", current_gene, "not found in Spatial object. Skipping SP plot.\n"))
    if (!"SP_ZScore" %in% colnames(section_bins)) {
      section_bins$SP_ZScore <- 0
    }
  }
  
  # --- 5.3 Merge and Plot ---
  if (current_gene %in% rownames(sc_obj_plot) || sp_found) {
    tissue_painted <- tissue_df %>%
      left_join(section_bins, by = c("Rounded_Section" = "Section"))
    
    p_tissue_sc <- ggplot(tissue_painted, aes(x = x, y = y, color = SC_ZScore)) +
      geom_point(size = 0.5, alpha = 0.8) +
      scale_color_viridis(option = "magma", name = "SC Expr (Z-Score)", limits = c(0, 2), oob = scales::squish) +
      coord_fixed(xlim = c(1, img_w), ylim = c(img_h, 1)) +
      scale_y_reverse() +
      theme_void() +
      labs(title = paste("Painted Tissue (Single-Cell):", current_gene))
    
    p_tissue_sp <- ggplot(tissue_painted, aes(x = x, y = y, color = SP_ZScore)) +
      geom_point(size = 0.5, alpha = 0.8) +
      scale_color_viridis(option = "magma", name = "Spatial Expr (Z-Score)", limits = c(0, 2), oob = scales::squish) +
      coord_fixed(xlim = c(1, img_w), ylim = c(img_h, 1)) +
      scale_y_reverse() +
      theme_void() +
      labs(title = paste("Painted Tissue (Spatial):", current_gene))
    
    # Perfectly align the plots using patchwork (3 panels)
    combined_plot <- (p_orig / p_tissue_sc / p_tissue_sp) + 
      plot_layout(guides = "collect") + 
      plot_annotation(title = paste("Expression Profile:", current_gene), 
                      theme = theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5))) & 
      theme(legend.position = "bottom")
    
    # Save the plot
    out_file <- paste0("../figures/centerline_projections/Spatial_Profile_", current_gene, ".png")
    ggsave(out_file, combined_plot, width = 10, height = 12, bg = "white")
    cat("Saved plot to:", out_file, "\n")
    
    # Print the plot to the notebook output
    print(combined_plot)
    
  } else {
    cat(paste("Gene", current_gene, "not found in either object. Skipping plot.\n"))
  }
}
```
```
