---
title: "3D UMAP Analysis (PCA + Spatial Pseudotime)"
output: html_document
date: "2026-01-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Seurat)
library(monocle3)
library(plotly)
library(dplyr)
library(ggplot2)
library(shiny)
library(sf)
library(viridis)
```

## 1. Setup & Load Data
Load the segmented Seurat object from the saved RDS file.

```{r setup_env}
if (!exists("visium_seg")) {
  message("Loading visium_seg from saved RDS...")
  vhd_list <- readRDS("/home/mattia/Thesis/03Slide4/Random data/VisiumHD_Drosophila_Resolution_List.rds")
  visium_seg <- vhd_list$seg
  rm(vhd_list)
  message(paste("Loaded visium_seg with", ncol(visium_seg), "cells."))
} else {
  message(paste("Found visium_seg in environment with", ncol(visium_seg), "cells."))
}
```

## 2. Prepare Data for Shiny App
Extract coordinates from the segmented object to use in the interactive selector.

```{r prepare_shiny_data}
# 1. Get Coordinates from Metadata (Robustly calculated in Seg_data.Rmd)
coords_seg <- visium_seg@meta.data[, c("x_centroid", "y_centroid")]
colnames(coords_seg) <- c("x", "y")
coords_seg$barcode <- rownames(coords_seg)

print(paste("Ready for Shiny: ", nrow(coords_seg), "cells."))
head(coords_seg)
```

## 2. Interactive Gut Subsetting

Run this chunk to launch the Shiny app and select the 3 guts.

```{r shiny_app}
# Define UI
ui <- fluidPage(
  titlePanel("Gut Subsetting Tool"),
  sidebarLayout(
    sidebarPanel(
      h4("Instructions"),
      p("1. Use the Lasso tool (top visual toolbar) to select points."),
      p("2. Click 'Save Gut X' to store the selection."),
      actionButton("save_gut_1", "Save Gut 1"),
      verbatimTextOutput("status_1"),
      hr(),
      actionButton("save_gut_2", "Save Gut 2"),
      verbatimTextOutput("status_2"),
      hr(),
      actionButton("done_btn", "DONE (Close & Save)"),
      class = "well"
    ),
    mainPanel(
      plotlyOutput("scatter_plot", height = "800px")
    )
  )
)

# Define Server
server <- function(input, output, session) {
  
  # Store selections
  vals <- reactiveValues(
    gut_1 = NULL,
    gut_2 = NULL
  )
  
  output$scatter_plot <- renderPlotly({
    plot_ly(coords_seg, x = ~x, y = ~y, key = ~barcode, type = "scatter", mode = "markers",
            marker = list(size = 3, color = 'gray', opacity = 0.5), source = "subset_source") %>%
      layout(dragmode = "lasso", title = "Select Gut Regions", yaxis = list(autorange = "reversed"))
  })
  
  # Save Gut 1
  observeEvent(input$save_gut_1, {
    event_data <- event_data("plotly_selected", source = "subset_source")
    if (!is.null(event_data)) {
      vals$gut_1 <- event_data$key
    }
  })
  output$status_1 <- renderText({ 
    if(is.null(vals$gut_1)) "Not saved" else paste(length(vals$gut_1), "cells saved")
  })
  
  # Save Gut 2
  observeEvent(input$save_gut_2, {
    event_data <- event_data("plotly_selected", source = "subset_source")
    if (!is.null(event_data)) {
      vals$gut_2 <- event_data$key
    }
  })
  output$status_2 <- renderText({ 
    if(is.null(vals$gut_2)) "Not saved" else paste(length(vals$gut_2), "cells saved")
  })
  
  # Done
  observeEvent(input$done_btn, {
    stopApp(list(gut_1 = vals$gut_1, gut_2 = vals$gut_2))
  })
}

# Run App
subset_res <- runApp(shinyApp(ui, server))

# Subset Objects
# DEBUG: Check if barcodes match
cat("--- Barcode Debugging ---\n")
cat("First 5 barcodes in object:\n")
print(head(Cells(visium_seg)))
cat("First 5 barcodes in Gut 1 selection (Raw):\n")
print(head(subset_res$gut_1))

# Helper to fix common selection issues
clean_barcodes <- function(selection, all_cells) {
  if (is.null(selection)) return(NULL)
  
  # 1. Clean whitespace
  selection <- trimws(selection)
  
  # 2. Try to Format (if they look like simple numbers)
  # The debug showed "1307", but object has "cellid_000001307-1".
  # We attempt to convert raw ID to formatted ID.
  normalized_selection <- tryCatch({
    sprintf("cellid_%09d-1", as.numeric(selection))
  }, warning = function(w) {
    return(selection) # If not numeric, keep original
  }, error = function(e) {
    return(selection)
  })
  
  # 3. Check intersection with BOTH raw and formatted
  # This allows it to work whether the selection is already correct or needs formatting
  valid_cells_raw <- intersect(selection, all_cells)
  valid_cells_fmt <- intersect(normalized_selection, all_cells)
  
  if (length(valid_cells_fmt) > length(valid_cells_raw)) {
    cat(paste("Formatting applied. Selected:", length(selection), "-> Valid:", length(valid_cells_fmt), "\n"))
    return(valid_cells_fmt)
  } else {
    cat(paste("Raw used. Selected:", length(selection), "-> Valid:", length(valid_cells_raw), "\n"))
    return(valid_cells_raw)
  }
}

cells_1 <- clean_barcodes(subset_res$gut_1, Cells(visium_seg))
cells_2 <- clean_barcodes(subset_res$gut_2, Cells(visium_seg))

if (length(cells_1) > 0) spe_gut_1 <- subset(visium_seg, cells = cells_1)
if (length(cells_2) > 0) spe_gut_2 <- subset(visium_seg, cells = cells_2)

print("Subsets created successfully (if valid cells were found).")
```

## 3. Analysis Function (PCA + Spatial Pseudotime)

We define a function to process each gut individually.

```{r analysis_function}
process_gut_3d <- function(seurat_obj, gut_label) {
  message(paste("Processing", gut_label, "..."))
  
  # --- A. Preprocessing & PCA (Seurat) ---
  seurat_obj <- NormalizeData(seurat_obj)
  seurat_obj <- FindVariableFeatures(seurat_obj)
  seurat_obj <- ScaleData(seurat_obj)
  
  # Prevent SVD error by dynamically setting npcs
  max_dims <- min(30, ncol(seurat_obj) - 1)
  seurat_obj <- RunPCA(seurat_obj, npcs = max_dims, verbose = FALSE)
  
  # --- B. Spatial Trajectory (Monocle3) ---
  # 1. Convert to CDS
  # (Simplified conversion wrapper)
  expr_matrix <- GetAssayData(seurat_obj, layer = "counts")
  cell_meta <- seurat_obj@meta.data
  gene_meta <- data.frame(gene_short_name = rownames(expr_matrix), row.names = rownames(expr_matrix))
  
  cds <- new_cell_data_set(expr_matrix, cell_metadata = cell_meta, gene_metadata = gene_meta)
  
  # 2. Inject Spatial Coordinates into 'UMAP' slot to force spatial trajectory
  # Use the robust centroids if available
  if (all(c("x_centroid", "y_centroid") %in% colnames(seurat_obj@meta.data))) {
    coords <- seurat_obj@meta.data[, c("x_centroid", "y_centroid")]
  } else {
    coords <- GetTissueCoordinates(seurat_obj)[, c(1,2)]
  }
  colnames(coords) <- c("x", "y")
  reducedDims(cds)$UMAP <- as.matrix(coords)
  
  # 3. Learn Graph
  cds <- cluster_cells(cds, reduction_method = "UMAP") 
  cds@clusters$UMAP$partitions[] <- factor("1")   # Merge partitions to bridge gap
  cds <- learn_graph(cds, use_partition = FALSE)
  
  # 4. Order Cells (Root = Leftmost / Min X)
  root_cells <- rownames(coords)[which.min(coords[, 1])]
  cds <- order_cells(cds, root_cells = root_cells)
  
  # 5. Extract Pseudotime & COLLAPSE FRACTURES
  # Using rank() instead of distance ensures that 'jumps' across empty gaps 
  # are ignored, focusing only on the sequence of existing cells.
  pt <- pseudotime(cds)
  pt_ranked <- rank(pt, ties.method = "min", na.last = "keep")
  pt_scaled <- (pt_ranked - min(pt_ranked, na.rm=T)) / (max(pt_ranked, na.rm=T) - min(pt_ranked, na.rm=T)) * 100
  
  # --- C. Assemble Data for 3D Plot ---
  embeddings <- Embeddings(seurat_obj, "pca")[, 1:2]
  
  bg_data <- data.frame(
    Barcode = rownames(embeddings),
    PC1 = embeddings[, 1],
    PC2 = embeddings[, 2],
    Pseudotime = pt_scaled,
    Gut = gut_label
  )
  
  return(bg_data)
}
```



## 5. Run Analysis Loop

```{r run_loop}
df_gut_1 <- process_gut_3d(spe_gut_1, "Gut 1")
df_gut_2 <- process_gut_3d(spe_gut_2, "Gut 2")

# Combine (Optional, but we plot separate 3D plots usually)
# all_guts3d <- rbind(df_gut_1, df_gut_2)
```

## 6. Technical QC Filtering
We apply standardized thresholds (Mito < 20%, Ribo < 30%, UMI > 200) to remove low-quality cells AFTER trajectory calculation to prevent graph fracturing.

```{r post_trajectory_filtering}
# Function to filter both the object and the plotting dataframe
filter_qc_post_trajectory <- function(seurat_obj, plotting_df, name) {
  if (is.null(seurat_obj) || is.null(plotting_df)) return(NULL)
  
  message(paste("\n--- Final QC Filtering:", name, "---"))
  
  # 1. Ensure QC metrics exist in object
  if (!"percent.mt" %in% colnames(seurat_obj@meta.data)) {
    seurat_obj[["percent.mt"]] <- PercentageFeatureSet(seurat_obj, pattern = "^mt:")
  }
  if (!"percent.ribo" %in% colnames(seurat_obj@meta.data)) {
    seurat_obj[["percent.ribo"]] <- PercentageFeatureSet(seurat_obj, pattern = "^Rp[LS]")
  }
  
  # 2. Identify passing cells
  count_var <- if("nCount_Segmented" %in% colnames(seurat_obj@meta.data)) "nCount_Segmented" else "nCount_Spatial"
  
  passing_cells <- rownames(seurat_obj@meta.data)[
    seurat_obj@meta.data$percent.mt < 20 & 
    seurat_obj@meta.data$percent.ribo < 30 & 
    seurat_obj@meta.data[[count_var]] > 1
  ]
  
  n_removed <- ncol(seurat_obj) - length(passing_cells)
  message(paste("Removed", n_removed, "low-quality cells."))
  
  # 3. Filter Object
  seurat_obj_filtered <- subset(seurat_obj, cells = passing_cells)
  
  # 4. Filter Plotting DF
  # (plotting_df uses 'Barcode' column for mapping)
  plotting_df_filtered <- plotting_df[plotting_df$Barcode %in% passing_cells, ]
  
  return(list(obj = seurat_obj_filtered, df = plotting_df_filtered))
}

# Run for all selected guts
if (exists("spe_gut_1") && exists("df_gut_1")) {
  res1 <- filter_qc_post_trajectory(spe_gut_1, df_gut_1, "Gut 1")
  spe_gut_1 <- res1$obj; df_gut_1 <- res1$df
}
if (exists("spe_gut_2") && exists("df_gut_2")) {
  res2 <- filter_qc_post_trajectory(spe_gut_2, df_gut_2, "Gut 2")
  spe_gut_2 <- res2$obj; df_gut_2 <- res2$df
}
```

## 7. Generate 3D Plots

### Gut 1 3D Map
X: PC1, Y: PC2, Z: Pseudotime (Spatial)
Color: Pseudotime

```{r plot_gut1}
plot_ly(df_gut_1, x = ~PC1, y = ~PC2, z = ~Pseudotime, 
        type = 'scatter3d', mode = 'markers',
        marker = list(
          size = 3, 
          color = ~Pseudotime, 
          colorscale = 'Magma', 
          reversescale = TRUE,
          colorbar = list(title = "Pseudotime")
        )) %>%
  layout(title = "Gut 1: 3D UMAP (PC1/PC2/Spatial)", 
         scene = list(
           xaxis = list(title = 'PC 1'),
           yaxis = list(title = 'PC 2'),
           zaxis = list(title = 'Spatial Pseudotime')
         ))
```

### Gut 2 3D Map

```{r plot_gut2}
plot_ly(df_gut_2, x = ~PC1, y = ~PC2, z = ~Pseudotime, 
        type = 'scatter3d', mode = 'markers',
        marker = list(
          size = 3, 
          color = ~Pseudotime, 
          colorscale = 'Magma', 
          reversescale = TRUE,
          colorbar = list(title = "Pseudotime")
        )) %>%
  layout(title = "Gut 2: 3D UMAP (PC1/PC2/Spatial)", 
         scene = list(
           xaxis = list(title = 'PC 1'),
           yaxis = list(title = 'PC 2'),
           zaxis = list(title = 'Spatial Pseudotime')
         ))
```

## 8. Label Transfer (All Guts)
We predict cell types using the single-cell reference. We define a helper function to avoid code duplication.

```{r label_transfer_all}
# 1. Load Reference Once
ref_path <- "/home/mattia/Thesis/Single-cell/seurat_ctrl_with_metadata.rds"

if (file.exists(ref_path)) {
  message("Loading Reference Dataset...")
  sc_ref <- readRDS(ref_path)
  
  # Ensure Reference uses SCT
  # Use names() check to be safe
  existing_assays <- tryCatch(names(sc_ref@assays), error = function(e) Seurat::Assays(sc_ref))
  if (!"SCT" %in% existing_assays) {
    message("SCTransforming Reference...")
    sc_ref <- SCTransform(sc_ref, verbose = FALSE)
  }
} else {
  stop(paste("Reference file not found at:", ref_path))
}

# 2. Define Transfer Function
transfer_labels <- function(query_obj, plotting_df, reference_obj) {
  
  # A. Prepare Query
  # Explicitly use Segmented assay and run PCA
  message("SCTransforming & PCA on Query...")
  query_obj <- SCTransform(query_obj, assay = "Segmented", verbose = FALSE)
  
  # Prevent SVD error by dynamically setting npcs
  max_dims_query <- min(50, ncol(query_obj) - 1)
  query_obj <- RunPCA(query_obj, npcs = max_dims_query, verbose = FALSE)
  
  # B. Find Anchors
  message("Finding Anchors...")
  anchors <- FindTransferAnchors(
    reference = reference_obj,
    query = query_obj,
    normalization.method = "SCT",
    reference.assay = "SCT",
    query.assay = "SCT",
    dims = 1:20
  )
  
  # C. Transfer Labels
  message("Transferring Labels...")
  predictions <- TransferData(
    anchorset = anchors,
    refdata = reference_obj@meta.data[["celltype"]],
    prediction.assay = TRUE,
    weight.reduction = query_obj[["pca"]],
    dims = 1:20
  )
  
  # D. Extract Winner with Confidence Filtering
  predicted_labels <- predictions@data
  if("max" %in% rownames(predicted_labels)) {
      predicted_labels <- predicted_labels[rownames(predicted_labels) != "max", ]
  }
  
  # Calculate confidence scores (max prediction score per cell)
  max_scores <- apply(predicted_labels, 2, max)
  max_idx <- apply(predicted_labels, 2, which.max)
  predicted_types <- rownames(predicted_labels)[max_idx]
  
  # Filter low-confidence predictions
  confidence_threshold <- 0.05
  low_confidence <- max_scores < confidence_threshold
  predicted_types[low_confidence] <- "Ambiguous"
  
  message(paste("  Filtered", sum(low_confidence), "low-confidence predictions (<", confidence_threshold, ")"))
  
  # E. Match to Plotting DataFrame
  target_barcodes <- as.character(plotting_df$Barcode)
  source_barcodes <- colnames(predicted_labels)
  
  # Handle potential NULL colnames
  if (is.null(source_barcodes)) source_barcodes <- names(predicted_labels)
  
  # Assign
  plotting_df$Predicted_Type <- predicted_types[match(target_barcodes, source_barcodes)]
  
  return(plotting_df)
}

# 3. Apply to All Guts
message("--- Processing Gut 1 ---")
df_gut_1 <- transfer_labels(spe_gut_1, df_gut_1, sc_ref)

message("--- Processing Gut 2 ---")
df_gut_2 <- transfer_labels(spe_gut_2, df_gut_2, sc_ref)

message("All Label Transfers Complete.")
```

### Gut 1 3D Map (Cell Type)

```{r plot_gut1_celltype}
plot_ly(df_gut_1, x = ~PC1, y = ~PC2, z = ~Pseudotime, 
      type = 'scatter3d', mode = 'markers',
      color = ~Predicted_Type, colors = "Set3", marker = list(size = 3)
    ) %>%
layout(title = "Gut 1: 3D UMAP (Cell Type)", 
       scene = list(xaxis = list(title = 'PC 1'), yaxis = list(title = 'PC 2'), zaxis = list(title = 'Pseudotime')))
```

### Gut 2 3D Map (Cell Type)

```{r plot_gut2_celltype}
plot_ly(df_gut_2, x = ~PC1, y = ~PC2, z = ~Pseudotime, 
      type = 'scatter3d', mode = 'markers',
      color = ~Predicted_Type, colors = "Set3", marker = list(size = 3)
    ) %>%
layout(title = "Gut 2: 3D UMAP (Cell Type)", 
       scene = list(xaxis = list(title = 'PC 1'), yaxis = list(title = 'PC 2'), zaxis = list(title = 'Pseudotime')))
```

## 9. Spatial Differential Expression Analysis
We merge the three gut sections and compare "Anterior" vs "Posterior" expression profiles within each cell type.

```{r spatial_de_analysis}
message("Starting Spatial DE Analysis...")

# --- Helper Function for DE ---
run_spatial_de <- function(seurat_obj, analysis_df, name) {
  
  message(paste0("\n=== Processing ", name, " ==="))
  
  # 1. Clean Object (Remove Images to prevent subset errors)
  # Iterate over image names to safely remove them
  for (img in Images(seurat_obj)) {
    seurat_obj[[img]] <- NULL
  }
  
  # 2. Join Layers if needed (Seurat v5)
  if (inherits(seurat_obj, "Seurat")) {
    seurat_obj <- JoinLayers(seurat_obj)
  }
  
  # 3. Map Metadata from Analysis DataFrame
  # Ensure we can match them (Barcodes should match)
  # Check alignment
  common_cells <- intersect(Cells(seurat_obj), analysis_df$Barcode)
  
  if (length(common_cells) < 10) {
    warning(paste("Very few matching cells for", name, "- Check Barcodes!"))
    return(NULL)
  }
  
  # Subset both to common cells to be safe
  seurat_obj <- subset(seurat_obj, cells = common_cells)
  analysis_df <- analysis_df[analysis_df$Barcode %in% common_cells, ]
  
  # Re-order df to match object
  analysis_df <- analysis_df[match(Cells(seurat_obj), analysis_df$Barcode), ]
  
  # Assign Metadata
  seurat_obj$Spatial_Pseudotime <- analysis_df$Pseudotime
  seurat_obj$Predicted_CellType <- analysis_df$Predicted_Type
  
  # 4. Define Spatial Regions (Threshold: 55)
  seurat_obj$Spatial_Region <- ifelse(seurat_obj$Spatial_Pseudotime < 55, "Anterior", "Posterior")
  
  # 5. DE Loop
  Idents(seurat_obj) <- seurat_obj$Predicted_CellType
  unique_types <- unique(na.omit(seurat_obj$Predicted_CellType))
  
  results_list <- list()
  
  for (ctype in unique_types) {
    # Subset to Cell Type
    sub_obj <- subset(seurat_obj, idents = ctype)
    
    # Count cells in Anterior/Posterior
    counts <- table(sub_obj$Spatial_Region)
    
    if (!is.na(counts["Anterior"]) && !is.na(counts["Posterior"]) && 
        counts["Anterior"] > 3 && counts["Posterior"] > 3) {
      
      message(paste("  Running DE for:", ctype, "(Ant:", counts["Anterior"], "Post:", counts["Posterior"], ")"))
      
      tryCatch({
        markers <- FindMarkers(
          sub_obj,
          ident.1 = "Anterior",
          ident.2 = "Posterior",
          group.by = "Spatial_Region",
          min.pct = 0.1,
          logfc.threshold = 0.25,
          verbose = FALSE
        )
        
        if (nrow(markers) > 0) {
          markers$CellType <- ctype
          markers$Gene <- rownames(markers)
          markers$Gut_Origin <- name
          results_list[[ctype]] <- markers
        }
        
      }, error = function(e) {
        message(paste("    Error:", e$message))
      })
    }
  }
  
  # 6. Save & Return
  if (length(results_list) > 0) {
    all_res <- do.call(rbind, results_list)
    
    # Calculate Specificity
    sig_res <- all_res[all_res$p_val_adj < 0.05, ]
    if (nrow(sig_res) > 0) {
      gene_counts <- table(sig_res$Gene)
      all_res$CellType_Count <- as.numeric(gene_counts[all_res$Gene])
      all_res$CellType_Count[is.na(all_res$CellType_Count)] <- 0
      all_res$Specificity <- ifelse(all_res$CellType_Count == 1, "Unique (1 CellType)", 
                                    paste0("Shared (", all_res$CellType_Count, " CellTypes)"))
    } else {
        all_res$Specificity <- "Not Significant"
    }

    write.csv(all_res, paste0("../Excels/Spatial_DE_", name, ".csv"))
    message(paste("  Saved results (with specificity) to ../Excels/Spatial_DE_", name, ".csv", sep=""))
    
    # NEW: Save merged guts for sc_spatial_dec analysis
    if (name == "Merged_Guts") {
      saveRDS(seurat_obj, "../Random data/merged_guts_spatial.rds")
      message("Saved merged guts to ../Random data/merged_guts_spatial.rds")
    }
    
    return(list(results = all_res, object = seurat_obj))
  } else {
    message("  No significant DE genes found.")
    return(NULL)
  }
}

# --- A. Run for Individual Guts ---
# Ensure metadata columns exist in individual dataframes as expected
res_gut1 <- run_spatial_de(spe_gut_1, df_gut_1, "Gut1")
res_gut2 <- run_spatial_de(spe_gut_2, df_gut_2, "Gut2")

# --- SAVE GUT 1 OBJECT FOR DEDICATED ANALYSIS ---
if (!is.null(res_gut1)) {
  saveRDS(res_gut1$object, file = "../Random data/gut1_processed_for_analysis.rds")
  message("Saved processed Gut 1 object to ../Random data/gut1_processed_for_analysis.rds")
}

# --- B. Run for Merged Dataset ---
message("\n=== Preparing Merged Dataset ===")
# Merge Objects
merged_guts <- merge(spe_gut_1, y = spe_gut_2, add.cell.ids = c("Gut1", "Gut2"))

# Prepare Merged Metadata DataFrame
# (Need to adjust barcodes in DF to match the merged object's "Origin_Barcode" format)
df_1_mod <- df_gut_1
df_1_mod$Unique_Barcode <- paste0("Gut1_", df_1_mod$Barcode)

df_2_mod <- df_gut_2
df_2_mod$Unique_Barcode <- paste0("Gut2_", df_2_mod$Barcode)

all_metadata <- rbind(df_1_mod, df_2_mod)
# IMPORTANT: For the merged run, we must use the Unique_Barcode as the 'Barcode' column
# because that's what matches Cells(merged_guts)
all_metadata_merged <- all_metadata
all_metadata_merged$Barcode <- all_metadata_merged$Unique_Barcode

# Run DE on Merged
# Capturing the returned list which contains both results and the processed object
out_merged <- run_spatial_de(merged_guts, all_metadata_merged, "Merged_Guts")

if (!is.null(out_merged)) {
    res_merged <- out_merged$results
    # UPDATE merged_guts with the version that has metadata added inside the function!
    merged_guts <- out_merged$object 
} else {
    res_merged <- NULL
}
```

## 10. Targeted Analysis: Unique Spatial Genes
We identify genes that exhibit spatial gradients (Anterior vs Posterior) in **exactly one cell type**. We focus specifically on **EB (Enteroblasts)**.

```{r analyze_unique_genes}
message("--- Analyzing Specificity ---")

if (exists("res_merged") && !is.null(res_merged)) {
  
  # 1. Filter for Unique Genes
  unique_genes <- res_merged[res_merged$Specificity == "Unique (1 CellType)", ]
  
  message(paste("Total Spatially Resovled Genes:", nrow(res_merged)))
  message(paste("Unique to a single CellType:", nrow(unique_genes)))
  
  # 2. Extract EB Specific Genes
  eb_unique <- unique_genes[unique_genes$CellType == "EB", ]
  
  message(paste("\n--- EB Specific Spatial Genes (", nrow(eb_unique), ") ---", sep=""))
  
  if (nrow(eb_unique) > 0) {
    # Sort by significance
    eb_unique <- eb_unique[order(eb_unique$p_val_adj), ]
    
    # Display clearly
    print(head(eb_unique[, c("Gene", "avg_log2FC", "p_val_adj")])) 
    # Actually, FindMarkers output direction (positive/negative) depends on ident.1 vs ident.2. 
    # Positive logFC = Higher in Anterior (ident.1), Negative = Higher in Posterior (ident.2)
    
    eb_unique$Direction <- ifelse(eb_unique$avg_log2FC > 0, "Upregulated in Anterior", "Upregulated in Posterior")
    
    print(head(eb_unique[, c("Gene", "Direction", "p_val_adj")], 20))
    
    # Save specifically for Research Agent
    write.csv(eb_unique, "../Excels/EB_Unique_Spatial_Genes.csv")
    message("Saved EB unique list to ../Excels/EB_Unique_Spatial_Genes.csv")
    
  } else {
    message("No unique spatial genes found for EB.")
  }
  
} else {
  warning("res_merged not found. Please run the Spatial DE Analysis chunk first.")
}
```

## 9. Visualizing Gene Expression (Delta)
We plot the expression of **Delta (dl)** on the 3D UMAP to visualize its spatial distribution.

```{r plot_delta_3d}
target_gene <- "alphaTry" # Delta

# Helper to plot gene expression
plot_gene_3d <- function(seurat_obj, plotting_df, gene, gut_name) {
  
  # Check if gene exists
  if (!gene %in% rownames(seurat_obj)) {
    message(paste("Gene", gene, "not found in", gut_name))
    return(NULL)
  }
  
  # Fetch data (robust method)
  expr_df <- FetchData(seurat_obj, vars = gene)
  colnames(expr_df) <- "Value"
  
  # Match cells
  common_cells <- intersect(rownames(expr_df), plotting_df$Barcode)
  
  if (length(common_cells) == 0) {
    warning(paste("No matching cells for", gut_name))
    return(NULL)
  }
  
  # Subset and align
  sub_df <- plotting_df[plotting_df$Barcode %in% common_cells, ]
  sub_expr <- expr_df[sub_df$Barcode, , drop = FALSE]
  
  sub_df$Expression <- sub_expr$Value
  
  # Plot
  p <- plot_ly(sub_df, x = ~PC1, y = ~PC2, z = ~Pseudotime, 
        type = 'scatter3d', mode = 'markers',
        marker = list(
          size = 3, 
          color = ~Expression, 
          colorscale = 'Viridis', 
          showscale = TRUE,
          colorbar = list(title = paste(gene, "Expr"))
        )
      ) %>%
  layout(title = paste(gut_name, ": 3D Expression of", gene), 
         scene = list(
           xaxis = list(title = 'PC 1'),
           yaxis = list(title = 'PC 2'),
           zaxis = list(title = 'Spatial Position')
         ))
         
  return(p)
}

# Plot for each gut
message("Plotting Delta (dl) for Gut 1...")
if(!is.null(p1 <- plot_gene_3d(spe_gut_1, df_gut_1, target_gene, "Gut 1"))) print(p1)

message("Plotting Delta (dl) for Gut 2...")
if(!is.null(p2 <- plot_gene_3d(spe_gut_2, df_gut_2, target_gene, "Gut 2"))) print(p2)
```

## 10. Collapsed Enterocyte Analysis
We group all Enterocyte subtypes (aEC, mEC, pEC, dEC) into a single "**Enterocyte**" label and re-run the Spatial DE Analysis. This increases statistical power for the general Enterocyte population and refines the specificity of gene markers for other cell types like EB.

```{r spatial_de_collapsed}
message("--- Starting Collapsed Enterocyte Analysis ---")

# 1. Prepare Collapsed Metadata
# We work with the consolidated 'all_metadata' from the merged analysis
all_metadata_collapsed <- all_metadata

# Define patterns to collapse
# We target standard Enterocyte abbreviations found in this dataset
# Define patterns to collapse
# We target standard Enterocyte abbreviations found in this dataset
ec_types <- c("aEC", "mEC", "pEC", "dEC", "daEC", "LFC", "Copper", "Iron")
# Alternatively, we can use pattern matching for any type ending in "EC"
# But we'll be explicit to avoid accidents (like 'dEC' vs 'dIEC' etc if distinct)

current_types <- as.character(all_metadata_collapsed$Predicted_Type)

# Collapse logic: replace if in list
current_types[current_types %in% ec_types] <- "Enterocyte"

# Also catch generic cases if label contains "Enterocyte"
current_types[grepl("Enterocyte", current_types, ignore.case = TRUE)] <- "Enterocyte"

# COLLAPSE 2: ISC and EB -> Progenitor
progenitors <- c("ISC", "EB")
current_types[current_types %in% progenitors] <- "Progenitor"

all_metadata_collapsed$Predicted_Type <- current_types

message("Cell Types after collapsing:")
print(table(all_metadata_collapsed$Predicted_Type))

# 2. Run Spatial DE (Collapsed)
# We reuse our robust helper function
res_merged_collapsed <- run_spatial_de(merged_guts, all_metadata_collapsed, "Merged_Guts_Collapsed_EC_Progenitor")

# 3. Analyze Specificity for Progenitor (Collapsed Context)
message("\n--- Analyzing Specificity (Collapsed Context) ---")

if (!is.null(res_merged_collapsed)) {
  
  # Filter for Unique Genes (Specificity column is auto-calculated by run_spatial_de)
  unique_genes_col <- res_merged_collapsed[res_merged_collapsed$Specificity == "Unique (1 CellType)", ]
  
  message(paste("Total Spatially Resolved Genes:", nrow(res_merged_collapsed)))
  message(paste("Unique to a single CellType:", nrow(unique_genes_col)))
  
  # Extract Progenitor Specific
  prog_unique_col <- unique_genes_col[unique_genes_col$CellType == "Progenitor", ]
  
  message(paste("Progenitor Specific Genes (Collapsed Context):", nrow(prog_unique_col)))
  
  if (nrow(prog_unique_col) > 0) {
    prog_unique_col <- prog_unique_col[order(prog_unique_col$p_val_adj), ]
    prog_unique_col$Direction <- ifelse(prog_unique_col$avg_log2FC > 0, "Upregulated in Anterior", "Upregulated in Posterior")
    
    print(head(prog_unique_col[, c("Gene", "Direction", "p_val_adj")], 20))
    
    write.csv(prog_unique_col, "../Excels/Progenitor_Unique_Spatial_Genes_Collapsed.csv")
    message("Saved Progenitor unique list to ../Excels/Progenitor_Unique_Spatial_Genes_Collapsed.csv")
    
  } else {
    message("No unique Progenitor genes found in this context.")
  }
}
```

## 11. Pseudotime Heatmap Analysis
We visualize gene expression changes along the spatial pseudotime trajectory of the gut.
1. **Global Heatmap**: Shows gene expression trends across the entire gut length, ignoring cell types.
2. **Progenitor Heatmap**: Specific to the pooled ISC/EB population.
3. **Enterocyte Heatmap**: Specific to the pooled Enterocyte population.

```{r pseudotime_heatmap}
message("--- Starting Pseudotime Heatmap Analysis (Binned) ---")

# Helper function to generate BINNED heatmap
# This aggregates cells into pseudotime bins (size 1) and plots the AVERAGE expression
plot_binned_heatmap <- function(seurat_object, subset_name, n_features = 50, bin_size = 1) {
  
  message(paste("Generating binned heatmap for:", subset_name))
  
  # 1. Create Bins
  # Floor the pseudotime to get integer bins (0-1 becomes 0, 1-2 becomes 1, etc.)
  seurat_object$Pseudotime_Bin <- floor(seurat_object$Spatial_Pseudotime / bin_size)
  
  # Ensure bins are factors and ordered
  bins <- sort(unique(seurat_object$Pseudotime_Bin))
  seurat_object$Pseudotime_Bin <- factor(seurat_object$Pseudotime_Bin, levels = bins)
  
  # 2. Find Variable Features (on single cells) to pick what to plot
  seurat_object <- FindVariableFeatures(seurat_object, selection.method = "vst", nfeatures = n_features, verbose = FALSE)
  top_genes <- VariableFeatures(seurat_object)
  
  # 3. Aggregate Data (Average Expression per Bin)
  # AverageExpression returns a list (for each assay). We take the default/SCT.
  # We group by the bins.
  Idents(seurat_object) <- seurat_object$Pseudotime_Bin
  
  # avg_exp is a matrix: Genes x Bins
  avg_exp <- AverageExpression(seurat_object, features = top_genes, return.seurat = FALSE, verbose = FALSE)
  # Extract matrix (usually first slot 'RNA' or 'SCT')
  mat <- avg_exp[[1]] 
  
  # 4. Scale the Averaged Data (Row-scaling/Z-score)
  # We do this manually since we are plotting a raw matrix now, not a Seurat object
  mat_scaled <- t(scale(t(mat)))
  
  # 5. Plot using pheatmap or heatmap()
  # We order genes by peak expression bin for visual flow
  peak_bin <- apply(mat_scaled, 1, which.max)
  mat_sorted <- mat_scaled[order(peak_bin), , drop = FALSE]
  
  pheatmap::pheatmap(mat_sorted, 
                     cluster_rows = FALSE, 
                     cluster_cols = FALSE,
                     show_rownames = TRUE, 
                     show_colnames = TRUE,
                     main = paste(subset_name, "- Binned Expression"),
                     fontsize_row = 8)
}

# --- 1. Global Heatmap (All Cells Aggregated by Bin) ---
# This merges all cell types together into the bins
plot_binned_heatmap(merged_guts, "Global Gut Trajectory (All Types Merged)")

# --- 2. Progenitor Heatmap ---
# Subset to Progenitors first
# Re-map the collapsed types if needed (safety check from previous chunk)
match_idx <- match(Cells(merged_guts), all_metadata_collapsed$Unique_Barcode)
if(!any(is.na(match_idx))) { 
    merged_guts$Collapsed_Type <- all_metadata_collapsed$Predicted_Type[match_idx]
}

prog_obj <- subset(merged_guts, subset = Collapsed_Type == "Progenitor")
plot_binned_heatmap(prog_obj, "Progenitors (ISC+EB)")

# --- 3. Enterocyte Heatmap ---
ec_obj <- subset(merged_guts, subset = Collapsed_Type == "Enterocyte")
plot_binned_heatmap(ec_obj, "Enterocytes (Collapsed)")

message("Binned Heatmaps Complete.")
```

## 12. Cell Type Specificity & Heatmap Gene Analysis
We identify highly specific markers for each cell type and then check if the spatially variable genes (from the heatmap) are cell-type specific.

```{r specificity_analysis}
message("--- Starting Cell Type Specificity Analysis ---")

# 1. Identify Markers for Each Cell Type (Original Labels)
Idents(merged_guts) <- merged_guts$Predicted_CellType
markers_original <- FindAllMarkers(merged_guts, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.5, verbose = FALSE)
write.csv(markers_original, "../Excels/Markers_All_Original_Types.csv")

# 2. Identify Markers for Collapsed Labels (Progenitor vs Enterocyte vs Others)
Idents(merged_guts) <- merged_guts$Collapsed_Type
markers_collapsed <- FindAllMarkers(merged_guts, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.5, verbose = FALSE)
write.csv(markers_collapsed, "../Excels/Markers_Collapsed_Types.csv")

message("Saved marker lists to ../Excels/")

# 3. Analyze Function: Cross-reference Heatmap Genes with Specificity
analyze_heatmap_genes <- function(seurat_object, subset_name, markers_df, bin_size=1) {
  
  message(paste("Analyzing heatmap genes for:", subset_name))
  
  # A. Re-identify the Top 50 Variable features used in the heatmap (same logic)
  seurat_object$Pseudotime_Bin <- floor(seurat_object$Spatial_Pseudotime / bin_size)
  bins <- sort(unique(seurat_object$Pseudotime_Bin))
  seurat_object$Pseudotime_Bin <- factor(seurat_object$Pseudotime_Bin, levels = bins)
  
  seurat_object <- FindVariableFeatures(seurat_object, selection.method = "vst", nfeatures = 50, verbose = FALSE)
  heatmap_genes <- VariableFeatures(seurat_object)
  
  # B. Calculate Peak Expression Bin
  Idents(seurat_object) <- seurat_object$Pseudotime_Bin
  avg_exp <- AverageExpression(seurat_object, features = heatmap_genes, return.seurat = FALSE, verbose = FALSE)[[1]]
  peak_bins <- bins[apply(avg_exp, 1, which.max)]
  
  # C. Check Specificity
  # A gene is "Specific" if it appears in the marker list for ANY cell type
  # We find which cell type it is a marker for (lowest p_val if multiple)
  
  results_df <- data.frame(Gene = heatmap_genes, Peak_Bin = peak_bins, Specificity = "Non-Specific", stringsAsFactors = FALSE)
  
  for (i in 1:nrow(results_df)) {
    g <- results_df$Gene[i]
    if (g %in% markers_df$gene) {
      # Find which cluster it marks
      g_markers <- markers_df[markers_df$gene == g, ]
      best_match <- g_markers[which.max(g_markers$avg_log2FC), "cluster"] # Pick strongest logFC
      results_df$Specificity[i] <- as.character(best_match)
    }
  }
  
  # Sort by Peak Bin
  results_df$Peak_Bin <- as.numeric(as.character(results_df$Peak_Bin))
  results_df <- results_df[order(results_df$Peak_Bin), ]
  
  return(results_df)
}

# --- Run Analysis ---

# A. Global Heatmap Genes -> Check against Original Cell Types
global_analysis <- analyze_heatmap_genes(merged_guts, "Global", markers_original)
print("Global Heatmap Genes (Top 10):")
print(head(global_analysis, 10))
write.csv(global_analysis, "../Excels/Global_Heatmap_Gene_Specificity.csv")

# B. Progenitor Heatmap Genes -> Check against Collapsed Types (Is it specific to Progenitors?)
# Re-subset
prog_obj <- subset(merged_guts, subset = Collapsed_Type == "Progenitor")
prog_analysis <- analyze_heatmap_genes(prog_obj, "Progenitors", markers_collapsed)
print("Progenitor Heatmap Genes (Top 10):")
print(head(prog_analysis, 10))
write.csv(prog_analysis, "../Excels/Progenitor_Heatmap_Gene_Specificity.csv")

# C. Enterocyte Heatmap Genes -> Check against Collapsed Types
ec_obj <- subset(merged_guts, subset = Collapsed_Type == "Enterocyte")
ec_analysis <- analyze_heatmap_genes(ec_obj, "Enterocytes", markers_collapsed)
print("Enterocyte Heatmap Genes (Top 10):")
print(head(ec_analysis, 10))
write.csv(ec_analysis, "../Excels/Enterocyte_Heatmap_Gene_Specificity.csv")

message("Specificity Analysis Complete.")

# --- Save Final Object for Comparison Scripts ---
out_dir <- "../Random data"
if (!dir.exists(out_dir)) dir.create(out_dir)
saveRDS(merged_guts, file.path(out_dir, "merged_guts.rds"))
message(paste("✓ Processed Seurat object saved to:", file.path(out_dir, "merged_guts.rds")))
```

## 13. WT Gut Selection & Processing (Same-Slide Comparison)
We select the WT guts from the same tissue slide and process them through the identical pipeline, producing a `merged_guts_wt.rds` object for direct WT vs Mutant comparison.

### WT Gut Selection (Shiny Lasso)

```{r wt_shiny_app}
message("--- Launching WT Gut Selection ---")

# Reuse the same coordinate data from the segmented object
ui_wt <- fluidPage(
  titlePanel("WT Gut Subsetting Tool"),
  sidebarLayout(
    sidebarPanel(
      h4("Instructions"),
      p("1. Use the Lasso tool to select WT gut regions."),
      p("2. Click 'Save WT Gut X' to store each selection."),
      actionButton("save_wt_gut_1", "Save WT Gut 1"),
      verbatimTextOutput("wt_status_1"),
      hr(),
      actionButton("save_wt_gut_2", "Save WT Gut 2"),
      verbatimTextOutput("wt_status_2"),
      hr(),
      actionButton("wt_done_btn", "DONE (Close & Save)"),
      class = "well"
    ),
    mainPanel(
      plotlyOutput("wt_scatter_plot", height = "800px")
    )
  )
)

server_wt <- function(input, output, session) {
  vals <- reactiveValues(gut_1 = NULL, gut_2 = NULL)
  
  output$wt_scatter_plot <- renderPlotly({
    plot_ly(coords_seg, x = ~x, y = ~y, key = ~barcode, type = "scatter", mode = "markers",
            marker = list(size = 3, color = 'gray', opacity = 0.5), source = "wt_source") %>%
      layout(dragmode = "lasso", title = "Select WT Gut Regions", yaxis = list(autorange = "reversed"))
  })
  
  observeEvent(input$save_wt_gut_1, {
    event_data <- event_data("plotly_selected", source = "wt_source")
    if (!is.null(event_data)) vals$gut_1 <- event_data$key
  })
  output$wt_status_1 <- renderText({
    if(is.null(vals$gut_1)) "Not saved" else paste(length(vals$gut_1), "cells saved")
  })
  
  observeEvent(input$save_wt_gut_2, {
    event_data <- event_data("plotly_selected", source = "wt_source")
    if (!is.null(event_data)) vals$gut_2 <- event_data$key
  })
  output$wt_status_2 <- renderText({
    if(is.null(vals$gut_2)) "Not saved" else paste(length(vals$gut_2), "cells saved")
  })
  
  observeEvent(input$wt_done_btn, {
    stopApp(list(gut_1 = vals$gut_1, gut_2 = vals$gut_2))
  })
}

wt_subset_res <- runApp(shinyApp(ui_wt, server_wt))

# Subset Objects (reuse barcode cleaning from earlier)
wt_cells_1 <- clean_barcodes(wt_subset_res$gut_1, Cells(visium_seg))
wt_cells_2 <- clean_barcodes(wt_subset_res$gut_2, Cells(visium_seg))

if (length(wt_cells_1) > 0) spe_wt_gut_1 <- subset(visium_seg, cells = wt_cells_1)
if (length(wt_cells_2) > 0) spe_wt_gut_2 <- subset(visium_seg, cells = wt_cells_2)

message(paste("WT Gut 1:", ncol(spe_wt_gut_1), "cells"))
message(paste("WT Gut 2:", ncol(spe_wt_gut_2), "cells"))
```

### WT Trajectory Analysis

```{r wt_trajectory}
# Run the same PCA + Spatial Pseudotime pipeline
df_wt_gut_1 <- process_gut_3d(spe_wt_gut_1, "WT Gut 1")
df_wt_gut_2 <- process_gut_3d(spe_wt_gut_2, "WT Gut 2")

message("WT trajectory analysis complete.")
```

### WT QC Filtering

```{r wt_qc_filtering}
if (exists("spe_wt_gut_1") && exists("df_wt_gut_1")) {
  res_wt1 <- filter_qc_post_trajectory(spe_wt_gut_1, df_wt_gut_1, "WT Gut 1")
  spe_wt_gut_1 <- res_wt1$obj; df_wt_gut_1 <- res_wt1$df
}
if (exists("spe_wt_gut_2") && exists("df_wt_gut_2")) {
  res_wt2 <- filter_qc_post_trajectory(spe_wt_gut_2, df_wt_gut_2, "WT Gut 2")
  spe_wt_gut_2 <- res_wt2$obj; df_wt_gut_2 <- res_wt2$df
}
```

### WT Label Transfer

```{r wt_label_transfer}
message("--- WT Label Transfer ---")
df_wt_gut_1 <- transfer_labels(spe_wt_gut_1, df_wt_gut_1, sc_ref)
df_wt_gut_2 <- transfer_labels(spe_wt_gut_2, df_wt_gut_2, sc_ref)
message("WT Label Transfer Complete.")
```

### WT Spatial DE & Merge

```{r wt_spatial_de}
message("--- WT Spatial DE Analysis ---")

# Individual gut DE
res_wt_gut1 <- run_spatial_de(spe_wt_gut_1, df_wt_gut_1, "WT_Gut1")
res_wt_gut2 <- run_spatial_de(spe_wt_gut_2, df_wt_gut_2, "WT_Gut2")

# Merge WT guts
merged_guts_wt <- merge(spe_wt_gut_1, y = spe_wt_gut_2, add.cell.ids = c("WTGut1", "WTGut2"))

# Prepare merged metadata
df_wt_1_mod <- df_wt_gut_1
df_wt_1_mod$Unique_Barcode <- paste0("WTGut1_", df_wt_1_mod$Barcode)

df_wt_2_mod <- df_wt_gut_2
df_wt_2_mod$Unique_Barcode <- paste0("WTGut2_", df_wt_2_mod$Barcode)

all_metadata_wt <- rbind(df_wt_1_mod, df_wt_2_mod)
all_metadata_wt_merged <- all_metadata_wt
all_metadata_wt_merged$Barcode <- all_metadata_wt_merged$Unique_Barcode

# Run DE on merged WT
out_merged_wt <- run_spatial_de(merged_guts_wt, all_metadata_wt_merged, "Merged_WT_Guts")

if (!is.null(out_merged_wt)) {
    merged_guts_wt <- out_merged_wt$object
}

# --- Save WT Object ---
saveRDS(merged_guts_wt, file.path(out_dir, "merged_guts_wt.rds"))
message(paste("✓ WT Seurat object saved to:", file.path(out_dir, "merged_guts_wt.rds")))
```
