---
title: "Cell Type Similarity Matrix (WT vs Mutant Single-Cell)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(Seurat)
library(dplyr)
library(ggplot2)
library(pheatmap)
library(viridis)
```

## 1. Load Single-Cell Data
We load the single-cell references used for spatial deconvolution in both conditions.
- WT: `seurat_ctrl_with_metadata.rds`
- Mutant: `ApcKO_merged_SC.rds`

```{r load_data}
path_wt <- "/home/mattia/Thesis/Single-cell/seurat_ctrl_with_metadata.rds"
path_mut <- "../Random data/ApcKO_merged_SC.rds"

if (!file.exists(path_wt)) stop("WT SC data not found at: ", path_wt)
if (!file.exists(path_mut)) stop("Mutant SC data not found at: ", path_mut)

message("Loading WT SC object...")
sc_wt <- readRDS(path_wt)

message("Loading Mutant SC object...")
sc_mut <- readRDS(path_mut)
```

## 2. Identify Common Cell Types & Variable Features
We align the two datasets to compute similarity strictly between shared cell types, avoiding artificial correlations from missing clusters.

```{r prep_data}
# Get common cell types
wt_celltypes <- unique(as.character(sc_wt$celltype))
mut_celltypes <- unique(as.character(sc_mut$celltype))
common_celltypes <- intersect(wt_celltypes, mut_celltypes)

# Exclude NAs or empty strings just in case
common_celltypes <- common_celltypes[!is.na(common_celltypes) & common_celltypes != ""]

message("Common cell types found: ", paste(common_celltypes, collapse = ", "))

# Filter objects to only common cell types
sc_wt <- subset(sc_wt, celltype %in% common_celltypes)
sc_mut <- subset(sc_mut, celltype %in% common_celltypes)

# Ensure layers are joined and data is normalized
sc_wt <- JoinLayers(sc_wt)
sc_mut <- JoinLayers(sc_mut)

sc_wt <- NormalizeData(sc_wt, verbose = FALSE)
sc_mut <- NormalizeData(sc_mut, verbose = FALSE)

# Get variable features for correlation
# We use the intersection of highly variable features from both datasets to ensure 
# we are correlating biologically informative genes rather than noise or housekeeping genes.
sc_wt <- FindVariableFeatures(sc_wt, nfeatures = 2000, verbose = FALSE)
sc_mut <- FindVariableFeatures(sc_mut, nfeatures = 2000, verbose = FALSE)

var_genes <- intersect(VariableFeatures(sc_wt), VariableFeatures(sc_mut))
message(paste("Using", length(var_genes), "shared highly variable genes for correlation."))
```

## 3. Calculate Average Expression Profiles

```{r calc_avg_exp}
# We use the 'data' slot (log-normalized) for average expression
avg_wt <- AverageExpression(sc_wt, group.by = "celltype", assays = "RNA", slot = "data")$RNA
avg_mut <- AverageExpression(sc_mut, group.by = "celltype", assays = "RNA", slot = "data")$RNA

# Ensure we only try to subset genes that actually exist in the matrices
valid_genes <- intersect(var_genes, intersect(rownames(avg_wt), rownames(avg_mut)))
if (length(valid_genes) < length(var_genes)) {
  message(paste("Dropped", length(var_genes) - length(valid_genes), "variable genes not present in RNA assays."))
}

# Seurat's AverageExpression sometimes sanitizes column names. 
# We intersect the columns to be completely safe.
valid_cols <- intersect(colnames(avg_wt), colnames(avg_mut))

# Filter to highly variable genes and matching columns
avg_wt <- avg_wt[valid_genes, valid_cols, drop = FALSE]
avg_mut <- avg_mut[valid_genes, valid_cols, drop = FALSE]
```

## 4. Compute Similarity Matrix
We compute the Spearman correlation between the expression profiles of WT cell types and Mutant cell types. Spearman is robust against expression outliers.

```{r similarity_matrix, fig.width=9, fig.height=8}
# Compute correlation matrix
cor_mat <- cor(as.matrix(avg_wt), as.matrix(avg_mut), method = "spearman")

# Order the matrix based on alphabetical cell type names
cor_mat <- cor_mat[sort(valid_cols), sort(valid_cols), drop = FALSE]

# Define row and column labels
rownames(cor_mat) <- paste0("WT_", rownames(cor_mat))
colnames(cor_mat) <- paste0("Mut_", colnames(cor_mat))

# Visualize with pheatmap
p_heatmap <- pheatmap(cor_mat,
         cluster_rows = TRUE, 
         cluster_cols = TRUE,
         display_numbers = round(cor_mat, 2),
         color = viridis(100),
         main = "Spearman Correlation: WT vs Mutant Cell Types",
         angle_col = 45,
         fontsize_number = 10,
         treeheight_row = 30,
         treeheight_col = 30)

# Save the plot automatically
out_dir <- "../figures/sc_comparisons"
if (!dir.exists(out_dir)) {
  dir.create(out_dir, recursive = TRUE)
}

png(file.path(out_dir, "CellType_Similarity_Matrix.png"), width = 900, height = 800, res = 120)
# pheatmap directly plots to device
pheatmap(cor_mat,
         cluster_rows = TRUE, 
         cluster_cols = TRUE,
         display_numbers = round(cor_mat, 2),
         color = viridis(100),
         main = "Spearman Correlation: WT vs Mutant Cell Types",
         angle_col = 45,
         fontsize_number = 10,
         treeheight_row = 30,
         treeheight_col = 30)
dev.off()

message("Matrix saved to figures/sc_comparisons/CellType_Similarity_Matrix.png")
```
