---
title: "Gene Expression Concordance Analysis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(Seurat)
library(dplyr)
library(ggplot2)
library(reshape2)
library(readr)
library(ggrepel)
library(clusterProfiler)
library(org.Dm.eg.db)
library(ggpubr)

# ------------------------------------------------------------------------------
# 1. Load Data
# ------------------------------------------------------------------------------

# Paths
path_111125ST <- "../../111125ST_def/Random data/merged_guts.rds"
path_03Slide4 <- "../Random data/merged_guts.rds"
path_seg_rds  <- "../Random data/VisiumHD_Drosophila_Resolution_List.rds"
path_mutant_csv <- "../Random data/Slide4_Segmented_Mutant_Selection.csv"
path_wt_csv <- "../Random data/Slide4_Segmented_WT_Selection.csv"

cat("Loading 111125ST_def merged_guts...\n")
merged_111125ST <- readRDS(path_111125ST)

cat("Loading 03Slide4 merged_guts...\n")
merged_03Slide4 <- readRDS(path_03Slide4)

# Create figures directory if it doesn't exist
if (!dir.exists("../figures/concordance")) dir.create("../figures/concordance", recursive = TRUE)
```

## 1. Global Concordance (111125ST vs 03Slide4)

Comparison of average gene expression across all cells in both datasets.

```{r global_concordance, fig.width=7, fig.height=6}
# Set identities to a single value for global average
Idents(merged_111125ST) <- "Global"
avg_111125ST_mat <- AverageExpression(merged_111125ST, assays = "Segmented", slot = "data")$Segmented
avg_111125ST <- data.frame(WT_111125ST = rowMeans(as.matrix(avg_111125ST_mat)), Gene = rownames(avg_111125ST_mat))

Idents(merged_03Slide4) <- "Global"
avg_03Slide4_mat <- AverageExpression(merged_03Slide4, assays = "Segmented", slot = "data")$Segmented
avg_03Slide4 <- data.frame(Slide4_Global = rowMeans(as.matrix(avg_03Slide4_mat)), Gene = rownames(avg_03Slide4_mat))

# Join datasets on common genes
concordance_df <- inner_join(avg_111125ST, avg_03Slide4, by = "Gene")

# Identify top 10 most different genes (outliers from diagonal)
concordance_df <- concordance_df %>%
    mutate(diff = abs(log1p(WT_111125ST) - log1p(Slide4_Global))) %>%
    arrange(desc(diff))

top10_global <- head(concordance_df, 10)

# Determine axis limits for 1:1 scaling
max_val <- max(log1p(concordance_df$WT_111125ST), log1p(concordance_df$Slide4_Global))

# Plot
p_global <- ggplot(concordance_df, aes(x = log1p(WT_111125ST), y = log1p(Slide4_Global))) +
    geom_point(alpha = 0.5, size = 0.5, color = "steelblue") +
    geom_abline(intercept = 0, slope = 1, color = "grey", linetype = "dashed") +
    geom_smooth(method = "lm", color = "darkred", linetype = "solid", size = 0.5) +
    geom_text_repel(data = top10_global, aes(label = Gene), size = 3, box.padding = 0.5) +
    stat_cor(method = "pearson", label.x = 0.5, label.y = max_val * 0.95) +
    labs(title = "Global Gene Expression Concordance",
         subtitle = "111125ST (WT) vs 03Slide4 (Apc Mutant/WT)",
         x = "log1p(Avg Expression) - 111125ST",
         y = "log1p(Avg Expression) - 03Slide4") +
    xlim(0, max_val) + ylim(0, max_val) +
    coord_fixed() +
    theme_pubr()

ggsave("../figures/concordance/Global_Concordance.png", p_global, width = 7, height = 7)
print(p_global)
```

## 2. Selection-Specific Concordance

For the mutant and WT selections, barcodes come from the **segmented** Seurat object, not from `merged_guts`. We load the segmented object and subset it.

```{r load_segmented}
cat("Loading segmented Seurat object...\n")
vhd_list <- readRDS(path_seg_rds)
seurat_seg <- vhd_list$seg
seurat_seg <- NormalizeData(seurat_seg, verbose = FALSE)
cat("Segmented object loaded:", ncol(seurat_seg), "cells\n")
```

### 2.1 111125ST vs Slide4 Segmented Mutant Selection

```{r mutant_concordance, fig.width=7, fig.height=6}
# Load Mutant barcodes
mutant_barcodes <- read_csv(path_mutant_csv)$barcode

# Subset segmented object
common_mutant <- intersect(mutant_barcodes, Cells(seurat_seg))
cat("Mutant barcodes matched:", length(common_mutant), "of", length(mutant_barcodes), "\n")

sub_mutant <- subset(seurat_seg, cells = common_mutant)

# Calculate average expression using rowMeans (no identity issues)
avg_mutant <- data.frame(
    Slide4_Mutant = rowMeans(as.matrix(GetAssayData(sub_mutant, layer = "data"))) * 2, # Multiply by 2 for visualization
    Gene = rownames(sub_mutant)
)

# Join with 111125ST
mutant_comp_df <- inner_join(avg_111125ST, avg_mutant, by = "Gene")

# Identify top 10 most different genes
mutant_comp_df <- mutant_comp_df %>%
    mutate(diff = abs(log1p(WT_111125ST) - log1p(Slide4_Mutant))) %>%
    arrange(desc(diff))

top10_mutant <- head(mutant_comp_df, 10)

# Axis limits
max_val_mut <- max(log1p(mutant_comp_df$WT_111125ST), log1p(mutant_comp_df$Slide4_Mutant))

# Plot
p_mutant <- ggplot(mutant_comp_df, aes(x = log1p(WT_111125ST), y = 2*log1p(Slide4_Mutant))) +
    geom_point(alpha = 0.5, size = 0.5, color = "indianred") +
    geom_abline(intercept = 0, slope = 1, color = "grey", linetype = "dashed") +
    geom_smooth(method = "lm", color = "black", linetype = "solid", size = 0.5) +
    geom_text_repel(data = top10_mutant, aes(label = Gene), size = 3, box.padding = 0.5) +
    stat_cor(method = "pearson", label.y = max_val_mut * 0.95) +
    labs(title = "Concordance: 111125ST vs Slide4 Mutant Selection",
         x = "log1p(Avg Expression) - 111125ST",
         y = "log1p(Avg Expression) - Mutant Selection") +
    xlim(0, max_val_mut) + ylim(0, max_val_mut) +
    coord_fixed(ratio = 1) +
    theme_pubr()

ggsave("../figures/concordance/Mutant_Selection_Concordance.png", p_mutant, width = 7, height = 7)
print(p_mutant)
```

### 2.2 111125ST vs Slide4 Segmented WT Selection

```{r wt_concordance, fig.width=7, fig.height=6}
# Load WT barcodes
wt_barcodes <- read_csv(path_wt_csv)$barcode

# Subset segmented object
common_wt <- intersect(wt_barcodes, Cells(seurat_seg))
cat("WT barcodes matched:", length(common_wt), "of", length(wt_barcodes), "\n")

sub_wt <- subset(seurat_seg, cells = common_wt)

# Calculate average expression using rowMeans
avg_wt_sel <- data.frame(
    Slide4_WT = rowMeans(as.matrix(GetAssayData(sub_wt, layer = "data"))) * 2, # Multiply by 2 for visualization
    Gene = rownames(sub_wt)
)

# Join with 111125ST
wt_comp_df <- inner_join(avg_111125ST, avg_wt_sel, by = "Gene")

# Identify top 10 most different genes
wt_comp_df <- wt_comp_df %>%
    mutate(diff = abs(log1p(WT_111125ST) - log1p(Slide4_WT))) %>%
    arrange(desc(diff))

top10_wt <- head(wt_comp_df, 10)

# Axis limits
max_val_wt <- max(log1p(wt_comp_df$WT_111125ST), log1p(wt_comp_df$Slide4_WT))

# Plot
p_wt <- ggplot(wt_comp_df, aes(x = log1p(WT_111125ST), y = 2*log1p(Slide4_WT))) +
    geom_point(alpha = 0.5, size = 0.5, color = "darkgreen") +
    geom_abline(intercept = 0, slope = 1, color = "grey", linetype = "dashed") +
    geom_smooth(method = "lm", color = "black", linetype = "solid", size = 0.5) +
    geom_text_repel(data = top10_wt, aes(label = Gene), size = 3, box.padding = 0.5) +
    stat_cor(method = "pearson", label.y = max_val_wt * 0.95) +
    labs(title = "Concordance: 111125ST vs Slide4 WT Selection",
         x = "log1p(Avg Expression) - 111125ST",
         y = "log1p(Avg Expression) - WT Selection") +
    xlim(0, max_val_wt) + ylim(0, max_val_wt) +
    coord_fixed(ratio = 1) +
    theme_pubr()

ggsave("../figures/concordance/WT_Selection_Concordance.png", p_wt, width = 7, height = 7)
print(p_wt)
```

## 3. GO Enrichment of Concordance Outliers

We identify the top 200 differentially expressed genes (100 Up, 100 Down) for both Mutant and WT selections relative to the 111125ST baseline and perform GO enrichment analysis.

```{r go_enrichment, fig.width=10, fig.height=12}
# 1. Identify Top 400 Up/Down for Mutant Selection
mutant_comp_df <- mutant_comp_df %>%
    mutate(log2FC = log2(Slide4_Mutant / WT_111125ST))

mutant_up_all <- mutant_comp_df %>% filter(is.finite(log2FC)) %>% arrange(desc(log2FC)) %>% head(400) %>% pull(Gene)
mutant_down_all <- mutant_comp_df %>% filter(is.finite(log2FC)) %>% arrange(log2FC) %>% head(400) %>% pull(Gene)

# 2. Identify Top 400 Up/Down for WT Selection
wt_comp_df <- wt_comp_df %>%
    mutate(log2FC = log2(Slide4_WT / WT_111125ST))

wt_up_all <- wt_comp_df %>% filter(is.finite(log2FC)) %>% arrange(desc(log2FC)) %>% head(400) %>% pull(Gene)
wt_down_all <- wt_comp_df %>% filter(is.finite(log2FC)) %>% arrange(log2FC) %>% head(400) %>% pull(Gene)

# 3. Define Mutant-Specific Sets
# Genes NOT upregulated in WT but ARE in mutant
mutant_spec_up <- setdiff(mutant_up_all, wt_up_all)

# Genes downregulated in mutant but NOT in WT
mutant_spec_down <- setdiff(mutant_down_all, wt_down_all)

# 4. Combine into list for compareCluster, including counts in names
genes_list <- list(mutant_spec_up, mutant_spec_down)
names(genes_list) <- c(
    paste0("Mutant_Specific_Up (n=", length(mutant_spec_up), ")"),
    paste0("Mutant_Specific_Down (n=", length(mutant_spec_down), ")")
)

# 5. Run compareCluster (BP - Biological Process)
cat("Running GO Enrichment (BP) for mutant-specific outliers...\n")
go_cluster <- compareCluster(geneCluster = genes_list, 
                             fun = "enrichGO", 
                             OrgDb = org.Dm.eg.db, 
                             keyType = "SYMBOL", 
                             ont = "BP", 
                             pvalueCutoff = 0.05)

# 6. Visualize
if (!is.null(go_cluster) && nrow(go_cluster) > 0) {
    p_go <- dotplot(go_cluster, showCategory = 10) +
        labs(title = "GO Enrichment: Mutant-Specific Outliers (Top 10 Terms)",
             subtitle = "Mutant-specific = Change observed in Mutant Selection but NOT in WT Selection") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))

    ggsave("../figures/concordance/Mutant_Specific_Outliers_GO_Dotplot.png", p_go, width = 8, height = 12)
    print(p_go)
} else {
    cat("No significant GO terms found for the mutant-specific outliers.\n")
}

# 7. Print the number of differential genes
cat("\nNumber of Mutant-Specific Up-regulated genes:", length(mutant_spec_up), "\n")
cat("Number of Mutant-Specific Down-regulated genes:", length(mutant_spec_down), "\n")
```
