---
title: "Target Gene Spatial Profile Comparison (WT vs Mutant)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(Seurat)
library(ggplot2)
library(dplyr)
library(patchwork)
library(mgcv) # For GAM smoothing
library(pheatmap)
library(zoo)
library(gridExtra)

# ==========================================
# CONFIGURATION
# ==========================================
# Change this variable to your gene of interest
target_gene <- "wg" 
```

## 1. Load Data
We load three `merged_guts.rds` objects: WT (03Slide4), Mutant (03Slide4), and WT (111125ST).

```{r load_data}
path_mutant <- "../Random data/merged_guts.rds"
path_wt <- "../Random data/merged_guts_wt.rds"
path_wt_111 <- "../../111125ST_def/Random data/merged_guts.rds"

if (!file.exists(path_mutant)) stop(paste("Mutant merged_guts.rds not found at", path_mutant))
if (!file.exists(path_wt)) stop(paste("WT merged_guts_wt.rds not found at", path_wt))
if (!file.exists(path_wt_111)) stop(paste("WT 111125ST merged_guts.rds not found at", path_wt_111))

message("Loading datasets...")
seurat_mutant <- readRDS(path_mutant)
seurat_wt <- readRDS(path_wt)
seurat_wt_111 <- readRDS(path_wt_111)

# Use the entire merged_guts objects directly without subsetting
gut_mutant <- seurat_mutant
gut_wt <- seurat_wt
gut_wt_111 <- seurat_wt_111

# Remove ambiguous cell types if the annotation exists
if ("Predicted_CellType" %in% colnames(gut_mutant@meta.data)) {
  gut_mutant <- subset(gut_mutant, cells = Cells(gut_mutant)[!grepl("ambiguous", gut_mutant$Predicted_CellType, ignore.case = TRUE)])
}
if ("Predicted_CellType" %in% colnames(gut_wt@meta.data)) {
  gut_wt <- subset(gut_wt, cells = Cells(gut_wt)[!grepl("ambiguous", gut_wt$Predicted_CellType, ignore.case = TRUE)])
}
if ("Predicted_CellType" %in% colnames(gut_wt_111@meta.data)) {
  gut_wt_111 <- subset(gut_wt_111, cells = Cells(gut_wt_111)[!grepl("ambiguous", gut_wt_111$Predicted_CellType, ignore.case = TRUE)])
}

message(paste("WT (03Slide4):", ncol(gut_wt), "cells"))
message(paste("Mutant (03Slide4):", ncol(gut_mutant), "cells"))
message(paste("WT (111125ST):", ncol(gut_wt_111), "cells"))
```

## 2. Extract Expression Data

```{r extract_expression}
# Ensure the default assay is set appropriately (usually SCT)
if ("SCT" %in% names(Assays(gut_mutant))) DefaultAssay(gut_mutant) <- "SCT"
if ("SCT" %in% names(Assays(gut_wt))) DefaultAssay(gut_wt) <- "SCT"
if ("SCT" %in% names(Assays(gut_wt_111))) DefaultAssay(gut_wt_111) <- "SCT"

# Check if gene exists in all datasets
gene_in_wt <- target_gene %in% rownames(gut_wt)
gene_in_mut <- target_gene %in% rownames(gut_mutant)
gene_in_wt111 <- target_gene %in% rownames(gut_wt_111)

if (!gene_in_wt) warning(paste(target_gene, "not found in WT (03Slide4) dataset."))
if (!gene_in_mut) warning(paste(target_gene, "not found in Mutant (03Slide4) dataset."))
if (!gene_in_wt111) warning(paste(target_gene, "not found in WT (111125ST) dataset."))
if (!gene_in_wt && !gene_in_mut && !gene_in_wt111) stop("Gene not found in any dataset.")

# Fetch Data
df_wt <- data.frame()
if (gene_in_wt) {
  df_wt <- FetchData(gut_wt, vars = c(target_gene, "Spatial_Pseudotime"))
  df_wt$Condition <- "WT (03Slide4)"
}

df_mut <- data.frame()
if (gene_in_mut) {
  df_mut <- FetchData(gut_mutant, vars = c(target_gene, "Spatial_Pseudotime"))
  df_mut$Condition <- "Mutant (03Slide4)"
}

df_wt111 <- data.frame()
if (gene_in_wt111) {
  df_wt111 <- FetchData(gut_wt_111, vars = c(target_gene, "Spatial_Pseudotime"))
  df_wt111$Condition <- "WT (111125ST)"
}

# Combine
plot_df <- rbind(df_wt, df_mut, df_wt111)
colnames(plot_df)[1] <- "Expression"

# Filter out NAs in Spatial_Pseudotime
plot_df <- plot_df %>% filter(!is.na(Spatial_Pseudotime))

# Optional: Add Log1p expression for better visibility if raw values are sparse
plot_df$Expression_Log <- log1p(plot_df$Expression)

# Set factor levels for consistent plotting order
plot_df$Condition <- factor(plot_df$Condition, levels = c("WT (111125ST)", "WT (03Slide4)", "Mutant (03Slide4)"))
```

## 3. Visualize Spatial Profiles

### Smoothed Expression Trend (GAM)
Using a Generalized Additive Model (GAM) to fit a smooth curve over the spatial pseudotime (anterior to posterior).

```{r plot_smooth, fig.width=9, fig.height=5}
p_smooth <- ggplot(plot_df, aes(x = Spatial_Pseudotime, y = Expression, color = Condition, fill = Condition)) +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), alpha = 0.2, size = 1.2) +
  scale_color_manual(values = c("WT (111125ST)" = "#4DAF4A", "WT (03Slide4)" = "#377EB8", "Mutant (03Slide4)" = "#E41A1C")) +
  scale_fill_manual(values = c("WT (111125ST)" = "#4DAF4A", "WT (03Slide4)" = "#377EB8", "Mutant (03Slide4)" = "#E41A1C")) +
  coord_cartesian(ylim = c(0, NA)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme_minimal() +
  labs(
    title = paste("Spatial Expression Profile:", target_gene),
    subtitle = "Smoothed trend across inferred spatial position",
    x = "Inferred Spatial Position (Pseudotime)",
    y = "Normalized Expression"
  ) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.line.y = element_line(color = "black"),
        axis.line.x = element_line(color = "black"),
        text = element_text(size = 14),
        plot.title = element_text(face = "bold"))

print(p_smooth)
```

### Scatter Plot with LOESS Smoothing
Showing the individual cellular expression values along with a local regression line.

```{r plot_scatter, fig.width=14, fig.height=5}
p_scatter <- ggplot(plot_df, aes(x = Spatial_Pseudotime, y = Expression, color = Condition)) +
  geom_point(alpha = 0.3, size = 1.5, stroke = 0) +
  geom_smooth(method = "loess", span = 0.3, color = "black", size = 1) +
  scale_color_manual(values = c("WT (111125ST)" = "#4DAF4A", "WT (03Slide4)" = "#377EB8", "Mutant (03Slide4)" = "#E41A1C")) +
  facet_wrap(~Condition, ncol = 3) +
  coord_cartesian(ylim = c(0, NA)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme_minimal() +
  labs(
    title = paste("Cell-Level Spatial Expression:", target_gene),
    x = "Inferred Spatial Position (Pseudotime)",
    y = "Normalized Expression"
  ) +
  theme(legend.position = "none",
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.line.y = element_line(color = "black"),
        axis.line.x = element_line(color = "black"),
        text = element_text(size = 14),
        strip.text = element_text(face = "bold", size = 12))

print(p_scatter)
```

### Overlaid Scatter Plot with LOESS Smoothing
Showing both conditions on the same axes to directly compare cell density and local expression trends.

```{r plot_scatter_overlaid, fig.width=9, fig.height=5}
p_scatter_overlaid <- ggplot(plot_df, aes(x = Spatial_Pseudotime, y = Expression, color = Condition, fill = Condition)) +
  geom_point(alpha = 0.3, size = 1.5, stroke = 0) +
  geom_smooth(method = "loess", span = 0.3, size = 1) +
  scale_color_manual(values = c("WT (111125ST)" = "#4DAF4A", "WT (03Slide4)" = "#377EB8", "Mutant (03Slide4)" = "#E41A1C")) +
  scale_fill_manual(values = c("WT (111125ST)" = "#4DAF4A", "WT (03Slide4)" = "#377EB8", "Mutant (03Slide4)" = "#E41A1C")) +
  coord_cartesian(ylim = c(0, NA)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme_minimal() +
  labs(
    title = paste("Overlaid Cell-Level Spatial Expression:", target_gene),
    x = "Inferred Spatial Position (Pseudotime)",
    y = "Normalized Expression"
  ) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.line.y = element_line(color = "black"),
        axis.line.x = element_line(color = "black"),
        text = element_text(size = 14),
        plot.title = element_text(face = "bold"))

print(p_scatter_overlaid)
```

## 4. Expression by Cell Type
Visualizing which inferred cell types primarily express the target gene.

```{r plot_celltype, fig.width=14, fig.height=6}
# Look for the cell type column
celltype_col <- "Predicted_CellType"
if (!celltype_col %in% colnames(gut_wt@meta.data)) {
  # Fallback if standard col is missing
  if ("celltype" %in% colnames(gut_wt@meta.data)) celltype_col <- "celltype"
  else if ("seurat_clusters" %in% colnames(gut_wt@meta.data)) celltype_col <- "seurat_clusters"
}

if (celltype_col %in% colnames(gut_wt@meta.data) && celltype_col %in% colnames(gut_mutant@meta.data)) {
  
  df_wt_ct <- data.frame(
    Expression = FetchData(gut_wt, vars = target_gene)[[1]],
    CellType = gut_wt@meta.data[[celltype_col]],
    Condition = "WT (03Slide4)"
  )
  
  df_mut_ct <- data.frame(
    Expression = FetchData(gut_mutant, vars = target_gene)[[1]],
    CellType = gut_mutant@meta.data[[celltype_col]],
    Condition = "Mutant (03Slide4)"
  )
  
  df_wt111_ct <- data.frame(
    Expression = FetchData(gut_wt_111, vars = target_gene)[[1]],
    CellType = gut_wt_111@meta.data[[celltype_col]],
    Condition = "WT (111125ST)"
  )
  
  plot_df_ct <- rbind(df_wt_ct, df_mut_ct, df_wt111_ct)
  plot_df_ct <- plot_df_ct[!is.na(plot_df_ct$CellType) & plot_df_ct$CellType != "", ]
  plot_df_ct$Condition <- factor(plot_df_ct$Condition, levels = c("WT (111125ST)", "WT (03Slide4)", "Mutant (03Slide4)"))
  
  p_vln <- ggplot(plot_df_ct, aes(x = CellType, y = Expression, fill = Condition)) +
    geom_violin(scale = "width", position = position_dodge(width = 0.8), color = "black", alpha = 0.7) +
    geom_boxplot(width = 0.2, position = position_dodge(width = 0.8), color = "black", outlier.shape = NA, alpha = 0.5) +
    scale_fill_manual(values = c("WT (111125ST)" = "#4DAF4A", "WT (03Slide4)" = "#377EB8", "Mutant (03Slide4)" = "#E41A1C")) +
    coord_cartesian(ylim = c(0, NA)) +
    theme_minimal() +
    labs(
      title = paste("Expression by Cell Type:", target_gene),
      x = "Predicted Cell Type",
      y = "Normalized Expression"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      axis.line = element_line(color = "black"),
      text = element_text(size = 14),
      plot.title = element_text(face = "bold")
    )
  
  print(p_vln)
} else {
  message("Cell type column not found. Skipping cell type plot.")
}
```
## 5. Spatial Expression Heatmap
Visualizing the spatial transitions across the gut for the selected target gene.

```{r spatial_heatmaps, fig.width=14, fig.height=5}
message("Preparing data for heatmaps...")

# Combine objects temporarily to bin pseudotime
combined <- merge(gut_wt, y = list(gut_mutant, gut_wt_111), add.cell.ids = c("WT_S4", "Mutant", "WT_111"))

# Set batch assignments based on prefix
combined$Batch <- dplyr::case_when(
  grepl("^WT_S4_", Cells(combined)) ~ "WT (03Slide4)",
  grepl("^Mutant_", Cells(combined)) ~ "Mutant (03Slide4)",
  grepl("^WT_111_", Cells(combined)) ~ "WT (111125ST)"
)

# We are only interested in the target_gene
genes_to_plot <- c(target_gene)

# 1. Binning by Spatial_Pseudotime (0-100)
# Ensure Bin is defined for all cells
combined$Bin <- cut(combined$Spatial_Pseudotime, breaks = 100, labels = FALSE)

# 2. Function to calculate smoothed expression averages per bin
get_smoothed_profiles <- function(seurat_obj, batch_name, genes, n_bins = 100, window = 5) {
  sub <- subset(seurat_obj, Batch == batch_name)
  if (ncol(sub) == 0) return(NULL)
  
  assay_to_use <- if ("SCT" %in% names(sub@assays)) "SCT" else DefaultAssay(sub)
  genes <- intersect(genes, rownames(sub[[assay_to_use]]))
  if (length(genes) == 0) return(NULL)
  
  exp_mat <- GetAssayData(sub, assay = assay_to_use, layer = "data")[genes, , drop = FALSE]
  
  bin_means <- t(apply(exp_mat, 1, function(x) {
    res <- tapply(x, sub$Bin, mean, na.rm = TRUE)
    full_res <- setNames(rep(NA, n_bins), 1:n_bins)
    full_res[names(res)] <- res
    return(full_res)
  }))
  
  bin_means <- t(apply(bin_means, 1, function(x) {
    if (all(is.na(x))) return(rep(0, n_bins))
    zoo::na.approx(x, xout = 1:n_bins, na.rm = FALSE, rule = 2)
  }))
  
  bin_smoothed <- t(apply(bin_means, 1, function(x) {
    zoo::rollapply(x, width = window, FUN = mean, fill = "extend")
  }))
  colnames(bin_smoothed) <- 1:n_bins
  rownames(bin_smoothed) <- genes
  return(bin_smoothed)
}

message("Calculating smoothed profiles...")
wt_profiles <- get_smoothed_profiles(combined, "WT (03Slide4)", genes_to_plot)
mutant_profiles <- get_smoothed_profiles(combined, "Mutant (03Slide4)", genes_to_plot)
wt111_profiles <- get_smoothed_profiles(combined, "WT (111125ST)", genes_to_plot)

# 4. Heatmap Data Preparation
scale_rows <- function(x) {
  res <- t(apply(x, 1, function(y) {
    if(sd(y) == 0) return(rep(0, length(y)))
    (y - mean(y)) / sd(y)
  }))
  rownames(res) <- rownames(x)
  return(res)
}

if (!is.null(wt_profiles) && !is.null(mutant_profiles) && !is.null(wt111_profiles)) {
  wt_plot <- scale_rows(wt_profiles)
  mutant_plot <- scale_rows(mutant_profiles)
  wt111_plot <- scale_rows(wt111_profiles)

  # Combine into a single matrix for plotting together
  combined_plot <- rbind(wt111_plot, wt_plot, mutant_plot)
  rownames(combined_plot) <- c(
    paste(target_gene, "WT (111125ST)"), 
    paste(target_gene, "WT (03Slide4)"), 
    paste(target_gene, "Mutant (03Slide4)")
  )

  # 5. Generate Heatmap
  col_palette <- colorRampPalette(c("#440154", "#21908C", "#FDE725"))(100) # Viridis
  break_val <- seq(-2, 2, length.out = 101)

  pheatmap(combined_plot, cluster_rows = FALSE, cluster_cols = FALSE, 
           show_colnames = FALSE, main = paste("Spatial Expression:", target_gene),
           color = col_palette, breaks = break_val)
} else {
  message(paste("Gene", target_gene, "not found or not enough data to create heatmap."))
}
```
