---
title: "Global Transcriptional Difference - 03Slide4 vs 111125ST_def"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(Seurat)
library(dplyr)
library(ggplot2)
library(viridis)
library(reshape2)
library(gridExtra)
library(patchwork)
library(scales)
library(zoo)

# Ensure output directory exists
if (!dir.exists("../figures/3D_glo")) dir.create("../figures/3D_glo", recursive = TRUE)
```

## 1. Load Data
This script compares the transcriptional similarity maps of two experiments. It expects `merged_guts.rds` to be present in the `Random data` folders of both `03Slide4` and `111125ST_def`.

```{r load_data}
path_slide4 <- "../Random data/merged_guts.rds"
path_slide6 <- "../../03Slide6/Random data/merged_guts.rds"
path_111125 <- "../../111125ST_def/Random data/merged_guts.rds"

if (!file.exists(path_slide4)) stop("Data for 03Slide4 not found at: ", path_slide4)
if (!file.exists(path_slide6)) stop("Data for 03Slide6 not found at: ", path_slide6)
if (!file.exists(path_111125)) stop("Data for 111125ST_def not found at: ", path_111125)

message("Loading 03Slide4 data...")
merged_slide4 <- readRDS(path_slide4)

message("Loading 03Slide6 data...")
merged_slide6 <- readRDS(path_slide6)

message("Loading 111125ST_def data...")
merged_111125 <- readRDS(path_111125)
```

## 2. Core Functions
Defining the functions for similarity calculation and difference plotting.

```{r functions}
# Internal function to calculate interpolated similarity matrix with Optional Smoothing
calculate_similarity_matrix <- function(seurat_obj, cells, n_bins = 100, var_features = NULL, smoothing_window = 1) {
    if (length(cells) == 0) return(NULL)
    
    sub <- subset(seurat_obj, cells = cells)
    
    if (is.null(var_features)) {
        sub <- FindVariableFeatures(sub, nfeatures = 2000, verbose = FALSE)
        var_features <- VariableFeatures(sub)
    }
    
    # Bin by Spatial_Pseudotime
    sub$Bin <- factor(cut(sub$Spatial_Pseudotime, breaks = n_bins, labels = FALSE), levels = 1:n_bins)
    
    # Calculate Mean Expression per Bin
    exp_mat <- GetAssayData(sub, layer = "data")[var_features, ]
    bin_means <- t(apply(exp_mat, 1, function(x) tapply(x, sub$Bin, mean, na.rm = TRUE)))
    
    # 1. Interpolate missing bins (handle gaps in data)
    bin_means <- t(apply(bin_means, 1, function(x) {
        if (all(is.na(x))) return(x)
        na.approx(x, na.rm = FALSE, rule = 2)
    }))
    
    # 2. Optional Smoothing across spatial bins (Moving Average)
    if (smoothing_window > 1) {
        bin_means <- t(apply(bin_means, 1, function(x) {
            # use rollmean from zoo
            as.numeric(rollmean(x, k = smoothing_window, fill = "extend"))
        }))
    }
    
    # Pairwise Spearman Correlation
    cor_mat <- cor(bin_means, method = "spearman", use = "pairwise.complete.obs")
    return(cor_mat)
}

# Function to plot the difference with Thresholding and Contrast control
plot_similarity_difference <- function(cor_mat1, cor_mat2, title = "Difference Map (Slide4 - 111125)", 
                                       n_bins = 100, diff_threshold = 0, contrast_limit = 0.5) {
    
    # Subtract matrices
    diff_mat <- cor_mat1 - cor_mat2
    
    # Transform to Diamonds
    n <- nrow(diff_mat)
    tri_data <- expand.grid(i = 1:n, j = 1:n) %>%
        filter(j >= i) %>%
        rowwise() %>%
        mutate(Difference = diff_mat[i, j]) %>%
        filter(!is.na(Difference))
    
    # Apply Threshold: Only keep differences above a certain level
    # Small differences are set to 0 (white)
    tri_data$Difference[abs(tri_data$Difference) < diff_threshold] <- 0
    
    poly_list <- lapply(1:nrow(tri_data), function(idx) {
        row <- tri_data[idx, ]
        cx <- (row$i + row$j) / 2
        cy <- (row$j - row$i) / 2
        data.frame(
            x = c(cx, cx + 0.5, cx, cx - 0.5),
            y = c(cy + 0.5, cy, cy - 0.5, cy),
            Difference = row$Difference,
            ID = idx
        )
    })
    poly_df <- do.call(rbind, poly_list)
    
    # Ensure limits are symmetric
    contrast_limit <- abs(contrast_limit)
    
    # Plot with divergent palette
    p <- ggplot(poly_df, aes(x = x, y = y, fill = Difference, group = ID)) +
        geom_polygon() +
        scale_fill_gradient2(
            high = "#B2182B", mid = "#F7F7F7", low = "#2166AC", 
            midpoint = 0, 
            limits = c(-contrast_limit, contrast_limit),
            oob = squish, # Rescale values outside limits to min/max
            name = "Diff\n(Mutant-WT)"
        ) +
        theme_minimal() +
        coord_fixed(ratio = 1) +
        scale_x_continuous(expand = c(0,0), limits = c(0.5, n + 0.5), breaks = seq(0, n, by = 20)) +
        scale_y_continuous(expand = c(0,0)) +
        labs(title = title, 
             subtitle = paste0("Difference Plot | Threshold: ", diff_threshold, 
                               " | Contrast Limit: ", contrast_limit),
             x = "Anatomical Position (Gut Bins)", 
             y = "Relative Distance") +
        theme(
            panel.grid = element_blank(),
            axis.text.y = element_blank(),
            axis.ticks.y = element_blank(),
            plot.title = element_text(face = "bold", size = 14)
        )
    
    # Add Region Labels
    region_breaks <- c(0, 15, 40, 55, 85, 100) * (n_bins / 100)
    region_mid <- (region_breaks[-1] + region_breaks[-length(region_breaks)]) / 2
    p <- p + annotate("text", x = region_mid, y = -2, label = c("R1", "R2", "R3", "R4", "R5"), fontface = "bold", size = 3.5) +
        annotate("segment", x = region_breaks, xend = region_breaks, y = -4, yend = 0, color = "grey20", linetype = "dotted")
        
    return(p)
}
```

## 3. Run Comparison
Adjust the parameters below to make differences more "striking".

```{r run_comparison, fig.width=10, fig.height=6}
# === ADJUSTABLE PARAMETERS ===
SMOOTHING_WINDOW <- 5      # Higher = smoother regional trends
DIFF_THRESHOLD   <- 0.05   # Ignore differences smaller than this
CONTRAST_LIMIT   <- 0.3   # Saturate color early (striking contrast)
# =============================

# Define common Variable Features for fair comparison
common_genes <- intersect(VariableFeatures(merged_slide4), VariableFeatures(merged_111125))
if (length(common_genes) < 500) {
    message("Few common variable features. Finding union...")
    common_genes <- union(VariableFeatures(merged_slide4), VariableFeatures(merged_111125))
}
common_genes <- common_genes[common_genes %in% rownames(merged_slide4) & common_genes %in% rownames(merged_111125)]

message(paste("Comparing using", length(common_genes), "genes..."))

# Filter Cells (Subset to specific guts and exclude Ambiguous)
message("Subsetting 03Slide4 to Gut 1...")
gut1_cells_slide4 <- grep("^Gut1_", Cells(merged_slide4), value = TRUE)
obj_slide4_gut1 <- subset(merged_slide4, cells = gut1_cells_slide4)
cells_slide4 <- Cells(obj_slide4_gut1)[!grepl("ambiguous", obj_slide4_gut1$Predicted_CellType, ignore.case = TRUE)]

message("Subsetting 03Slide6 to Gut 1 (WT+Cas9/GFP)...")
gut1_cells_slide6 <- grep("^Gut1_", Cells(merged_slide6), value = TRUE)
obj_slide6_gut1 <- subset(merged_slide6, cells = gut1_cells_slide6)
cells_slide6 <- Cells(obj_slide6_gut1)[!grepl("ambiguous", obj_slide6_gut1$Predicted_CellType, ignore.case = TRUE)]

message("Subsetting 111125ST_def to Gut 3...")
gut3_cells_111125 <- grep("^Gut3_", Cells(merged_111125), value = TRUE)
obj_111125_gut3 <- subset(merged_111125, cells = gut3_cells_111125)
cells_111125 <- Cells(obj_111125_gut3)[!grepl("ambiguous", obj_111125_gut3$Predicted_CellType, ignore.case = TRUE)]

# Calculate Matrices
message("Calculating similarity matrix for 03Slide4 (Gut 1)...")
sim_slide4 <- calculate_similarity_matrix(obj_slide4_gut1, cells_slide4, n_bins = 100, 
                                          var_features = common_genes, 
                                          smoothing_window = SMOOTHING_WINDOW)

message("Calculating similarity matrix for 111125ST_def (Gut 3)...")
sim_111125 <- calculate_similarity_matrix(obj_111125_gut3, cells_111125, n_bins = 100, 
                                          var_features = common_genes,
                                          smoothing_window = SMOOTHING_WINDOW)

# Plot Difference
p_diff <- plot_similarity_difference(sim_slide4, sim_111125, 
                                     title = "Difference Map (Mutant Gut 1 - WT Gut 3)",
                                     n_bins = 100,
                                     diff_threshold = DIFF_THRESHOLD,
                                     contrast_limit = CONTRAST_LIMIT)

# Save and Display
ggsave("../figures/3D_glo/Global_Difference_Map_Striking.png", p_diff, width = 10, height = 6)
p_diff
```

## 4. Gene Expression Across Gut Length
This section allows for the visualization of a specific gene's expression profile across the anterior-posterior axis, comparing the Mutant (Gut 1) and WT (Gut 3) directly.

```{r gene_expression_spatial, fig.width=10, fig.height=5}
# Define the target gene
target_gene <- "nkd" # Change this to the gene of interest (e.g., "wg", "dpp", "Dl")

# Function to extract and bin expression (Robust to missing genes)
get_binned_expression <- function(seurat_obj, cells_to_use, gene, label, n_bins = 100, smoothing = 3) {
    
    # Handle Missing Genes by returning a 0-expression profile
    if (!(gene %in% rownames(seurat_obj))) {
        message(paste("! Note: Gene", gene, "not found in", label, "- plotting 0."))
        return(data.frame(
            Bin = 1:n_bins,
            Expression = 0,
            Condition = label
        ))
    }

    sub <- subset(seurat_obj, cells = cells_to_use)
    
    # Bin by Spatial_Pseudotime
    sub$Bin <- factor(cut(sub$Spatial_Pseudotime, breaks = n_bins, labels = FALSE), levels = 1:n_bins)
    
    # Extract expression (ensure we use the data layer for normalized values)
    assay_to_use <- DefaultAssay(sub)
    expr <- GetAssayData(sub, assay = assay_to_use, layer = "data")[gene, ]
    bin_means <- tapply(expr, sub$Bin, mean, na.rm = TRUE)
    
    # Interpolate missing
    if (any(is.na(bin_means))) {
        bin_means <- as.numeric(na.approx(bin_means, na.rm = FALSE, rule = 2))
    }
    
    # Smooth
    if (smoothing > 1) {
         bin_means <- as.numeric(rollmean(bin_means, k = smoothing, fill = "extend"))
    }
    
    data.frame(
        Bin = 1:n_bins,
        Expression = bin_means,
        Condition = label
    )
}

# Check if gene exists in AT LEAST one dataset to avoid empty plotting
any_present <- target_gene %in% rownames(merged_slide4) || 
               target_gene %in% rownames(merged_slide6) || 
               target_gene %in% rownames(merged_111125)

if (any_present) {
    # Get data for all three conditions
    df_mutant   <- get_binned_expression(obj_slide4_gut1, cells_slide4, target_gene, "Mutant (03Slide4, Gut 1)", smoothing = SMOOTHING_WINDOW)
    df_wt_cas9  <- get_binned_expression(obj_slide6_gut1, cells_slide6, target_gene, "WT+Cas9/GFP (03Slide6, Gut 1)", smoothing = SMOOTHING_WINDOW)
    df_wt       <- get_binned_expression(obj_111125_gut3, cells_111125, target_gene, "WT (111125ST, Gut 3)", smoothing = SMOOTHING_WINDOW)
    
    plot_data <- rbind(df_mutant, df_wt_cas9, df_wt)
    
    # Plot
    p_gene <- ggplot(plot_data, aes(x = Bin, y = Expression, color = Condition)) +
        geom_line(size = 1.2) +
        scale_color_manual(values = c(
            "Mutant (03Slide4, Gut 1)" = "#E41A1C", 
            "WT+Cas9/GFP (03Slide6, Gut 1)" = "#4DAF4A",
            "WT (111125ST, Gut 3)" = "#377EB8"
        )) +
        theme_minimal() +
        labs(title = paste("Spatial Expression Profile:", target_gene),
             subtitle = "Comparison across Anterior-Posterior Axis",
             x = "Anatomical Position (Gut Bins, 1=Anterior, 100=Posterior)",
             y = "Mean Normalized Expression (Smoothed)") +
        theme(
            plot.title = element_text(face = "bold", size = 14),
            legend.position = "top",
            legend.title = element_blank(),
            panel.grid.minor = element_blank()
        ) +
         geom_vline(xintercept = c(15, 40, 55, 85), color = "grey50", linetype = "dashed", alpha = 0.5)
    
    print(p_gene)
    
    # Save the plot
    ggsave(paste0("../figures/3D_glo/Gene_Spatial_", target_gene, ".png"), p_gene, width = 10, height = 5)
    
} else {
    message(paste("Warning: Gene", target_gene, "not found in ANY of the datasets."))
}
```
## 5. Collapsed-Gut Comparison (Experiment Averages)
This section averages all individual guts for each specimen to show spatial trends at the experiment level.

```{r collapsed_gut_profiles, fig.width=10, fig.height=5}
if (any_present) {
    message("Calculating Collapsed-Gut profiles...")
    
    # 1. Prepare Cell Lists for Collapsed Guts (Exclude Ambiguous)
    get_collapsed_cells <- function(obj) {
        Cells(obj)[!grepl("ambiguous", obj$Predicted_CellType, ignore.case = TRUE)]
    }
    
    cells_coll_s4 <- get_collapsed_cells(merged_slide4)
    cells_coll_s6 <- get_collapsed_cells(merged_slide6)
    cells_coll_11 <- get_collapsed_cells(merged_111125)
    
    # 2. Extract Profiles
    df_coll_mutant <- get_binned_expression(merged_slide4, cells_coll_s4, target_gene, "Mutant (Collapsed)", smoothing = SMOOTHING_WINDOW)
    df_coll_cas9   <- get_binned_expression(merged_slide6, cells_coll_s6, target_gene, "WT+Cas9/GFP (Collapsed)", smoothing = SMOOTHING_WINDOW)
    df_coll_wt     <- get_binned_expression(merged_111125, cells_coll_11, target_gene, "WT (Collapsed)", smoothing = SMOOTHING_WINDOW)
    
    plot_data_coll <- rbind(df_coll_mutant, df_coll_cas9, df_coll_wt)
    
    # 3. Plot
    p_coll <- ggplot(plot_data_coll, aes(x = Bin, y = Expression, color = Condition)) +
        geom_line(size = 1.2) +
        scale_color_manual(values = c(
            "Mutant (Collapsed)" = "#E41A1C", 
            "WT+Cas9/GFP (Collapsed)" = "#4DAF4A",
            "WT (Collapsed)" = "#377EB8"
        )) +
        theme_minimal() +
        labs(title = paste("Spatial Expression Profile (Collapsed Guts):", target_gene),
             subtitle = "Averaged across all guts per experiment",
             x = "Anatomical Position (Gut Bins, 1=Anterior, 100=Posterior)",
             y = "Mean Normalized Expression (Smoothed)") +
        theme(
            plot.title = element_text(face = "bold", size = 14),
            legend.position = "top",
            legend.title = element_blank(),
            panel.grid.minor = element_blank()
        ) +
         geom_vline(xintercept = c(15, 40, 55, 85), color = "grey50", linetype = "dashed", alpha = 0.5)
    
    print(p_coll)
    
    # Save the plot
    ggsave(paste0("../figures/3D_glo/Gene_Spatial_Collapsed_", target_gene, ".png"), p_coll, width = 10, height = 5)
    
} else {
    message("Gene not present for collapsed comparison.")
}
```
