---
title: "Deepened GO Analysis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(Seurat)
library(dplyr)
library(ggplot2)
library(clusterProfiler)
library(org.Dm.eg.db)
library(stringr)
library(reshape2)
library(rioja)
library(dendextend)
library(ggsci)

# ------------------------------------------------------------------------------
# 1. Load Data
# ------------------------------------------------------------------------------
if (exists("merged_guts")) {
    cat("Using existing 'merged_guts' object from environment.\n")
    sub_guts <- merged_guts
} else {
    path_def <- "../../111125ST_def/Random data/merged_guts.rds"
    path_norm <- "../Random data/merged_guts.rds"
    if (file.exists(path_norm)) {
        cat("Loading merged_guts from", path_norm, "\n")
        sub_guts <- readRDS(path_norm)
    } else if (file.exists(path_def)) {
        cat("Loading merged_guts from", path_def, "\n")
        sub_guts <- readRDS(path_def)
    } else {
        stop("merged_guts not found. Please ensure it is loaded or available in '../Random data/merged_guts.rds'")
    }
}

# Ensure figures directory exists
if (!dir.exists("../figures/3D_glo")) dir.create("../figures/3D_glo", recursive=TRUE, showWarnings=FALSE)
```

## 1. Differential Gene Ontology Analysis for Specific Cell Types

We analyze target cell types: **aEC, mEC, Copper, LFC, pEC**.

```{r cell_type_go, fig.width=10, fig.height=12}
# Target cell types
target_cts <- c("aEC", "A3_aEC", "ItC", "Copper", "LFC", "Iron", "pEC")

# Check metadata column for cell types
# Based on Trajectory.Rmd, Most_Likely_CellType or Predicted_CellType is used.
if ("Most_Likely_CellType" %in% colnames(sub_guts@meta.data)) {
    ct_col <- "Most_Likely_CellType"
} else if ("Predicted_CellType" %in% colnames(sub_guts@meta.data)) {
    ct_col <- "Predicted_CellType"
} else {
    # Fallback to current Idents
    ct_col <- "Idents"
}

cat("Using metadata column:", ct_col, "\n")

# Create a custom detailed cell type column
sub_guts$Detailed_CellType <- as.character(sub_guts@meta.data[[ct_col]])

# Split mECs based on Zip42C.1 and Zip42C.2 expression
mEC_idx <- which(sub_guts$Detailed_CellType == "mEC")
if (length(mEC_idx) > 0) {
    expr_data <- GetAssayData(sub_guts, layer = "data")
    zip42_expr <- rep(0, length(mEC_idx))
    if ("Zip42C.1" %in% rownames(expr_data)) zip42_expr <- zip42_expr + expr_data["Zip42C.1", mEC_idx]
    if ("Zip42C.2" %in% rownames(expr_data)) zip42_expr <- zip42_expr + expr_data["Zip42C.2", mEC_idx]
    
    sub_guts$Detailed_CellType[mEC_idx[zip42_expr > 0]] <- "ItC"
    sub_guts$Detailed_CellType[mEC_idx[zip42_expr == 0]] <- "A3_aEC"
}

# Split LFCs based on CG17571 expression
lfc_idx <- which(sub_guts$Detailed_CellType == "LFC")
if (length(lfc_idx) > 0) {
    expr_data <- GetAssayData(sub_guts, layer = "data")
    cg17571_expr <- rep(0, length(lfc_idx))
    if ("CG17571" %in% rownames(expr_data)) cg17571_expr <- expr_data["CG17571", lfc_idx]
    
    sub_guts$Detailed_CellType[lfc_idx[cg17571_expr > 0]] <- "Iron"
    sub_guts$Detailed_CellType[lfc_idx[cg17571_expr == 0]] <- "LFC"
}

# Subset for target cell types
sub_ct <- subset(sub_guts, cells = Cells(sub_guts)[sub_guts$Detailed_CellType %in% target_cts])
Idents(sub_ct) <- factor(sub_ct$Detailed_CellType, levels = target_cts)

cat("\n============================================\n")
cat("CELL COUNTS FOR SPATIAL GO ANALYSIS:\n")
print(table(Idents(sub_ct)))
cat("============================================\n\n")

# 1.1 Find Markers
cat("Finding markers for target cell types...\n")
ct_markers <- FindAllMarkers(sub_ct, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)

# Special handling for aEC vs pEC to find distinguishing terms
# We run FindMarkers directly between them and add these to the pool
cat("Finding differential markers between aEC and pEC...\n")
a_vs_p <- FindMarkers(sub_ct, ident.1 = "aEC", ident.2 = "pEC", only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
p_vs_a <- FindMarkers(sub_ct, ident.1 = "pEC", ident.2 = "aEC", only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)

a_vs_p$gene <- rownames(a_vs_p)
a_vs_p$cluster <- "aEC"
p_vs_a$gene <- rownames(p_vs_a)
p_vs_a$cluster <- "pEC"

# Combine with main markers
ct_markers_combined <- rbind(ct_markers, a_vs_p, p_vs_a)

# 1.2 Run GO Enrichment
cat("Running compareCluster for Cell Type GO enrichment...\n")
# Increase to top 300 genes to ensure enough results for aEC/pEC
top_ct_genes <- ct_markers_combined %>% group_by(cluster) %>% top_n(n = 300, wt = avg_log2FC)
ct_gene_list <- split(top_ct_genes$gene, top_ct_genes$cluster)

ct_go_res <- compareCluster(geneCluster = ct_gene_list, fun = "enrichGO", 
                            OrgDb = org.Dm.eg.db, keyType = "SYMBOL", 
                            ont = "BP", pAdjustMethod = "BH", pvalueCutoff = 0.2) # Loosen cutoff for selection

# 1.3 Process and Plot
    # We avoid simplify() if it merges terms across aEC and pEC too aggressively, 
    # instead we select top 7 per cluster from the raw results.
    ct_go_all <- ct_go_res@compareClusterResult
    
    # Identify top 7 terms per cell type based on p.adjust
    top_terms_by_cluster <- ct_go_all %>%
        filter(p.adjust < 0.1) %>% # Set a reasonable significance threshold
        group_by(Cluster) %>%
        arrange(p.adjust) %>%
        slice_head(n = 7) %>%
        mutate(Rank = row_number()) %>%
        ungroup()
    
    # Ensure aEC and pEC have 7 terms by picking from higher p-values if needed
    if (nrow(top_terms_by_cluster %>% filter(Cluster == "aEC")) < 7 || 
        nrow(top_terms_by_cluster %>% filter(Cluster == "pEC")) < 7) {
        cat("Note: Loosening p-value filter to ensure 7 terms for aEC/pEC.\n")
        top_terms_by_cluster <- ct_go_all %>%
            group_by(Cluster) %>%
            arrange(p.adjust) %>%
            slice_head(n = 7) %>%
            mutate(Rank = row_number()) %>%
            ungroup()
    }
    
    # SPECIFIC FILTERING
    ct_plot_data_specific <- ct_go_all %>% 
        inner_join(top_terms_by_cluster %>% dplyr::select(Cluster, Description, Rank), 
                   by = c("Cluster", "Description"))
    
    ct_plot_data_specific$GeneRatio_num <- sapply(ct_plot_data_specific$GeneRatio, function(x) {
        parts <- as.numeric(strsplit(x, "/")[[1]])
        if(length(parts) == 2) parts[1] / parts[2] else 0
    })
    
    ct_plot_data_specific$Description <- str_wrap(ct_plot_data_specific$Description, width = 45)
    ct_plot_data_specific$Cluster <- factor(ct_plot_data_specific$Cluster, levels = target_cts)
    
    term_order_ct_specific <- top_terms_by_cluster %>% 
        arrange(factor(Cluster, levels=target_cts), Rank) %>%
        pull(Description) %>%
        unique() %>%
        str_wrap(width = 45)
    
    ct_plot_data_specific$Description <- factor(ct_plot_data_specific$Description, levels = rev(term_order_ct_specific))
    
    p_ct_go_specific <- ggplot(ct_plot_data_specific, aes(x = Cluster, y = Description, size = GeneRatio_num, color = p.adjust)) +
        geom_point() +
        scale_color_viridis_c(option = "magma", direction = -1, name = "P_adj") +
        theme_minimal() +
        labs(size = "Gene ratio", title = "Cell Type Specific GO Enrichment (Specific)", y = "", x = "") +
        theme(axis.text.x = element_text(size=12, face="bold", angle = 90, vjust = 0.5, hjust=1),
              axis.text.y = element_text(size=10),
              panel.grid.major.x = element_line(linetype="dotted", color="grey80"),
              panel.border = element_rect(color="black", fill=NA))
              
    ggsave("../figures/3D_glo/CellType_GO_Dotplot.png", p_ct_go_specific, width = 10, height = 12)
    print(p_ct_go_specific)
    
    # SHARED FILTERING: Top 5 terms across all
    top_terms_shared <- ct_go_all %>% group_by(Cluster) %>% arrange(p.adjust) %>% slice_head(n = 5) %>% mutate(Rank = row_number()) %>% ungroup()
    top_descriptions_shared <- unique(top_terms_shared$Description)
    ct_plot_data_shared <- ct_go_all %>% filter(Description %in% top_descriptions_shared)
    
    ct_plot_data_shared$GeneRatio_num <- sapply(ct_plot_data_shared$GeneRatio, function(x) {
        parts <- as.numeric(strsplit(x, "/")[[1]])
        if(length(parts) == 2) parts[1] / parts[2] else 0
    })
    
    ct_plot_data_shared$Description <- str_wrap(ct_plot_data_shared$Description, width = 45)
    ct_plot_data_shared$Cluster <- factor(ct_plot_data_shared$Cluster, levels = target_cts)
    
    term_ordering_shared <- top_terms_shared %>% group_by(Description) %>% slice_min(Rank, n=1, with_ties=FALSE) %>% ungroup() %>% arrange(factor(Cluster, levels=target_cts), Rank)
    term_order_shared <- str_wrap(term_ordering_shared$Description, width = 45)
    ct_plot_data_shared$Description <- factor(ct_plot_data_shared$Description, levels = rev(term_order_shared))
    
    p_ct_go_shared <- ggplot(ct_plot_data_shared, aes(x = Cluster, y = Description, size = GeneRatio_num, color = p.adjust)) +
        geom_point() +
        scale_color_viridis_c(option = "mako", direction = -1, name = "P_adj") +
        theme_minimal() +
        labs(size = "Gene ratio", title = "Cell Type Shared GO Enrichment (Top 5 Terms)", y = "", x = "") +
        theme(axis.text.x = element_text(size=12, face="bold", angle = 90, vjust = 0.5, hjust=1),
              axis.text.y = element_text(size=10),
              panel.grid.major.x = element_line(linetype="dotted", color="grey80"),
              panel.border = element_rect(color="black", fill=NA))
              
    ggsave("../figures/3D_glo/CellType_GO_Shared_Dotplot.png", p_ct_go_shared, width = 10, height = 12)
    print(p_ct_go_shared)
```

## 2. Constrained Hierarchical Clustering of Bins by GO Terms (2.5% Resolution)

We apply 40 bins (2.5% resolution) and run GO on all expressed genes per bin.

```{r bin_go_clustering, fig.width=12, fig.height=15}
# 2.1 Binning (40 bins = 2.5% resolution)
n_bins <- 40
sub_guts$Bin <- factor(cut(sub_guts$Spatial_Pseudotime, breaks = n_bins, labels = FALSE), levels = 1:n_bins)

# 2.2 Collect genes per bin
# "Run GO on all genes for each bin" -> We take genes with positive mean expression in the bin
# To keep it robust, we take the top N expressed genes in each bin as the 'active' set.
cat("Extracting active gene sets per bin (top 200 genes by mean expression)...\n")
exp_mat <- GetAssayData(sub_guts, layer = "data")
bin_gene_lists <- list()

for (i in 1:n_bins) {
    cells_in_bin <- Cells(sub_guts)[sub_guts$Bin == i]
    if (length(cells_in_bin) > 5) { # Minimum cells required
        bin_means <- rowMeans(exp_mat[, cells_in_bin])
        # Take top 200 genes for enrichment test
        active_genes <- names(sort(bin_means, decreasing = TRUE))[1:200]
        bin_gene_lists[[paste0("B", i)]] <- active_genes
    }
}

# 2.3 Run Bin GO Enrichment
cat("Running compareCluster for Bins (this may take a few minutes)...\n")
bin_go_res <- compareCluster(geneCluster = bin_gene_lists, fun = "enrichGO", 
                             OrgDb = org.Dm.eg.db, keyType = "SYMBOL", 
                             ont = "BP", pAdjustMethod = "BH", pvalueCutoff = 0.5)

# 2.4 Profile Matrix and Clustering
if (!is.null(bin_go_res) && nrow(bin_go_res@compareClusterResult) > 0) {
    
    # Create profile matrix: Bins (cols) x GO Terms (rows)
    # Value: -log10(p.adjust)
    res_df <- bin_go_res@compareClusterResult
    res_df$negLogP <- -log10(res_df$p.adjust)
    
    # Pivot to wide format for clustering
    profile_mat <- acast(res_df, Description ~ Cluster, value.var = "negLogP", fill = 0)
    
    # Handle bin ordering (B1, B2, ..., B40)
    bin_order <- paste0("B", 1:n_bins)
    available_bins <- intersect(bin_order, colnames(profile_mat))
    profile_mat <- profile_mat[, available_bins]
    
    # 2.5 Constrained Hierarchical Clustering (CONISS) on GO profiles
    cat("Performing CONISS on GO profiles...\n")
    dist_mat <- dist(t(profile_mat), method = "euclidean")
    go_hc <- chclust(dist_mat, method = "coniss")
    
    # 2.6 Constrained Dendrogram Visualization (Similar to reference)
    k_to_use <- 5 # Matching user image
    dend <- as.dendrogram(go_hc)
    dend <- color_branches(dend, k = k_to_use)
    cluster_labels <- cutree(go_hc, k = k_to_use)
    
    # Dutta Regions Math for Annotation
    dutta_breaks <- c(0, 15, 39, 53, 85, 100) * (n_bins / 100)
    display_breaks <- dutta_breaks + 0.5
    display_mids <- (display_breaks[-1] + display_breaks[-length(display_breaks)]) / 2
    region_labs <- c("R1", "R2", "R3", "R4", "R5")
    
    bg_colors <- scales::hue_pal()(k_to_use)
    cluster_rle <- rle(cluster_labels)
    ends <- cumsum(cluster_rle$lengths)
    starts <- c(1, ends[-length(ends)] + 1)
    
    rects_df <- data.frame(
        xmin = starts - 0.5,
        xmax = ends + 0.5,
        ymin = -Inf,
        ymax = max(go_hc$height) * 1.1,
        fill = factor(cluster_rle$values)
    )
    
    p_dend <- ggplot(as.ggdend(dend)) +
        geom_rect(data = rects_df, aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=fill), alpha = 0.2, inherit.aes = FALSE) +
        scale_fill_manual(values = bg_colors, guide = "none") +
        geom_vline(xintercept = display_breaks, color = "black", linetype = "dotted", size = 0.8) +
        annotate("text", x = display_mids, y = max(go_hc$height) * 1.05, 
                 label = region_labs, fontface = "bold", size = 5) +
        theme_minimal() +
        labs(title = "Constrained Hierarchical Clustering of Bins (GO Profile Basis)",
             subtitle = paste("CONISS Clustering | 40 Bins (2.5%) | K =", k_to_use),
             y = "Height (Dissimilarity)", x = "Anatomical Bin Position (Anterior -> Posterior)") +
        theme(axis.text.x = element_blank(), panel.grid = element_blank())

    ggsave("../figures/3D_glo/Bin_GO_Dendrogram.png", p_dend, width = 10, height = 6)
    print(p_dend)
    
    # 2.7 High-Res Comparative GO Map (Dotplot)
    # Pick top 2-3 terms per domain (R1-R5) or just interesting terms
    # We'll pick top 5 terms per cluster (k_to_use)
    res_df$Domain <- paste0("Domain_", cluster_labels[match(res_df$Cluster, available_bins)])
    
    top_bin_terms <- res_df %>%
        group_by(Domain) %>%
        arrange(p.adjust) %>%
        slice_head(n = 5) %>%
        pull(Description) %>%
        unique()
    
    bin_plot_data <- res_df[res_df$Description %in% top_bin_terms, ]
    bin_plot_data$Cluster <- factor(bin_plot_data$Cluster, levels = available_bins)
    
    # Numeric GeneRatio
    bin_plot_data$GeneRatio_num <- sapply(bin_plot_data$GeneRatio, function(x) {
        p <- as.numeric(strsplit(x, "/")[[1]])
        p[1] / p[2]
    })
    
    bin_plot_data$Description <- str_wrap(bin_plot_data$Description, width = 50)
    
    # Order Y-axis by spatial appearance
    term_order_bin <- bin_plot_data %>% 
        group_by(Description) %>% 
        summarize(min_bin = min(as.numeric(Cluster))) %>%
        arrange(min_bin) %>%
        pull(Description)
    bin_plot_data$Description <- factor(bin_plot_data$Description, levels = rev(term_order_bin))

    p_bin_dot <- ggplot(bin_plot_data, aes(x = Cluster, y = Description, size = GeneRatio_num, color = negLogP)) +
        geom_point() +
        scale_color_viridis_c(option = "magma", name = "-log10(P_adj)") +
        theme_minimal() +
        labs(size = "Gene ratio", title = "GO Profiles Along Trajectory (40 Bins)", y = "", x = "Spatial Bin") +
        theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8),
              axis.text.y = element_text(size = 9),
              panel.border = element_rect(color = "black", fill = NA))
    
    ggsave("../figures/3D_glo/HighRes_Bin_GO_Map.png", p_bin_dot, width = 12, height = 10)
    print(p_bin_dot)
    
} else {
    cat("Warning: No significant Bin GO enrichment found.\n")
}
```

## 3. Differential Gene Ontology Analysis for Single Cell Object

We load the single cell object downstream to obtain the GO enrichment for cell types that might be dropped out in the spatial data (like ZIP42C for mEC).

```{r sc_cell_type_go, fig.width=10, fig.height=12}
# 1. Ensure required packages for DGE and GO analysis are loaded
library(clusterProfiler)
library(enrichplot)
library(org.Dm.eg.db)
library(Seurat)

# --- Load Single Cell Data ---
sc_obj_sec3 <- readRDS("../Random data/sc_lineage_labeled.rds")

# Map clusters to expected names
sc_obj_sec3$Detailed_CellType <- as.character(sc_obj_sec3$Combined_Cluster)
sc_obj_sec3$Detailed_CellType[sc_obj_sec3$Detailed_CellType == "mEC_Zip_pos"] <- "ItC"
sc_obj_sec3$Detailed_CellType[sc_obj_sec3$Detailed_CellType == "mEC_Zip_neg"] <- "A3_aEC"

# Set the active identity
Idents(sc_obj_sec3) <- "Detailed_CellType"

# --- A. Differential Gene Expression (DGE) ---
# Find markers comparing Copper cells to ItC on single cell data
dge_results <- FindMarkers(sc_obj_sec3, 
                           ident.1 = "Copper", 
                           ident.2 = "ItC", 
                           test.use = "wilcox", 
                           min.pct = 0.25, 
                           logfc.threshold = 0.25)

# Filter for statistically significant markers (relax to p_val < 0.05 to capture more ItC genes)
sig_markers <- dge_results[dge_results$p_val < 0.05, ]

# Separate genes upregulated in Copper cells vs those upregulated in ItC
genes_up_copper <- rownames(sig_markers[sig_markers$avg_log2FC > 0, ])
genes_up_itc <- rownames(sig_markers[sig_markers$avg_log2FC < 0, ])

# --- B. Gene Identifier Conversion ---
entrez_up_copper <- mapIds(org.Dm.eg.db, 
                           keys = genes_up_copper, 
                           column = "ENTREZID", 
                           keytype = "SYMBOL", 
                           multiVals = "first")

entrez_up_itc <- mapIds(org.Dm.eg.db, 
                        keys = genes_up_itc, 
                        column = "ENTREZID", 
                        keytype = "SYMBOL", 
                        multiVals = "first")

# Clean up any NAs that failed to map
entrez_up_copper <- na.omit(entrez_up_copper)
entrez_up_itc <- na.omit(entrez_up_itc)

# --- C. Differential GO Enrichment Analysis ---
gene_clusters <- list(
  "Copper Cells" = entrez_up_copper, 
  "Interstitial Cells (ItC)" = entrez_up_itc
)

# Define the background universe (all expressed genes in the single cell object)
# This drastically increases statistical power by removing unexpressed genes from the multiple-testing penalty
expressed_genes <- rownames(GetAssayData(sc_obj_sec3, layer = "data"))
universe_entrez <- na.omit(mapIds(org.Dm.eg.db, 
                                  keys = expressed_genes, 
                                  column = "ENTREZID", 
                                  keytype = "SYMBOL", 
                                  multiVals = "first"))

# We set pvalueCutoff to 1.0 to guarantee we can retrieve the top 15 terms even if they have lower significance
comp_go <- compareCluster(geneCluster = gene_clusters, 
                          fun = "enrichGO", 
                          universe = universe_entrez,
                          OrgDb = org.Dm.eg.db, 
                          ont = "BP", 
                          pAdjustMethod = "BH", 
                          pvalueCutoff = 1.0, 
                          qvalueCutoff = 1.0, 
                          readable = TRUE)

# Simplify to remove redundant terms and prevent one cluster from overwhelming the plot
comp_go <- simplify(comp_go, cutoff = 0.7, by = "pvalue", select_fun = min)

# Force extract top 15 terms per cluster by pvalue
library(dplyr)
top15_terms <- comp_go@compareClusterResult %>%
    group_by(Cluster) %>%
    arrange(pvalue) %>%
    slice_head(n = 15) %>%
    pull(Description) %>%
    unique()

# --- D. Visualization ---
p_sec3 <- dotplot(comp_go, showCategory = top15_terms, title = "Differential GO: Copper Cells vs ItC (Single Cell Data)")

# Ensure the output directory exists
dir.create("../figures/3D_glo", recursive = TRUE, showWarnings = FALSE)

# Save the plot
ggsave("../figures/3D_glo/Sec3_Copper_vs_ItC_GO.png", p_sec3, width = 10, height = 12)
print(p_sec3)

# --- E. mEC vs Copper Cells (Ion, Transport, Assembly) ---
cat("\nComparing mEC vs Copper cells...\n")
dge_mec_cu <- tryCatch(FindMarkers(sc_obj_sec3, ident.1 = "A3_aEC", ident.2 = "Copper", test.use = "wilcox", min.pct = 0.1, logfc.threshold = 0.1), error = function(e) NULL)

if(!is.null(dge_mec_cu)) {
    sig_mec_cu <- dge_mec_cu[dge_mec_cu$p_val < 0.05, ]
    genes_up_mec <- rownames(sig_mec_cu[sig_mec_cu$avg_log2FC > 0, ])
    genes_up_cu2 <- rownames(sig_mec_cu[sig_mec_cu$avg_log2FC < 0, ])

    entrez_up_mec <- na.omit(mapIds(org.Dm.eg.db, keys = genes_up_mec, column = "ENTREZID", keytype = "SYMBOL", multiVals = "first"))
    entrez_up_cu2 <- na.omit(mapIds(org.Dm.eg.db, keys = genes_up_cu2, column = "ENTREZID", keytype = "SYMBOL", multiVals = "first"))

    gene_clusters_mec_cu <- list("mEC" = entrez_up_mec, "Copper Cells" = entrez_up_cu2)

    comp_mec_cu <- compareCluster(geneCluster = gene_clusters_mec_cu, fun = "enrichGO", OrgDb = org.Dm.eg.db, ont = "BP", pAdjustMethod = "BH", pvalueCutoff = 1.0, qvalueCutoff = 1.0, readable = TRUE)
    comp_mec_cu <- simplify(comp_mec_cu, cutoff = 0.7, by = "pvalue", select_fun = min)

    library(stringr)
    filtered_terms_mec <- comp_mec_cu@compareClusterResult %>%
        filter(str_detect(tolower(Description), "ion|transport|assembly")) %>%
        group_by(Cluster) %>%
        arrange(pvalue) %>%
        slice_head(n = 5) %>%
        pull(Description) %>%
        unique()

    if(length(filtered_terms_mec) > 0) {
        p_mec_cu <- dotplot(comp_mec_cu, showCategory = filtered_terms_mec, title = "mEC vs Copper: Ion, Transport, Assembly")
        ggsave("../figures/3D_glo/Sec3_mEC_vs_Copper_Transport_GO.png", p_mec_cu, width = 10, height = 12)
        print(p_mec_cu)
    } else {
        cat("\nNo ion/transport/assembly terms found for mEC vs Copper\n")
    }
}

# --- F. LFC vs Iron Cells ---
cat("\nComparing LFC vs Iron cells...\n")
# Ensure LFC cells are split by CG17571 to define Iron cells
lfc_idx_sc3 <- which(sc_obj_sec3$Detailed_CellType == "LFC")
if (length(lfc_idx_sc3) > 0) {
    expr_data_sc3 <- GetAssayData(sc_obj_sec3, layer = "data")
    cg17571_expr_sc3 <- rep(0, length(lfc_idx_sc3))
    if ("CG17571" %in% rownames(expr_data_sc3)) {
        cg17571_expr_sc3 <- expr_data_sc3["CG17571", lfc_idx_sc3]
    }
    sc_obj_sec3$Detailed_CellType[lfc_idx_sc3[cg17571_expr_sc3 > 0]] <- "Iron"
    sc_obj_sec3$Detailed_CellType[lfc_idx_sc3[cg17571_expr_sc3 == 0]] <- "LFC"
}
Idents(sc_obj_sec3) <- "Detailed_CellType"

dge_lfc_fe <- tryCatch(FindMarkers(sc_obj_sec3, ident.1 = "LFC", ident.2 = "Iron", test.use = "wilcox", min.pct = 0.1, logfc.threshold = 0.1), error = function(e) NULL)

if(!is.null(dge_lfc_fe)) {
    sig_lfc_fe <- dge_lfc_fe[dge_lfc_fe$p_val < 0.05, ]
    genes_up_lfc <- rownames(sig_lfc_fe[sig_lfc_fe$avg_log2FC > 0, ])
    genes_up_fe <- rownames(sig_lfc_fe[sig_lfc_fe$avg_log2FC < 0, ])
    
    entrez_up_lfc <- na.omit(mapIds(org.Dm.eg.db, keys = genes_up_lfc, column = "ENTREZID", keytype = "SYMBOL", multiVals = "first"))
    entrez_up_fe <- na.omit(mapIds(org.Dm.eg.db, keys = genes_up_fe, column = "ENTREZID", keytype = "SYMBOL", multiVals = "first"))
    
    gene_clusters_lfc_fe <- list("LFC" = entrez_up_lfc, "Iron Cells" = entrez_up_fe)
    
    comp_lfc_fe <- compareCluster(geneCluster = gene_clusters_lfc_fe, fun = "enrichGO", OrgDb = org.Dm.eg.db, ont = "BP", pAdjustMethod = "BH", pvalueCutoff = 1.0, qvalueCutoff = 1.0, readable = TRUE)
    comp_lfc_fe <- simplify(comp_lfc_fe, cutoff = 0.7, by = "pvalue", select_fun = min)
    
    top15_lfc_fe <- comp_lfc_fe@compareClusterResult %>%
        group_by(Cluster) %>%
        arrange(pvalue) %>%
        slice_head(n = 15) %>%
        pull(Description) %>%
        unique()
    
    if(length(top15_lfc_fe) > 0) {
        p_lfc_fe <- dotplot(comp_lfc_fe, showCategory = top15_lfc_fe, title = "Differential GO: LFC vs Iron Cells")
        ggsave("../figures/3D_glo/Sec3_LFC_vs_Iron_GO.png", p_lfc_fe, width = 10, height = 12)
        print(p_lfc_fe)
    }
}

```

```{r sc_html_plot, fig.width=12, fig.height=20}
# --- G. Interactive HTML Plot for all ItC and Copper GO terms ---
cat("\nGenerating Interactive HTML Plot for top 50 Copper vs ItC GO terms...\n")
library(plotly)
library(htmlwidgets)
library(dplyr)
library(stringr)

# Filter for significance, then extract top 50 terms per cluster
significant_terms <- comp_go@compareClusterResult %>% filter(p.adjust < 0.05)

if(nrow(significant_terms) > 0) {
    top50_html_terms <- significant_terms %>%
        group_by(Cluster) %>%
        arrange(pvalue) %>%
        slice_head(n = 50) %>%
        pull(Description) %>%
        unique()
    
    # Wrap text to prevent horizontal overlap, and subset comp_go
    comp_go_sig <- comp_go
    comp_go_sig@compareClusterResult <- significant_terms
    comp_go_sig@compareClusterResult$Description <- str_wrap(comp_go_sig@compareClusterResult$Description, width = 50)
    top50_html_terms_wrapped <- str_wrap(top50_html_terms, width = 50)

    p_all <- dotplot(comp_go_sig, showCategory = top50_html_terms_wrapped, title = "Top 50 Significant GO Terms: Copper vs ItC") +
        theme(axis.text.y = element_text(size = 10))
    
    # Increase height based on number of terms (approx 35px per term to completely prevent vertical overlap)
    n_terms <- length(top50_html_terms_wrapped)
    p_all_html <- ggplotly(p_all, height = max(800, n_terms * 35))
    
    saveWidget(p_all_html, file = "../figures/3D_glo/Sec3_Copper_vs_ItC_Top50_GO.html", selfcontained = TRUE)
    cat("Interactive HTML plot saved to ../figures/3D_glo/Sec3_Copper_vs_ItC_Top50_GO.html\n")
} else {
    cat("No significant GO terms remaining for HTML plot.\n")
}
```

```{r}
# --- H. Venn Diagram of Copper vs ItC Shared and Unique Genes ---
cat("\nGenerating Venn Diagram for Copper vs ItC genes...\n")
```

```{r sc_venn_diagram, fig.width=6.5, fig.height=4}
library(ggVennDiagram)

expr_copper <- GetAssayData(sc_obj_sec3, layer = "data")[, sc_obj_sec3$Detailed_CellType == "Copper", drop = FALSE]
expr_itc <- GetAssayData(sc_obj_sec3, layer = "data")[, sc_obj_sec3$Detailed_CellType == "ItC", drop = FALSE]

pct_copper <- rowSums(expr_copper > 0) / ncol(expr_copper)
pct_itc <- rowSums(expr_itc > 0) / ncol(expr_itc)

shared_genes <- rownames(expr_copper)[pct_copper > 0.25 & pct_itc > 0.25]
shared_genes <- setdiff(shared_genes, c(genes_up_copper, genes_up_itc))

venn_list <- list(
    "ItC" = c(genes_up_itc, shared_genes),
    "Copper Cells" = c(genes_up_copper, shared_genes)
)

p_venn <- ggVennDiagram(venn_list, label_alpha = 0) + 
    scale_fill_gradient(low = "white", high = "red") + 
    labs(title = "Gene Expression Overlap: ItC vs Copper Cells",
         subtitle = "Intersection = Shared highly expressed genes\nSides = Significantly upregulated markers") +
    coord_flip() + 
    scale_x_continuous(expand = expansion(mult = 0.3)) + 
    scale_y_continuous(expand = expansion(mult = 0.3)) +
    theme(legend.position = "bottom", aspect.ratio = 0.6) 

ggsave("../figures/3D_glo/Sec3_Copper_vs_ItC_Venn.png", p_venn, width = 8, height = 5.5)
print(p_venn)
```

```{r sc_venn_diagram_strict, fig.width=6.5, fig.height=4}
# --- I. Strict Venn Diagram based strongly on Log2 Fold Change (Bimod Test) ---
cat("\nGenerating Strict Venn Diagram for Copper vs ItC genes (bimod test, log2FC > 1.0, 2-fold)...\n")

# We use the 'bimod' Likelihood-ratio test (a classic scRNA-seq method) and a strict 2-fold change cutoff (logfc.threshold = 1.0)
# We also enforce adjusted p-value < 0.05 for ultimate stringency.
dge_strict <- FindMarkers(sc_obj_sec3, ident.1 = "Copper", ident.2 = "ItC", test.use = "bimod", min.pct = 0.25, logfc.threshold = 2.0)
sig_strict <- dge_strict[dge_strict$p_val_adj < 0.05, ]

genes_up_copper_strict <- rownames(sig_strict[sig_strict$avg_log2FC > 0, ])
genes_up_itc_strict <- rownames(sig_strict[sig_strict$avg_log2FC < 0, ])

# Shared genes logically include anything expressed broadly that didn't make the strict marker cutoff
shared_genes_strict <- rownames(expr_copper)[pct_copper > 0.25 & pct_itc > 0.25]
shared_genes_strict <- setdiff(shared_genes_strict, c(genes_up_copper_strict, genes_up_itc_strict))

venn_list_strict <- list(
    "ItC" = c(genes_up_itc_strict, shared_genes_strict),
    "Copper Cells" = c(genes_up_copper_strict, shared_genes_strict)
)

p_venn_strict <- ggVennDiagram(venn_list_strict, label_alpha = 0) + 
    scale_fill_gradient(low = "white", high = "red") + 
    labs(title = "Strict Expression Overlap: ItC vs Copper Cells",
         subtitle = "Intersection = Shared highly expressed genes\nSides = Strict Upregulated Markers (Bimod test, log2FC > 1, 2-fold)") +
    coord_flip() + 
    scale_x_continuous(expand = expansion(mult = 0.3)) + 
    scale_y_continuous(expand = expansion(mult = 0.3)) +
    theme(legend.position = "bottom", aspect.ratio = 0.6) 

ggsave("../figures/3D_glo/Sec3_Copper_vs_ItC_Venn_Strict.png", p_venn_strict, width = 8, height = 5.5)
print(p_venn_strict)
```

```{r sc_strict_marker_dotplot, fig.width=12, fig.height=4}
# --- J. Interactive HTML DotPlot of ALL Extreme Marker Genes ---
cat("\nGenerating Interactive HTML DotPlot for ALL extreme markers of Copper and ItC cells...\n")
library(plotly)
library(htmlwidgets)

# Order ALL strict markers by absolute log2FC
top_copper <- rownames(sig_strict[sig_strict$avg_log2FC > 0, ])
top_copper <- top_copper[order(sig_strict[top_copper, "avg_log2FC"], decreasing = TRUE)]

top_itc <- rownames(sig_strict[sig_strict$avg_log2FC < 0, ])
top_itc <- top_itc[order(sig_strict[top_itc, "avg_log2FC"], decreasing = FALSE)]

all_markers <- c(top_copper, top_itc)

if(length(all_markers) > 0) {
    # Temporarily set identity to plot just these two cell types clearly
    Idents(sc_obj_sec3) <- "Detailed_CellType"
    
    chunk_size <- 20
    dotplot_list <- list()
    
    for(i in seq(1, length(all_markers), by = chunk_size)) {
        end_idx <- min(i + chunk_size - 1, length(all_markers))
        genes_chunk <- all_markers[i:end_idx]
        part_num <- ceiling(i / chunk_size)
        
        p_dot <- DotPlot(sc_obj_sec3, features = genes_chunk, idents = c("Copper", "ItC")) + 
            theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
                  axis.text.y = element_text(size = 9)) +
            labs(title = paste0("All Strict Markers: Copper vs ItC (2-fold change) (Part ", part_num, ")"))
        
        # Save to list and print to the R environment
        dotplot_list[[part_num]] <- p_dot
        print(p_dot)
        
        # Convert to plotly and save as a separate, uncrowded HTML file
        p_dot_html <- ggplotly(p_dot, height = 300)
        html_file <- paste0("../figures/3D_glo/Sec3_Copper_vs_ItC_Strict_DotPlot_Part_", part_num, ".html")
        saveWidget(p_dot_html, file = html_file, selfcontained = TRUE)
        cat(sprintf("Saved HTML Part %d to %s\n", part_num, html_file))
    }
} else {
    cat("No markers survived the strict 4-fold cutoff to plot.\n")
}
```

```{r sc_custom_marker_dotplot, fig.width=5, fig.height=10}
# --- J2. Custom 20-Gene DotPlot (Copper vs ItC) ---
cat("\nGenerating Custom Top 20 Marker DotPlot for Copper and ItC...\n")

# Top 10 ItC markers
top_itc_10 <- head(genes_up_itc_strict[order(sig_strict[genes_up_itc_strict, "avg_log2FC"], decreasing = FALSE)], 10)

# Copper markers: 5 literature + 5 top (using harmonized uppercase names)
copper_lit <- c("LAB", "VHA16-1", "VHA100-4", "CTR1A", "CTR1B", "MTNA", "MTNB", "MTNC")
avail_lit <- intersect(copper_lit, genes_up_copper_strict)
selected_lit <- head(avail_lit, 5)

# Fill the rest with top markers not already selected
top_copper_ranked <- genes_up_copper_strict[order(sig_strict[genes_up_copper_strict, "avg_log2FC"], decreasing = TRUE)]
top_copper_ranked <- setdiff(top_copper_ranked, selected_lit)
selected_top_copper <- head(top_copper_ranked, 10 - length(selected_lit))

final_copper_10 <- c(selected_lit, selected_top_copper)
custom_20_markers <- c(final_copper_10, top_itc_10)

if(length(custom_20_markers) > 0) {
    Idents(sc_obj_sec3) <- "Detailed_CellType"
    p_dot_custom <- DotPlot(sc_obj_sec3, features = custom_20_markers, idents = c("Copper", "ItC")) + 
        coord_flip() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12, face = "bold"),
              axis.text.y = element_text(size = 12)) +
        labs(title = "Top 20 Markers: Copper (Literature + Top) vs ItC (Top)")
    
    ggsave("../figures/3D_glo/Sec3_Copper_vs_ItC_Custom20_DotPlot.png", p_dot_custom, width = 5, height = 10, bg = "white")
    print(p_dot_custom)
} else {
    cat("No markers available for custom plot.\n")
}
```

```{r sc_gene_module_go, fig.width=10, fig.height=12}
# --- K. Gene Module GO Enrichment (Top 50 Markers per Cell Type) ---
# Instead of relying on GSEA or variable-size marker lists,
# we force a fixed module of exactly 50 genes per cell type,
# ranked by log2FC from the bimod test (each vs all other cells).
# This puts Copper and ItC on perfectly equal statistical footing.
cat("\nCalculating top 50 gene modules for Copper and ItC...\n")

Idents(sc_obj_sec3) <- "Detailed_CellType"
module_size <- 200

# Copper module: top 50 genes upregulated in Copper vs all other cells
dge_copper <- FindMarkers(sc_obj_sec3, ident.1 = "Copper", test.use = "bimod", min.pct = 0.1, logfc.threshold = 0)
dge_copper <- dge_copper[dge_copper$avg_log2FC > 0 & dge_copper$p_val_adj < 0.05, ]
dge_copper <- dge_copper[order(dge_copper$avg_log2FC, decreasing = TRUE), ]
copper_module <- head(rownames(dge_copper), module_size)

# ItC module: top 50 genes upregulated in ItC vs all other cells
dge_itc <- FindMarkers(sc_obj_sec3, ident.1 = "ItC", test.use = "bimod", min.pct = 0.1, logfc.threshold = 0)
dge_itc <- dge_itc[dge_itc$avg_log2FC > 0 & dge_itc$p_val_adj < 0.05, ]
dge_itc <- dge_itc[order(dge_itc$avg_log2FC, decreasing = TRUE), ]
itc_module <- head(rownames(dge_itc), module_size)

cat(sprintf("Copper module: %d genes (requested %d)\n", length(copper_module), module_size))
cat(sprintf("ItC module:    %d genes (requested %d)\n", length(itc_module), module_size))
cat(sprintf("\nCopper top 10: %s\n", paste(head(copper_module, 10), collapse = ", ")))
cat(sprintf("ItC top 10:    %s\n", paste(head(itc_module, 10), collapse = ", ")))

# Map to Entrez IDs
entrez_copper_mod <- na.omit(mapIds(org.Dm.eg.db, keys = copper_module, column = "ENTREZID", keytype = "SYMBOL", multiVals = "first"))
entrez_itc_mod <- na.omit(mapIds(org.Dm.eg.db, keys = itc_module, column = "ENTREZID", keytype = "SYMBOL", multiVals = "first"))

cat(sprintf("\nCopper module mapped to Entrez: %d\n", length(entrez_copper_mod)))
cat(sprintf("ItC module mapped to Entrez:    %d\n", length(entrez_itc_mod)))

# Run GO enrichment on each module independently using compareCluster
module_clusters <- list(
    "Copper (Top 50)" = entrez_copper_mod,
    "ItC (Top 50)" = entrez_itc_mod
)

comp_go_modules <- compareCluster(geneCluster = module_clusters,
                                  fun = "enrichGO",
                                  universe = universe_entrez,
                                  OrgDb = org.Dm.eg.db,
                                  ont = "BP",
                                  pAdjustMethod = "BH",
                                  pvalueCutoff = 0.05,
                                  qvalueCutoff = 0.05,
                                  readable = TRUE)

if(nrow(comp_go_modules@compareClusterResult) > 0) {
    # Extract top 15 per cluster
    top15_module_terms <- comp_go_modules@compareClusterResult %>%
        group_by(Cluster) %>%
        arrange(p.adjust) %>%
        slice_head(n = 15) %>%
        pull(Description) %>%
        unique()
    
    cat(sprintf("\nTotal significant GO terms: %d\n", nrow(comp_go_modules@compareClusterResult)))
    cat(sprintf("Showing top terms: %d\n", length(top15_module_terms)))
    
    # Wrap descriptions
    comp_go_modules@compareClusterResult$Description <- stringr::str_wrap(comp_go_modules@compareClusterResult$Description, width = 50)
    top15_wrapped <- stringr::str_wrap(top15_module_terms, width = 50)
    
    p_mod_go <- dotplot(comp_go_modules, showCategory = top15_wrapped,
                        title = "GO Enrichment: Top 50 Gene Modules (p.adjust < 0.05)") +
        theme(axis.text.y = element_text(size = 9, lineheight = 0.8))
    
    ggsave("../figures/3D_glo/Sec3_Copper_vs_ItC_Module_GO.png", p_mod_go, width = 10, height = 10)
    print(p_mod_go)
} else {
    cat("No GO terms survived p.adjust < 0.05. Try increasing module_size to 100.\n")
}
```

```{r sc_venn_diagram_strict_lfc_fe, fig.width=6.5, fig.height=4}
# --- L. Strict Venn Diagram based strongly on Log2 Fold Change (LFC vs Iron) ---
cat("\nGenerating Strict Venn Diagram for LFC vs Iron cells (bimod test, log2FC > 1.0, 2-fold)...\n")

dge_strict_fe <- FindMarkers(sc_obj_sec3, ident.1 = "LFC", ident.2 = "Iron", test.use = "bimod", min.pct = 0.25, logfc.threshold = 1.0)
sig_strict_fe <- dge_strict_fe[dge_strict_fe$p_val_adj < 0.05, ]

genes_up_lfc_strict <- rownames(sig_strict_fe[sig_strict_fe$avg_log2FC > 0, ])
genes_up_fe_strict <- rownames(sig_strict_fe[sig_strict_fe$avg_log2FC < 0, ])

expr_lfc <- GetAssayData(sc_obj_sec3, layer = "data")[, sc_obj_sec3$Detailed_CellType == "LFC", drop = FALSE]
expr_fe <- GetAssayData(sc_obj_sec3, layer = "data")[, sc_obj_sec3$Detailed_CellType == "Iron", drop = FALSE]

pct_lfc <- rowSums(expr_lfc > 0) / ncol(expr_lfc)
pct_fe <- rowSums(expr_fe > 0) / ncol(expr_fe)

shared_genes_strict_fe <- rownames(expr_lfc)[pct_lfc > 0.25 & pct_fe > 0.25]
shared_genes_strict_fe <- setdiff(shared_genes_strict_fe, c(genes_up_lfc_strict, genes_up_fe_strict))

venn_list_strict_fe <- list(
    "Iron Cells" = c(genes_up_fe_strict, shared_genes_strict_fe),
    "LFC" = c(genes_up_lfc_strict, shared_genes_strict_fe)
)

p_venn_strict_fe <- ggVennDiagram(venn_list_strict_fe, label_alpha = 0) + 
    scale_fill_gradient(low = "white", high = "red") + 
    labs(title = "Strict Expression Overlap: LFC vs Iron Cells",
         subtitle = "Intersection = Shared highly expressed genes\nSides = Extreme Upregulated Markers (Bimod test, log2FC > 1, 2-fold)") +
    coord_flip() + 
    scale_x_continuous(expand = expansion(mult = 0.3)) + 
    scale_y_continuous(expand = expansion(mult = 0.3)) +
    theme(legend.position = "bottom", aspect.ratio = 0.6) 

ggsave("../figures/3D_glo/Sec3_LFC_vs_Iron_Venn_Strict.png", p_venn_strict_fe, width = 8, height = 5.5)
print(p_venn_strict_fe)
```

```{r sc_strict_marker_dotplot_lfc_fe, fig.width=12, fig.height=4}
# --- M. Interactive HTML DotPlot of ALL Extreme Marker Genes (LFC vs Iron) ---
cat("\nGenerating Interactive HTML DotPlot for ALL extreme markers of LFC and Iron cells...\n")

top_lfc <- genes_up_lfc_strict[order(sig_strict_fe[genes_up_lfc_strict, "avg_log2FC"], decreasing = TRUE)]
top_fe <- genes_up_fe_strict[order(sig_strict_fe[genes_up_fe_strict, "avg_log2FC"], decreasing = FALSE)]

all_markers_fe <- c(top_lfc, top_fe)

if(length(all_markers_fe) > 0) {
    Idents(sc_obj_sec3) <- "Detailed_CellType"
    chunk_size <- 20
    
    for(i in seq(1, length(all_markers_fe), by = chunk_size)) {
        end_idx <- min(i + chunk_size - 1, length(all_markers_fe))
        genes_chunk <- all_markers_fe[i:end_idx]
        part_num <- ceiling(i / chunk_size)
        
        p_dot <- DotPlot(sc_obj_sec3, features = genes_chunk, idents = c("LFC", "Iron")) + 
            theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
                  axis.text.y = element_text(size = 9)) +
            labs(title = paste0("Extreme Markers: LFC vs Iron (Part ", part_num, ")"))
        
        print(p_dot)
        
        p_dot_html <- ggplotly(p_dot, height = 300)
        html_file <- paste0("../figures/3D_glo/Sec3_LFC_vs_Iron_Strict_DotPlot_Part_", part_num, ".html")
        saveWidget(p_dot_html, file = html_file, selfcontained = TRUE)
        cat(sprintf("Saved HTML Part %d to %s\n", part_num, html_file))
    }
} else {
    cat("No markers survived the strict 4-fold cutoff to plot.\n")
}
```

```{r sc_custom_marker_dotplot_lfc_fe, fig.width=5, fig.height=10}
# --- M2. Custom 20-Gene DotPlot (LFC vs Iron) ---
cat("\nGenerating Custom Top 20 Marker DotPlot for LFC and Iron...\n")

top_lfc_10 <- head(genes_up_lfc_strict[order(sig_strict_fe[genes_up_lfc_strict, "avg_log2FC"], decreasing = TRUE)], 10)
top_fe_10 <- head(genes_up_fe_strict[order(sig_strict_fe[genes_up_fe_strict, "avg_log2FC"], decreasing = FALSE)], 10)

custom_20_fe_markers <- c(top_lfc_10, top_fe_10)

if(length(custom_20_fe_markers) > 0) {
    Idents(sc_obj_sec3) <- "Detailed_CellType"
    p_dot_custom_fe <- DotPlot(sc_obj_sec3, features = custom_20_fe_markers, idents = c("LFC", "Iron")) + 
        coord_flip() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12, face = "bold"),
              axis.text.y = element_text(size = 12)) +
        labs(title = "Top 20 Markers: LFC vs Iron")
    
    ggsave("../figures/3D_glo/Sec3_LFC_vs_Iron_Custom20_DotPlot.png", p_dot_custom_fe, width = 5, height = 10, bg = "white")
    print(p_dot_custom_fe)
} else {
    cat("No markers available for custom plot.\n")
}
```

```{r sc_gene_module_go_lfc_fe, fig.width=10, fig.height=12}
# --- N. Gene Module GO Enrichment (Top 50 Markers per Cell Type) for LFC vs Iron ---
cat("\nCalculating top 50 gene modules for LFC and Iron cells...\n")

Idents(sc_obj_sec3) <- "Detailed_CellType"
module_size <- 50

dge_lfc_mod <- FindMarkers(sc_obj_sec3, ident.1 = "LFC", test.use = "bimod", min.pct = 0.1, logfc.threshold = 0)
dge_lfc_mod <- dge_lfc_mod[dge_lfc_mod$avg_log2FC > 0 & dge_lfc_mod$p_val_adj < 0.05, ]
dge_lfc_mod <- dge_lfc_mod[order(dge_lfc_mod$avg_log2FC, decreasing = TRUE), ]
lfc_module <- head(rownames(dge_lfc_mod), module_size)

dge_fe_mod <- FindMarkers(sc_obj_sec3, ident.1 = "Iron", test.use = "bimod", min.pct = 0.1, logfc.threshold = 0)
dge_fe_mod <- dge_fe_mod[dge_fe_mod$avg_log2FC > 0 & dge_fe_mod$p_val_adj < 0.05, ]
dge_fe_mod <- dge_fe_mod[order(dge_fe_mod$avg_log2FC, decreasing = TRUE), ]
fe_module <- head(rownames(dge_fe_mod), module_size)

cat(sprintf("LFC module:  %d genes\n", length(lfc_module)))
cat(sprintf("Iron module: %d genes\n", length(fe_module)))

entrez_lfc_mod <- na.omit(mapIds(org.Dm.eg.db, keys = lfc_module, column = "ENTREZID", keytype = "SYMBOL", multiVals = "first"))
entrez_fe_mod <- na.omit(mapIds(org.Dm.eg.db, keys = fe_module, column = "ENTREZID", keytype = "SYMBOL", multiVals = "first"))

module_clusters_fe <- list(
    "LFC (Top 50)" = entrez_lfc_mod,
    "Iron Cells (Top 50)" = entrez_fe_mod
)

comp_go_modules_fe <- compareCluster(geneCluster = module_clusters_fe,
                                  fun = "enrichGO",
                                  universe = universe_entrez,
                                  OrgDb = org.Dm.eg.db,
                                  ont = "BP",
                                  pAdjustMethod = "BH",
                                  pvalueCutoff = 0.05,
                                  qvalueCutoff = 0.05,
                                  readable = TRUE)

if(!is.null(comp_go_modules_fe) && nrow(comp_go_modules_fe@compareClusterResult) > 0) {
    top15_module_terms_fe <- comp_go_modules_fe@compareClusterResult %>%
        group_by(Cluster) %>%
        arrange(p.adjust) %>%
        slice_head(n = 15) %>%
        pull(Description) %>%
        unique()
    
    comp_go_modules_fe@compareClusterResult$Description <- stringr::str_wrap(comp_go_modules_fe@compareClusterResult$Description, width = 50)
    top15_wrapped_fe <- stringr::str_wrap(top15_module_terms_fe, width = 50)
    
    p_mod_go_fe <- dotplot(comp_go_modules_fe, showCategory = top15_wrapped_fe,
                        title = "GO Enrichment: Top 50 Gene Modules (p.adjust < 0.05)") +
        theme(axis.text.y = element_text(size = 9, lineheight = 0.8))
    
    ggsave("../figures/3D_glo/Sec3_LFC_vs_Iron_Module_GO.png", p_mod_go_fe, width = 10, height = 10)
    print(p_mod_go_fe)
} else {
    cat("No GO terms survived p.adjust < 0.05.\n")
}
```

```{r sc_global_similarity_heatmap, fig.width=8, fig.height=7}
# --- O. Global Transcriptomic Similarity Matrix (Pearson) ---
cat("\nGenerating Global Transcriptomic Similarity Matrix (Pearson)...\n")

heatmap_cts <- c("aEC", "A3_aEC", "ItC", "Copper", "LFC", "Iron", "pEC")
sc_obj_heatmap <- subset(sc_obj_sec3, Detailed_CellType %in% heatmap_cts)
sc_obj_heatmap$Detailed_CellType <- factor(sc_obj_heatmap$Detailed_CellType, levels = heatmap_cts)

# Calculate Average Expression for the target cell types manually to avoid Seurat layer warnings
expr_mat <- GetAssayData(sc_obj_heatmap, layer = "data")
if(is.null(expr_mat) || nrow(expr_mat) == 0) {
    expr_mat <- GetAssayData(sc_obj_heatmap, slot = "data") # Fallback for older Seurat objects
}

avg_expr <- sapply(levels(sc_obj_heatmap$Detailed_CellType), function(ct) {
    cells <- colnames(sc_obj_heatmap)[sc_obj_heatmap$Detailed_CellType == ct]
    if (length(cells) == 0) return(rep(0, nrow(expr_mat)))
    if (length(cells) == 1) return(expr_mat[, cells])
    Matrix::rowMeans(expr_mat[, cells])
})
avg_expr <- as.matrix(avg_expr)

# Remove genes with zero variance across all cell types before correlation
var_genes <- apply(avg_expr, 1, var)
avg_expr_filtered <- avg_expr[var_genes > 0, ]

# Compute Pearson correlation matrix
cor_mat <- cor(avg_expr_filtered, method = "pearson")

# Ensure pheatmap is available
if(!requireNamespace("pheatmap", quietly = TRUE)) {
    install.packages("pheatmap", repos = "http://cran.us.r-project.org")
}
library(pheatmap)

# Create color palette similar to reference (blue to white to red)
heat_cols <- colorRampPalette(c("navy", "white", "firebrick3"))(100)

p_heat <- pheatmap::pheatmap(cor_mat, 
                             color = heat_cols, 
                             display_numbers = FALSE, 
                             fontsize = 12, 
                             main = "Global Transcriptomic Similarity Matrix (Pearson)",
                             angle_col = 45,
                             treeheight_row = 30,
                             treeheight_col = 30)

# Save the plot with explicitly white background to prevent transparency rendering as black
ggsave("../figures/3D_glo/Sec3_Global_Similarity_Heatmap.png", plot = p_heat$gtable, width = 8, height = 7, bg = "white")
```


## 4. Single Cell GO Enrichment (500-Gene Pipeline)

This section loads single-cell data and runs a comprehensive GO enrichment pipeline
with marker injection for under-represented cell types, publication-quality filtering
(p.adjust < 0.05), and specificity filtering (terms present in 1-3 cell types only).

```{r sc_go_500, fig.width=10, fig.height=14}
cat("\n\n--- Starting Single Cell Data Analysis (500-Gene Pipeline) ---\n")
sc_obj <- readRDS("../Random data/sc_lineage_labeled.rds")

# Ensure cell types are detailed
sc_obj$Detailed_CellType <- as.character(sc_obj$Combined_Cluster)
sc_obj$Detailed_CellType[sc_obj$Detailed_CellType == "mEC_Zip_pos"] <- "ItC"
sc_obj$Detailed_CellType[sc_obj$Detailed_CellType == "mEC_Zip_neg"] <- "A3_aEC"

# Split LFC by CG17571
lfc_idx_sc <- which(sc_obj$Detailed_CellType == "LFC")
if (length(lfc_idx_sc) > 0) {
    expr_data_sc <- GetAssayData(sc_obj, layer = "data")
    cg17571_expr_sc <- rep(0, length(lfc_idx_sc))
    if ("CG17571" %in% rownames(expr_data_sc)) {
        cg17571_expr_sc <- expr_data_sc["CG17571", lfc_idx_sc]
    }
    sc_obj$Detailed_CellType[lfc_idx_sc[cg17571_expr_sc > 0]] <- "Iron"
    sc_obj$Detailed_CellType[lfc_idx_sc[cg17571_expr_sc == 0]] <- "LFC"
}

target_cts_sc <- c("aEC", "ItC", "A3_aEC", "Copper", "Iron", "LFC", "pEC")

sub_sc <- subset(sc_obj, cells = Cells(sc_obj)[sc_obj$Detailed_CellType %in% target_cts_sc])
Idents(sub_sc) <- factor(sub_sc$Detailed_CellType, levels = target_cts_sc)

cat("\nSC CELL COUNTS:\n")
print(table(Idents(sub_sc)))

# --- Find Markers ---
cat("Finding markers for SC cell types...\n")
sc_markers <- FindAllMarkers(sub_sc, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)

cat("Finding differential SC markers between aEC and pEC...\n")
sc_a_vs_p <- FindMarkers(sub_sc, ident.1 = "aEC", ident.2 = "pEC", only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
sc_p_vs_a <- FindMarkers(sub_sc, ident.1 = "pEC", ident.2 = "aEC", only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)

sc_a_vs_p$cluster <- "aEC"
sc_a_vs_p$gene <- rownames(sc_a_vs_p)
sc_p_vs_a$cluster <- "pEC"
sc_p_vs_a$gene <- rownames(sc_p_vs_a)

sc_markers_combined <- rbind(sc_markers, sc_a_vs_p, sc_p_vs_a)

# --- Multi-strategy marker injection for mEC_ZIP42C_neg ---
cat("\n--- Marker injection for mEC_ZIP42C_neg ---\n")
neg_markers <- sc_markers_combined %>% filter(cluster == "A3_aEC")
cat(paste0("  Base markers for mEC_ZIP42C_neg: ", nrow(neg_markers), "\n"))

if (nrow(neg_markers) < 100) {
    # Strategy 1: vs ItC
    cat("  Injecting markers: mEC_ZIP42C_neg vs ItC...\n")
    neg_vs_itc <- tryCatch(
        FindMarkers(sub_sc, ident.1 = "A3_aEC", ident.2 = "ItC", 
                    only.pos = TRUE, min.pct = 0.1, logfc.threshold = 0.1),
        error = function(e) data.frame()
    )
    if (nrow(neg_vs_itc) > 0) {
        neg_vs_itc$cluster <- "A3_aEC"
        neg_vs_itc$gene <- rownames(neg_vs_itc)
        sc_markers_combined <- rbind(sc_markers_combined, neg_vs_itc)
        cat(paste0("    Added ", nrow(neg_vs_itc), " markers\n"))
    }
    
    # Strategy 2: vs Copper
    cat("  Injecting markers: mEC_ZIP42C_neg vs Copper...\n")
    neg_vs_cu <- tryCatch(
        FindMarkers(sub_sc, ident.1 = "A3_aEC", ident.2 = "Copper",
                    only.pos = TRUE, min.pct = 0.1, logfc.threshold = 0.1),
        error = function(e) data.frame()
    )
    if (nrow(neg_vs_cu) > 0) {
        neg_vs_cu$cluster <- "A3_aEC"
        neg_vs_cu$gene <- rownames(neg_vs_cu)
        sc_markers_combined <- rbind(sc_markers_combined, neg_vs_cu)
        cat(paste0("    Added ", nrow(neg_vs_cu), " markers\n"))
    }
}

# Deduplicate
sc_markers_combined <- sc_markers_combined %>% distinct(gene, cluster, .keep_all = TRUE)

# --- Run GO Enrichment (500 genes) ---
cat("\nRunning compareCluster for SC Cell Type GO enrichment (500 genes)...\n")
top_sc_genes <- sc_markers_combined %>% group_by(cluster) %>% top_n(n = 500, wt = avg_log2FC)
sc_gene_list <- split(top_sc_genes$gene, top_sc_genes$cluster)

cat("Gene counts per cell type:\n")
print(lengths(sc_gene_list))

sc_go_res <- compareCluster(geneCluster = sc_gene_list, fun = "enrichGO", 
                            OrgDb = org.Dm.eg.db, keyType = "SYMBOL", 
                            ont = "BP", pAdjustMethod = "BH", pvalueCutoff = 0.2)

sc_go_all <- sc_go_res@compareClusterResult

# --- SPECIFIC plot: publication-quality, 1-3 celltype specificity ---
cat("\n--- Building SPECIFIC plot ---\n")

# Count number of cell types per term
term_ct_count <- sc_go_all %>%
    filter(p.adjust < 0.05) %>%
    group_by(Description) %>%
    summarise(n_cts = n_distinct(Cluster)) %>%
    ungroup()

# Keep only terms in 1-3 celltypes (not ubiquitous)
specific_terms <- term_ct_count %>% filter(n_cts <= 3) %>% pull(Description)
cat(paste0("  Terms passing specificity filter (1-3 celltypes): ", length(specific_terms), "\n"))

sc_go_specific <- sc_go_all %>%
    filter(p.adjust < 0.05, Description %in% specific_terms)

top_terms_specific <- sc_go_specific %>%
    group_by(Cluster) %>%
    arrange(p.adjust) %>%
    slice_head(n = 7) %>%
    mutate(Rank = row_number()) %>%
    ungroup()

# If any cluster has <7, fill from all significant
for (ct in target_cts_sc) {
    ct_terms <- top_terms_specific %>% filter(Cluster == ct)
    if (nrow(ct_terms) < 7) {
        cat(paste0("  ", ct, " has only ", nrow(ct_terms), " specific terms, filling...\n"))
        extra <- sc_go_all %>%
            filter(Cluster == ct, p.adjust < 0.05, !Description %in% ct_terms$Description) %>%
            arrange(p.adjust) %>%
            slice_head(n = 7 - nrow(ct_terms)) %>%
            mutate(Rank = nrow(ct_terms) + row_number())
        top_terms_specific <- rbind(top_terms_specific, extra)
    }
}

sc_plot_data <- sc_go_all %>%
    inner_join(top_terms_specific %>% dplyr::select(Cluster, Description, Rank),
               by = c("Cluster", "Description"))

sc_plot_data$GeneRatio_num <- sapply(sc_plot_data$GeneRatio, function(x) {
    parts <- as.numeric(strsplit(x, "/")[[1]])
    if(length(parts) == 2) parts[1] / parts[2] else 0
})

sc_plot_data$Description <- str_wrap(sc_plot_data$Description, width = 45)
sc_plot_data$Cluster <- factor(sc_plot_data$Cluster, levels = target_cts_sc)

term_order_sc <- top_terms_specific %>%
    arrange(factor(Cluster, levels = target_cts_sc), Rank) %>%
    pull(Description) %>%
    unique() %>%
    str_wrap(width = 45)

sc_plot_data$Description <- factor(sc_plot_data$Description, levels = rev(unique(term_order_sc)))

p_sc_go <- ggplot(sc_plot_data, aes(x = Cluster, y = Description, size = GeneRatio_num, color = p.adjust)) +
    geom_point() +
    scale_color_viridis_c(option = "magma", direction = -1, name = "P_adj") +
    theme_minimal() +
    labs(size = "Gene ratio", title = "SC Cell Type GO Enrichment (Specific, 500 genes)", y = "", x = "") +
    theme(axis.text.x = element_text(size=12, face="bold", angle = 90, vjust = 0.5, hjust=1),
          axis.text.y = element_text(size=10),
          panel.grid.major.x = element_line(linetype="dotted", color="grey80"),
          panel.border = element_rect(color="black", fill=NA))

ggsave("../figures/3D_glo/SC_CellType_GO_Dotplot.png", p_sc_go, width = 10, height = 14)
print(p_sc_go)

# --- SHARED plot ---
cat("\n--- Building SHARED plot ---\n")
top_terms_shared <- sc_go_all %>%
    filter(p.adjust < 0.05) %>%
    group_by(Cluster) %>%
    arrange(p.adjust) %>%
    slice_head(n = 5) %>%
    mutate(Rank = row_number()) %>%
    ungroup()
top_descriptions_shared <- unique(top_terms_shared$Description)
sc_plot_shared <- sc_go_all %>% filter(Description %in% top_descriptions_shared, p.adjust < 0.05)

sc_plot_shared$GeneRatio_num <- sapply(sc_plot_shared$GeneRatio, function(x) {
    parts <- as.numeric(strsplit(x, "/")[[1]])
    if(length(parts) == 2) parts[1] / parts[2] else 0
})

sc_plot_shared$Description <- str_wrap(sc_plot_shared$Description, width = 45)
sc_plot_shared$Cluster <- factor(sc_plot_shared$Cluster, levels = target_cts_sc)

term_ordering_shared <- top_terms_shared %>%
    group_by(Description) %>%
    slice_min(Rank, n=1, with_ties=FALSE) %>%
    ungroup() %>%
    arrange(factor(Cluster, levels=target_cts_sc), Rank)
term_order_shared <- str_wrap(term_ordering_shared$Description, width = 45)
sc_plot_shared$Description <- factor(sc_plot_shared$Description, levels = rev(term_order_shared))

p_sc_go_shared <- ggplot(sc_plot_shared, aes(x = Cluster, y = Description, size = GeneRatio_num, color = p.adjust)) +
    geom_point() +
    scale_color_viridis_c(option = "mako", direction = -1, name = "P_adj") +
    theme_minimal() +
    labs(size = "Gene ratio", title = "SC Cell Type GO Enrichment (Shared Terms)", y = "", x = "") +
    theme(axis.text.x = element_text(size=12, face="bold", angle = 90, vjust = 0.5, hjust=1),
          axis.text.y = element_text(size=10),
          panel.grid.major.x = element_line(linetype="dotted", color="grey80"),
          panel.border = element_rect(color="black", fill=NA))

ggsave("../figures/3D_glo/SC_CellType_GO_Shared_Dotplot.png", p_sc_go_shared, width = 10, height = 12)
print(p_sc_go_shared)

# --- Pairwise: ItC vs mEC_ZIP42C_neg ---
cat("\n\n--- Pairwise Differential GO: ItC vs mEC_ZIP42C_neg ---\n")
# Find markers upregulated in pos
pw_pos <- sc_markers_combined %>% filter(cluster == "ItC") %>% top_n(n = 300, wt = avg_log2FC) %>% pull(gene)
pw_neg <- sc_markers_combined %>% filter(cluster == "A3_aEC") %>% top_n(n = 300, wt = avg_log2FC) %>% pull(gene)

pw_gene_list <- list("ItC" = pw_pos, "A3_aEC" = pw_neg)

pw_go_res <- compareCluster(geneCluster = pw_gene_list, fun = "enrichGO",
                             OrgDb = org.Dm.eg.db, keyType = "SYMBOL",
                             ont = "BP", pAdjustMethod = "BH", pvalueCutoff = 0.2)

if (!is.null(pw_go_res) && nrow(pw_go_res@compareClusterResult) > 0) {
    pw_df <- pw_go_res@compareClusterResult
    pw_top <- pw_df %>% filter(p.adjust < 0.05) %>% group_by(Cluster) %>% arrange(p.adjust) %>% slice_head(n = 10) %>% ungroup()
    
    if (nrow(pw_top) > 0) {
        pw_top$GeneRatio_num <- sapply(pw_top$GeneRatio, function(x) {
            parts <- as.numeric(strsplit(x, "/")[[1]])
            if(length(parts) == 2) parts[1] / parts[2] else 0
        })
        pw_top$Description <- str_wrap(pw_top$Description, width = 50)
        
        p_pw <- ggplot(pw_top, aes(x = Cluster, y = reorder(Description, -p.adjust), 
                                    size = GeneRatio_num, color = p.adjust)) +
            geom_point() +
            scale_color_viridis_c(option = "magma", direction = -1, name = "P_adj") +
            theme_minimal() +
            labs(title = "Pairwise GO: ItC vs mEC_ZIP42C_neg", y = "", x = "", size = "Gene ratio") +
            theme(axis.text.y = element_text(size = 9),
                  panel.border = element_rect(color = "black", fill = NA))
        
        ggsave("../figures/3D_glo/SC_Pairwise_ZIP42C_GO.png", p_pw, width = 10, height = 8)
        print(p_pw)
    }
}

# --- Pairwise: Iron vs LFC_CG17571_neg ---
cat("\n--- Pairwise Differential GO: Iron vs LFC_CG17571_neg ---\n")
pw_pos_cu <- sc_markers_combined %>% filter(cluster == "Iron") %>% top_n(n = 300, wt = avg_log2FC) %>% pull(gene)
pw_neg_cu <- sc_markers_combined %>% filter(cluster == "LFC") %>% top_n(n = 300, wt = avg_log2FC) %>% pull(gene)

pw_gene_list_cu <- list("Iron" = pw_pos_cu, "LFC" = pw_neg_cu)

pw_go_res_cu <- compareCluster(geneCluster = pw_gene_list_cu, fun = "enrichGO", 
                                OrgDb = org.Dm.eg.db, keyType = "SYMBOL", 
                                ont = "BP", pAdjustMethod = "BH", pvalueCutoff = 0.5)
    
if (!is.null(pw_go_res_cu) && nrow(pw_go_res_cu@compareClusterResult) > 0) {
    pw_df_cu <- pw_go_res_cu@compareClusterResult
    pw_top_cu <- pw_df_cu %>% filter(p.adjust < 0.05) %>% group_by(Cluster) %>% arrange(p.adjust) %>% slice_head(n = 10) %>% ungroup()
    
    if (nrow(pw_top_cu) > 0) {
        pw_top_cu$GeneRatio_num <- sapply(pw_top_cu$GeneRatio, function(x) {
            parts <- as.numeric(strsplit(x, "/")[[1]])
            if(length(parts) == 2) parts[1] / parts[2] else 0
        })
        pw_top_cu$Description <- str_wrap(pw_top_cu$Description, width = 50)
        
        p_pw_cu <- ggplot(pw_top_cu, aes(x = Cluster, y = reorder(Description, -p.adjust),
                                          size = GeneRatio_num, color = p.adjust)) +
            geom_point() +
            scale_color_viridis_c(option = "magma", direction = -1, name = "P_adj") +
            theme_minimal() +
            labs(title = "Pairwise GO: Iron vs LFC_CG17571_neg", y = "", x = "", size = "Gene ratio") +
            theme(axis.text.y = element_text(size = 9),
                  panel.border = element_rect(color = "black", fill = NA))
        
        ggsave("../figures/3D_glo/SC_Pairwise_CG17571_GO.png", p_pw_cu, width = 10, height = 8)
        print(p_pw_cu)
    }
}

# --- Spatial module scoring for ZIP42C markers ---
cat("\n--- Spatial Module Scoring for ZIP42C markers ---\n")
if (exists("sub_guts")) {
    top_pos_genes <- sc_markers_combined %>% filter(cluster == "ItC") %>% top_n(n = 50, wt = avg_log2FC) %>% pull(gene)
    top_neg_genes <- sc_markers_combined %>% filter(cluster == "A3_aEC") %>% top_n(n = 50, wt = avg_log2FC) %>% pull(gene)
    
    # Filter to genes present in spatial data
    spatial_genes <- rownames(sub_guts)
    pos_genes_spatial <- intersect(top_pos_genes, spatial_genes)
    neg_genes_spatial <- intersect(top_neg_genes, spatial_genes)
    
    cat(paste0("  ItC genes in spatial data: ", length(pos_genes_spatial), "\n"))
    cat(paste0("  mEC_ZIP42C_neg genes in spatial data: ", length(neg_genes_spatial), "\n"))
    
    if (length(pos_genes_spatial) >= 5 && length(neg_genes_spatial) >= 5) {
        sub_guts <- AddModuleScore(sub_guts, features = list(pos_genes_spatial), name = "ZIP42C_Pos_Module")
        sub_guts <- AddModuleScore(sub_guts, features = list(neg_genes_spatial), name = "ZIP42C_Neg_Module")
        
        if ("Spatial_Pseudotime" %in% colnames(sub_guts@meta.data)) {
            library(patchwork)
            plot_pos <- ggplot(sub_guts@meta.data, aes(x = x_centroid, y = y_centroid, color = ZIP42C_Pos_Module1)) +
                geom_point(size = 0.5) + scale_color_viridis_c(option = "inferno") +
                theme_minimal() + coord_fixed() + labs(title = "ItC Signature (Spatial)", color="Score")
            
            plot_neg <- ggplot(sub_guts@meta.data, aes(x = x_centroid, y = y_centroid, color = ZIP42C_Neg_Module1)) +
                geom_point(size = 0.5) + scale_color_viridis_c(option = "inferno") +
                theme_minimal() + coord_fixed() + labs(title = "mEC ZIP42C Neg Signature (Spatial)", color="Score")
        
            p_spatial <- plot_pos + plot_neg
            ggsave("../figures/3D_glo/Spatial_ZIP42C_Modules.png", p_spatial, width = 14, height = 6)
            print(p_spatial)
            cat("Spatial module maps saved to Spatial_ZIP42C_Modules.png!\n")
        }
    }
}

cat("Finished SC Analysis!\n")

# --- SC GO Subgroup Plots ---
cat("\n--- Building Subgroup GO Plots (Prioritizing Cell-Type Specificity) ---\n")

# Group 1: A3_aEC, ItC, Copper
group1_cts <- c("A3_aEC", "ItC", "Copper")
g1_all <- sc_go_all %>% filter(Cluster %in% group1_cts, p.adjust < 0.05)
g1_counts <- g1_all %>% group_by(Description) %>% summarise(n_cts_sub = n())

g1_terms <- g1_all %>%
    left_join(g1_counts, by = "Description") %>%
    group_by(Cluster) %>%
    arrange(n_cts_sub, p.adjust) %>%  # Sort by uniqueness in subgroup first, then significance
    slice_head(n = 7) %>%
    ungroup()

if(nrow(g1_terms) > 0) {
    g1_terms$GeneRatio_num <- sapply(g1_terms$GeneRatio, function(x) {
        parts <- as.numeric(strsplit(x, "/")[[1]])
        if(length(parts) == 2) parts[1] / parts[2] else 0
    })
    g1_terms$Description <- str_wrap(g1_terms$Description, width = 50)
    g1_terms$Cluster <- factor(g1_terms$Cluster, levels = group1_cts)
    
    g1_order <- g1_terms %>% arrange(Cluster, n_cts_sub, p.adjust) %>% pull(Description) %>% unique()
    g1_terms$Description <- factor(g1_terms$Description, levels = rev(g1_order))
    
    p_g1 <- ggplot(g1_terms, aes(x = Cluster, y = Description, size = GeneRatio_num, color = p.adjust)) +
        geom_point() +
        scale_color_viridis_c(option = "magma", direction = -1, name = "P_adj") +
        theme_minimal() +
        labs(title = "Top Differentiating GO Terms: A3_aEC, ItC, Copper", y = "", x = "", size = "Gene ratio") +
        theme(axis.text.x = element_text(size = 12, face = "bold"),
              axis.text.y = element_text(size = 9),
              panel.border = element_rect(color = "black", fill = NA))
    
    ggsave("../figures/3D_glo/SC_GO_Subgroup1.png", p_g1, width = 9, height = 12, bg = "white")
    print(p_g1)
}

# Group 2: LFC, Iron, pEC
group2_cts <- c("LFC", "Iron", "pEC")
g2_all <- sc_go_all %>% filter(Cluster %in% group2_cts, p.adjust < 0.05)
g2_counts <- g2_all %>% group_by(Description) %>% summarise(n_cts_sub = n())

g2_terms <- g2_all %>%
    left_join(g2_counts, by = "Description") %>%
    group_by(Cluster) %>%
    arrange(n_cts_sub, p.adjust) %>%
    slice_head(n = 7) %>%
    ungroup()

if(nrow(g2_terms) > 0) {
    g2_terms$GeneRatio_num <- sapply(g2_terms$GeneRatio, function(x) {
        parts <- as.numeric(strsplit(x, "/")[[1]])
        if(length(parts) == 2) parts[1] / parts[2] else 0
    })
    g2_terms$Description <- str_wrap(g2_terms$Description, width = 50)
    g2_terms$Cluster <- factor(g2_terms$Cluster, levels = group2_cts)
    
    g2_order <- g2_terms %>% arrange(Cluster, n_cts_sub, p.adjust) %>% pull(Description) %>% unique()
    g2_terms$Description <- factor(g2_terms$Description, levels = rev(g2_order))
    
    p_g2 <- ggplot(g2_terms, aes(x = Cluster, y = Description, size = GeneRatio_num, color = p.adjust)) +
        geom_point() +
        scale_color_viridis_c(option = "magma", direction = -1, name = "P_adj") +
        theme_minimal() +
        labs(title = "Top Differentiating GO Terms: LFC, Iron, pEC", y = "", x = "", size = "Gene ratio") +
        theme(axis.text.x = element_text(size = 12, face = "bold"),
              axis.text.y = element_text(size = 9),
              panel.border = element_rect(color = "black", fill = NA))
    
    ggsave("../figures/3D_glo/SC_GO_Subgroup2.png", p_g2, width = 9, height = 12, bg = "white")
    print(p_g2)
}

## 5. Curated Mid-Level GO Enrichment

#**Algorithm**:

#1. For each cell type independently, run `enrichGO` on its top marker genes.
#2. Constrain gene-set size (`minGSSize=5, maxGSSize=500`) to exclude hyper-broad parent terms while retaining mid-level processes.
#3. Apply `simplify(cutoff=0.5)` via GOSemSim to collapse semantically redundant terms into a single representative.
#4. Take the top 10 most significant non-redundant terms per cell type.
#5. Pool into a single dotplot.


cat("\n\n========== CURATED MID-LEVEL GO ENRICHMENT ==========\n")

if (!requireNamespace("GOSemSim", quietly = TRUE)) {
    BiocManager::install("GOSemSim")
}
library(GOSemSim)

# --- Build per-celltype gene lists from existing markers ---
# Reuse sc_markers_combined (the 500-gene pipeline markers with neg injection)
curated_genes <- sc_markers_combined %>% 
    group_by(cluster) %>% 
    top_n(n = 500, wt = avg_log2FC) %>% 
    ungroup()

# Rename clusters for visualization
curated_genes <- curated_genes %>%
    mutate(cluster = case_when(
        cluster == "ItC" ~ "ItC (ZIP42C pos mEC)",
        cluster == "A3_aEC" ~ "mEC",
        TRUE ~ as.character(cluster)
    ))

curated_gene_list <- split(curated_genes$gene, curated_genes$cluster)

# Update target_cts_sc for consistent ordering in Sections 5 and 6 plots
target_cts_sc <- c("aEC", "ItC (ZIP42C pos mEC)", "mEC", "Copper", "Iron", "LFC", "pEC")

cat("Gene counts per cell type for curated analysis:\n")
print(lengths(curated_gene_list))

# --- Per-celltype enrichGO + simplify ---
curated_results <- list()

for (ct in names(curated_gene_list)) {
    cat(paste0("\n--- Processing: ", ct, " ---\n"))
    
    ego <- enrichGO(
        gene         = curated_gene_list[[ct]],
        OrgDb        = org.Dm.eg.db,
        keyType      = "SYMBOL",
        ont          = "BP",
        pAdjustMethod = "BH",
        pvalueCutoff = 0.05,
        qvalueCutoff = 0.1,
        minGSSize    = 5,
        maxGSSize    = 500
    )
    
    if (is.null(ego) || nrow(ego@result %>% filter(pvalue < 0.05)) == 0) {
        cat(paste0("  No significant terms at pvalue<0.05 for ", ct, ", relaxing cutoff...\n"))
        ego <- enrichGO(
            gene         = curated_gene_list[[ct]],
            OrgDb        = org.Dm.eg.db,
            keyType      = "SYMBOL",
            ont          = "BP",
            pAdjustMethod = "none",
            pvalueCutoff = 0.2,
            qvalueCutoff = 0.5,
            minGSSize    = 3,
            maxGSSize    = 600
        )
    }
    
    if (!is.null(ego) && nrow(ego@result) > 0) {
        ego_simple <- simplify(ego, cutoff = 0.5, by = "pvalue", select_fun = min)
        
        df <- ego_simple@result %>% 
            filter(pvalue < 0.05) %>%
            arrange(pvalue) %>% 
            head(10)
        
        if (nrow(df) < 5) {
            ego_simple2 <- simplify(ego, cutoff = 0.7, by = "pvalue", select_fun = min)
            df <- ego_simple2@result %>% 
                filter(pvalue < 0.05) %>%
                arrange(pvalue) %>% 
                head(10)
        }
        
        if (nrow(df) > 0) {
            df$Cluster <- ct
            curated_results[[ct]] <- df
            cat(paste0("  Kept ", nrow(df), " terms after simplification\n"))
        } else {
            cat(paste0("  No terms survived for ", ct, "\n"))
        }
    } else {
        cat(paste0("  enrichGO returned no results for ", ct, "\n"))
    }
}

# --- Combine and plot ---
curated_all <- bind_rows(curated_results)

if (nrow(curated_all) > 0) {
    curated_all$GeneRatio_num <- sapply(curated_all$GeneRatio, function(x) {
        parts <- as.numeric(strsplit(x, "/")[[1]])
        if(length(parts) == 2) parts[1] / parts[2] else 0
    })
    
    term_sharing <- curated_all %>% group_by(Description) %>% summarise(n_cts = n_distinct(Cluster))
    cat("\n--- Term sharing distribution ---\n")
    print(table(term_sharing$n_cts))
    
    curated_all$Description <- str_wrap(curated_all$Description, width = 45)
    curated_all$Cluster <- factor(curated_all$Cluster, levels = target_cts_sc)
    
    curated_all <- curated_all %>% 
        group_by(Cluster) %>% 
        mutate(Rank = row_number()) %>% 
        ungroup()
    
    term_order_curated <- curated_all %>% 
        arrange(factor(Cluster, levels = target_cts_sc), Rank) %>% 
        pull(Description) %>% 
        unique()
    
    curated_all$Description <- factor(curated_all$Description, levels = rev(term_order_curated))
    
    all_curated_terms <- unique(as.character(curated_all$Description))
    
    curated_compare <- compareCluster(geneCluster = curated_gene_list, fun = "enrichGO",
                                      OrgDb = org.Dm.eg.db, keyType = "SYMBOL",
                                      ont = "BP", pAdjustMethod = "BH", 
                                      pvalueCutoff = 0.5, minGSSize = 5, maxGSSize = 500)
    
    cross_data <- curated_compare@compareClusterResult %>%
        mutate(Description = str_wrap(Description, width = 45)) %>%
        filter(Description %in% all_curated_terms, pvalue < 0.05)
    
    cross_data$GeneRatio_num <- sapply(cross_data$GeneRatio, function(x) {
        parts <- as.numeric(strsplit(x, "/")[[1]])
        if(length(parts) == 2) parts[1] / parts[2] else 0
    })
    
    cross_data$Cluster <- factor(cross_data$Cluster, levels = target_cts_sc)
    cross_data$Description <- factor(cross_data$Description, levels = rev(term_order_curated))
    cross_data <- cross_data %>% filter(!is.na(Description))
    
    p_curated <- ggplot(cross_data, aes(x = Cluster, y = Description, 
                                         size = GeneRatio_num, color = p.adjust)) +
        geom_point() +
        scale_color_viridis_c(option = "magma", direction = -1, name = "Adjusted\np-value") +
        scale_size_continuous(range = c(2, 8), name = "Gene Ratio") +
        theme_minimal(base_size = 12) +
        labs(title = "Curated GO Biological Processes",
             subtitle = "Per-celltype enrichment with semantic deduplication (simplify, cutoff=0.5)",
             y = "", x = "") +
        theme(
            axis.text.x = element_text(size = 12, face = "bold", angle = 90, vjust = 0.5, hjust = 1),
            axis.text.y = element_text(size = 10),
            panel.grid.major.x = element_line(linetype = "dotted", color = "grey80"),
            panel.border = element_rect(color = "black", fill = NA),
            plot.title = element_text(face = "bold", size = 14),
            plot.subtitle = element_text(size = 10, color = "grey40")
        )
    
    ggsave("../figures/3D_glo/SC_Curated_GO_Dotplot.png", p_curated, width = 12, height = 16)
    print(p_curated)
    cat("Curated GO dotplot saved!\n")
    
    summary_tbl <- curated_all %>% group_by(Cluster) %>% summarise(n_terms = n(), .groups = "drop")
    print(summary_tbl)
    
} else {
    cat("No curated GO terms were found across any cell type.\n")
}

cat("\n========== CURATED GO ANALYSIS COMPLETE ==========\n")

# ========== BROAD VIEW: high-level umbrella terms ==========
cat("\n\n========== BROAD VIEW GO ENRICHMENT ==========\n")

broad_results <- list()

for (ct in names(curated_gene_list)) {
    cat(paste0("\n--- Broad Processing: ", ct, " ---\n"))
    
    ego_broad <- enrichGO(
        gene         = curated_gene_list[[ct]],
        OrgDb        = org.Dm.eg.db,
        keyType      = "SYMBOL",
        ont          = "BP",
        pAdjustMethod = "BH",
        pvalueCutoff = 0.05,
        qvalueCutoff = 0.1,
        minGSSize    = 50,
        maxGSSize    = 1000
    )
    
    if (is.null(ego_broad) || nrow(ego_broad@result %>% filter(pvalue < 0.05)) == 0) {
        cat(paste0("  No significant broad terms for ", ct, ", relaxing...\n"))
        ego_broad <- enrichGO(
            gene         = curated_gene_list[[ct]],
            OrgDb        = org.Dm.eg.db,
            keyType      = "SYMBOL",
            ont          = "BP",
            pAdjustMethod = "none",
            pvalueCutoff = 0.2,
            qvalueCutoff = 0.5,
            minGSSize    = 30,
            maxGSSize    = 1500
        )
    }
    
    if (!is.null(ego_broad) && nrow(ego_broad@result) > 0) {
        ego_broad_simple <- simplify(ego_broad, cutoff = 0.4, by = "pvalue", select_fun = min)
        
        df_broad <- ego_broad_simple@result %>% 
            filter(pvalue < 0.05) %>%
            arrange(pvalue) %>% 
            head(10)
        
        if (nrow(df_broad) < 5) {
            ego_broad_simple2 <- simplify(ego_broad, cutoff = 0.6, by = "pvalue", select_fun = min)
            df_broad <- ego_broad_simple2@result %>% 
                filter(pvalue < 0.05) %>%
                arrange(pvalue) %>% 
                head(10)
        }
        
        if (nrow(df_broad) > 0) {
            df_broad$Cluster <- ct
            broad_results[[ct]] <- df_broad
            cat(paste0("  Kept ", nrow(df_broad), " broad terms\n"))
        }
    }
}

broad_all <- bind_rows(broad_results)

if (nrow(broad_all) > 0) {
    broad_all$GeneRatio_num <- sapply(broad_all$GeneRatio, function(x) {
        parts <- as.numeric(strsplit(x, "/")[[1]])
        if(length(parts) == 2) parts[1] / parts[2] else 0
    })
    
    broad_all$Description <- str_wrap(broad_all$Description, width = 45)
    broad_all$Cluster <- factor(broad_all$Cluster, levels = target_cts_sc)
    
    broad_all <- broad_all %>% 
        group_by(Cluster) %>% 
        mutate(Rank = row_number()) %>% 
        ungroup()
    
    term_order_broad <- broad_all %>% 
        arrange(factor(Cluster, levels = target_cts_sc), Rank) %>% 
        pull(Description) %>% 
        unique()
    
    broad_all$Description <- factor(broad_all$Description, levels = rev(term_order_broad))
    
    all_broad_terms <- unique(as.character(broad_all$Description))
    
    broad_compare <- compareCluster(geneCluster = curated_gene_list, fun = "enrichGO",
                                     OrgDb = org.Dm.eg.db, keyType = "SYMBOL",
                                     ont = "BP", pAdjustMethod = "BH", 
                                     pvalueCutoff = 0.5, minGSSize = 50, maxGSSize = 1000)
    
    cross_broad <- broad_compare@compareClusterResult %>%
        mutate(Description = str_wrap(Description, width = 45)) %>%
        filter(Description %in% all_broad_terms, p.adjust < 0.05)
    
    cross_broad$GeneRatio_num <- sapply(cross_broad$GeneRatio, function(x) {
        parts <- as.numeric(strsplit(x, "/")[[1]])
        if(length(parts) == 2) parts[1] / parts[2] else 0
    })
    
    cross_broad$Cluster <- factor(cross_broad$Cluster, levels = target_cts_sc)
    cross_broad$Description <- factor(cross_broad$Description, levels = rev(term_order_broad))
    cross_broad <- cross_broad %>% filter(!is.na(Description))
    
    p_broad <- ggplot(cross_broad, aes(x = Cluster, y = Description, 
                                        size = GeneRatio_num, color = p.adjust)) +
        geom_point() +
        scale_color_viridis_c(option = "magma", direction = -1, name = "Adjusted\np-value") +
        scale_size_continuous(range = c(2, 8), name = "Gene Ratio") +
        theme_minimal(base_size = 12) +
        labs(title = "Broad GO Biological Processes",
             subtitle = "High-level umbrella terms (minGSSize=50, maxGSSize=1000, simplify=0.4)",
             y = "", x = "") +
        theme(
            axis.text.x = element_text(size = 12, face = "bold", angle = 90, vjust = 0.5, hjust = 1),
            axis.text.y = element_text(size = 10),
            panel.grid.major.x = element_line(linetype = "dotted", color = "grey80"),
            panel.border = element_rect(color = "black", fill = NA),
            plot.title = element_text(face = "bold", size = 14),
            plot.subtitle = element_text(size = 10, color = "grey40")
        )
    
    ggsave("../figures/3D_glo/SC_Broad_GO_Dotplot.png", p_broad, width = 12, height = 16)
    print(p_broad)
    cat("Broad GO dotplot saved!\n")
}

cat("\n========== ALL CURATED GO ANALYSES COMPLETE ==========\n")

# ========== SPECIFIC VIEW: very granular metabolic processes ==========
cat("\n\n========== SPECIFIC METABOLIC GO ENRICHMENT ==========\n")

specific_results <- list()

for (ct in names(curated_gene_list)) {
    cat(paste0("\n--- Specific Processing: ", ct, " ---\n"))
    
    ego_spec <- enrichGO(
        gene         = curated_gene_list[[ct]],
        OrgDb        = org.Dm.eg.db,
        keyType      = "SYMBOL",
        ont          = "BP",
        pAdjustMethod = "BH",
        pvalueCutoff = 0.05,
        qvalueCutoff = 0.1,
        minGSSize    = 3,
        maxGSSize    = 500
    )
    
    if (is.null(ego_spec) || nrow(ego_spec@result %>% filter(p.adjust < 0.05)) == 0) {
        cat(paste0("  No significant specific terms for ", ct, ", relaxing maxGSSize...\n"))
        ego_spec <- enrichGO(
            gene         = curated_gene_list[[ct]],
            OrgDb        = org.Dm.eg.db,
            keyType      = "SYMBOL",
            ont          = "BP",
            pAdjustMethod = "BH",
            pvalueCutoff = 0.1,
            qvalueCutoff = 0.2,
            minGSSize    = 3,
            maxGSSize    = 800
        )
    }
    
    if (!is.null(ego_spec) && nrow(ego_spec@result) > 0) {
        # Keep unsimplified results to rescue metal/transport terms
        unsimplified_df <- ego_spec@result %>% filter(p.adjust < 0.05)
        
        ego_spec_simple <- simplify(ego_spec, cutoff = 0.7, by = "p.adjust", select_fun = min)
        
        df_spec <- ego_spec_simple@result %>% 
            filter(p.adjust < 0.05) %>%
            arrange(p.adjust) %>% 
            head(15)
        
        if (nrow(df_spec) < 5) {
            ego_spec_simple2 <- simplify(ego_spec, cutoff = 0.85, by = "p.adjust", select_fun = min)
            df_spec <- ego_spec_simple2@result %>% 
                filter(p.adjust < 0.05) %>%
                arrange(p.adjust) %>% 
                head(15)
        }
        
        # Rescue: if simplify collapsed away key metal/transport terms, force them back
        metal_pattern <- "iron|copper|zinc|metal|ferr|siderophore|transition metal"
        metal_terms <- unsimplified_df %>% 
            filter(str_detect(tolower(Description), metal_pattern)) %>%
            arrange(p.adjust) %>% head(3)
        
        if (nrow(metal_terms) > 0) {
            new_metal <- metal_terms %>% filter(!Description %in% df_spec$Description)
            if (nrow(new_metal) > 0) {
                df_spec <- rbind(df_spec, new_metal) %>% 
                    distinct(Description, .keep_all = TRUE) %>%
                    arrange(p.adjust) %>% 
                    head(15)
                cat(paste0("  Rescued ", nrow(new_metal), " metal/transport terms for ", ct, "\n"))
            }
        }
        
        df_spec <- head(df_spec, 15)
        
        if (nrow(df_spec) > 0) {
            df_spec$Cluster <- ct
            specific_results[[ct]] <- df_spec
            cat(paste0("  Kept ", nrow(df_spec), " specific terms\n"))
        }
    }
}

specific_all <- bind_rows(specific_results)

if (nrow(specific_all) > 0) {
    specific_all$GeneRatio_num <- sapply(specific_all$GeneRatio, function(x) {
        parts <- as.numeric(strsplit(x, "/")[[1]])
        if(length(parts) == 2) parts[1] / parts[2] else 0
    })
    
    specific_all$Description <- str_wrap(specific_all$Description, width = 45)
    specific_all$Cluster <- factor(specific_all$Cluster, levels = target_cts_sc)
    
    specific_all <- specific_all %>% 
        group_by(Cluster) %>% 
        mutate(Rank = row_number()) %>% 
        ungroup()
    
    term_order_spec <- specific_all %>% 
        arrange(factor(Cluster, levels = target_cts_sc), Rank) %>% 
        pull(Description) %>% 
        unique()
    
    specific_all$Description <- factor(specific_all$Description, levels = rev(term_order_spec))
    
    all_spec_terms <- unique(as.character(specific_all$Description))
    
    spec_compare <- compareCluster(geneCluster = curated_gene_list, fun = "enrichGO",
                                    OrgDb = org.Dm.eg.db, keyType = "SYMBOL",
                                    ont = "BP", pAdjustMethod = "BH", 
                                    pvalueCutoff = 0.5, minGSSize = 3, maxGSSize = 200)
    
    cross_spec <- spec_compare@compareClusterResult %>%
        mutate(Description = str_wrap(Description, width = 45)) %>%
        filter(Description %in% all_spec_terms, p.adjust < 0.05)
    
    cross_spec$GeneRatio_num <- sapply(cross_spec$GeneRatio, function(x) {
        parts <- as.numeric(strsplit(x, "/")[[1]])
        if(length(parts) == 2) parts[1] / parts[2] else 0
    })
    
    cross_spec$Cluster <- factor(cross_spec$Cluster, levels = target_cts_sc)
    cross_spec$Description <- factor(cross_spec$Description, levels = rev(term_order_spec))
    cross_spec <- cross_spec %>% filter(!is.na(Description))
    
    p_specific <- ggplot(cross_spec, aes(x = Cluster, y = Description, 
                                          size = GeneRatio_num, color = p.adjust)) +
        geom_point() +
        scale_color_viridis_c(option = "magma", direction = -1, name = "Adjusted\np-value") +
        scale_size_continuous(range = c(2, 8), name = "Gene Ratio") +
        theme_minimal(base_size = 12) +
        labs(title = "Specific Metabolic GO Processes",
             subtitle = "Fine-grained metabolic terms (minGSSize=3, maxGSSize=200)",
             y = "", x = "") +
        theme(
            axis.text.x = element_text(size = 12, face = "bold", angle = 90, vjust = 0.5, hjust = 1),
            axis.text.y = element_text(size = 10),
            panel.grid.major.x = element_line(linetype = "dotted", color = "grey80"),
            panel.border = element_rect(color = "black", fill = NA),
            plot.title = element_text(face = "bold", size = 14),
            plot.subtitle = element_text(size = 10, color = "grey40")
        )
    
    ggsave("../figures/3D_glo/SC_Specific_Metabolic_GO_Dotplot.png", p_specific, width = 12, height = 16)
    print(p_specific)
    cat("Specific metabolic GO dotplot saved!\n")
}

cat("\n========== ALL GO ANALYSES COMPLETE ==========\n")
```

## 6. Targeted Metal/Ion Transport GO Activity

This section focuses exclusively on ItC, Copper, and Iron cells — the three metal-handling cell types of the midgut — to extract GO terms related to copper, iron, zinc, and generic ion/metal transport.

```{r metal_ion_go, fig.width=10, fig.height=14}
cat("\n\n========== METAL/ION TRANSPORT GO ACTIVITY ==========\n")

# Subset curated_gene_list to only the 3 metal-relevant cell types
metal_cts <- c("ItC (ZIP42C pos mEC)", "Copper", "Iron")
metal_gene_list <- curated_gene_list[metal_cts]
cat("Gene counts for metal cell types:\n")
print(lengths(metal_gene_list))

# Run a dedicated compareCluster with small gene-set sizes to capture specific metal terms
metal_compare <- compareCluster(geneCluster = metal_gene_list, fun = "enrichGO",
                                OrgDb = org.Dm.eg.db, keyType = "SYMBOL",
                                ont = "BP", pAdjustMethod = "BH", 
                                pvalueCutoff = 1.0, minGSSize = 3, maxGSSize = 500)

metal_df <- metal_compare@compareClusterResult

# Filter for metal/ion related terms that are significant
metal_pattern <- "iron|copper|zinc|metal|ferr|siderophore|transition metal|ion transport|cation|anion transport|calcium|potassium|sodium|proton|transmembrane transport|homeostasis|detoxif|oxidat|redox|v-type|vacuolar"
metal_filtered <- metal_df %>% 
    filter(str_detect(tolower(Description), metal_pattern), p.adjust < 0.05)

cat(paste0("Total significant metal/ion terms found: ", nrow(metal_filtered), "\n"))

if (nrow(metal_filtered) > 0) {
    # Pick top 20 unique terms per cell type, prioritising cell-type specificity
    metal_counts <- metal_filtered %>% group_by(Description) %>% summarise(n_cts = n())
    
    metal_top <- metal_filtered %>%
        left_join(metal_counts, by = "Description") %>%
        group_by(Cluster) %>%
        arrange(n_cts, p.adjust) %>%
        slice_head(n = 20) %>%
        ungroup()
    
    # Keep only the selected terms, but show cross-enrichment across all 3 cell types
    selected_terms <- unique(metal_top$Description)
    plot_data_metal <- metal_filtered %>% filter(Description %in% selected_terms)
    
    # Parse GeneRatio
    plot_data_metal$GeneRatio_num <- sapply(plot_data_metal$GeneRatio, function(x) {
        parts <- as.numeric(strsplit(x, "/")[[1]])
        if(length(parts) == 2) parts[1] / parts[2] else 0
    })
    
    plot_data_metal$Description <- str_wrap(plot_data_metal$Description, width = 55)
    plot_data_metal$Cluster <- factor(plot_data_metal$Cluster, levels = metal_cts)
    
    # Order terms by cell type where they are most significant
    term_ordering_df <- plot_data_metal %>%
        group_by(Description) %>%
        arrange(p.adjust) %>%
        slice_head(n = 1) %>%
        ungroup() %>%
        arrange(factor(Cluster, levels = metal_cts), p.adjust)
    
    term_order_metal <- term_ordering_df$Description
    plot_data_metal$Description <- factor(plot_data_metal$Description, levels = rev(term_order_metal))
    
    p_metal <- ggplot(plot_data_metal, aes(x = Cluster, y = Description, size = GeneRatio_num, color = p.adjust)) +
        geom_point() +
        scale_color_viridis_c(option = "magma", direction = -1, name = "Adjusted\np-value") +
        theme_minimal(base_size = 12) +
        labs(title = "Metal & Ion Transport GO Activity",
             subtitle = "ItC, Copper and Iron cells — significant terms only (p.adjust < 0.05)",
             size = "Gene Ratio", y = "", x = "") +
        theme(
            axis.text.x = element_text(size = 12, face = "bold"),
            axis.text.y = element_text(size = 9),
            panel.grid.major.x = element_line(linetype = "dotted", color = "grey80"),
            panel.border = element_rect(color = "black", fill = NA)
        )
    
    ggsave("../figures/3D_glo/SC_Metal_Ion_GO_Dotplot.png", p_metal, width = 10, height = 14, bg = "white")
    print(p_metal)
    cat("Metal/Ion GO dotplot saved to SC_Metal_Ion_GO_Dotplot.png!\n")
} else {
    cat("No significant metal/ion GO terms found.\n")
}

cat("\n========== SCRIPT FULLY COMPLETE ==========\n")
```
